### 2018-08-13 paper/design.tex
migrating from gitlab ugp:af90fa20

\section{Design considerations}
The development of our tool was guided by experts in urban sciences and
sociology whose main concern is to identify similar groups and how they evolved
over time, with as much detail as possible.

Practically, beyond the basis requirement that our method needs to be able to
work with non-geographically harmonized data, we divided the other requirements
into two different categories: \textit{Method}, the practical aspects of the
clustering method and classification, and \textit{Interface}, concerning the
visual representation of the information.

\begin{enumerate}
    \item[M1]{\textbf{Geographical information}: The clustering method needs to
    consider the available geographical information along with the data
    associated to the region.}

    \item[M2]{\textbf{Parameter configuration}: The configuration parameters for
    the clustering method should be configurable by the user.}

    \item[I1]{\textbf{Temporal evolution}: The evolution of each region over
    time can be inferred by its associated clusters. This evolution needs to be
    easily represented. }

    \item[I2]{\textbf{Cluster characteristics}:  The visual representation of
    each cluster should easily convey its relevant characteristics, including
    both geographic and content information.}

    \item[I3]{\textbf{Details of the changes}: Once a geographic region is
    selected, the interface should clearly convey how the associated information
    changed over time.}
\end{enumerate}

### 2018-08-13 paper/discussion.tex
migrating from gitlab ugp:af90fa20

\section{Discussion and limitations}

The objective of this work was to leverage a graph based data representation
with visual tools to allow for the exploration of geographically inconsistent
census data. While we successfully replicated and corroborated results from the
literature, this method still has significant limitations.

Removing the geographical normalization/interpolation step greatly reduces the
amount of work necessary, but the method still requires consistent variables
across the years. Matching the fields of the public census can be trivial for
some aspects, like Age, but challenging for others, like Income. The divulged
income ranges vary over time, the actual value changes due to inflation and
other factors, and so on. Moreover, some fields were not considered in earlier
censuses, such as Race in Canada, or Hispanic population in the USA, hampering its
use when they are available. We matched some aspects, but a deeper demographic
analysis would greatly benefit from all available information.

Another limitation is the lack of control on how much the geographical
information will impact the clustering result. While the adopted method met our
needs for this work, a configurable control would add another dimension to the
exploration, allowing for more intra-cluster variance to obtain more 'compact'
clusters. We explored changing the number of content based augmented edges, but
this proved to be unreliable and hard to interpret. The~\emph{ClustGeo}
method~\cite{Chavent2017} can be a viable option for this, allowing a graph
based input and a hierarchical output, combined using a single mixing
parameter. Alternatively, one could cluster the changes~\cite{bian2018survey}
instead of the stable states.

There are also technological limitations, such as memory use on the
visualization client. To allow for changes on the CTs over the years, we use a
geographic file that contains all possible intersections, which can grow rather
large if the original city was expansive and contained several CTs, like NYC or
LA. However, the most significant technological limitation relates to parameters
that are not immediately interactive, such as the clustering configuration.
Since the clustering is computationally expensive and performed on the server,
which allows for cached results, some changes can take a few minutes to be
considered, removing any possibility of a continuous exploration.

Indeed, the cognitive load on the user is already significant, as we compromised
simplicity for accuracy. While other works labelled the clusters, as 'young
urban', 'struggling', and so on~\cite{Delmelle2016,Delmelle2017}, we show the
statistical characteristics of the clusters, which are harder to interpret, as
the data may have subtle nuances that labels would otherwise hide. This also led
to a crowded interface, mitigated somewhat the use of pop-up panels and collapsible
sections. For some cities, especially if they are small and stable, the panels
can appear redundant, but each provide a different way to interact with the
information that can ease the exploration process for larger and dynamic cities.

### 2018-08-13 paper/intro.tex
migrating from gitlab ugp:af90fa20

% General context + motivation
\IEEEPARstart{U}{rban} sciences are blooming thanks to a renewed interest in
understanding and improving the urban environment. Visual analytics is following
this trend, fueled by new public datasets that encompass progressively more of
our daily lives~\cite{Chen2017}. There is no shortage of methods to explore
mobility patterns~\cite{VonLandesberger2016}, social media~\cite{Chen2017},
traffic~\cite{chen2015survey}, and so on, providing experts, planners, policy
makers, and the general population with deeper insights about their cities.

These new datasets usually contain GPS coordinates for the records, leading to
\emph{point-based} data. Combined with the corresponding timestamps, this data
is easily suitable for longitudinal analysis. But most demographic datasets are
\emph{region-based}, where the measurements are associated with pre-defined
regions, not only for an additional level of privacy protection, but because
some measurements only make sense over a defined area. Census data is a classic
example of this format, with datasets available from 1790 onwards for the
US~\cite{nhgis}. Despite this unmatched temporal availability, longitudinal
analyses of census data are often restricted in time, especially when smaller
tabulation areas are considered, such as census tracts (CT) or dissemination
areas, which evolve to reflect changes in population density, leading to
geographic inconsistencies across time, and the traditional time-series based
approach is no longer viable.

However, these analyses are necessary to understand the urban environment.
Indeed, two different regions can have similar average income for a given year,
while one is experiencing a process of economic improvement and the other one
impoverishment. Quite obviously, a single snapshot cannot be used to identify
gentrification, migration, education changes, or any of the relevant processes
that happen over time.

To overcome these inconsistencies, the traditional approach is the
\emph{geographical harmonization} of the data, the interpolation of the
measurements into a common set of
regions\cite{Logan2014,Hallisey2017,Allen2018}, so that each variable can be
represented using time-series. This is laborious work that inevitably introduces
some amount of error~\cite{Logan2016}, even when additional data is
provided~\cite{eicher2001dasymetric}. Nevertheless, this step is considered
mandatory in the current literature: \emph{"(...) tract-by-tract comparison is
not possible unless data from 2000 is interpolated to 2010 boundaries
(...)"}~\cite{Dmowska2017}.

The main contribution of this application paper is an visualization-based
alternative to the geographical harmonization, a combination of established
graph based processing and information visualization techniques allowing
tract-by-tract comparison, the identification and visualization of patterns of
demographic evolution without geographic harmonization, effectively removing one
of the most challenging problems in longitudinal demographic analysis. We also
include illustrative scenarios and our prototype is available at
\url{http://uoft.me/piccard}, including more than fourty regions in the US and
Canada.

% We adopt census data as familiar operational
% context, but the proposed methodology is suitable for most region based
% measurements.

### 2018-08-13 paper/letter/letter.tex
migrating from gitlab ugp:af90fa20

\documentclass{article}
\usepackage[a4paper, total={6in, 8in}]{geometry}
\usepackage{tcolorbox}
\usepackage{amssymb}

\newcommand{\reply}[1]{\begin{tcolorbox}\noindent\textbf{Comment:}\\#1\hfill\end{tcolorbox}}
\newcommand{\didThis}[1]{\underline{#1}\marginpar{\checkmark}}

\title{Previous Reviews}

\begin{document}
\maketitle
A previous version of this paper was submitted to IEEE VAST 2018. While the full
text of the reviews is attached, we would like to overview the changes made,
leveraging the summary review:

\begin{itemize}
    \item{Evaluation: no evaluation of the proposed clustering methods (R3), the details
    of the domain expert interview are missing. (R1, R3)

    \reply{We did not introduce a new clustering method. The original paper
    contains such evaluations, we added that explicitly on the text. All the
    communication with the experts is included, verbatim, in the supplementary
    material, with the exception of Patrick Adler, which is a member of the
    Urban Genome Project, but did not collaborate directly in this work.}}

    \item{Design: the consideration and final choice of design are not clearly discussed.
    Specially, the color usage is confusing. (R2, R3, R4).

    }

    \item{Motivation: the authors fail to clearly demonstrate the motivation of their
    work, probably due to their writing style. (R2, R4)}

    \item{Presentation: no technical contribution clearly stated, writing style not
    qualified as a good research paper, paper organization can be improved (R2, R4)}
\end{itemize}

\section{Reviewer 1 - score 4/5}

\begin{itemize}
\item{\textbf{Paper type}\\Application}
\item{\textbf{Expertise}\\Expert}
\item{\textbf{Overall Rating}

    4 - Accept

    The paper should be accepted with some minor
    revisions.

    Once these have been completed it will meet the quality standard.}

\item{\textbf{Supplemental Materials}\\Acceptable}

\item{\textbf{Justification}

    the paper presents an interesting work on visualizing demographic evolution using
    geographically inconsistent census data. The strength of this paper is putting all
    the interesting pieces to present something an application that does show some
    useful insights into the data across time. The weakness of this paper is lack of
    anything new in particular and focused on a very specific type of demographic
    region (CT).
}
\item{\textbf{The Review}

    the paper presents an interesting work on visualizing demographic evolution using
    geographically inconsistent census data.

    Overall the paper reads well and it is easy to read and follow all the technical
    work. Most of the diagrams are useful. The paper tackles on a difficult problem of
    visualizing temporal data where each temporal piece may not be consistent with
    other pieces.

    The strength of this paper is putting all the interesting pieces to present
    something an application that does show some useful insights into the data across
    time. The weakness of this paper is lack of anything new in particular and focused
    on a very specific type of demographic region (CT). In summary, the work uses a
    spatio-temporal graph based approach and perform clustering based on the
    attributes of each node where each node represents a CT for a given year. The
    edges are added based on neighboring CT's or if they are same CT's but different
    years. This approach seems reasonable and has been adopted to model such problems
    in the past. The authors then perform clustering an introduce trajectories which
    all seems useful to the end-uses. Overall, the work seems useful to the community
    as it put together many ideas in the same application and it seems to work well
    with specific datasets and problem in hand.

    The weakness of the paper is very limited evaluation as mostly it was feedback
    from experts and the user-study lacked details. Also, most of the work is
    performed using CT as the region in census data, it would have been nice to see if
    this work is also applicable to other census regions as well.

    \reply{The specific region used for the analysis is irrelevant to the
    method, we could have used mixed regions for instance. We adopted CTs
    because it is the most finely grained data available including more than
    population count. Additionally, other census areas do not change as
    frequently as CTs. }

    \textbf{Revision Required:}
    Figure 2: while useful could use more detail and better explanation of the
    algorithm, it could do a better job in showing clustering and explaing the diagram
    since this is the figure that explains the entire paper.
}
\end{itemize}

\section{Reviewer 2 - score 2/5}
\begin{itemize}

\item{\textbf{Paper type}\\Application}

\item{\textbf{Expertise}\\Knowledgeable}

\item{\textbf{Overall Rating}

    2 - Reject

    The paper is not ready for publication in VAST / TVCG.

    The work may have some value but the paper requires major revisions or
    additional work that are beyond the scope of the conference review cycle to meet
    the quality standard. Without this I am not going to be able to return a score of
    '4 - Accept'.}

\item{\textbf{Supplemental Materials}\\Acceptable}

\item{\textbf{Justification}

    In this paper, the authors propose a method for longitudinal data analysis that
    avoids the geographical harmonization. The paper focuses on an interesting topic.
    The weakness of the paper is that the motivation and design requirements are not
    clearly demonstrated, and the design process is not justified.}

\item{\textbf{The Review}

    In this paper, the authors propose a method for longitudinal data analysis that
    avoids the geographical harmonization. Specifically, they adopt an improved
    clustering algorithm and a graph based visualization to help users identify
    clusters and the demographical details of each cluster. Though this paper is
    interesting to read, it still has several issues to be addressed. The authors fail
    to clearly demonstrate the motivation and the design requirements of the proposed
    work. A number of important details are also missing such as the discussion about
    design choices. In addition, the paper is not easy to follow in many sections such
    as the description of the teaser. I feel these issues can be hardly fully
    addressed in a minor revision and I suggest the authors to keep improving the
    work.

    My first concern is that the authors fail to clearly demonstrate the motivation of
    their work. The authors mentioned that longitudinal analysis of census data is
    missing and its first step is geographical harmonization that is difficult.
    Therefore, they propose a method that analyze geographical data without the need
    of geographical harmonization. However, they fail to demonstrate that why the
    longitudinal analysis of census data is needed and important. In addition, they
    also fail to illustrate why it is important and challenging for omitting the
    process of geographical harmonization in the data analysis. The authors need to
    explicitly describe the motivation of the proposed work.

    In addition, the authors are suggested to clearly discuss about the design
    considerations. For example, “Details of the changes” is not clear enough. Many
    questions can be asked: What kind of information the authors want to address for
    the changes? The geographic information or the content information or both? The
    authors need to clearly describe the design considerations.

    Moreover, the design choices are also not discussed. The authors need to discuss
    how they make the final design decisions and if there are any alternative design
    choices. For example, the authors use stacked bar charts to encode the information
    of each aspect, in which the rectangle width represents the percentage of the
    corresponding variable. Many questions can be asked: Why the author choose the
    stacked bar charts? Do they consider other design choices such as pie charts? Why
    they use the length of rectangle to represent percentage? Do they consider color
    saturation or hue? The authors need to clearly discuss the reasons they finalize
    the current design.

    The writings are hard to follow sometimes. For example, the description of the
    teaser is really difficult to understand. Many questions can be asked: What does
    each part represent in the figure? How to tell “While the whole city follows this
    trend”? What regions are “these regions”? What are the “relevant clusters” and why
    they are “visible”? The authors need to clearly explain the figures and improve
    the writing.

    In summary, the authors are suggested to revised the paper and improve their work
    to address the above issues.
}
\end{itemize}

\section{Reviewer 3 - score 3/5}
\begin{itemize}

\item{\textbf{Paper type}\\Application}

\item{\textbf{Expertise}\\Expert}

\item{\textbf{Overall Rating}

    3.5 - Between Possible Accept and Accept}

\item{\textbf{Supplemental Materials}\\Acceptable with minor revisions (specify
revisions in The Review section)}

\item{\textbf{Justification}

    This paper targets an interesting research question on analyzing the evolution of
    geographically inconsistent data. The major technical contributions of this paper
    are
    \begin{enumerate} 
        \item{A graph based spatio-temporal clustering method;}
        \item{an enhanced box plot visualization to better support comparison;}
        \item{the heatmap-like map view showing the spatial information of clusters.}
    \end{enumerate}
    \noindent Another big bonus is that the authors provided an interactive demo
    online. Still, there are several issues on 
    \begin{enumerate} 
        \item{the lack of evaluation of the proposed clustering method;} 
        \item{the design choice, mainly for the color usage;}
        \item{the lack of details of expert review process.}
    \end{enumerate}
    \noindent Detailed reviews can be found in the review section.}

\item{\textbf{The Review}

    In this paper, the authors proposed a visual analytics system to explore and
    analyze the evolution of geographically inconsistent data. Several visualization
    techniques, such as enhanced box plots, Sankey diagram, heatmap-like map
    visualization with small multiple settings, are used in the system. In general,
    the author tackles on an interesting and not easy research question, and the paper
    is well written and easy to follow. The case studies and expert interviews have
    clearly shown its usefulness and effectiveness.

    The major contributions have been mentioned above in justification section, here
    list the details of the major concerns.
    \begin{enumerate}

    \item{There is no proper evaluation of the proposed clustering methods.

    In Section 4.2, the authors proposed a graph based geographic content clustering,
    which considers spatial, temporal, and similarity information among regions. Here,
    it will be worth to conducting comparisons with some traditional methods, for
    example, k-means purely based on histogram distances or weighted distance by
    considering spatio-temporal information (specific for the first one, since I think
    there is a high correlation of spatio-temporal information with the histogram
    distance).
    \reply{TODO}
    }

    \item{The color usage for the whole visual analytic system is sometimes confusing.

    In the application, the authors use colors in different ways across
    visualizations. In the Sankey graph, the color is used to encode the categories.
    In the map view, additional opacity is used here (and blending is optionally used)
    together with the same color encoding as the one in the Sankey graph. From the
    right figure in Figure 6, it is clear that the color is not distinguishable (this
    is also related to the clustering itself since the three clusters are too small
    compared to the others). What makes it worse is the color usage in the box plot
    view to encode the relevance, which is very confusing even with the legend shown
    on top since there is some color overlap with the categorical encoding (e.g., the
    green and grape color).

    \reply{Colour assignment was indeed challenge for this work, mostly because
    it needs to be used, but there isn't a definitive way to actually do it.
    That was also the reason for artificially limiting the maxium number of
    clusters to eight. There is indeed overlap between the colours used for
    relevance and the clusters, mostly because the clusters used virtually all
    of the colour space. However, we believe that their different meanings are
    clear, since none of the experts, nor the other reviewers, mentioned any
    confusion in this specific aspect. It is not ideal, but we believe it is a
    acceptable approximation.}}

    \item{The details of the domain expert interview is missing in the paper.  
    I really appreciate that the authors attached the detailed expert review comments
    and the online demo is available. However, in the main paper, there is a very
    little description of how the expert interviews were conducted. Since there are
    still plenty of space, I would suggest the authors to add more details in Section
    6.}
    \end{enumerate}
    \textbf{Minor issue}:
    The demo video can be improved by providing a more comprehensive walkthrough of a
    case study instead of fragmented videos to explain the functionalities.
}
\end{itemize}

\section{Reviewer 4 - score 2/5}

\begin{itemize}

\item{\textbf{Paper type}\\Technique}

\item{\textbf{Expertise}\\Expert}

\item{\textbf{Overall Rating}

    2 - Reject

    The paper is not ready for publication in VAST / TVCG.

    The work may have some value but the paper requires major revisions or
    additional work that are beyond the scope of the conference review cycle to meet
    the quality standard. Without this I am not going to be able to return a score of
    '4 - Accept'.}

\item{\textbf{Supplemental Materials}\\Acceptable with minor revisions (specify revisions in The Review section)}

\item{\textbf{Justification}

    This paper introduces a combined graph clustering and geospatial visualization
    technique to illustrate the city census data that is potentially inconsistent over
    the space dimension. The technique is generally OK, though not entirely new, and
    can be said as appropriate for the problem attacked. The major strength is the
    detailed case studies on two north American cities and it is appreciated that the
    authors attach a full record of their expert study. On the other hand, the paper
    in its current shape suffers significantly from presentation issues, i.e., beyond
    minor grammar errors, most technical content would benefit from a thorough re-
    written. There is no explicit contribution statement. It seems the contribution on
    visualization is rather limited. The authors are also encouraged to re-submit to a
    GIS venue.

\item{\textbf{The Review}

    This paper proposed a framework to visualize the spatiotemporal population
    clusters detected from the census data. A graph representation is built where the
    nodes are fine-grained census tracts (CT) and edges are computed from the
    similarity of census attributes of related CTs. The graph is clustered by an
    existing watershed cut algorithm and then undergo a minor refinement. The derived
    spatiotemporal clusters are displayed on a geospatial map and their dynamics over
    time are visualized in a standard Sankey diagram. The work is well evaluated with
    census data in LA and Toronto. The feedbacks from five experts are also reported.
    While the technical side of this work does have some merit, there are at least two
    major flaws or technical incompleteness in the current presentation.

    First, I missed the technical contribution in this paper. In the first paragraph
    of Sec. 4, it is said ``beyond the main contributions'', what are these
    contributions? In fact, I see rather limited contribution in the visualization
    side. Besides the map design, the only ingredient of information visualization
    seems to be the Sankey diagram. I look at this diagram multiple times and could
    not sense what is different from a basic D3 example available on the website. The
    analytics part is similar, the SMM algorithm is not new, and the graph
    representation and clustering seems straightforward. There are little details
    discussed (in fact, should be some!). For example, why to connect the adjacent or
    overlapping CTs under different time affinities, how the clustering parameter is
    chosen? What are the nature of the resulting clusters? To this end, what is the
    specialized design for inconsistent census data?

    \reply{Following the guidelines, an application paper doesn't need to
    include new techniques. Indeed, the objective of this work was to leverage
    "commonplace" VIS methods to provide a more efficient solution to a common
    issue in another field. Indeed, we provide a viable alternative to time
    consuming and challenging problem that was believed to be unavoidable.}}

    Second, the writing style and paper organization is hardly a qualified
    research paper on visualization. Many key technical content are missing. The
    introduction and related work are fairly short, thus do not motivate this
    work well. For example, in the intro, I would like to see, what is the
    research problem, what are the challenges, and how you solve it with
    exquisite designs and/or algorithms. The authors only write one plain
    sentence there: ``We use a graph based spatio- temporal representation,
    combined with an improved clustering algorithm'', could be used for 100+
    papers on the similar topic. The related work cites an appropriate number of
    literature, but fails to discuss their pros and cons in order to position
    this work. The entire cluster of work by Andrienko et al. are missing while
    it is said: ``it is reasonable to argue that the availability and quality of
    the dataset provided by the Taxi and Limousine Commission of New York
    contributed to its centrality in this field''. The most related work on
    visualizing census data is also missing. In the visualization design part
    (Section 4), the authors spend several paragraphs on the coloring choice,
    which is good. Only one sentence is used to describe the Sankey diagram,
    which might be the only information visualization design applied in this
    work. Note that the graph representation has been the highlight of this
    work. No design rationale is discussed here. For example, what if I use a
    simple alternative of multiple/stacked line charts to represent the dynamics
    of spatiotemporal clusters? In the evaluation part, while I appreciate the
    detailed case study on LA and Toronto, the user feedback seems to be
    problematic given that the users are only given the video, not the
    interactive tool to try themselves! From the preview video, it is hard to
    imagine a real system exists for this work.

    \reply{Unfortunately the reviewer doesn't mention which is the "most related
    work". I did review some works from Andrienko's group, after all they are
    fairly known for this, but some were not included because they are not
    relevant to the current problem. TODO I included more text expliciting
    exactly what was accomplished in this work, which hopefully will make that
    distinction clearer. \\ As the others reviewers mentioned, the tool was
    available not only for the experts, but for the reviewers as well, on an
    anonymous site.}

    There are some other minor issues which the authors can address in any later
    versions.

    \begin{itemize}
        \item{The design consideration of M1 \& M2 are too general, could be
        said for most urban visual analytics methods}
        \item{Last paragraph of Section 4.2, why augment two edges, any quantitative
        evident/support from data?}
        \item{Figure 4 could be put at page 4 for better viewing experience}
        \item{Section 4.6, Sect. to Sec., on top left to on the top left}
        \item{It will be helpful to put a sample census data in the paper, e.g., a table}
        \item{The Sankey diagram in Figure 6 is cluttered, any solution?}
    \end{itemize}

    Overall, the authors have studied a valid problem with an appropriate
    solution and well-conducted case study, but the contribution on
    visualization is vague and many key technical content is missing. The
    writing quality and depth still have much room to improve. For these
    reasons, I return a reject score for now. A last point, if there is little
    new contribution on visualization, would this paper fit better to a GIS
    venue?

    \reply{We considered a GIS venue for this work, but the technical part is
    fairly advanced to a general public, requiring significant tangents to
    explain all the involved concepts. Further publications on applied geography
    and demographics venues are planned in the scope of the Urban Genome
    Project, which will cite this publication for the technical details.}}

\end{itemize}

\subsection{The Summary Review}

\textbf{Summary Rating}

    Possible Accept

    The paper is not acceptable in its current state, but might be made
    acceptable with significant revisions within the conference review cycle.

    If the specified revisions are addressed fully and effectively I may be able
    to return a score of '4 - Accept'.

\section{The Summary Review}
    The reviewers have divided opinions on this paper.

    On the positive side, all reviewers recognize that the authors studied an
    interesting research problem and proposed appropriate/valid solutions to
    that problem (R1, R2, R3, R4). Some finds the proposed techniques to have
    enough contribution (R3), while some others comment these techniques as lack
    of novelty (R1, R4). Most reviewers are convinced on the usefulness of the
    approach as demonstrated in the application of the case studies (R1, R3,
    R4). R3 especially likes the released online demo which helps to grasp the
    main functionality of the technique.

    On the negative side, several concerns are raised and suggestions for
    revision are made, which the authors must address within the conference
    cycle if the paper is conditionally accepted.

    \begin{itemize}
    \item{Evaluation: no evaluation of the proposed clustering methods (R3), the details
    of the domain expert interview are missing. (R1, R3)}

    \item{Design: the consideration and final choice of design are not clearly discussed.
    Specially, the color usage is confusing. (R2, R3, R4)}

    \item{Motivation: the authors fail to clearly demonstrate the motivation of their
    work, probably due to their writing style. (R2, R4)}

    \item{Presentation: no technical contribution clearly stated, writing style not
    qualified as a good research paper, paper organization can be improved (R2, R4)}
    \end{itemize}

    More detailed comments can be found below.

    After discussion, we conclude that a borderline summary rating is appropriate. We
    do hope the authors can address all of our concerns in the conference cycle
    through significant efforts.

\end{document}

### 2018-08-13 paper/method.tex
migrating from gitlab ugp:af90fa20

\section{Visualizing the demographic spatio-temporal evolution}
Beyond the objective of allowing the study of inconsistent data, our method
includes incremental developments in most steps of the analysis, from data
representation to the visualization method for the clusters.
Figure~\ref{fig:overview} presents an overview of the processing steps of the
proposed method.

\begin{figure}
    \centering 
    \includegraphics[width=0.95\linewidth]{overview.pdf}
    \caption{Overview of the proposed method. A graph is generated combining the
        original census data, encoding the changing geographical information.
        The graph is partitioned into an hierarchy~\cite{markus2017}. The
        characteristics and evolution of the clusters are then visually
        represented.
        \label{fig:overview}}
\end{figure}

\subsection{Census methodology and data representation}

Census data is disseminated in a tabulated form for aggregation areas: whole
country, state/province, metropolitan region, and so on. To provide as much
detail as possible, we focus on the smallest region with available data:
\emph{census tracts} (CT). They are usually defined to maintain the anonymity of
the population, leading to a population count in the order of thousands in
densely populated areas. Physical barriers are usually adopted as borders, so
these regions can change because of new roads, construction or removal of high
density buildings, and so on. Some census entities also consider demographic
characteristics, aiming to establish the CTs as a cohesive unit. Therefore, CTs
are the least geographically stable tabulation area.

Each CT is associated with a series of variables, with counts derived from the
census questionnaires, covering several aspects of the demographic
characteristics of the population. Some questions allow for multiple choices or
open answers, that are then tabulated into the most frequent categories. Since
the census is often used to direct government initiatives, which variables are
measured/disseminated is dependent on administrative interests, the general
understanding of the population, and current customs. For instance, income is
disseminated with a finer tabulation in the lower portion than on the higher.

To match these variables over time and allow for direct comparison across
different census years, we aggregated similar ones (e.g. White, Black, Asian,
Other) into \emph{aspects} (e.g. Race), encoding the distribution of that facet
of the population. In this convention, we refer to the composing parts of an
aspect as a \emph{part} or the traditional \emph{variable}. Internally, the
aspects are represented using normalized histograms. This normalized
representation is crucial for the comparison between inconsistent regions.

In our graph based representation, each CT of each census year is represented as
a node, and edges are placed between nodes if the corresponding CTs share
geographic borders in the same year. Further, edges are placed between nodes if
the corresponding CTs belong to sequential years and there is geographical
overlap between them. This approach leads to a single graph representing the
whole spatio-temporal space of the data. Our objective then becomes to identify
partitions of this graph such that the nodes of each partition are more similar
between themselves than to the other nodes. This representation is not the only
option, nor unique, but it allows the use of existing graph-based methods for
the other steps.

\subsection{Geographic content clustering}
To partition the graph we must first establish a distance function between the
nodes, measuring the data similarity. This similarity is then associated with
the edges, leading to a weighted dynamic graph. Every node has a collection of
histograms, each representing the distribution of certain aspect in the
population.

Let $\G=(\V,\E)$ be a graph, where
$\V=\{\vertice_1,\vertice_2,\dots,\vertice_n\}$ is the set of nodes and
$\E=\{(\vertice_i,\vertice_j), i\not=j\text{ and }i,j \in [1,n]\}$ is the set of
edges. A function $\Hist$ associates each node to a set of $K$ histograms. We
define the distance $\D$ between two nodes $\vertice_i$ and $\vertice_j$ as:

\begin{equation}
    \D(\vertice_i,\vertice_j)=\sum_{k \in [1,K]}{\weight_k\, d(\Hist_k(\vertice_i),\Hist_k(\vertice_j))}
    \label{eq:dist}
\end{equation}

\noindent where $d$ is a distance metric between histograms and $\weight$ is a
sequence of non-negative weights associated with each aspect,
$\sum_{k\in[1,K]}{\weight_k}=1$. While any histogram metric can be used, we
adopted a euclidean distance between the vectors, because it led to reasonable
results with reduced computational cost. Therefore the distance between two
nodes is defined as the weighted average distance between its associated
histograms, where the weights can be adjusted by the user.

Once the distances are associated to the edges, we use watershed
cuts~\cite{Cousty2009} to create an initial clustering, which is then refined
into a hierarchy using the Sorted Maximal Matching (SMM)~\cite{markus2017} with
median linkage. The initial watershed step is performed to create an initial
clustering and reduce the running time of the SMM. For completeness, we briefly
review this method, but we refer the reader to the original
paper~\cite{markus2017} for more details, including a complete performance
evaluation using several metrics.

We included two application-specific parameters: the maximum number of clusters
to be shown and a distance threshold. Contrarily to the original SMM, which
merges all clusters in all steps, we only merge two clusters where the distance
is above the threshold after we reach the maximum number of displayed clusters.
Without this restriction, significantly different clusters would be merged
early, leading to increased intra cluster variance and the disappearance of
small outlier regions. Further, after the maximum number of clusters is reached,
we create one step of the hierarchy for each merge, leading to a binary
partition tree. In this structure we can directly access a result with an
arbitrary number of clusters.

Each resulting cluster is contiguous in the graph. This means that two similar,
but non-contiguous, sets of CTs will be classified into two different clusters,
which can be counterintuitive. To overcome this issue, we \emph{augment} the
graph with two new edges per node from a nearest neighbors
graph~\cite{scikit-learn} using only the distances between the histograms. These
edges connect nodes with similar content, if they are not already connected,
providing a path for the algorithm to group similar nodes. Theoretically, adding
more of these content based edges could be used to decrease the impact of the
spatio-temporal edges, controling the balance between content and topology in
the result. In practice, the effect is dependent on the data itself, and the
results are not consistent, or predictable, across different cities. We fixed it
at two edges because it was the lowest number that empircally led to consistent
clusters, but we believe that this idea warrants further investigation, as an
alternative to mixing parameters in the distance metric~\cite{Chavent2017}.

\subsection{Cluster characterization and variable relevance}
\label{sec:relevance}
The composition of each cluster is determined by simple statistical measures,
considering each aspect separately. We compute the minimum, maximum, median,
25\%, and 75\% quantiles for each part of each aspect for all clusters in the
hierarchy. While interpreting these values is more complex than interpreting
just the average, they provide far more information about the underlying
distribution.

We also use these statistical measurements to discover what characterizes each
cluster, that is, what makes it different from the others.  We define the
\emph{relevance} of a part of an aspect based on the distance between the
interquantile ranges (IQR) of the clusters in the same hierarchical level. If
the IQRs overlap for all clusters, that variable is not relevant to the
characterization of the cluster, but if the IQRs are distant, it means that this
specific range of values is something that only occurs in this cluster.

\subsection{Clusters and trajectories}
While the partition of the data into different clusters helps the user to
understand what demographic groups exist and where they are, we are also
interested in the changes in those groups. To this end, we introduce the concept
of \emph{trajectories}, which are composed by regions that are classified into
the same sequence of clusters over the considered period. This enables direct
access to regions that evolved in the same manner. While the interface provides
access to data by census tract, the trajectories are the main unit of
exploration in this work.

\subsection{Colors}
\label{sec:colors}
As illustrated by Figure~\ref{fig:ui} and further explored in the next
subsection, our proposed interface heavily relies on color to express
cluster-related information. We adopted this convention because colors can be
used in all our visual tools in a coherent manner. However, this also introduced
significant challenges. The first is the limit on the number of clusters that
can be visually represented.  We limited the number of clusters to eight because
this was the largest number of colors that we could reliably use, derived from
the 8-class Dark2 set from ColorBrewer~\cite{ColorBrewer}.

While we can reasonably limit the number of clusters, there are far more
possible trajectories. And the color associated with each trajectory should bear
some resemblance to the clusters included in it. Therefore, we were left with a
conundrum: \emph{Should we associate each trajectory with a unique color, which
the user probably cannot distinguish, or should we use a reduced set of colors
and associate the same color to different trajectories?} Since there are
advantages and disadvantages for each of those options, we adopted both. The
user can control which color policy is used via a checkbox in the configuration
panel, on the top left of the interface.

By default, the interface adopts a simplified color scheme, where a trajectory
is painted in the same color of a cluster if the regions were associated to that
cluster for \emph{all} times; in a slightly less saturated version of that color
if the regions were associated to that cluster for \emph{the simple majority} of
the time, and gray otherwise. In this mode, the colors will mostly represent
stability, immediately identifying the regions that were consistently associated
with each cluster. It also easily identifies volatile regions, painted gray.

When this simplified color scheme is disabled, each trajectory will be painted
using the average of the colors of the involved clusters, in the LAB color
space. In this mode, the map becomes more similar to a heatmap, where stronger
presence of a color indicates more temporal affinity to the cluster. Volatile
regions will also tend to be displayed in gray, as the average of three or more
colors.

\begin{figure}
    \centering 
    \includegraphics[width=0.95\linewidth]{colours.pdf}
    \caption{Different color schemes for Chicago with four clusters. Left:
    simplified, right: average color.\label{fig:colour}}
\end{figure}

While both approaches will use more than eight colors, in practice this is not
as significant because most cities can be explained using less than eight
clusters.  In fact, articles in the literature usually employ from two to five,
which are fairly stable across time. For the more dynamic scenarios, user
interaction can be used to alleviate the shortcomings of both approaches.

\subsection{User interface}
The initial interface is illustrated in Figure~\ref{fig:ui}. Since demographic
data can be nuanced, with intricate interconnections, we decided against
presenting the interface using synthetic data, considering instead data from the
Chicago region between 1970 and 2010, using previous published studies as
corroboration. This region is known for its entrenched racial divide and the
emergence of a \emph{'young urban'} population with a higher education
level~\cite{Delmelle2016,Delmelle2017}. More details about this scenario are
presented in Section~\ref{sec:study}.

\begin{figure*}
    \centering 
    \includegraphics[width=0.95\linewidth]{ui.pdf}
    \caption{Initial interface of our method showing the demographic evolution of Chicago. 
        \textbf{(a)}: Configuration panel with the current clustering parameters and controls.
        \textbf{(b)}: Cluster overview illustrating the most relevant aspect for each cluster. 
        \textbf{(c)}: Trajectories overview and the general evolution of the population, geographical information, and how it changed. 
        \textbf{(d)}: Details of the selected trajectories, including precise geographic locations, population numbers, and the composition of the aspects.\label{fig:ui}}
\end{figure*}

The configuration panel, on top left in Figure~\ref{fig:ui}, displays which
aspects were considered and their weights (following Equation~\ref{eq:dist}). It
also includes other configuration options that can be altered without
re-processing the data, such as the number of clusters and the color option. The
gear button allows access to the other configuration options that do require
further processing, such as changing location, aspects, and weights. This panel
also includes the configuration of the selection mode for the trajectories,
which allows the user to set, add, or remove the next selected trajectories to
the current selection. This feature enables the analysis of complex sets of
trajectories.

The cluster overview panel, on the bottom left in Figure~\ref{fig:ui}, displays
a brief summary of each cluster, based on the distance between the IQRs, as
detailed in Section~\ref{sec:relevance}. The \emph{View all} button opens a new
panel where all aspects are represented, while the chevron at the side of the
color lets the user expand each cluster separately. While the standard approach
to represent cluster characteristics is to use parallel
coordinates~\cite{johansson2005revealing,guo2006visualization}, this
representation occupies screen space proportional to the number of variables and
can get cluttered with a higher number of clusters, or when the clusters are not
well defined for multiple variables. To save space and leverage the familiarity
scientists have with statistical tools, we opted to use boxplots to properly
convey the distribution of each variable in the current cluster. However, a
simple boxplot would not include information about the other clusters, forcing
the user to mentally compare them to find what is relevant.

We adopt an \emph{enhanced} version of the traditional boxplot, which includes
the minimum, maximum, 25\% and 75\% quantiles for the current cluster, but also
the IQRs for the other clusters, in slightly larger and faded black rectangles.
We also color the current IQR according to its relevance. While there might be
some degree of similarity between the color schema for relevance and for
trajectory identification, none of the experts consulted reported confusion.
These simple changes allow the user to easily understand the composition of the
cluster and how it relates to the others. Violin plots~\cite{hintze1998violin}
could also be used, but with increased potential for cluttering the
representation of the other clusters, hindering the comparisons.

For instance, the boxplot that summarizes the purple cluster, detailed on
Figure~\ref{fig:boxplot}, illustrates that this cluster is best defined by the
proportion of the population with four or more years of college. The user can
quickly see that this is relevant because the corresponding IQR is colored with
the highest relevance present in the legend. It is also clear that, while this
cluster includes CTs that have between 10\% to 90\% of people in this variable,
approximately, half of them have about 60\% of the population with four or more
years of college. Since all the other IQRs are well separated, this is a
defining characteristic of this cluster. By clicking on the colored bar above
the boxplot, the user can select all trajectories that contain this cluster at
any point in time.

\begin{figure}
    \centering 
    \includegraphics[width=0.9\linewidth]{boxplot.pdf}
    \caption{Enhanced boxplot of the clusters' characteristics allows a quick
    comparison to the other clusters.\label{fig:boxplot}}
\end{figure}

The other clusters identified on Figure~\ref{fig:ui} correspond to higher
concentration of people that identify as Black in the green cluster, people that
identify as "Asian, Hawaiian, other pacific islander" in the pink cluster, and
people that identify as White in the orange cluster. From these plots, it is
clear that the city is indeed racially divided~\cite{Delmelle2016}, with several
CTs that are almost exclusively occupied by people of the same racial category.

The trajectories overview aims to convey basic information about the
trajectories present in the data, where they are, and what changes are involved.
This is done using three sub panels. The first, on the left, contains a Sankey
diagram illustrating the evolution of the clusters over time. The widths are
proportional to the population involved, the colors follow a policy detailed in
Section~\ref{sec:colors}. A stacked graph could also be used to represent the
proportions of each cluster~\cite{Valdivia2015} with less clutter, since the
transitions between clusters would not be represented. However, this is only
viable if more temporal steps are available, making the plot smoother. Another
option to remove clutter is to remove portions of this plot when trajectories
are selected, but this would change the layout and compromise the user's mental
map.

In our example in Figure~\ref{fig:ui}, we can see that the total population of
Chicago is decreasing. Additionally, the orange and green clusters contain most
of the population and are fairly stable over time. The pink cluster is small and
mostly stable. The purple cluster is increasing, mostly by incorporating areas
that were previously orange. Since the purple group corresponds to the emergent
'young urban' group, this corroborates the findings of
Delmelle~\cite{Delmelle2016,Delmelle2017}. This diagram can also be used to
select specific trajectories, by clicking on the bands, or all trajectories that
contain a specific cluster at a specific time, by clicking on the rectangles.

The next sub panel, in the top middle of Figure~\ref{fig:ui}, is a transition
matrix between the clusters. It indicates the percentage of the population whose
area changed in each pattern. This kind of table can be found in the related
literature~\cite{Delmelle2016}, so it is familiar to the advanced users, and it
not only informs the proportional changes, but allows the selection of the
corresponding trajectories for further analysis.

Contrary to the trajectories plot, this representation is more Markovian, where
only the current and next state are considered. This panel also enables easier
access to trajectories with specific changes, by clicking on the corresponding
percentage values. The combo box allows the user to refine the transitions, from
'Any', which includes all transitions between years, to specific transitions, to
changes from the first year to the last year. In this example, approximately
99\% of the population in areas classified as green were also in areas
classified as green in the next year, while 1\% changed to purple at some point
and another 1\% to orange. The total is over 100\% due to rounding errors.
Regions changed from the orange to the green cluster for 6\% of its population,
1\% to pink, and 9\% to purple. This further corroborates the fact that most of
the growth of the purple cluster came from the orange cluster. Additionally, the
lack of transitions is also relevant, for instance, no CT changed from majority
of Black population (green) to Asian population (pink), and no CT with
significant Asian population had significant increase in education levels
(purple).

The panel in the top right of Figure~\ref{fig:ui} is a map of the region under
analysis, summarizing the geographical evolution of the clusters over time.  The
colors are derived from the clusters involved in each trajectory as detailed in
Section~\ref{sec:colors}, which are consistent across the linked views.

The bottom part of the interface contains the details for the selected
trajectories, or for the whole city if nothing is selected, as illustrated in
Figure~\ref{fig:ui}. This panel contains two main regions: the small multiple
maps, depicting the clusters at each year, and the stacked bar plots that
summarize the overall composition of these regions. Some finer localization
information is lost using small multiples, such as small border changes, but
that information is available at the larger map. All the maps are linked with
syncronized navigation, and the use of small multiples allows the exploration of
each temporal census individually, and its comparison to the others, with
minimal interaction.

In this example, the maps show the transition from orange to green and purple in
several regions over time. Clicking on a region in these maps will bring up a
new panel with the original census data of this specific region. The actual
population numbers are below the maps, and they confirm the notion provided by
the Sankey diagram that the total population is indeed decreasing.

Each aspect is represented by a stacked bar plot, where the width of each
rectangle corresponds to the average percentage of that variable over the
considered period. We chose stacked bar plots to represent the composition of
the regions because they can accurately and succinctly inform the proportions of
each aspect, without any interaction. In this case, about half of the people in
Chicago in the considered period are married, and the percentage that are
Widowers or Divorced is roughly similar. About half of the population work in
Administrative jobs, a third never completed high-school, approximately half
have gross family income below 50,000USD per year. The vast majority identify as
white. Placing the mouse over one of the bars will open a small panel with the
temporal evolution of that specific variable, and clicking on the chevron on the
right side expands the corresponding aspect, showing details of the temporal
evolution of each variable and also the corresponding IQRs for the whole city.

### 2018-08-13 paper/related.tex
migrating from gitlab ugp:af90fa20

\section{Related Work}

Since our problem emcompasses several fields, we divided this section into
specific subproblems: \emph{longitudinal demographic studies}, describing the
traditional approach to perform longitudinal studies; \emph{data
representation}, exploring how evolving geographic data can be represented for
processing; \emph{Data clustering}, briefly reviewing existing clustering
methods; and \emph{cluster characterization}, exploring how the clusters can be
visually summarized.

\subsection{Longitudinal demographic studies}
Census data is used not only to discover demographic
patterns~\cite{Firebaugh2016}, but to correlate demographic charateristics to
other measurements~\cite{diez1997neighborhood}. However, longitudinal studies
are rare: \emph{"(...) One of the most challenging and fascinating areas in
spatial statistics is the synthesis of spatial data collected at different
spatial scales(...)"}~\cite{gotway2002combining}.

While CT level data is readily available for the US since 1910~\cite{nhgis},
most studies consider the period between 1970 and 2010, using pre-harmonized
data~\cite{Logan2014,nhgis}. Despite the inherent
errors~\cite{Logan2016,Hallisey2017}, this dataset became the standard source
for longitudinal demographic data, with similar efforts appearing in other
countries~\cite{Liu2015,Lee2015,Allen2018}. This result was significant for the
field, but it also restricts the useable data, since new datasets need to be
similarly processed.

Another option considers the use of grid data~\cite{Dmowska2017,Dmowska2018},
where small rectangular areas are used, in an approach similar to satellite
imagery. Beyond the increased spatial accuracy, this approach does not require
complex harmonizations when new data is considered. However, demographic data is
usually not available in this format, especially from older sources, and the
conversion from tabulation areas can introduce significant errors.

In the proposed methodology, we avoid the harmonization by considering each
measurement using its actual geographic region. It does not require the regions
to be consistent across time because they are already represented as different
entities.

\subsection{Data representation}
Most data is represented in tabular form, where the rows and columns have
coherent definitions. For example, consider a table with rain measurements over
time, with the rows representing different locations and the columns different
times. This representation can also be interpreted as a collection of
time-series, one for each location. Geographic data followed this format, only
including an additional field that describes the associated geographic area.
Following the example, the data would now represent the amount of rain for a
given region and time. As long each region remains the same, the data is
coherent and can be interpreted again as a collection of time-series.

In the proposed method, we remove the requirement for consistency in the
measurement regions by leveraging a graph-based representation, where each
region in time corresponds to a different node. Instead of a collection of
time-series, the data is represented as a dynamic graph. Graph based
representation of geographic information is fairly well explored in the
literature, as a basis for topological methods for event
detection~\cite{Doraiswamy2014}, leveraging signal processing on
graphs~\cite{shuman2013emerging,sandryhaila2013discrete} to find patterns and
outliers~\cite{Valdivia2015,Dias2015,Alce2018}. Graphs are well suited to
represent trajectories as
well~\cite{VonLandesberger2016,Huang2016,chen2015survey}, allowing the use of
graph visualization methods~\cite{Vehlow2015,Beck2014}.

Graphs were used to represent census data for clustering purposes
before~\cite{Dias2015,Setiadi2017}, but these works did not explore temporal
evolution, where graphs are particular powerful as they allow a natural
representation of inconsistent regions, with both spatial and temporal
connections. Note that there are other possible representations that have
similar properties, but we adopted graphs to allow the use of the existing
literature and methods.

\subsection{Data clustering}
Data clustering is one of the elementary processes for data analysis,
simplifying the data into a smaller number of homogeneous sets that can be
interpreted in the same way. While there is no shortage of contributions for
this problem~\cite{Fahad2014}, most applications still rely on
k-means~\cite{jain2010data,Delmelle2016} and, to a lesser extent, Self
Organizing Maps~\cite{Delmelle2017,Ling2016}.

However, a method for geographic data analysis should not ignore the geographic
component of the data. One straightforward option, for agglomerative
methods~\cite{han2001spatial}, is to consider only nearby clusters for
merging~\cite{Chavent2017}, which can also be done for
k-means~\cite{soor2018extending}. Alternatively, the spatial distance could be
directly added to the inter-cluster metric~\cite{Chavent2017} via a mixing
parameter, which adds flexibility to the method, but introduces the problem of
finding the correct application-dependent values.

Indeed, one crucial step in most clustering algorithms is the definition of the
number of clusters. We sidestep this problem by considering hierarchical
methods~\cite{soille2012morphological}, where the result is not a partition of
the data, but a tree of partitions. This approach is interesting for interactive
methods, because it allows the user to change the number of displayed clusters
with minimal processing. Since our data is represented as a graph, one option
would be the watershed cuts algorithm~\cite{Cousty2009}, inspired by the well
known image processing segmentation and equally prone to oversegmentation.
Considering that the processing time is also a relevant factor, we opted for an
heuristic variation of the maximum weighted matching algorithm called
\emph{sorted maximal matching}~\cite{markus2017}, which merges clusters based on
the weights of the edges between pairs of clusters.

\subsection{Cluster characterization}
Visually representing evolving spatial data is a challenging old
problem~\cite{monmonier1990strategies,andrienko2003exploratory,ferreira2015visual,Zheng2016}.
Most geographic data is naturally bidimensional and maps work well in this
case~\cite{Zheng2016,ward2015interactive}, but the temporal dimension cannot be
so naturally represented. One straightforward option is to leverage
tridimensional plots~\cite{andrienko2014visualization,Tominski2012a}, but this
can lead to visual obstructions or scaling problems unless a tridimensional
display device is used. Animation can also be explored in some specific
cases~\cite{buschmann2014real}, but it is not a general approach. Glyphs can
also be used~\cite{seebacher2017visual,Andrienko2017}, but this may lead to
cluttering when many small regions are present. A simpler, well adopted, option
is to display a map that corresponds to a subset of the temporal information,
allowing the user to change the time with an associated
control~\cite{Chen2017,Valdivia2015,Alce2018,Doraiswamy2014}. Small multiples
can be used~\cite{VonLandesberger2016}, but only when there are few temporal
snapshots. However, none of these options is suitable to represent many
variables at the same time.

Using data clustering, we can represent the region's cluster instead of all the
its variables~\cite{Alce2018,Valdivia2015,VonLandesberger2016}. While this
simplifies the geographic portion of the visualization, it introduces the
problem of how to summarize the contents of each cluster. One traditional
approach is to use parallel coordinates
plot~\cite{ferreira2015urbane,LI2018,johansson2005revealing,guo2006visualization},
but these they can get cluttered representing similar clusters over several
variables. Further, for demographic applications, the clusters are usually
strongly characterized by a small subset of
values~\cite{Delmelle2016,Delmelle2017}. Therefore, in the proposed method, we
identify the variables that are most relevant to the characterization of each
cluster. The distribution of values on that variable is then represented using a
boxplot, a well known statistical plot displaying basic properties of the
distributions.

### 2018-08-13 paper/study.tex
migrating from gitlab ugp:af90fa20

\section{Illustrative scenarios}
\label{sec:study}

We used decennial census data from the United States~\cite{nhgis} and
Canada\footnote{\url{http://datacentre.chass.utoronto.ca/census/}}, tabulated by
CTs,  from 1970 to 2010. The prototype allows access to 40 regions, 28 in the US
and 12 in Canada. Due to the high number of CTs, New York City was split into
its boroughs.

We used five aspects for the USA: Education level, Family income, Marital
status, Occupation, and Race; and seven for Canada: Age, Education level, Home
language, Household Income, Marital status, Occupation, Place of birth, and
Religion.  While our method does not require geographic harmonization, it
requires matching variables over time. The supplementary material contains the
details of which census columns were used for each aspect. Income is slightly
inaccurate, even though we did correct for inflation. We grouped the original ranges into
three larger ranges, but they do not match precisely.

% \subsection{Chicago}
% \begin{figure*}
%     \centering
%  \includegraphics[width=0.85\linewidth]{newTeaser.pdf}
%  \caption{Boxplot representation for some of the clusters found, characterized
%  by White population (orange), high education (purple), and Black population
%  (green). The trajectories plot represents the associated population, showing an
%  increase in the purple cluster. The plot on the right compares the intra
%  quantile ranges for \emph{Four or more years of College}. While there was an
%  increase in the whole city (gray band), it was sharper in this
%  region.\label{fig:teaser}}
% \end{figure*}

\subsection{Los Angeles}
We selected a region around the metropolitan area of Los Angeles (LA). This
region follows urban density, not administrative boundaries. The summary of the
results using all aspects with equal weights and eight clusters is illustrated
in Figure~\ref{fig:la}, while the statistical description of the clusters is
illustrated in Figure~\ref{fig:laAll}, where the most relevant aspect of each
cluster is highlighted. From the trajectories plot, we can see that there is a
large but shrinking cluster, depicted in green, one increasing cluster in
purple, an almost constant orange cluster, a smaller but increasing pink
cluster, and three other small clusters. The corresponding map illustrates where
these clusters are located, and that they are somewhat geographically stable,
with some movement between the green, orange, and purple clusters.

\begin{figure}
    \centering 
    \includegraphics[width=0.95\linewidth]{LA.pdf}
    \caption{Result for Los Angeles with 8 clusters, including three small and
        ephemeral clusters. Cluster characterization is displayed
        in Figure~\ref{fig:laAll}.\label{fig:la}}
\end{figure}

From Figure~\ref{fig:laAll}, we can see that the green cluster is characterized by
a high percentage of White population, low percentage of population in the
lowest income bracket, mostly administrative occupations, and about 30\% of the
population with four or more years of college. The orange cluster is
characterized by a high percentage of Black population, with few people in the
highest range of income and education. The purple cluster corresponds to a high
concentration of "Other" in race, which includes Hispanic for this dataset, high
concentration of Laborers, and low education and income. The pink cluster
contains a high percentage of Asian population and about 30\% of the population
with four or more years of college. The light green cluster contains very few
people in the lower income bracket, mostly White population, with the highest
percentage of population with four or more years of college, working
administrative jobs, and a high concentration of singles. The yellow cluster
represents Black population, with higher level of education and income, mostly
working administrative jobs. The brown cluster represent a majority of single
population, working administrative jobs with mostly low income. The dark gray
cluster is characterized by all its population in the lowest income bracket, low
education level, with a majority of White population.

\begin{figure*}
    \centering 
    \includegraphics[width=0.95\linewidth]{laAll.pdf}
    \caption{Full characterization of the eight clusters found for LA. The red
     rectangles indicate the most relevant aspects for each
     cluster.\label{fig:laAll}}
\end{figure*}

Further, the larger clusters, green, orange, and purple, present a significant
intra-cluster variance in most variables, as indicated by the maximum and
minimum whiskers of the boxplots. For instance, while fifty percent of the CTs
in the green cluster have between 20\% and 40\% of people in the lowest income
bracket, that cluster also includes CTs where none and all the population
belongs to that bracket. This might indicate that this cluster could be further
divided, and currently represents different groups of people that are not
different enough to be separated into clusters in this level of the hierarchy.

While the larger clusters are expected, the light green, brown, and dark gray
clusters are more surprising. These clusters are small and ephemeral, including
only a few CTs, and are peculiarly different enough to be separated into their
own clusters, at this level of the hierarchy. While these small clusters might
be interesting enough to warrant further study, they might suppress larger, but
more subtle, clusters from appearing in this interface,  due to the limited
number of clusters.

The orange area in the map in Figure~\ref{fig:la} presents movement, indicated by
the presence of green and purple tones mixed with the orange, which may warrant
further exploration. By clicking on the orange bar above the boxplot, we select
all trajectories that contain the orange cluster. The corresponding details are
illustrated in Figure~\ref{fig:labh}. This shows a location change, where the
orange cluster is progressively replaced by the purple cluster on its east side,
and in turn expanding to the west. Interestingly, the population increased, the
racial profile changed, but the distribution of income was reasonably stable,
with a higher amount of the population in the lowest income range and very few
people in the highest income range. Indeed, the income  difference is
significant when compared to the city-wide distribution.

\begin{figure}
    \centering 
    \includegraphics[width=0.95\linewidth]{laBH.pdf}
    \caption{Top: Geographic changes in the majority Black population cluster
    (orange) and Laborers cluster (green). Bottom: Income evolution for this
    region (black) and the whole city (gray).\label{fig:labh}}
\end{figure}

A portion of this region is classified into the green cluster in 2010,
indicating a majority white population. To further understand that change, we
clear the current selection, and select all regions that changed from orange in
1970 to green in 2010, using the transition matrix. A portion of the resulting
region, near the Florence-Graham region, is depicted in Figure~\ref{fig:laVol},
along with the temporal evolution of Race. Despite this difference, the other
aspects are similar to the ones from the region in Figure~\ref{fig:labh}, with
slightly lower income and education profiles. While the racial aspect changed
considerably, the economic and educational aspects stayed the same.

\begin{figure}
    \centering 
    \includegraphics[width=0.95\linewidth]{laVol.pdf}
    \caption{Details for a volatile region contained in the area of
         Figure~\ref{fig:labh}. This region went from Black to Hispanic to
        White.\label{fig:laVol}}
\end{figure}

\subsection{Toronto}

We considered a region that is approximately the administrative border of the
current city of Toronto and all seven available aspects. The results are
summarized in Figure~\ref{fig:to}, considering eight clusters.

\begin{figure}
    \centering 
    \includegraphics[width=0.95\linewidth]{to.pdf}
    \caption{Results for Toronto, with eight clusters.\label{fig:to}}
\end{figure}

The population with low percentage of University degrees is represented in
orange, mostly anglophone population in green, Asian immigrants in yellow, high
percentage of income in the highest bracket in purple, high percentage of Jewish
people in light green and brown, high percentage of Eastern Non-Christian
religion in pink, and high concentration of single people in dark gray.

From the trajectories plot, we can see that Toronto is more dynamic than both
Chicago and LA, with one cluster constantly shrinking, and others growing more
prominent. In the 1970s, the city was divided into four clusters: low number of
university degrees, Jewish population, majority anglophones, and high income,
with the latter being rather small. There is some variance in the other
characteristics of the orange cluster, but none as significant as the education
level, especially when compared to the other years. While the orange cluster
persists until 2011, the other clusters present have similar education profiles
to one another, leading the clustering method to use other aspects to
characterize them.

Toronto's dynamism also appears in that we see the influx of Asian population
starting in the 1980s and building thereafter, as represented by the yellow and
pink clusters. While both include a high percentage of people born in Asia, the
pink region is more defined by religion, with low percentage of university
degrees, and contains the lowest percentage  of people in the highest income
bracket for these clusters; the yellow is less defined by religion, and has
higher education and income, geographically corresponding to the Markham region,
know for its Chinese population. A similar division also happens for the two
Jewish clusters, where the light green cluster has lower education and income
levels than the brown cluster. The purple cluster of high income is somewhat
stable. This cluster includes until 2011 the Bridle Path neighborhood, known for
its wealthy population.  In 2011, however, it was classified into the yellow
cluster that includes Asian immigration, since about 35\% of the population for
this CT were born in Asia.  It maintained its income profile, with 85\% with 90k
CAD or more.

The grey regions of the map are another indication of Toronto's volatility over
this period. While there is a cluster that is painted dark grey, annotated by a
red circle in Figure~\ref{fig:to}, the light gray regions on the north part of the
city indicate volatility. Here the same region was classified into three or more
clusters for this five census period. Using the 'Add' mode for the trajectory
selection, we select these volatile regions. A subset of the details is
illustrated in Figure~\ref{fig:toVol}.

\begin{figure}
    \centering 
    \includegraphics[width=0.95\linewidth]{toVol.pdf}
    \caption{Details for some regions of Toronto that were classified into 3 or
         more clusters over time.\label{fig:toVol}}
\end{figure}

These regions account for about 5\% of Toronto's population. The whole region
was classified into the orange cluster in 1971 (low level of university degrees).
By 1991, most of the region was classified into the green cluster, representing
anglophone population, mostly Canadian born, with a higher level of education.
As the corresponding plot indicates, this trend in increasing education is
city-wide, but this region has people with better education than most.

In 2001, the purple cluster of high income annexes neighboring parts of the
volatile region, and the Asian born population increases sharply, as illustrated
by the appearance of the yellow cluster.  This cluster, once again, indicates
well educated, higher income, and about 30\%-50\% Asian born population. By
2011, the yellow cluster increased considerably, annexing parts of the high
income purple cluster, including the neighboring Bridle Path area.

While Toronto is more dynamic than Los Angeles, possibly due to size
differences, the volatile regions shown in Figure~\ref{fig:toVol} did not change as
quickly or dramatically as the ones shown in Figure~\ref{fig:laVol}, which involved
twice as many people. We found this trend to be related to the countries
themselves, Canadian cities have larger areas undergoing slow, gradual changes,
whereas American cities have more general stability, but quicker changes in
smaller scales. The supplementary material contains brief summaries of all the
regions accessible in this prototype.

\section{Expert feedback}
%This could be a big step toward better understanding of evolving urban morphology.
To assess the proposed method, we contacted academic and industry experts in
sociology and urban sciences. The supplementary material contains all our
communication with five experts, identified from A to E.

Their response was positive,  mentioning that the prototype allows them to
analyze census data without the additional work of obtaining and cleaning the
data (A, B, E), and it allows the inclusion of geographic visual analysis tools
in their research process (D). It enables the users to tell different stories
about neighborhoods/cities and their changes (A), visualize the relationship
between key urban variables over time (D), offering a quick way to identify
particular neighborhoods that one may be interested in studying more in depth
around a particular issue or efficiently understanding the context of an area
(E).  Indeed, the experts identified gentrification processes in Manhattan (B)
and Dallas (E), reinforced a hypothesis for occupational clustering (D), and
highlighted how the method can be used to compare neighborhoods and cities (A).
These remarks attest that we satisfied the design requirements regarding the
existing demographic groups and their changes.

While the interface was "easy to navigate" (B), it was also considered
"overwhelming" (A), "intimidating" (E), and "tricky to interpret" (C), possible
side-effects of our effort to increase  representational accuracy, where we
avoided using simplified labels. Specifically, while identifying clusters by
their most relevant variables was welcome, the overlap of information from
different clusters in the boxplot was "a bit confusing" (C). While the map of
trajectories was mentioned as a "good summary map", how it related to the
clustering method was unclear (C). While we provide different options on how the
colors are used on this map, both are sub optimal, reliably representing several
distinct entities using colors is still an open problem in visualization.

The experts also mentioned the poor responsiveness of the method when changes in
the clustering parameters required server-side processing (B,D). Indeed, the
current implementation can take a few minutes to cluster regions with high
number of CTs, like Los Angeles or Brooklyn.

Most of the experts demonstrated interest in using our method in their research
(A, B, D, E), aiming to use the census data as a backdrop for other datasets,
providing demographic context. They also mentioned the need to export subsets of
data, plots, and maps to be used in reports and publications (C, D, E).

More importantly, while these experts were aware that our method does not
perform geographic normalization, none of them mention it. We did not
specifically ask if this difference led to unexpected results, but rather
inquired if they found interesting insights, leading them to consider the
interface, its options, and results as a whole. Therefore the method was
neutrally compared against other, traditionally normalized results. We interpret
the fact that most of them were interested in the next steps as an affirmation
of the accuracy of the method.

### 2018-08-13 paper/vis.tex
migrating from gitlab ugp:af90fa20

% $Id: template.tex 11 2007-04-03 22:25:53Z jpeltier $

%\documentclass{vgtc}                          % final (conference style)
\documentclass[review]{vgtc}                 % review
%\documentclass[widereview]{vgtc}             % wide-spaced review
%\documentclass[preprint]{vgtc}               % preprint
%\documentclass[electronic]{vgtc}             % electronic version

%% Uncomment one of the lines above depending on where your paper is
%% in the conference process. ``review'' and ``widereview'' are for review
%% submission, ``preprint'' is for pre-publication, and the final version
%% doesn't use a specific qualifier. Further, ``electronic'' includes
%% hyperreferences for more convenient online viewing.

%% Please use one of the ``review'' options in combination with the
%% assigned online id (see below) ONLY if your paper uses a double blind
%% review process. Some conferences, like IEEE Vis and InfoVis, have NOT
%% in the past.

%% Figures should be in CMYK or Grey scale format, otherwise, colour 
%% shifting may occur during the printing process.

%% These few lines make a distinction between latex and pdflatex calls and they
%% bring in essential packages for graphics and font handling.
%% Note that due to the \DeclareGraphicsExtensions{} call it is no longer necessary
%% to provide the the path and extension of a graphics file:
%% \includegraphics{diamondrule} is completely sufficient.
%%
\ifpdf%                                % if we use pdflatex
  \pdfoutput=1\relax                   % create PDFs from pdfLaTeX
  \pdfcompresslevel=9                  % PDF Compression
  \pdfoptionpdfminorversion=7          % create PDF 1.7
  \ExecuteOptions{pdftex}
  \usepackage{graphicx}                % allow us to embed graphics files
  \DeclareGraphicsExtensions{.pdf,.png,.jpg,.jpeg} % for pdflatex we expect .pdf, .png, or .jpg files
\else%                                 % else we use pure latex
  \ExecuteOptions{dvips}
  \usepackage{graphicx}                % allow us to embed graphics files
  \DeclareGraphicsExtensions{.eps}     % for pure latex we expect eps files
\fi%

%% it is recomended to use ``\autoref{sec:bla}'' instead of ``Fig.~\ref{sec:bla}''
\graphicspath{{figures/}{pictures/}{images/}{./}} % where to search for the images

\newcommand{\squishlist}{
 \begin{list}{$\bullet$}
  { \setlength{\itemsep}{0pt}
     \setlength{\parsep}{3pt}
     \setlength{\topsep}{3pt}
     \setlength{\partopsep}{0pt}
     \setlength{\leftmargin}{1.5em}
     \setlength{\labelwidth}{1em}
     \setlength{\labelsep}{0.5em} } }

\newcommand{\squishlisttwo}{
 \begin{list}{$\bullet$}
  { \setlength{\itemsep}{0pt}
    \setlength{\parsep}{0pt}
    \setlength{	opsep}{0pt}
    \setlength{\partopsep}{0pt}
    \setlength{\leftmargin}{2em}
    \setlength{\labelwidth}{1.5em}
    \setlength{\labelsep}{0.5em} } }

\newcommand{\squishend}{
  \end{list}  }

\usepackage{algpseudocode} 
\usepackage{algorithm}
\usepackage{amsmath}
\usepackage{microtype}                 % use micro-typography (slightly more compact, better to read)
\PassOptionsToPackage{warn}{textcomp}  % to address font issues with \textrightarrow
\usepackage{textcomp}                  % use better special symbols
\usepackage{mathptmx}                  % use matching math font
\usepackage{times}                     % we use Times as the main font
\renewcommand*\ttdefault{txtt}         % a nicer typewriter font
\usepackage{cite}                      % needed to automatically sort the references
%% We encourage the use of mathptmx for consistent usage of times font
%% throughout the proceedings. However, if you encounter conflicts
%% with other math-related packages, you may want to disable it.
\usepackage{color}
\newcommand{\G}{\ensuremath{G}}
\newcommand{\V}{\ensuremath{V}}
\newcommand{\vertice}{\ensuremath{v}}
\newcommand{\weight}{\ensuremath{w}}
\newcommand{\E}{\ensuremath{E}}
\newcommand{\Hist}{\ensuremath{H}}
\newcommand{\D}{\ensuremath{D}}
\newcommand{\relevance}{\ensuremath{R}}
\newcommand{\cluster}{\ensuremath{\mathcal{C}}}
\newcommand{\watershed}{\ensuremath{\psi}}

\newcommand{\lookDan}[1]{\marginpar{---}{\color{blue}{#1}}}

\DeclareMathOperator*{\median}{\ensuremath{\text{median}}}

%% If you are submitting a paper to a conference for review with a double
%% blind reviewing process, please replace the value ``0'' below with your
%% OnlineID. Otherwise, you may safely leave it at ``0''.
\onlineid{1101}

%% declare the category of your paper, only shown in review mode
\vgtccategory{Research}

%% allow for this line if you want the electronic option to work properly
\vgtcinsertpkg

%% In preprint mode you may define your own headline.
%\preprinttext{To appear in an IEEE VGTC sponsored conference.}

%% Paper title.

\title{Visualizing demographic evolution using geographically inconsistent census data}

%% This is how authors are specified in the conference style

%% Author and Affiliation (single author).
%%\author{Roy G. Biv\thanks{e-mail: roy.g.biv@aol.com}}
%%\affiliation{\scriptsize Allied Widgets Research}

%% Author and Affiliation (multiple authors with single affiliations).
%%\author{Roy G. Biv\thanks{e-mail: roy.g.biv@aol.com} %
%%\and Ed Grimley\thanks{e-mail:ed.grimley@aol.com} %
%%\and Martha Stewart\thanks{e-mail:martha.stewart@marthastewart.com}}
%%\affiliation{\scriptsize Martha Stewart Enterprises \\ Microsoft Research}

\author{Fabio Dias\thanks{e-mail: fabio.dias@utoronto.ca} %
\and Daniel Silver\thanks{e-mail: dsilver@utsc.utoronto.ca}}
\affiliation{\scriptsize University of Toronto}

%% A teaser figure can be included as follows, but is not recommended since
%% the space is now taken up by a full width abstract.
\teaser{
  \centering
 \includegraphics[width=0.85\linewidth]{newTeaser.pdf}
 \caption{Higher increase in the education level (purple cluster) in Chicago
  between 1970 and 2010. While the whole city follows this trend, the change was
  far more pronounced in these regions. The relevant clusters defined by Black
  (green) and White (orange) majority of population are also visible. 
  \label{fig:teaser}}}

%% Abstract section.
\abstract{Census measurements provide reliable demographic data going back
centuries. However, their analysis is often hampered by the lack of geographical
consistency across time. We propose a visual analytics system that enables the
exploration of geographically inconsistent data. Our method also includes
incremental developments in the representation, clustering, and visual
exploration of census data, allowing an easier  understanding of the demographic
groups present in a city and their evolution over time. We present the feedback
of experts in urban sciences and sociology, along with illustrative scenarios in
the USA and Canada.}

  % end of abstract

%% ACM Computing Classification System (CCS). 
%% See <http://www.acm.org/about/class> for details.
%% We recommend the 2012 system <http://www.acm.org/about/class/class/2012>
%% For the 2012 system use the ``\CCScatTwelve'' which command takes four arguments.
%% The 1998 system <http://www.acm.org/about/class/class/2012> is still possible
%% For the 1998 system use the ``\CCScat'' which command takes four arguments.
%% In both cases the last two arguments (1998) or last three (2012) can be empty.

\CCScatlist{ %TODO
  \CCScatTwelve{Human-centered computing}{Visu\-al\-iza\-tion}{Visualization application domains}{Visual analytics};
  \CCScatTwelve{Mathematics of computing}{Probability and statistics}{Statistical paradigms}{Exploratory data analysis
  }
}

%\CCScatlist{
  %\CCScat{H.5.2}{User Interfaces}{User Interfaces}{Graphical user interfaces (GUI)}{};
  %\CCScat{H.5.m}{Information Interfaces and Presentation}{Miscellaneous}{}{}
%}

%% Copyright space is enabled by default as required by guidelines.
%% It is disabled by the 'review' option or via the following command:
% \nocopyrightspace

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%% START OF THE PAPER %%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\usepackage{lineno}
\linenumbers
\begin{document}
%% The ``\maketitle'' command must be the first command after the
%% ``\begin{document}'' command. It prepares and prints the title block.

%% the only exception to this rule is the \firstsection command
\firstsection{Introduction}
\maketitle
\input{intro}
%% \section{Introduction} %for journal use above \firstsection{..} instead

\input{related}

\input{design}

\input{method}

\input{study}

\input{discussion}

\section{Conclusions}
Our objective was to allow for the exploration of census data without
geographical normalization, an approach that is usually not considered:
\emph{"(...) tract-by-tract comparison is not possible unless data from 2000 is
interpolated to 2010 boundaries (...)"}~\cite{Dmowska2017}. Our method was able
to corroborate previous findings from the specialized literature, with an
increased level of detail due to our data representation and visualization
choices. The feedback from experts was remarkably positive, most of them were
able to extract insight from the prototype and demonstrated interest in using it
on their research efforts.

While we focused on census data in this work, the proposed method can be used
with most similarly area-based datasets. Further, while there is interest in the
use of census data, it is mostly in combination with other, larger and more
dynamic, datasets. We postulate that, while census data is interesting, the
amount of effort required to obtain and prepare it often discourages its use as
complementary information, reinforcing our hypothesis that there is need for
methods that are able to deal with geographically inconsistent data.

%% if specified like this the section will be ommitted in review mode
\acknowledgments{This research was supported by a University of Toronto
Connaught Global Challenge grant and is part of the Urban Genome Project.}
\bibliographystyle{abbrv-doi-hyperref-narrow}

\bibliography{rev}
\end{document}

### 2018-08-13 sup/sup.tex
migrating from gitlab ugp:af90fa20

\documentclass[a4paper]{article}  
\usepackage{url}
\usepackage{hyperref}
\title{Supplementary material}
\author{}
\date{}
\ifpdf%                                % if we use pdflatex
  \pdfoutput=1\relax                   % create PDFs from pdfLaTeX
  \pdfcompresslevel=9                  % PDF Compression
  \pdfoptionpdfminorversion=7          % create PDF 1.7
  \ExecuteOptions{pdftex}
  \usepackage{graphicx}                % allow us to embed graphics files
  \DeclareGraphicsExtensions{.pdf,.png,.jpg,.jpeg} % for pdflatex we expect .pdf, .png, or .jpg files
\else%                                 % else we use pure latex
  \ExecuteOptions{dvips}
  \usepackage{graphicx}                % allow us to embed graphics files
  \DeclareGraphicsExtensions{.eps}     % for pure latex we expect eps files
\fi%
\graphicspath{{figures/}{pictures/}{images/}{./}} % where to search for the images

\newcommand{\question}[1]{\smallskip\noindent\emph{#1}}
\newcommand{\squishlist}{
 \begin{list}{$\bullet$}
  { \setlength{\itemsep}{0pt}
     \setlength{\parsep}{3pt}
     \setlength{\topsep}{3pt}
     \setlength{\partopsep}{0pt}
     \setlength{\leftmargin}{1.5em}
     \setlength{\labelwidth}{1em}
     \setlength{\labelsep}{0.5em} } }

\newcommand{\squishlisttwo}{
 \begin{list}{$\bullet$}
  { \setlength{\itemsep}{0pt}
    \setlength{\parsep}{0pt}
    \setlength{	opsep}{0pt}
    \setlength{\partopsep}{0pt}
    \setlength{\leftmargin}{2em}
    \setlength{\labelwidth}{1.5em}
    \setlength{\labelsep}{0.5em} } }

\newcommand{\squishend}{
  \end{list}  }

\begin{document}

\maketitle

\section{Expert comments}

% We sent feedback requests to several researchers involved in Sociology and Urban
% sciences. Not all answered before the submission deadline, but we will keep an
% updated version of this document at:

% \href{http://ug.daniels.utoronto.ca/~diasf/CTEvo/sup.pdf}{\url{http://ug.daniels.utoronto.ca/~diasf/CTEvo/sup.pdf}}

% Last edit: \today

\subsection{Original request}
As part of our efforts in the Urban Genome Project of the University of Toronto,
we are trying to build infrastructure to help understand recurrent and
distinctive patterns of urban change.

This effort has led to a prototype interface to allow the visual exploration of
census data. While we plan to add additional data sources, the objective of this
prototype is to facilitate interpretation of census data, minimizing the need
for specialised analytical tools. Hopefully this will allow non-experts to be
able to gain insight from it, but we believe that it can be helpful to experts
as well.

This prototype is part of an article, to be submitted to the next IEEE VIS
conference, and we are humbly requesting feedback on how effective it actually
is.

The documentation that briefly explains what we did is available at:

\href{http://ug.daniels.utoronto.ca/~diasf/CTEvo/doc.html}{\url{http://ug.daniels.utoronto.ca/~diasf/CTEvo/doc.html}}

The video guides demonstrating how to use the tool are available at:

\href{https://www.youtube.com/watch?v=OZFUA0ThEaY&list=PLBd6KRI4PG-RcxAX35OfHd1DKl3qWfXRV}{youtube}

And the prototype itself is available at:

\href{http://ug.daniels.utoronto.ca/~diasf/CTEvo/}{\url{http://ug.daniels.utoronto.ca/~diasf/CTEvo/}}

The prototype should work on all major browsers, but in our experience Chrome
seems to offer the best performance.

Should you chose to provide feedback, we will add it, verbatim, to the
supplementary material of the paper (along with this text), and excerpts of it
may be used in the paper itself. If you do not wish for some information to be
publicly available, including your name/affiliation, please let us know.

Specifically, we would like more information regarding:
\squishlist
    \item{Does this prototype make the exploration of census data easier?}
    \item{By using it, did you find anything interesting? Either new insight or something that demonstrates/disproves an established insight? }
    \item{What questions would you be interested in pursuing further with this tool? }
    \item{What would you add/remove/change on it?}
\squishend

However, please do not refrain from providing further comments, any and all
feedback is welcome.

% In order to be included in the IEEE VIS conference
% submission, we would need your feedback by March 21.  We recognize this is short
% notice, and definitely would welcome your feedback even if it comes later.  In
% addition, if you know of others who might be interested and willing to provide
% feedback, please feel free to forward this message to them.

\subsection{A - Cary Wu}
\href{https://soci.ubc.ca/persons/cary-wu/}{\url{https://soci.ubc.ca/persons/cary-wu/}}
% \subsection{Expert A}
\emph{Doctoral candidate in Sociology, UBC.}

\question{Does this prototype make the exploration of census data easier?}

It definitely does! With the visualizations, we can tell many different stories
about neighborhoods/cities and their changes over time without doing the “dirty
work” of analyzing raw/big census data.

\question{By using it, did you find anything interesting? Either new insight or
something that demonstrates/disproves an established insight? }

This program demonstrates how specific combinations of local characteristics can
create distinctive or comparable scenes. As such, neighborhoods and cities can
be studied and compared in a way that the unit of sociological/urban analysis is
scenes.

\question{What questions would you be interested in pursuing further with this tool?}

We can perhaps develop measures based on cluster changes that could capture
whether a city or a neighborhood has become more diverse or homogeneous, and
also in terms of different aspects (e.g., race/ethnicity, SES, religion, and
immigration). It will also be interesting to explore whether different clusters
would shape other aspects of the neighborhoods and cities, such as trust, voting
pattern, and population movement.

\question{What would you add/remove/change on it?}

Instead of presenting everything on one overwhelming webpage (!), I would maybe
create multiple pages for different sections (e.g., configuration, choose
different cities), and embed them into one home page. Or maybe use popups for
results?

Overall, it is a very cool program, and I can see huge potential as a tool for
survey data visualization. For example, we can develop similar tools to
visualize the U.S. General Social Survey across different states from 1972-2016.

\subsection{B - Ethan A. Fosse}
% \href{https://scholar.harvard.edu/ethanfosse}{\url{https://scholar.harvard.edu/ethanfosse}}
% \subsection{Expert B}
\emph{Research Associate in  Sociology, Harvard}

Very cool visualization tool! To answer your questions:

\question{Does this prototype make the exploration of census data easier?}

Yes, absolutely. Census data can be difficult to download and clean, especially
for people who are not familiar with statistical software packages.

\question{By using it, did you find anything interesting? Either new insight or
something that demonstrates/disproves an established insight?}

I focused on Manhattan, since this is near where I currently live (Jersey City,
New Jersey). The visualization tool confirmed the gentrification of Manhattan,
with the exception of some areas (e.g., Chinatown and parts of the Lower East
Side). I'd like to know more about who lives in the Lower East Side. Is there a
conscious effort to resist gentrification in that part of the city, or is it
just a consequence of limited public transportation to the area, or is something
else going on?

\question{What questions would you be interested in pursuing further with this tool? }

I would like to see other data incorporated into the tool -- e.g., "cultural"
data involving consumption practices. Of course, it's no easy feat to obtain
these data, let alone merge into a geographic information system.

\question{What would you add/remove/change on it?}

Overall I enjoyed using the visualization tool and I found it easy to navigate.
I have two suggestions. First, I would work on speeding up the code that loads
the tool. Even after several minutes I could not load the Brooklyn data on
various browsers, for example. Second, as far as changes to the interface, my
only recommendation is to place the choice of various locations in a more
prominent place. Typically I think of the "Gear" icon as representing more
technical options, but the first change I made was altering the location.

\subsection{C - Fernando Calderón Figueroa}
\emph{Sociology Department, University of Toronto}

\textbf{Overall assessment}

As a first impression, I think the tool needs a brief summary of what it does
and how can people play with it. It took me a while between reading the
documentation and playing with it to make sense of what was going on underneath.
Maybe having a 5-minute tutorial would solve the problem, with someone (a voice
and captions) showing the user what’s going on in each section of the interface.
This has helped me in the past dealing with new software.

I really like the idea of the colours having the additional tag of the feature
that defines them more clearly on the left. However, once one starts playing
around with the platform using several variables at the same time, things become
tricky to interpret. For example, in Vancouver, things seem to change
dramatically from all green to all orange. This is highlighted by the graph of
trajectories in which the population involved in each cluster is nested in
colours. Once one gets into the data, the changes don’t look so sharp after all
(maybe in education and income). This may not be a problem of the platform as it
is, but there may be a warning regarding that the “severity” of changes is not
necessarily the same across cities.

Related to my previous point, in some cities the interpretation of the patterns
is clearer than in others. I think the platform works great in Montreal, for
instance. With a little knowledge about the city, the clustering makes a lot of
sense in how the composition has evolved.

\textbf{Some specifics}

\begin{itemize}
    \item{For the boxplots, overlapping the gray ones in the back to show the
            difference with other clusters is a bit confusing, especially for
            those with no colour in them. Maybe a different colour or more
            transparency would help. }

    \item{I have trouble making sense of the top-right map. In the documentation
            it’s described as “the geographic positions of the trajectories,
            considering all years (the other maps show the clustering results
            for specific years).” While this is clear, it seems a little tricky
            to interpret. Is it the clustering adding all years together? Is it
            calculated as an average trajectory? How is that interpreted?
            Overall, I thought this to be a good summary map, but it has such an
            importance in the interface that I think obscures the fact that the
            maps below show, as a whole, how the trajectories move over time.}

    \item{If, as a user, I wanted to get a simple output for a presentation, it
            would be ideal to get the sequence of maps to show the changes over
            time. I could get a screenshot but providing it as a potential
            product would be very useful. A more sophisticated version would
            include a GIF image, but that may be harder to do and not so popular
            for other users beyond me.}
\end{itemize}

\subsection{D - Patrick Adler}
\emph{Research Associate at the Martin Prosperity Institute, Canada.}

\question{Does this prototype make the exploration of census data easier?}

The prototype allows me to visualize the relationship between key urban
variables over time and easily explore  what variables distinguish spaces from
each other. As someone who can’t make my own maps, I usually have to wait until
the end of a project before I start to commission visuals. I only have the
luxury of visualizing data that I fully understand (or think I fully understand
at least).  This tool allows me to use visualization as a tool in my own
research process.

\question{By using it, did you find anything interesting? Either new insight or
 something that demonstrates/disproves an established insight? What questions
 would you be interested in pursuing further with this tool? }

The spatial analysis  supports speculative reasoning from [1] about the return
to central cities and central business districts of moneyed, educated, knowledge
workers. They hypothesized that this transition happened over the past 20 years
and  that it is more complete in a set of metros with high levels of
‘post-industrialism’. The tools helps to verify this analysis and show that
occupational clustering is robust in many cities, even when orthogonal measures
are added.  This tool can be used to formally extend the analysis to the full
range of metros.

\noindent [1]: \href{https://www.tandfonline.com/doi/abs/10.1080/07352166.2017.1360743}{\url{https://www.tandfonline.com/doi/abs/10.1080/07352166.2017.1360743}}

\question{What would you add/remove/change on it?}
\begin{enumerate}
  \item{More summary statistics about the clusters would be very useful (density,
  population, distance from transit etc)}
  \item{A tool that would allow us to  point to a cluster and find similar clusters in
  other neighborhoods would be most useful.}
  \item{It would be great to have a quick way to snap pictures, as well as a
  tool to export data in to CSV}
  \item{We would appreciate the ability to restrict the analysis to an interval
  within. For the analysis above 2000-2010 might be useful}
  \item{Processing speed is an issue that I'm sure will be resolved in time}
\end{enumerate}

\subsection{E - James Murdoch}%, 
\emph{Policy Analyst, 2M Research}

Wow!! This is quite an undertaking and I definitely see the potential of this
tool for understanding neighborhood dynamics overtime. Here are some uses I see

It offers a quick way to identify particular neighborhoods (i.e. Census tracts)
that one may be interested in studying more in depth around a particular issue.
A good example is gentrification. I was able to use the tool to identify several
neighborhoods in Dallas that I know have gentrified over the years by looking at
how race/education/income changed since 1970. You can also easily see when the
neighborhood “tipped” as the cluster color will change. That is quite a feat as
without this tool just getting all the necessary data is a lot of work. Plus
being able to see a neighborhood change from black dominated to white dominated,
for example, is really powerful. Its really a nice tool for efficiently
understanding the context of an area. In a program evaluation of some kind of
place-based intervention this tool would be a great way to produce data on the
context in which the intervention is occurring. I could also see it facilitating
descriptive analysis that provides some evidence of how an intervention may
impact a particular place as the tool allows for the analysis of trends. I think
it could be an excellent planning tool for local governments. As more data is
added, the tool will be able to quickly show clusters of particular types of
economic activity, cultural activity, types of residents and consumers, and many
other things that planners may be interested in. It could also serve as a
communication tool for more participatory planning. For example, you could take
an asset-based approach where you highlight the resources in the neighborhood
with the tool and ask residents to react, offer insight, and make suggestions
how to capitalize on what already exists in a neighborhood or city.

Some questions:
\begin{itemize}
\item{Do you think users will ever be able to import their own data? For
example, what if someone had violent crime data from the local police department
that they want to overlay with the other data in the tool. Will that be possible
at some point?}

\item{I know you can’t export raw data, but what about the ability to save a
customized map or some type of data graphic? That would be super useful for
people that need to write reports or do a presentation.}

\item{Any ambitions to include social media info or other “big data” sources
similar to that in here?}
\end{itemize}

A couple criticisms:

\begin{itemize}

  \item{Obviously the tool is still buggy. I had to restart several times. Also
  it would be nice if the big map in the top right also moved the smaller maps
  that correspond to different time points just below it. I notice that moving
  one of the smaller maps moves all the other smaller maps, but moving the big
  map doesn’t do anything. Also I didn’t see a way to enlarge the map. I’d like
  to be able to have a full screen view of the map that I can toggle when
  needed.}

  \item{It is a little intimidating at first. I might recommend starting with a
  simpler default (only one variable, less than 5 clusters, etc.) and allow
  people to work up in complexity if they want to. I found it a bit hard to get
  in to it as I had to really think about what I was looking at first.}
\end{itemize}

I really must commend you on your work. This is an amazing tool that you have
put together. Great job and thanks for sharing this with me!! Please keep me
updated as you develop it further. I could definitely see a use for it in the
work I do both in my personal research and at my work at 2M.

\clearpage
\section{Variable matching}

The following listing contains the fields that were added to form the aspects.
If the name of a field starts with "(-)", it means that this field was
subtracted.

\subsection{Canada}
\subsubsection{Home language}
\paragraph{English}
\begin{itemize}
   \item{\textbf{1971}:  Home language: English;}
   \item{\textbf{1981}:  Home language - - English home language; Home language - - English home language; Home language - - English home language;}
   \item{\textbf{1991}:  English, home language;}
   \item{\textbf{2001}:  English;}
   \item{\textbf{2011}:  Detailed language spoken most often at home - Both sexes / Total population excluding institutional residents, Both sexes / Single responses, Both sexes / English, Both sexes;}
\end{itemize}

\paragraph{French}
\begin{itemize}
   \item{\textbf{1971}:  Home language: French;}
   \item{\textbf{1981}:  Home language - - French home language; Home language - - French home language; Home language - - French home language;}
   \item{\textbf{1991}:  French, home language;}
   \item{\textbf{2001}:  French;}
   \item{\textbf{2011}:  Detailed language spoken most often at home - Both sexes / Total population excluding institutional residents, Both sexes / Single responses, Both sexes / French, Both sexes;}
\end{itemize}

\paragraph{Other}
\begin{itemize}
   \item{\textbf{1971}:  Home language: other;}
   \item{\textbf{1981}:  Home language - - home language same as mother tongue;}
   \item{\textbf{1991}:  Non-official languages, home language;}
   \item{\textbf{2001}:  Non-official languages;}
   \item{\textbf{2011}:  Detailed language spoken most often at home - Both sexes / Total population excluding institutional residents, Both sexes / Single responses, Both sexes / Non-official languages, Both sexes;}
\end{itemize}

\subsubsection{Population}
\paragraph{Total}
\begin{itemize}
   \item{\textbf{1971}:  Total population;}
   \item{\textbf{1981}:  Population, 1981;}
   \item{\textbf{1991}:  Total population;}
   \item{\textbf{2001}:  Population, 2001 - 100\% Data;}
   \item{\textbf{2011}:  Population and dwelling counts / Population, 2011;}
\end{itemize}

\subsubsection{Age}
\paragraph{0-4}
\begin{itemize}
   \item{\textbf{1971}:  Population: males: 0-4 years of age; Population: females: 0-4 years of age;}
   \item{\textbf{1981}:  Males, 0 - 4 years; Females, 0 - 4 years;}
   \item{\textbf{1991}:  Males, 0 - 4 years; Females, 0 - 4 years;}
   \item{\textbf{2001}:  0-4; 0-4;}
   \item{\textbf{2011}:  Age \& Sex - Both sexes / Total population by age groups, Both sexes / 0 to 4 years, Both sexes;}
\end{itemize}

\paragraph{5-14}
\begin{itemize}
   \item{\textbf{1971}:  Population: males: 10-14 years of age; Population: females: 10-14 years of age; Population: males: 5-9 years of age; Population: females: 5-9 years of age;}
   \item{\textbf{1981}:  Males, 10 - 14 years; Females, 10 - 14 years; Males, 5 - 9 years; Females, 5 - 9 years;}
   \item{\textbf{1991}:  Males, 10 - 14 years; Females, 10 - 14 years; Males, 5 - 9 years; Females, 5 - 9 years;}
   \item{\textbf{2001}:  10-14; 10-14; 5-9; 5-9;}
   \item{\textbf{2011}:  Age \& Sex - Both sexes / Total population by age groups, Both sexes / 10 to 14 years, Both sexes; Age \& Sex - Both sexes / Total population by age groups, Both sexes / 5 to 9 years, Both sexes;}
\end{itemize}

\paragraph{15-24}
\begin{itemize}
   \item{\textbf{1971}:  Population: males: 15 years of age; Population: males: 16 years of age; Population: males: 17 years of age; Population: males: 18-19 years of age; Population: females: 15 years of age; Population: females: 16 years of age; Population: females: 17 years of age; Population: females: 18-19 years of age; Population: males: 20-24 years of age; Population: females: 20-24 years of age;}
   \item{\textbf{1981}:  Males, 15 - 19 years; Females, 15 - 19 years; Males, 20 - 24 years; Females, 20 - 24 years;}
   \item{\textbf{1991}:  Males, 15 - 19 years; Females, 15 - 19 years; Males, 20 - 24 years; Females, 20 - 24 years;}
   \item{\textbf{2001}:  15-19; 15-19; 20-24; 20-24;}
   \item{\textbf{2011}:  Age \& Sex - Both sexes / Total population by age groups, Both sexes / 15 to 19 years, Both sexes; Age \& Sex - Both sexes / Total population by age groups, Both sexes / 20 to 24 years, Both sexes;}
\end{itemize}

\paragraph{25-34}
\begin{itemize}
   \item{\textbf{1971}:  Population: males: 25-29 years of age; Population: males: 30-34 years of age; Population: females: 25-29 years of age; Population: females: 30-34 years of age;}
   \item{\textbf{1981}:  Males, 25 - 34 years; Females, 25 - 34 years;}
   \item{\textbf{1991}:  Males, 25 - 29 years; Males, 30 - 34 years; Females, 25 - 29 years; Females, 30 - 34 years;}
   \item{\textbf{2001}:  25-29; 30-34; 25-29; 30-34;}
   \item{\textbf{2011}:  Age \& Sex - Both sexes / Total population by age groups, Both sexes / 25 to 29 years, Both sexes; Age \& Sex - Both sexes / Total population by age groups, Both sexes / 30 to 34 years, Both sexes;}
\end{itemize}

\paragraph{35-44}
\begin{itemize}
   \item{\textbf{1971}:  Population: males: 35-39 years of age; Population: males: 40-44 years of age; Population: females: 35-39 years of age; Population: females: 40-44 years of age;}
   \item{\textbf{1981}:  Males, 35 - 44 years; Females, 35 - 44 years;}
   \item{\textbf{1991}:  Males, 35 - 39 years; Males, 40 - 44 years; Females, 35 - 39 years; Females, 40 - 44 years;}
   \item{\textbf{2001}:  35-39; 40-44; 35-39; 40-44;}
   \item{\textbf{2011}:  Age \& Sex - Both sexes / Total population by age groups, Both sexes / 35 to 39 years, Both sexes; Age \& Sex - Both sexes / Total population by age groups, Both sexes / 40 to 44 years, Both sexes;}
\end{itemize}

\paragraph{45-54}
\begin{itemize}
   \item{\textbf{1971}:  Population: males: 45-49 years of age; Population: males: 50-54 years of age; Population: females: 45-49 years of age; Population: females: 50-54 years of age;}
   \item{\textbf{1981}:  Males, 45 - 54 years; Females, 45 - 54 years;}
   \item{\textbf{1991}:  Males, 45 - 49 years; Males, 50 - 54 years; Females, 45 - 49 years; Females, 50 - 54 years;}
   \item{\textbf{2001}:  45-49; 50-54; 45-49; 50-54;}
   \item{\textbf{2011}:  Age \& Sex - Both sexes / Total population by age groups, Both sexes / 45 to 49 years, Both sexes; Age \& Sex - Both sexes / Total population by age groups, Both sexes / 50 to 54 years, Both sexes;}
\end{itemize}

\paragraph{55-64}
\begin{itemize}
   \item{\textbf{1971}:  Population: males: 55-59 years of age; Population: males: 60-64 years of age; Population: females: 55-59 years of age; Population: females: 60-64 years of age;}
   \item{\textbf{1981}:  Males, 55 - 64 years; Females, 55 - 64 years;}
   \item{\textbf{1991}:  Males, 55 - 59 years; Males, 60 - 64 years; Females, 55 - 59 years; Females, 60 - 64 years;}
   \item{\textbf{2001}:  55-59; 60-64; 55-59; 60-64;}
   \item{\textbf{2011}:  Age \& Sex - Both sexes / Total population by age groups, Both sexes / 55 to 59 years, Both sexes; Age \& Sex - Both sexes / Total population by age groups, Both sexes / 60 to 64 years, Both sexes;}
\end{itemize}

\paragraph{65 and over}
\begin{itemize}
   \item{\textbf{1971}:  Population: males: 65-69 years of age; Population: males: 70 years of age and over; Population: females: 65-69 years of age; Population: females: 70 years of age and over;}
   \item{\textbf{1981}:  Males, 65 - 69 years; Males, 70 years and over; Females, 65 - 69 years; Females, 70 years and over;}
   \item{\textbf{1991}:  Males, 65 - 74 years; Males, 75 - 84 years; Males, 85 years and over; Females, 65 - 74 years; Females, 75 - 84 years; Females, 85 years and over;}
   \item{\textbf{2001}:  65-69; 70-74; 75-79; 80-84; 85+; 65-69; 70-74; 75-79; 80-84; 85+;}
   \item{\textbf{2011}:  Age \& Sex - Both sexes / Total population by age groups, Both sexes / 65 to 69 years, Both sexes; Age \& Sex - Both sexes / Total population by age groups, Both sexes / 70 to 74 years, Both sexes; Age \& Sex - Both sexes / Total population by age groups, Both sexes / 75 to 79 years, Both sexes; Age \& Sex - Both sexes / Total population by age groups, Both sexes / 80 to 84 years, Both sexes; Age \& Sex - Both sexes / Total population by age groups, Both sexes / 85 years and over, Both sexes;}
\end{itemize}

\subsubsection{Education level}
\paragraph{HS or below}
\begin{itemize}
   \item{\textbf{1971}:  Pop 5 yrs+: highest level schooling: grade 11-13: no other t; Pop 5 yrs+: highest level schooling: some university: no oth; Pop 5 yrs+: highest level schooling: less than grade 9; Pop 5 yrs+: highest level schooling: grade 9-10: no other tr; (-)  Population: males: 10-14 years of age; (-)  Population: females: 10-14 years of age; (-)  Population: males: 5-9 years of age; (-)  Population: females: 5-9 years of age;}
   \item{\textbf{1981}:  Highest level of schooling: grades 9-13 with secondary certificate; Highest level of schooling: other non-university education only(11)- without certificate; Highest level of schooling: grades 9-13 without secondary certificate;}
   \item{\textbf{1991}:  Population 15+ years - Grades 9-13 - With secondary certificate; Population 15+ years - Other non-university - Without certificate; Population 15+ years - University - Without degree; Population 15+ years - University - Without certificate; Population 15+ years - Highest level of schooling, less than grade 9; Population 15+ years - Grades 9-13 - Without secondary certificate;}
   \item{\textbf{2001}:  With high school graduation certificate; Without certificate or diploma; Without degree; Without certificate or diploma; Less than grade 9; Without high school graduation certificate;}
   \item{\textbf{2011}:  Education - Both sexes / Total population aged 15 years and over by highest certificate, diploma or degree, Both sexes / High school diploma or equivalent, Both sexes; 
                         Education - Both sexes / Total population aged 15 years and over by highest certificate, diploma or degree, Both sexes / No certificate, diploma or degree, Both sexes;}
\end{itemize}

\paragraph{Trades or college}
\begin{itemize}
   \item{\textbf{1971}:  Pop 5 yrs+: highest level schooling: grade 9-10: \& other tra; Pop 5 yrs+: highest level schooling: grade 11-13: \& other tr; Pop 5 yrs+: highest level schooling: some university: \& othe;}
   \item{\textbf{1981}:  Highest level of schooling: trades certificate or diploma; Highest level of schooling: other non-university education only- with certificate(12);}
   \item{\textbf{1991}:  Population 15+ years - Trades certificate or diploma; Population 15+ years - Other non-university - With certificate;}
   \item{\textbf{2001}:  Trades certificate or diploma; With certificate or diploma;}
   \item{\textbf{2011}:  Education - Both sexes / Total population aged 15 years and over by highest certificate, diploma or degree, Both sexes / Postsecondary certificate, diploma or degree, Both sexes / Apprenticeship or trades certificate or diploma, Both sexes; Education - Both sexes / Total population aged 15 years and over by highest certificate, diploma or degree, Both sexes / Postsecondary certificate, diploma or degree, Both sexes / College, CEGEP or other non-university certificate or diploma, Both sexes;}
\end{itemize}

\paragraph{University}
\begin{itemize}
   \item{\textbf{1971}:  Pop 5 yrs+: highest level schooling: univ degree: no other t; Pop 5 yrs+: highest level schooling: univ degree: \& other tr;}
   \item{\textbf{1981}:  Highest level of schooling: university(13)- with degree;}
   \item{\textbf{1991}:  Population 15+ years - University - With certificate; Population 15+ years - University - With degree;}
   \item{\textbf{2001}:  With certificate or diploma; With bachelors degree or higher;}
   \item{\textbf{2011}:  Education - Both sexes / Total population aged 15 years and over by highest certificate, diploma or degree, Both sexes / Postsecondary certificate, diploma or degree, Both sexes / University certificate or diploma below bachelor level, Both sexes; 
                         Education - Both sexes / Total population aged 15 years and over by highest certificate, diploma or degree, Both sexes / Postsecondary certificate, diploma or degree, Both sexes / University certificate, diploma or degree at bachelor level or above, Both sexes;}
\end{itemize}

\subsubsection{Occupation}
\paragraph{Managerial and business support}
\begin{itemize}
   \item{\textbf{1971}:  Managerial, admin \& related occupations: males; Managerial, administrative \& related occupations: females; Clerical \& related: males; Clerical \& related occupations: females;}
   \item{\textbf{1981}:  Occupation major groups(18): males - managerial, administrative and related occupations; Occupation major groups(18): females - managerial, administrative and related occupations; Occupation major groups(18): males - clerical and related occupations; Occupation major groups(18): females - clerical and related occupations;}
   \item{\textbf{1991}:  Males, Managerial, administrative and related occupations; Females, Managerial, administrative and related occupations; Males, Clerical and related occupations; Females, Clerical and related occupations;}
   \item{\textbf{2001}:  A Management occupations; B Business, finance and administration occupations;}
   \item{\textbf{2011}:  Occupation - Both sexes / Total labour force population aged 15 years and over by occupation - National Occupational Classification (NOC) 2011, Both sexes / All occupations, Both sexes / 0 Management occupations, Both sexes; Occupation - Both sexes / Total labour force population aged 15 years and over by occupation - National Occupational Classification (NOC) 2011, Both sexes / All occupations, Both sexes / 1 Business, finance and administration occupations, Both sexes;}
\end{itemize}

\paragraph{Culture}
\begin{itemize}
   \item{\textbf{1971}:  Technological, social, religious, artistic occupations: male; Technological, social, religious, artistic occupations: fema; Teaching \& related occupations: males; Teaching \& related occupations: females;}
   \item{\textbf{1981}:  Occupation major groups(18): males - teaching and related occupations; Occupation major groups(18): males - technological, social, religious, artistic and related occupations (21); Occupation major groups(18): females - teaching and related occupations; Occupation major groups(18): females - technological, social, religious, artistic \& related occupations(21);}
   \item{\textbf{1991}:  Males, Occupations in nat. sciences, engineering and math.; ales, Occupations in social sciences and related fields; Males, Occupations in religion; Males, Teaching and related occupations; Males, Artistic, literary, recreational and related occs.; Females, Occupations in nat. sciences, engineering and math.; Females, Occupations in social sciences and related fields; Females, Occupations in religion; Females, Teaching and related occupations; Females, Artistic, literary, recreational and related occs.;}
   \item{\textbf{2001}:  C Natural and applied sciences and related occupations; E Occupations in social science, education, government service and religion; F Occupations in art, culture, recreation and sport;}
   \item{\textbf{2011}:  Occupation - Both sexes / Total labour force population aged 15 years and over by occupation - National Occupational Classification (NOC) 2011, Both sexes / All occupations, Both sexes / 2 Natural and applied sciences and related occupations, Both sexes; Occupation - Both sexes / Total labour force population aged 15 years and over by occupation - National Occupational Classification (NOC) 2011, Both sexes / All occupations, Both sexes / 4 Occupations in education, law and social, community and government services, Both sexes; Occupation - Both sexes / Total labour force population aged 15 years and over by occupation - National Occupational Classification (NOC) 2011, Both sexes / All occupations, Both sexes / 5 Occupations in art, culture, recreation and sport, Both sexes;}
\end{itemize}

\paragraph{Health}
\begin{itemize}
   \item{\textbf{1971}:  Medicine \& health occupations: males; Medicine \& health occupations: females;}
   \item{\textbf{1981}:  Occupation major groups(18): males - occupations in medicine and health; Occupation major groups(18): females - occupations in medicine and health;}
   \item{\textbf{1991}:  Males, Occupations in medicine and health; Females, Occupations in medicine and health;}
   \item{\textbf{2001}:  D Health occupations;}
   \item{\textbf{2011}:  Occupation - Both sexes / Total labour force population aged 15 years and over by occupation - National Occupational Classification (NOC) 2011, Both sexes / All occupations, Both sexes / 3 Health occupations, Both sexes;}
\end{itemize}

\paragraph{Sales and service}
\begin{itemize}
   \item{\textbf{1971}:  Sales occupations: males; Service occupations: males; Sales occupations: females; Service occupations: females;}
   \item{\textbf{1981}:  Occupation major groups(18): males - sales occupations; Occupation major groups(18): males - service occupations; Occupation major groups(18): females - sales occupations; Occupation major groups(18): females - service occupations;}
   \item{\textbf{1991}:  Males, Sales occupations; Males, Service occupations; Females, Sales occupations; Females, Service occupations;}
   \item{\textbf{2001}:  G Sales and service occupations;}
   \item{\textbf{2011}:  Occupation - Both sexes / Total labour force population aged 15 years and over by occupation - National Occupational Classification (NOC) 2011, Both sexes / All occupations, Both sexes / 6 Sales and service occupations, Both sexes;}
\end{itemize}

\paragraph{Trades and construction}
\begin{itemize}
   \item{\textbf{1971}:  Construction trade occupations: males; Transport equipment operating occupations: males; Construction trade occupations: females; Transport equipment operating occupations: females;}
   \item{\textbf{1981}:  Occupation major groups(18): males - construction trades occupations; Occupation major groups(18): males - transport equipment operating occupations;}
   \item{\textbf{1991}:  Males, Construction trades occupations; Males, Transport equipment operating occupations; Females, Construction trades occupations; Females, Transport equipment operating occupations;}
   \item{\textbf{2001}:  H Trades, transport and equipment operators and related occupations;}
   \item{\textbf{2011}:  Occupation - Both sexes / Total labour force population aged 15 years and over by occupation - National Occupational Classification (NOC) 2011, Both sexes / All occupations, Both sexes / 7 Trades, transport and equipment operators and related occupations, Both sexes;}
\end{itemize}

\paragraph{Primary}
\begin{itemize}
   \item{\textbf{1971}:  Fishing \& hunting, forestry, mining, etc. occupations: males; Fishing \& hunting, forestry, mining, etc occupations: female;}
   \item{\textbf{1981}:  Occupation major groups(18): males - primary occupations(22); Occupation major groups(18): females - primary occupations(22);}
   \item{\textbf{1991}:  Males, Farming, horticultural and animal husbandry occs.; Males, Fishing, trapping and related occupations; Males, Forestry and logging occupations; Males, Mining \& quarrying incl. oil \& gas field occs.; Females, Farming, horticultural and animal husbandry occs.; Females, Fishing, trapping and related occupations; Females, Forestry and logging occupations; Females, Mining \& quarrying incl. oil \& gas field occs.;}
   \item{\textbf{2001}:  I Occupations unique to primary industry;}
   \item{\textbf{2011}:  Occupation - Both sexes / Total labour force population aged 15 years and over by occupation - National Occupational Classification (NOC) 2011, Both sexes / All occupations, Both sexes / 8 Natural resources, agriculture and related production occupations, Both sexes;}
\end{itemize}

\paragraph{Manufacturing}
\begin{itemize}
   \item{\textbf{1971}:  Processing occupations: males; Machining and related occupations: males; Processing occupations: females; Machining and related occupations: females;}
   \item{\textbf{1981}:  Occupation major groups(18): males - processing occupations; Occupation major groups(18): males - machining, product fabricating, assembling, and repairing occupations(23); Occupation major groups(18): females - processing occupations; Occupation major groups(18): females - machining, product fabricating, assemblng \& repairing occs(23);}
   \item{\textbf{1991}:  Males, Processing occupations; Males, Machining and related occupations; Males, Product fabricating, assembling \& repairing occs.; Males, Material handling and related occupations, n.e.c.; Males, Other crafts and equipment operating occupations; Females, Product fabricating, assembling \& repairing occs.; Females, Material handling and related occupations, n.e.c.; Females, Other crafts and equipment operating occupations;}
   \item{\textbf{2001}:  J Occupations unique to processing, manufacturing and utilities;}
   \item{\textbf{2011}:  Occupation - Both sexes / Total labour force population aged 15 years and over by occupation - National Occupational Classification (NOC) 2011, Both sexes / All occupations, Both sexes / 9 Occupations in manufacturing and utilities, Both sexes;}
\end{itemize}

\subsubsection{Religion}
\paragraph{Catholic}
\begin{itemize}
   \item{\textbf{1971}:  Religion: Roman Catholic; Religion: Ukrainian Catholic;}
   \item{\textbf{1981}:  Religion - Catholic;}
   \item{\textbf{1991}:  Religion, Catholic, total;}
   \item{\textbf{2001}:  Roman Catholic; Ukrainian Catholic;}
   \item{\textbf{2011}:  Religion - Both sexes / Total population in private households by religion, Both sexes / Christian, Both sexes / Catholic, Both sexes;}
\end{itemize}

\paragraph{Protestant}
\begin{itemize}
   \item{\textbf{1971}:  Religion: Adventist; Religion: Anglican; Religion: Baptist; Religion: United Church; Religion: Presbyterian; Religion: Lutheran; Religion: Pentecostal; Religion: Mennonite; Religion: Jehovahs Witness; Religion: Salvation Army; Religion: Mormon; Religion: Hutterite; Religion: Christian \& Missionary Alliance; Religion: Christian Reformed; Religion: Churches of Christ Disciples;}
   \item{\textbf{1981}:  Religion - Protestant;}
   \item{\textbf{1991}:  Religion, Protestant, total;}
   \item{\textbf{2001}:  Salvation Army; Jehovahs Witnesses; United Church; Anglican; Baptist; Lutheran; Protestant not included elsewhere; Presbyterian; Pentecostal; Mennonite; Adventist; Church of Jesus Christ of Latter-day Saints (Mormons); Hutterite; Christian and Missionary Alliance; Christian Reformed Church; Methodist; Brethren in Christ;}
   \item{\textbf{2011}:  Religion - Both sexes / Total population in private households by religion, Both sexes / Christian, Both sexes / Anglican, Both sexes; Religion - Both sexes / Total population in private households by religion, Both sexes / Christian, Both sexes / Baptist, Both sexes; Religion - Both sexes / Total population in private households by religion, Both sexes / Christian, Both sexes / Lutheran, Both sexes; Religion - Both sexes / Total population in private households by religion, Both sexes / Christian, Both sexes / Pentecostal, Both sexes; Religion - Both sexes / Total population in private households by religion, Both sexes / Christian, Both sexes / Presbyterian, Both sexes; Religion - Both sexes / Total population in private households by religion, Both sexes / Christian, Both sexes / United Church, Both sexes;}
\end{itemize}

\paragraph{Jewish}
\begin{itemize}
   \item{\textbf{1971}:  Religion: Jewish;}
   \item{\textbf{1981}:  Religion - Jewish;}
   \item{\textbf{1991}:  Other religions - Jewish;}
   \item{\textbf{2001}:  Jewish;}
   \item{\textbf{2011}:  Religion - Both sexes / Total population in private households by religion, Both sexes / Jewish, Both sexes;}
\end{itemize}

\paragraph{Eastern Orthodox}
\begin{itemize}
   \item{\textbf{1971}:  Religion: Greek Orthodox;}
   \item{\textbf{1981}:  Religion - Eastern Orthodox;}
   \item{\textbf{1991}:  Other religions - Eastern Orthodox;}
   \item{\textbf{2001}:  Greek Orthodox; Ukrainian Orthodox; Serbian Orthodox; Orthodox not included elsewhere;}
   \item{\textbf{2011}:  Religion - Both sexes / Total population in private households by religion, Both sexes / Christian, Both sexes / Christian Orthodox, Both sexes;}
\end{itemize}

\paragraph{Eastern non-christian}
\begin{itemize}
   \item{\textbf{1971}:  Religion: Buddhist \& Confucian;}
   \item{\textbf{1981}:  Religion - eastern non-cristian;}
   \item{\textbf{1991}:  Other religions - Islam; Other religions - Buddhist; Other religions - Hindu; Other religions - Sikh;}
   \item{\textbf{2001}:  Muslim; Buddhist; Hindu; Sikh;}
   \item{\textbf{2011}:  Religion - Both sexes / Total population in private households by religion, Both sexes / Buddhist, Both sexes; Religion - Both sexes / Total population in private households by religion, Both sexes / Sikh, Both sexes; Religion - Both sexes / Total population in private households by religion, Both sexes / Muslim, Both sexes; Religion - Both sexes / Total population in private households by religion, Both sexes / Hindu, Both sexes;}
\end{itemize}

\paragraph{No preference}
\begin{itemize}
   \item{\textbf{1971}:  Religion: no religion;}
   \item{\textbf{1981}:  Religion - no religious preference;}
   \item{\textbf{1991}:  No religious affiliation;}
   \item{\textbf{2001}:  No religion;}
   \item{\textbf{2011}:  Religion - Both sexes / Total population in private households by religion, Both sexes / No religious affiliation, Both sexes;}
\end{itemize}

\paragraph{Other}
\begin{itemize}
   \item{\textbf{1971}:  Religion: other;}
   \item{\textbf{1981}:  Religion - other;}
   \item{\textbf{1991}:  Other religions - Other (other religions);}
   \item{\textbf{2001}:  Aboriginal spirituality; Pagan; Christian not included elsewhere; Non-denominational; Evangelical Missionary Church;}
   \item{\textbf{2011}:  Religion - Both sexes / Total population in private households by religion, Both sexes / Other religions, Both sexes; Religion - Both sexes / Total population in private households by religion, Both sexes / Traditional (Aboriginal) Spirituality, Both sexes; Religion - Both sexes / Total population in private households by religion, Both sexes / Christian, Both sexes / Other Christian, Both sexes;}
\end{itemize}

\subsubsection{Household Income (2015)}
\paragraph{Less than 30k}
\begin{itemize}
   \item{\textbf{1971}:  Households: income from all sources: under \$3,000; Households: income from all sources: \$3,000-\$4,999;}
   \item{\textbf{1981}:  Private household income - all households- under \$5,000; Private household income - all households- \$ 5,000-\$ 9,999;}
   \item{\textbf{1991}:  Under \$10,000, household income; \$10,000 - \$14,999, household income; \$15,000 - \$19,999, household income;}
   \item{\textbf{2001}:  Under \$10,000; \$ 10,000 - \$19,999; \$ 20,000 - \$29,999;}
   \item{\textbf{2011}:  Income of households in 2010 / Household total income in 2010 of private households / Under \$5,000; Income of households in 2010 / Household total income in 2010 of private households / \$5,000 to \$9,999; Income of households in 2010 / Household total income in 2010 of private households / \$10,000 to \$14,999; Income of households in 2010 / Household total income in 2010 of private households / \$15,000 to \$19,999; Income of households in 2010 / Household total income in 2010 of private households / \$20,000 to \$29,999;}
\end{itemize}

\paragraph{30-90k}
\begin{itemize}
   \item{\textbf{1971}:  Households: income from all sources: \$5,000-\$6,999; Households: income from all sources: \$7,000-\$9,999; Households: income from all sources: \$10,000-\$14,999;}
   \item{\textbf{1981}:  Private household income - all households- \$10,000-\$14,999; Private household income - all households- \$15,000-\$19,999; Private household income - all households- \$20,000-\$24,999; Private household income - all households- \$25,000-\$29,999;}
   \item{\textbf{1991}:  \$20,000 - \$29,999, household income; \$30,000 - \$39,999, household income; \$40,000 - \$49,999, household income; \$50,000 - \$59,999, household income;}
   \item{\textbf{2001}:  \$ 30,000 - \$39,999; \$ 40,000 - \$49,999; \$ 60,000 - \$69,999; \$ 50,000 - \$59,999;}
   \item{\textbf{2011}:  Income of households in 2010 / Household total income in 2010 of private households / \$30,000 to \$39,999; Income of households in 2010 / Household total income in 2010 of private households / \$40,000 to \$49,999; Income of households in 2010 / Household total income in 2010 of private households / \$50,000 to \$59,999; Income of households in 2010 / Household total income in 2010 of private households / \$60,000 to \$79,999;}
\end{itemize}

\paragraph{More than 90k}
\begin{itemize}
   \item{\textbf{1971}:  Households: income from all sources: \$15,000 \& over;}
   \item{\textbf{1981}:  Private household income - all households- \$30,000-\$39,000; Private household income - all households- \$40,000 \& over;}
   \item{\textbf{1991}:  \$60,000 - \$69,999, household income; \$70,000 and over, household income;}
   \item{\textbf{2001}:  \$ 70,000 - \$79,999; \$ 80,000 - \$89,999; \$ 90,000 - \$99,999; \$100,000 and over;}
   \item{\textbf{2011}:  Income of households in 2010 / Household total income in 2010 of private households / \$80,000 to \$99,999; Income of households in 2010 / Household total income in 2010 of private households / \$100,000 to \$124,999; Income of households in 2010 / Household total income in 2010 of private households / \$125,000 to \$149,999; Income of households in 2010 / Household total income in 2010 of private households / \$150,000 and over;}
\end{itemize}

\subsubsection{Marital status}
\paragraph{Single}
\begin{itemize}
   \item{\textbf{1971}:  Marital status: single: total;}
   \item{\textbf{1981}:  Marital status: single, total;}
   \item{\textbf{1991}:  Single (never married) persons 15 years of age and over;}
   \item{\textbf{2001}:  Never legally married (single);}
   \item{\textbf{2011}:  Marital Status - Both sexes / Total population 15 years and over by marital status, Both sexes / Not married and not living with a common-law partner, Both sexes / Single (never legally married), Both sexes;}
\end{itemize}

\paragraph{Married (incl. separated)}
\begin{itemize}
   \item{\textbf{1971}:  Marital status: married: total; Marital status: separated: total;}
   \item{\textbf{1981}:  Marital status: married;}
   \item{\textbf{1991}:  Legally married (and not separated); Legally married and separated;}
   \item{\textbf{2001}:  Legally married (and not separated); Separated, but still legally married;}
   \item{\textbf{2011}:  Marital Status - Both sexes / Total population 15 years and over by marital status, Both sexes / Married or living with a common-law partner, Both sexes; Marital Status - Both sexes / Total population 15 years and over by marital status, Both sexes / Not married and not living with a common-law partner, Both sexes / Separated, Both sexes;}
\end{itemize}

\paragraph{Widowed}
\begin{itemize}
   \item{\textbf{1971}:  Marital status: widowed: total;}
   \item{\textbf{1981}:  Marital status: widowed;}
   \item{\textbf{1991}:  Widowed;}
   \item{\textbf{2001}:  Widowed;}
   \item{\textbf{2011}:  Marital Status - Both sexes / Total population 15 years and over by marital status, Both sexes / Not married and not living with a common-law partner, Both sexes / Widowed, Both sexes;}
\end{itemize}

\paragraph{Divorced}
\begin{itemize}
   \item{\textbf{1971}:  Marital status: divorced: total;}
   \item{\textbf{1981}:  Marital status: divorced;}
   \item{\textbf{1991}:  Divorced;}
   \item{\textbf{2001}:  Divorced;}
   \item{\textbf{2011}:  Marital Status - Both sexes / Total population 15 years and over by marital status, Both sexes / Not married and not living with a common-law partner, Both sexes / Divorced, Both sexes;}
\end{itemize}

\subsubsection{Place of birth}
\paragraph{Canada}
\begin{itemize}
   \item{\textbf{1971}:  Birthplace: born in Canada;}
   \item{\textbf{1981}:  Place of birth - born in Canada;}
   \item{\textbf{1991}:  Non-immigrant population, total;}
   \item{\textbf{2001}:  Non-immigrant population;}
   \item{\textbf{2011}:  Immigrant status and selected places of birth - Both sexes / Total population in private households by immigrant status and selected places of birth, Both sexes / Non-immigrants, Both sexes;}
\end{itemize}

\paragraph{Europe}
\begin{itemize}
   \item{\textbf{1971}:  Birthplace: Belgium \& Luxembourg; Birthplace: France; Birthplace: Germany; Birthplace: Netherlands; Birthplace: other western Europe; Birthplace: Republic of Ireland; Birthplace: United Kingdom; Birthplace: other northern Europe; Birthplace: Greece; Birthplace: Italy; Birthplace: Spain \& Portugal; Birthplace: other southern Europe; Birthplace: Poland; Birthplace: U.S.S.R.; Birthplace: other eastern Europe;}
   \item{\textbf{1981}:  Place of birth - - United Kingdom; Place of birth - - other European;}
   \item{\textbf{1991}:  United Kingdom, place of birth; Other Europe, place of birth;}
   \item{\textbf{2001}:  United Kingdom; Italy; Poland; Germany; Portugal; Netherlands; Greece; France; Yugoslavia; Romania; Ukraine; Hungary; Russian Federation; Croatia; Ireland, Republic of (EIRE); Bosnia and Herzegovina; Austria; Switzerland; Belgium;}
   \item{\textbf{2011}:  Immigrant status and selected places of birth - Both sexes / Total population in private households by immigrant status and selected places of birth, Both sexes / Immigrants, Both sexes / Europe, Both sexes;}
\end{itemize}

\paragraph{Asia}
\begin{itemize}
   \item{\textbf{1971}:  Birthplace: China; Birthplace: India \& Pakistan; Birthplace: other Asia;}
   \item{\textbf{1981}:  Place of birth - - Asia;}
   \item{\textbf{1991}:  India, place of birth; Other Asia, place of birth;}
   \item{\textbf{2001}:  China, Peoples Republic of; India; Hong Kong, Special Administrative Region; Philippines; Viet Nam; Sri Lanka; Pakistan; Iran; Korea, South; Lebanon; Taiwan; Iraq; Afghanistan; Bangladesh; Malaysia; Cambodia;}
   \item{\textbf{2011}:  Immigrant status and selected places of birth - Both sexes / Total population in private households by immigrant status and selected places of birth, Both sexes / Immigrants, Both sexes / Asia, Both sexes;}
\end{itemize}

\paragraph{Africa}
\begin{itemize}
   \item{\textbf{1971}:  Birthplace: Africa;}
   \item{\textbf{1981}: }
   \item{\textbf{1991}:  Africa, place of birth;}
   \item{\textbf{2001}:  Egypt; South Africa, Republic of; Morocco; Kenya; Tanzania, United Republic of; Algeria;}
   \item{\textbf{2011}:  Immigrant status and selected places of birth - Both sexes / Total population in private households by immigrant status and selected places of birth, Both sexes / Immigrants, Both sexes / Africa, Both sexes;}
\end{itemize}

\paragraph{Americas}
\begin{itemize}
   \item{\textbf{1971}:  Birthplace: United States; Birthplace: West Indies \& Latin America;}
   \item{\textbf{1981}:  Place of birth - - United States of America; Place of birth - - other America;}
   \item{\textbf{1991}:  United States of America, place of birth; Central and South America, place of birth; Caribbean and Bermuda, place of birth;}
   \item{\textbf{2001}:  United States; Jamaica; Guyana; Trinidad and Tobago; Haiti; El Salvador; Mexico; Chile;}
   \item{\textbf{2011}:  Immigrant status and selected places of birth - Both sexes / Total population in private households by immigrant status and selected places of birth, Both sexes / Immigrants, Both sexes / Americas, Both sexes;}
\end{itemize}

\paragraph{Other}
\begin{itemize}
   \item{\textbf{1971}:  Birthplace: all other;}
   \item{\textbf{1981}: }
   \item{\textbf{1991}:  Oceania \& Other, place of birth;}
   \item{\textbf{2001}:  Fiji; All other places of birth;}
   \item{\textbf{2011}:  Immigrant status and selected places of birth - Both sexes / Total population in private households by immigrant status and selected places of birth, Both sexes / Immigrants, Both sexes / Oceania and other, Both sexes;}
\end{itemize}

\subsection{USA}
\subsubsection{Population}
\paragraph{Total}
\begin{itemize}
   \item{\textbf{1970}:  CEB001: Male - White; CEB010: Female - White; CEB002: Male - Negro; CEB011: Female - Negro; CEB003: Male - Indian; CEB012: Female - Indian; CEB004: Male - Japanese; CEB013: Female - Japanese; CEB005: Male - Chinese; CEB014: Female - Chinese; CEB006: Male - Filipino; CEB015: Female - Filipino; CEB007: Male - Hawaiian; CEB016: Female - Hawaiian; CEB008: Male - Korean; CEB017: Female - Korean; CEB009: Male - Other; CEB018: Female - Other;}
   \item{\textbf{1980}:  C9D001: White; C9D002: Black; C9D003: American Indian, Eskimo, and Aleut: American Indian; C9D004: American Indian, Eskimo, and Aleut: Eskimo; C9D005: American Indian, Eskimo, and Aleut: Aleut; C9D006: Asian and Pacific Islander: Japanese; C9D007: Asian and Pacific Islander: Chinese; C9D008: Asian and Pacific Islander: Filipino; C9D009: Asian and Pacific Islander: Korean; C9D010: Asian and Pacific Islander: Asian Indian; C9D011: Asian and Pacific Islander: Vietnamese; C9D012: Asian and Pacific Islander: Hawaiian; C9D013: Asian and Pacific Islander: Guamanian; C9D014: Asian and Pacific Islander: Samoan; C9D015: Other;}
   \item{\textbf{1990}:  EUZ001: White (800-869, 971); EUZ002: Black (870-934, 972); EUZ003: American Indian, Eskimo, or Aleut (000-599, 935-970, 973-975): American Indian (000-599, 973); EUZ004: American Indian, Eskimo, or Aleut (000-599, 935-970, 973-975): Eskimo (935-940, 974); EUZ005: American Indian, Eskimo, or Aleut (000-599, 935-970, 973-975): Aleut (941-970, 975); EUZ006: Asian or Pacific Islander (600-699, 976-985): Asian (600-652, 976, 977, 979-982, 985): Chinese (605-607, 976); EUZ007: Asian or Pacific Islander (600-699, 976-985): Asian (600-652, 976, 977, 979-982, 985): Filipino (608, 977); EUZ008: Asian or Pacific Islander (600-699, 976-985): Asian (600-652, 976, 977, 979-982, 985): Japanese (611, 981); EUZ009: Asian or Pacific Islander (600-699, 976-985): Asian (600-652, 976, 977, 979-982, 985): Asian Indian (600, 982); EUZ010: Asian or Pacific Islander (600-699, 976-985): Asian (600-652, 976, 977, 979-982, 985): Korean (612, 979); EUZ011: Asian or Pacific Islander (600-699, 976-985): Asian (600-652, 976, 977, 979-982, 985): Vietnamese (619, 980); EUZ012: Asian or Pacific Islander (600-699, 976-985): Asian (600-652, 976, 977, 979-982, 985): Cambodian (604); EUZ013: Asian or Pacific Islander (600-699, 976-985): Asian (600-652, 976, 977, 979-982, 985): Hmong (609); EUZ014: Asian or Pacific Islander (600-699, 976-985): Asian (600-652, 976, 977, 979-982, 985): Laotian (613); EUZ015: Asian or Pacific Islander (600-699, 976-985): Asian (600-652, 976, 977, 979-982, 985): Thai (618); EUZ016: Asian or Pacific Islander (600-699, 976-985): Asian (600-652, 976, 977, 979-982, 985): Other Asian (601-603, 610, 614-617, 620-652, 985); EUZ017: Asian or Pacific Islander (600-699, 976-985): Pacific Islander (653-699, 978, 983, 984): Polynesian (653-659, 978, 983): Hawaiian (653, 654, 978); EUZ018: Asian or Pacific Islander (600-699, 976-985): Pacific Islander (653-699, 978, 983, 984): Polynesian (653-659, 978, 983): Samoan (655, 983); EUZ019: Asian or Pacific Islander (600-699, 976-985): Pacific Islander (653-699, 978, 983, 984): Polynesian (653-659, 978, 983): Tongan (657); EUZ020: Asian or Pacific Islander (600-699, 976-985): Pacific Islander (653-699, 978, 983, 984): Polynesian (653-659, 978, 983): Other Polynesian (656, 658, 659); EUZ021: Asian or Pacific Islander (600-699, 976-985): Pacific Islander (653-699, 978, 983, 984): Micronesian (660-675, 984): Guamanian (660, 984); EUZ022: Asian or Pacific Islander (600-699, 976-985): Pacific Islander (653-699, 978, 983, 984): Micronesian (660-675, 984): Other Micronesian (661-675); EUZ023: Asian or Pacific Islander (600-699, 976-985): Pacific Islander (653-699, 978, 983, 984): Micronesian (660-675, 984): Melanesian (676-680); EUZ024: Asian or Pacific Islander (600-699, 976-985): Pacific Islander (653-699, 978, 983, 984): Micronesian (660-675, 984): Pacific Islander, not specified (681-699); EUZ025: Asian or Pacific Islander (600-699, 976-985): Pacific Islander (653-699, 978, 983, 984): Micronesian (660-675, 984): Other race (700-799, 986-999);}
   \item{\textbf{2000}:  FL9003: American Indian and Alaska Native alone; FL9002: Black or African American alone; FL9001: White alone; FL9004: Asian alone; FL9005: Native Hawaiian and Other Pacific Islander alone; FL9006: Some other race alone;}
   \item{\textbf{2010}:  JMCE003: Population of one race: White; JMCE004: Population of one race: Black or African American; JMCE005: Population of one race: American Indian and Alaska Native; JMCE006: Population of one race: Asian alone; JMCE007: Population of one race: Native Hawaiian and Other Pacific Islander; JMCE008: Population of one race: Some other race;}
\end{itemize}

\subsubsection{Race}
\paragraph{White}
\begin{itemize}
   \item{\textbf{1970}:  CEB001: Male - White; CEB010: Female - White;}
   \item{\textbf{1980}:  C9D001: White;}
   \item{\textbf{1990}:  ET4001: White - Male - Under 1 year; ET4002: White - Male - 1 and 2 years; ET4003: White - Male - 3 and 4 years; ET4004: White - Male - 5 years; ET4005: White - Male - 6 years; ET4006: White - Male - 7 to 9 years; ET4007: White - Male - 10 and 11 years; ET4008: White - Male - 12 and 13 years; ET4009: White - Male - 14 years; ET4010: White - Male - 15 years; ET4011: White - Male - 16 years; ET4012: White - Male - 17 years; ET4013: White - Male - 18 years; ET4014: White - Male - 19 years; ET4015: White - Male - 20 years; ET4016: White - Male - 21 years; ET4017: White - Male - 22 to 24 years; ET4018: White - Male - 25 to 29 years; ET4019: White - Male - 30 to 34 years; ET4020: White - Male - 35 to 39 years; ET4021: White - Male - 40 to 44 years; ET4022: White - Male - 45 to 49 years; ET4023: White - Male - 50 to 54 years; ET4024: White - Male - 55 to 59 years; ET4025: White - Male - 60 and 61 years; ET4026: White - Male - 62 to 64 years; ET4027: White - Male - 65 to 69 years; ET4028: White - Male - 70 to 74 years; ET4029: White - Male - 75 to 79 years; ET4030: White - Male - 80 to 84 years; ET4031: White - Male - 85 years and over; ET4032: White - Female - Under 1 year; ET4033: White - Female - 1 and 2 years; ET4034: White - Female - 3 and 4 years; ET4035: White - Female - 5 years; ET4036: White - Female - 6 years; ET4037: White - Female - 7 to 9 years; ET4038: White - Female - 10 and 11 years; ET4039: White - Female - 12 and 13 years; ET4040: White - Female - 14 years; ET4041: White - Female - 15 years; ET4042: White - Female - 16 years; ET4043: White - Female - 17 years; ET4044: White - Female - 18 years; ET4045: White - Female - 19 years; ET4046: White - Female - 20 years; ET4047: White - Female - 21 years; ET4048: White - Female - 22 to 24 years; ET4049: White - Female - 25 to 29 years; ET4050: White - Female - 30 to 34 years; ET4051: White - Female - 35 to 39 years; ET4052: White - Female - 40 to 44 years; ET4053: White - Female - 45 to 49 years; ET4054: White - Female - 50 to 54 years; ET4055: White - Female - 55 to 59 years; ET4056: White - Female - 60 and 61 years; ET4057: White - Female - 62 to 64 years; ET4058: White - Female - 65 to 69 years; ET4059: White - Female - 70 to 74 years; ET4060: White - Female - 75 to 79 years; ET4061: White - Female - 80 to 84 years; ET4062: White - Female - 85 years and over;}
   \item{\textbf{2000}:  FL9001: White alone;}
   \item{\textbf{2010}:  JMCE003: Population of one race: White;}
\end{itemize}

\paragraph{Black}
\begin{itemize}
   \item{\textbf{1970}:  CEB002: Male - Negro; CEB011: Female - Negro;}
   \item{\textbf{1980}:  C9D002: Black;}
   \item{\textbf{1990}:  ET4063: Black - Male - Under 1 year; ET4064: Black - Male - 1 and 2 years; ET4065: Black - Male - 3 and 4 years; ET4066: Black - Male - 5 years; ET4067: Black - Male - 6 years; ET4068: Black - Male - 7 to 9 years; ET4069: Black - Male - 10 and 11 years; ET4070: Black - Male - 12 and 13 years; ET4071: Black - Male - 14 years; ET4072: Black - Male - 15 years; ET4073: Black - Male - 16 years; ET4074: Black - Male - 17 years; ET4075: Black - Male - 18 years; ET4076: Black - Male - 19 years; ET4077: Black - Male - 20 years; ET4078: Black - Male - 21 years; ET4079: Black - Male - 22 to 24 years; ET4080: Black - Male - 25 to 29 years; ET4081: Black - Male - 30 to 34 years; ET4082: Black - Male - 35 to 39 years; ET4083: Black - Male - 40 to 44 years; ET4084: Black - Male - 45 to 49 years; ET4085: Black - Male - 50 to 54 years; ET4086: Black - Male - 55 to 59 years; ET4087: Black - Male - 60 and 61 years; ET4088: Black - Male - 62 to 64 years; ET4089: Black - Male - 65 to 69 years; ET4090: Black - Male - 70 to 74 years; ET4091: Black - Male - 75 to 79 years; ET4092: Black - Male - 80 to 84 years; ET4093: Black - Male - 85 years and over; ET4094: Black - Female - Under 1 year; ET4095: Black - Female - 1 and 2 years; ET4096: Black - Female - 3 and 4 years; ET4097: Black - Female - 5 years; ET4098: Black - Female - 6 years; ET4099: Black - Female - 7 to 9 years; ET4100: Black - Female - 10 and 11 years; ET4101: Black - Female - 12 and 13 years; ET4102: Black - Female - 14 years; ET4103: Black - Female - 15 years; ET4104: Black - Female - 16 years; ET4105: Black - Female - 17 years; ET4106: Black - Female - 18 years; ET4107: Black - Female - 19 years; ET4108: Black - Female - 20 years; ET4109: Black - Female - 21 years; ET4110: Black - Female - 22 to 24 years; ET4111: Black - Female - 25 to 29 years; ET4112: Black - Female - 30 to 34 years; ET4113: Black - Female - 35 to 39 years; ET4114: Black - Female - 40 to 44 years; ET4115: Black - Female - 45 to 49 years; ET4116: Black - Female - 50 to 54 years; ET4117: Black - Female - 55 to 59 years; ET4118: Black - Female - 60 and 61 years; ET4119: Black - Female - 62 to 64 years; ET4120: Black - Female - 65 to 69 years; ET4121: Black - Female - 70 to 74 years; ET4122: Black - Female - 75 to 79 years; ET4123: Black - Female - 80 to 84 years; ET4124: Black - Female - 85 years and over;}
   \item{\textbf{2000}:  FL9002: Black or African American alone;}
   \item{\textbf{2010}:  JMCE004: Population of one race: Black or African American;}
\end{itemize}

\paragraph{Indian, eskimo and aleut}
\begin{itemize}
   \item{\textbf{1970}:  CEB003: Male - Indian; CEB012: Female - Indian;}
   \item{\textbf{1980}:  C9D003: American Indian, Eskimo, and Aleut: American Indian; C9D004: American Indian, Eskimo, and Aleut: Eskimo; C9D005: American Indian, Eskimo, and Aleut: Aleut;}
   \item{\textbf{1990}:  ET4125: American Indian, Eskimo, or Aleut - Male - Under 1 year; ET4126: American Indian, Eskimo, or Aleut - Male - 1 and 2 years; ET4127: American Indian, Eskimo, or Aleut - Male - 3 and 4 years; ET4128: American Indian, Eskimo, or Aleut - Male - 5 years; ET4129: American Indian, Eskimo, or Aleut - Male - 6 years; ET4130: American Indian, Eskimo, or Aleut - Male - 7 to 9 years; ET4131: American Indian, Eskimo, or Aleut - Male - 10 and 11 years; ET4132: American Indian, Eskimo, or Aleut - Male - 12 and 13 years; ET4133: American Indian, Eskimo, or Aleut - Male - 14 years; ET4134: American Indian, Eskimo, or Aleut - Male - 15 years; ET4135: American Indian, Eskimo, or Aleut - Male - 16 years; ET4136: American Indian, Eskimo, or Aleut - Male - 17 years; ET4137: American Indian, Eskimo, or Aleut - Male - 18 years; ET4138: American Indian, Eskimo, or Aleut - Male - 19 years; ET4139: American Indian, Eskimo, or Aleut - Male - 20 years; ET4140: American Indian, Eskimo, or Aleut - Male - 21 years; ET4141: American Indian, Eskimo, or Aleut - Male - 22 to 24 years; ET4142: American Indian, Eskimo, or Aleut - Male - 25 to 29 years; ET4143: American Indian, Eskimo, or Aleut - Male - 30 to 34 years; ET4144: American Indian, Eskimo, or Aleut - Male - 35 to 39 years; ET4145: American Indian, Eskimo, or Aleut - Male - 40 to 44 years; ET4146: American Indian, Eskimo, or Aleut - Male - 45 to 49 years; ET4147: American Indian, Eskimo, or Aleut - Male - 50 to 54 years; ET4148: American Indian, Eskimo, or Aleut - Male - 55 to 59 years; ET4149: American Indian, Eskimo, or Aleut - Male - 60 and 61 years; ET4150: American Indian, Eskimo, or Aleut - Male - 62 to 64 years; ET4151: American Indian, Eskimo, or Aleut - Male - 65 to 69 years; ET4152: American Indian, Eskimo, or Aleut - Male - 70 to 74 years; ET4153: American Indian, Eskimo, or Aleut - Male - 75 to 79 years; ET4154: American Indian, Eskimo, or Aleut - Male - 80 to 84 years; ET4155: American Indian, Eskimo, or Aleut - Male - 85 years and over; ET4156: American Indian, Eskimo, or Aleut - Female - Under 1 year; ET4157: American Indian, Eskimo, or Aleut - Female - 1 and 2 years; ET4158: American Indian, Eskimo, or Aleut - Female - 3 and 4 years; ET4159: American Indian, Eskimo, or Aleut - Female - 5 years; ET4160: American Indian, Eskimo, or Aleut - Female - 6 years; ET4161: American Indian, Eskimo, or Aleut - Female - 7 to 9 years; ET4162: American Indian, Eskimo, or Aleut - Female - 10 and 11 years; ET4163: American Indian, Eskimo, or Aleut - Female - 12 and 13 years; ET4164: American Indian, Eskimo, or Aleut - Female - 14 years; ET4165: American Indian, Eskimo, or Aleut - Female - 15 years; ET4166: American Indian, Eskimo, or Aleut - Female - 16 years; ET4167: American Indian, Eskimo, or Aleut - Female - 17 years; ET4168: American Indian, Eskimo, or Aleut - Female - 18 years; ET4169: American Indian, Eskimo, or Aleut - Female - 19 years; ET4170: American Indian, Eskimo, or Aleut - Female - 20 years; ET4171: American Indian, Eskimo, or Aleut - Female - 21 years; ET4172: American Indian, Eskimo, or Aleut - Female - 22 to 24 years; ET4173: American Indian, Eskimo, or Aleut - Female - 25 to 29 years; ET4174: American Indian, Eskimo, or Aleut - Female - 30 to 34 years; ET4175: American Indian, Eskimo, or Aleut - Female - 35 to 39 years; ET4176: American Indian, Eskimo, or Aleut - Female - 40 to 44 years; ET4177: American Indian, Eskimo, or Aleut - Female - 45 to 49 years; ET4178: American Indian, Eskimo, or Aleut - Female - 50 to 54 years; ET4179: American Indian, Eskimo, or Aleut - Female - 55 to 59 years; ET4180: American Indian, Eskimo, or Aleut - Female - 60 and 61 years; ET4181: American Indian, Eskimo, or Aleut - Female - 62 to 64 years; ET4182: American Indian, Eskimo, or Aleut - Female - 65 to 69 years; ET4183: American Indian, Eskimo, or Aleut - Female - 70 to 74 years; ET4184: American Indian, Eskimo, or Aleut - Female - 75 to 79 years; ET4185: American Indian, Eskimo, or Aleut - Female - 80 to 84 years; ET4186: American Indian, Eskimo, or Aleut - Female - 85 years and over;}
   \item{\textbf{2000}:  FL9003: American Indian and Alaska Native alone;}
   \item{\textbf{2010}:  JMCE005: Population of one race: American Indian and Alaska Native;}
\end{itemize}

\paragraph{Asian, Hawaiian, other pacific islander}
\begin{itemize}
   \item{\textbf{1970}:  CEB004: Male - Japanese; CEB005: Male - Chinese; CEB006: Male - Filipino; CEB007: Male - Hawaiian; CEB008: Male - Korean; CEB013: Female - Japanese; CEB014: Female - Chinese; CEB015: Female - Filipino; CEB016: Female - Hawaiian; CEB017: Female - Korean;}
   \item{\textbf{1980}:  C9D006: Asian and Pacific Islander: Japanese; C9D007: Asian and Pacific Islander: Chinese; C9D008: Asian and Pacific Islander: Filipino; C9D009: Asian and Pacific Islander: Korean; C9D010: Asian and Pacific Islander: Asian Indian; C9D011: Asian and Pacific Islander: Vietnamese; C9D012: Asian and Pacific Islander: Hawaiian; C9D013: Asian and Pacific Islander: Guamanian; C9D014: Asian and Pacific Islander: Samoan;}
   \item{\textbf{1990}:  ET4187: Asian or Pacific Islander - Male - Under 1 year; ET4188: Asian or Pacific Islander - Male - 1 and 2 years; ET4189: Asian or Pacific Islander - Male - 3 and 4 years; ET4190: Asian or Pacific Islander - Male - 5 years; ET4191: Asian or Pacific Islander - Male - 6 years; ET4192: Asian or Pacific Islander - Male - 7 to 9 years; ET4193: Asian or Pacific Islander - Male - 10 and 11 years; ET4194: Asian or Pacific Islander - Male - 12 and 13 years; ET4195: Asian or Pacific Islander - Male - 14 years; ET4196: Asian or Pacific Islander - Male - 15 years; ET4197: Asian or Pacific Islander - Male - 16 years; ET4198: Asian or Pacific Islander - Male - 17 years; ET4199: Asian or Pacific Islander - Male - 18 years; ET4200: Asian or Pacific Islander - Male - 19 years; ET4201: Asian or Pacific Islander - Male - 20 years; ET4202: Asian or Pacific Islander - Male - 21 years; ET4203: Asian or Pacific Islander - Male - 22 to 24 years; ET4204: Asian or Pacific Islander - Male - 25 to 29 years; ET4205: Asian or Pacific Islander - Male - 30 to 34 years; ET4206: Asian or Pacific Islander - Male - 35 to 39 years; ET4207: Asian or Pacific Islander - Male - 40 to 44 years; ET4208: Asian or Pacific Islander - Male - 45 to 49 years; ET4209: Asian or Pacific Islander - Male - 50 to 54 years; ET4210: Asian or Pacific Islander - Male - 55 to 59 years; ET4211: Asian or Pacific Islander - Male - 60 and 61 years; ET4212: Asian or Pacific Islander - Male - 62 to 64 years; ET4213: Asian or Pacific Islander - Male - 65 to 69 years; ET4214: Asian or Pacific Islander - Male - 70 to 74 years; ET4215: Asian or Pacific Islander - Male - 75 to 79 years; ET4216: Asian or Pacific Islander - Male - 80 to 84 years; ET4217: Asian or Pacific Islander - Male - 85 years and over; ET4218: Asian or Pacific Islander - Female - Under 1 year; ET4219: Asian or Pacific Islander - Female - 1 and 2 years; ET4220: Asian or Pacific Islander - Female - 3 and 4 years; ET4221: Asian or Pacific Islander - Female - 5 years; ET4222: Asian or Pacific Islander - Female - 6 years; ET4223: Asian or Pacific Islander - Female - 7 to 9 years; ET4224: Asian or Pacific Islander - Female - 10 and 11 years; ET4225: Asian or Pacific Islander - Female - 12 and 13 years; ET4226: Asian or Pacific Islander - Female - 14 years; ET4227: Asian or Pacific Islander - Female - 15 years; ET4228: Asian or Pacific Islander - Female - 16 years; ET4229: Asian or Pacific Islander - Female - 17 years; ET4230: Asian or Pacific Islander - Female - 18 years; ET4231: Asian or Pacific Islander - Female - 19 years; ET4232: Asian or Pacific Islander - Female - 20 years; ET4233: Asian or Pacific Islander - Female - 21 years; ET4234: Asian or Pacific Islander - Female - 22 to 24 years; ET4235: Asian or Pacific Islander - Female - 25 to 29 years; ET4236: Asian or Pacific Islander - Female - 30 to 34 years; ET4237: Asian or Pacific Islander - Female - 35 to 39 years; ET4238: Asian or Pacific Islander - Female - 40 to 44 years; ET4239: Asian or Pacific Islander - Female - 45 to 49 years; ET4240: Asian or Pacific Islander - Female - 50 to 54 years; ET4241: Asian or Pacific Islander - Female - 55 to 59 years; ET4242: Asian or Pacific Islander - Female - 60 and 61 years; ET4243: Asian or Pacific Islander - Female - 62 to 64 years; ET4244: Asian or Pacific Islander - Female - 65 to 69 years; ET4245: Asian or Pacific Islander - Female - 70 to 74 years; ET4246: Asian or Pacific Islander - Female - 75 to 79 years; ET4247: Asian or Pacific Islander - Female - 80 to 84 years; ET4248: Asian or Pacific Islander - Female - 85 years and over;}
   \item{\textbf{2000}:  FL9004: Asian alone; FL9005: Native Hawaiian and Other Pacific Islander alone;}
   \item{\textbf{2010}:  JMCE006: Population of one race: Asian alone; JMCE007: Population of one race: Native Hawaiian and Other Pacific Islander;}
\end{itemize}

\paragraph{Other}
\begin{itemize}
   \item{\textbf{1970}:  CEB009: Male - Other; CEB018: Female - Other;}
   \item{\textbf{1980}:  C9D015: Other;}
   \item{\textbf{1990}:  ET4249: Other race - Male - Under 1 year; ET4250: Other race - Male - 1 and 2 years; ET4251: Other race - Male - 3 and 4 years; ET4252: Other race - Male - 5 years; ET4253: Other race - Male - 6 years; ET4254: Other race - Male - 7 to 9 years; ET4255: Other race - Male - 10 and 11 years; ET4256: Other race - Male - 12 and 13 years; ET4257: Other race - Male - 14 years; ET4258: Other race - Male - 15 years; ET4259: Other race - Male - 16 years; ET4260: Other race - Male - 17 years; ET4261: Other race - Male - 18 years; ET4262: Other race - Male - 19 years; ET4263: Other race - Male - 20 years; ET4264: Other race - Male - 21 years; ET4265: Other race - Male - 22 to 24 years; ET4266: Other race - Male - 25 to 29 years; ET4267: Other race - Male - 30 to 34 years; ET4268: Other race - Male - 35 to 39 years; ET4269: Other race - Male - 40 to 44 years; ET4270: Other race - Male - 45 to 49 years; ET4271: Other race - Male - 50 to 54 years; ET4272: Other race - Male - 55 to 59 years; ET4273: Other race - Male - 60 and 61 years; ET4274: Other race - Male - 62 to 64 years; ET4275: Other race - Male - 65 to 69 years; ET4276: Other race - Male - 70 to 74 years; ET4277: Other race - Male - 75 to 79 years; ET4278: Other race - Male - 80 to 84 years; ET4279: Other race - Male - 85 years and over; ET4280: Other race - Female - Under 1 year; ET4281: Other race - Female - 1 and 2 years; ET4282: Other race - Female - 3 and 4 years; ET4283: Other race - Female - 5 years; ET4284: Other race - Female - 6 years; ET4285: Other race - Female - 7 to 9 years; ET4286: Other race - Female - 10 and 11 years; ET4287: Other race - Female - 12 and 13 years; ET4288: Other race - Female - 14 years; ET4289: Other race - Female - 15 years; ET4290: Other race - Female - 16 years; ET4291: Other race - Female - 17 years; ET4292: Other race - Female - 18 years; ET4293: Other race - Female - 19 years; ET4294: Other race - Female - 20 years; ET4295: Other race - Female - 21 years; ET4296: Other race - Female - 22 to 24 years; ET4297: Other race - Female - 25 to 29 years; ET4298: Other race - Female - 30 to 34 years; ET4299: Other race - Female - 35 to 39 years; ET4300: Other race - Female - 40 to 44 years; ET4301: Other race - Female - 45 to 49 years; ET4302: Other race - Female - 50 to 54 years; ET4303: Other race - Female - 55 to 59 years; ET4304: Other race - Female - 60 and 61 years; ET4305: Other race - Female - 62 to 64 years; ET4306: Other race - Female - 65 to 69 years; ET4307: Other race - Female - 70 to 74 years; ET4308: Other race - Female - 75 to 79 years; ET4309: Other race - Female - 80 to 84 years; ET4310: Other race - Female - 85 years and over;}
   \item{\textbf{2000}:  FL9006: Some other race alone;}
   \item{\textbf{2010}:  JMCE008: Population of one race: Some other race;}
\end{itemize}

\subsubsection{Marital status}
\paragraph{Married (inc. Sep.)}
\begin{itemize}
   \item{\textbf{1970}:  CII001: Male - 14-19 years - Now married (excludes separated); CII006: Male - 20-24 years - Now married (excludes separated); CII011: Male - 25-34 years - Now married (excludes separated); CII016: Male - 35-44 years - Now married (excludes separated); CII021: Male - 45-54 years - Now married (excludes separated); CII026: Male - 55-64 years - Now married (excludes separated); CII031: Male - 65 years and over - Now married (excludes separated); CII036: Female - 14-19 years - Now married (excludes separated); CII041: Female - 20-24 years - Now married (excludes separated); CII046: Female - 25-34 years - Now married (excludes separated); CII051: Female - 35-44 years - Now married (excludes separated); CII056: Female - 45-54 years - Now married (excludes separated); CII061: Female - 55-64 years - Now married (excludes separated); CII066: Female - 65 years and over - Now married (excludes separated); CII004: Male - 14-19 years - Separated; CII009: Male - 20-24 years - Separated; CII014: Male - 25-34 years - Separated; CII019: Male - 35-44 years - Separated; CII024: Male - 45-54 years - Separated; CII029: Male - 55-64 years - Separated; CII034: Male - 65 years and over - Separated; CII039: Female - 14-19 years - Separated; CII044: Female - 20-24 years - Separated; CII049: Female - 25-34 years - Separated; CII054: Female - 35-44 years - Separated; CII059: Female - 45-54 years - Separated; CII064: Female - 55-64 years - Separated; CII069: Female - 65 years and over - Separated;}
   \item{\textbf{1980}:  C7F002: Male - Now married, except separated; C7F007: Female - Now married, except separated; C7F003: Male - Separated; C7F008: Female - Separated;}
   \item{\textbf{1990}:  ET6002: Male - Now married, except separated; ET6007: Female - Now married, except separated; ET6003: Male - Separated; ET6008: Female - Separated;}
   \item{\textbf{2000}:  GIZ002: Male - Now married; GIZ006: Female - Now married;}
   \item{\textbf{2010}:  J0WE018: Male: Now married; J0WE111: Female: Now married;}
\end{itemize}

\paragraph{Widowed}
\begin{itemize}
   \item{\textbf{1970}:  CII002: Male - 14-19 years - Widowed; CII007: Male - 20-24 years - Widowed; CII012: Male - 25-34 years - Widowed; CII017: Male - 35-44 years - Widowed; CII022: Male - 45-54 years - Widowed; CII027: Male - 55-64 years - Widowed; CII032: Male - 65 years and over - Widowed; CII037: Female - 14-19 years - Widowed; CII042: Female - 20-24 years - Widowed; CII047: Female - 25-34 years - Widowed; CII052: Female - 35-44 years - Widowed; CII057: Female - 45-54 years - Widowed; CII062: Female - 55-64 years - Widowed; CII067: Female - 65 years and over - Widowed;}
   \item{\textbf{1980}:  C7F004: Male - Widowed; C7F009: Female - Widowed;}
   \item{\textbf{1990}:  ET6004: Male - Widowed; ET6009: Female - Widowed;}
   \item{\textbf{2000}:  GIZ003: Male - Widowed; GIZ007: Female - Widowed;}
   \item{\textbf{2010}:  J0WE065: Male: Widowed; J0WE158: Female: Widowed;}
\end{itemize}

\paragraph{Divorced}
\begin{itemize}
   \item{\textbf{1970}:  CII003: Male - 14-19 years - Divorced; CII008: Male - 20-24 years - Divorced; CII013: Male - 25-34 years - Divorced; CII018: Male - 35-44 years - Divorced; CII023: Male - 45-54 years - Divorced; CII028: Male - 55-64 years - Divorced; CII033: Male - 65 years and over - Divorced; CII038: Female - 14-19 years - Divorced; CII043: Female - 20-24 years - Divorced; CII048: Female - 25-34 years - Divorced; CII053: Female - 35-44 years - Divorced; CII058: Female - 45-54 years - Divorced; CII063: Female - 55-64 years - Divorced; CII068: Female - 65 years and over - Divorced;}
   \item{\textbf{1980}:  C7F005: Male - Divorced; C7F010: Female - Divorced;}
   \item{\textbf{1990}:  ET6005: Male - Divorced; ET6010: Female - Divorced;}
   \item{\textbf{2000}:  GIZ004: Male - Divorced; GIZ008: Female - Divorced;}
   \item{\textbf{2010}:  J0WE080: Male: Divorced; J0WE173: Female: Divorced;}
\end{itemize}

\paragraph{Never married}
\begin{itemize}
   \item{\textbf{1970}:  CII005: Male - 14-19 years - Never married; CII010: Male - 20-24 years - Never married; CII015: Male - 25-34 years - Never married; CII020: Male - 35-44 years - Never married; CII025: Male - 45-54 years - Never married; CII030: Male - 55-64 years - Never married; CII035: Male - 65 years and over - Never married; CII040: Female - 14-19 years - Never married; CII045: Female - 20-24 years - Never married; CII050: Female - 25-34 years - Never married; CII055: Female - 35-44 years - Never married; CII060: Female - 45-54 years - Never married; CII065: Female - 55-64 years - Never married; CII070: Female - 65 years and over - Never married;}
   \item{\textbf{1980}:  C7F001: Male - Single; C7F006: Female - Single;}
   \item{\textbf{1990}:  ET6001: Male - Never married; ET6006: Female - Never married;}
   \item{\textbf{2000}:  GIZ001: Male - Never married; GIZ005: Female - Never married;}
   \item{\textbf{2010}:  J0WE003: Male: Never married; J0WE096: Female: Never married;}
\end{itemize}

\subsubsection{Education level}
\paragraph{Not completed High School}
\begin{itemize}
   \item{\textbf{1970}:  C06001: No school years completed (includes nursery and kindergarten); C06002: Elementary: 1-4 years; C06003: Elementary: 5-6 years; C06004: Elementary: 7 years; C06005: Elementary: 8 years; C06006: High school: 1-3 years;}
   \item{\textbf{1980}:  DHS001: Elementary (0-8 years) through high school (1-3 years);}
   \item{\textbf{1990}:  E33001: Less than 9th grade; E33002: 9th to 12th grade, no diploma;}
   \item{\textbf{2000}:  GKT001: Male - No schooling completed; GKT002: Male - Nursery to 4th grade; GKT003: Male - 5th and 6th grade; GKT004: Male - 7th and 8th grade; GKT005: Male - 9th grade; GKT006: Male - 10th grade; GKT007: Male - 11th grade; GKT008: Male - 12th grade, no diploma; GKT017: Female - No schooling completed; GKT018: Female - Nursery to 4th grade; GKT019: Female - 5th and 6th grade; GKT020: Female - 7th and 8th grade; GKT021: Female - 9th grade; GKT022: Female - 10th grade; GKT023: Female - 11th grade; GKT024: Female - 12th grade, no diploma;}
   \item{\textbf{2010}:  JN9E003: Male: No schooling completed; JN9E004: Male: Nursery to 4th grade; JN9E005: Male: 5th and 6th grade; JN9E006: Male: 7th and 8th grade; JN9E007: Male: 9th grade; JN9E008: Male: 10th grade; JN9E009: Male: 11th grade; JN9E010: Male: 12th grade, no diploma; JN9E020: Female: No schooling completed; JN9E021: Female: Nursery to 4th grade; JN9E022: Female: 5th and 6th grade; JN9E023: Female: 7th and 8th grade; JN9E024: Female: 9th grade; JN9E025: Female: 10th grade; JN9E026: Female: 11th grade; JN9E027: Female: 12th grade, no diploma;}
\end{itemize}

\paragraph{High School}
\begin{itemize}
   \item{\textbf{1970}:  C06007: High school: 4 years;}
   \item{\textbf{1980}:  DHS002: High school, 4 years;}
   \item{\textbf{1990}:  E33003: High school graduate (includes equivalency);}
   \item{\textbf{2000}:  GKT009: Male - High school graduate (includes equivalency); GKT025: Female - High school graduate (includes equivalency);}
   \item{\textbf{2010}:  JN9E011: Male: High school graduate, GED, or alternative; JN9E028: Female: High school graduate, GED, or alternative;}
\end{itemize}

\paragraph{College 1-3 years}
\begin{itemize}
   \item{\textbf{1970}:  C06008: College: 1-3 years;}
   \item{\textbf{1980}:  DHS003: College: 1-3 years;}
   \item{\textbf{1990}:  E33004: Some college, no degree; E33005: Associate degree;}
   \item{\textbf{2000}:  GKT010: Male - Some college, less than 1 year; GKT011: Male - Some college, 1 or more years, no degree; GKT012: Male - Associate degree; GKT026: Female - Some college, less than 1 year; GKT027: Female - Some college, 1 or more years, no degree; GKT028: Female - Associate degree;}
   \item{\textbf{2010}:  JN9E012: Male: Some college, less than 1 year; JN9E013: Male: Some college, 1 or more years, no degree; JN9E014: Male: Associate's degree; JN9E029: Female: Some college, less than 1 year; JN9E030: Female: Some college, 1 or more years, no degree; JN9E031: Female: Associate's degree;}
\end{itemize}

\paragraph{College 4+ years}
\begin{itemize}
   \item{\textbf{1970}:  C06009: College: 4; C06010: College: 5 years or more;}
   \item{\textbf{1980}:  DHS004: College: 4 years; DHS005: College: 5 or more years;}
   \item{\textbf{1990}:  E33006: Bachelor's degree; E33007: Graduate or professional degree;}
   \item{\textbf{2000}:  GKT013: Male - Bachelor's degree; GKT014: Male - Master's degree; GKT015: Male - Professional school degree; GKT016: Male - Doctorate degree; GKT029: Female - Bachelor's degree; GKT030: Female - Master's degree; GKT031: Female - Professional school degree; GKT032: Female - Doctorate degree;}
   \item{\textbf{2010}:  JN9E015: Male: Bachelor's degree; JN9E016: Male: Master's degree; JN9E017: Male: Professional school degree; JN9E018: Male: Doctorate degree; JN9E032: Female: Bachelor's degree; JN9E033: Female: Master's degree; JN9E034: Female: Professional school degree; JN9E035: Female: Doctorate degree;}
\end{itemize}

\subsubsection{Family income (2010)}
\paragraph{Less than 50k}
\begin{itemize}
   \item{\textbf{1970}:  C3T001: Under \$1000 (includes \$1-\$999, none, and loss); C3T002: \$1000-\$1999; C3T003: \$2000-\$2999; C3T004: \$3000-\$3999; C3T005: \$4000-\$4999; C3T006: \$5000-\$5999; C3T007: \$6000-\$6999; C3T008: \$7000-\$7999; C3T009: \$8000-\$8999;}
   \item{\textbf{1980}:  DIK001: Less than \$2,500; DIK002: \$2,500-\$4,999; DIK003: \$5,000-\$7,499; DIK004: \$7,500-\$9,999; DIK005: \$10,000-\$12,499; DIK006: \$12,500-\$14,999; DIK007: \$15,000-\$17,499; DIK008: \$17,500-\$19,999;}
   \item{\textbf{1990}:  E0Q001: Less than \$5,000; E0Q002: \$5,000 to \$9,999; E0Q003: \$10,000 to \$12,499; E0Q004: \$12,500 to \$14,999; E0Q005: \$15,000 to \$17,499; E0Q006: \$17,500 to \$19,999; E0Q007: \$20,000 to \$22,499; E0Q008: \$22,500 to \$24,999; E0Q009: \$25,000 to \$27,499; E0Q010: \$27,500 to \$29,999;}
   \item{\textbf{2000}:  GNN001: Less than \$10,000; GNN002: \$10,000 to \$14,999; GNN003: \$15,000 to \$19,999; GNN004: \$20,000 to \$24,999; GNN005: \$25,000 to \$29,999; GNN006: \$30,000 to \$34,999; GNN007: \$35,000 to \$39,999;}
   \item{\textbf{2010}:  JPNE002: Less than \$10,000; JPNE003: \$10,000 to \$14,999; JPNE004: \$15,000 to \$19,999; JPNE005: \$20,000 to \$24,999; JPNE006: \$25,000 to \$29,999; JPNE007: \$30,000 to \$34,999; JPNE008: \$35,000 to \$39,999; JPNE009: \$40,000 to \$44,999; JPNE010: \$45,000 to \$49,999;}
\end{itemize}

\paragraph{50k-100k}
\begin{itemize}
   \item{\textbf{1970}:  C3T010: \$9000-\$9999; C3T011: \$10000-\$11999; C3T012: \$12000-\$14999;}
   \item{\textbf{1980}:  DIK009: \$20,000-\$22,499; DIK010: \$22,500-\$24,999; DIK011: \$25,000-\$27,499; DIK012: \$27,500-\$29,999; DIK013: \$30,000-\$34,999;}
   \item{\textbf{1990}:  E0Q011: \$30,000 to \$32,499; E0Q012: \$32,500 to \$34,999; E0Q013: \$35,000 to \$37,499; E0Q014: \$37,500 to \$39,999; E0Q015: \$40,000 to \$42,499; E0Q016: \$42,500 to \$44,999; E0Q017: \$45,000 to \$47,499; E0Q018: \$47,500 to \$49,999; E0Q019: \$50,000 to \$54,999; E0Q020: \$55,000 to \$59,999;}
   \item{\textbf{2000}:  GNN008: \$40,000 to \$44,999; GNN009: \$45,000 to \$49,999; GNN010: \$50,000 to \$59,999; GNN011: \$60,000 to \$74,999;}
   \item{\textbf{2010}:  JPNE011: \$50,000 to \$59,999; JPNE012: \$60,000 to \$74,999; JPNE013: \$75,000 to \$99,999;}
\end{itemize}

\paragraph{More than 100k}
\begin{itemize}
   \item{\textbf{1970}:  C3T013: \$15000-\$24999; C3T014: \$25000-\$49999; C3T015: \$50000 and over;}
   \item{\textbf{1980}:  DIK014: \$35,000-\$39,999; DIK015: \$40,000-\$49,999; DIK016: \$50,000-\$74,999; DIK017: \$75,000 or more;}
   \item{\textbf{1990}:  E0Q021: \$60,000 to \$74,999; E0Q022: \$75,000 to \$99,999; E0Q023: \$100,000 to \$124,999; E0Q024: \$125,000 to \$149,999; E0Q025: \$150,000 or more;}
   \item{\textbf{2000}:  GNN012: \$75,000 to \$99,999; GNN013: \$100,000 to \$124,999; GNN014: \$125,000 to \$149,999; GNN015: \$150,000 to \$199,999; GNN016: \$200,000 or more;}
   \item{\textbf{2010}:  JPNE014: \$100,000 to \$124,999; JPNE015: \$125,000 to \$149,999; JPNE016: \$150,000 to \$199,999; JPNE017: \$200,000 or more;}
\end{itemize}

\subsubsection{Occupation}
\paragraph{Administrative}
\begin{itemize}
   \item{\textbf{1970}:  C27001: Professional, technical and kindred workers: Engineers, technical; C27002: Professional, technical and kindred workers: Physicians, dentists, and related practitioners; C27003: Professional, technical and kindred workers: Medical and other health workers, except practitioners; C27004: Professional, technical and kindred workers: Teachers, elementary and secondary schools; C27005: Professional, technical and kindred workers: Technical; C27006: Professional, technical and kindred workers: Other professional workers; C27007: Managers and administrators, except farm: Salaried: Manufacturing; C27008: Managers and administrators, except farm: Salaried: Retail trade; C27009: Managers and administrators, except farm: Salaried: Other industries; C27010: Managers and administrators, except farm: Self-employed: Retail trade; C27011: Managers and administrators, except farm: Self-employed: Other industries; C27015: Clerical and kindred workers: Bookkeepers; C27016: Clerical and kindred workers: Secretaries, stenographers, and typists; C27017: Clerical and kindred workers: Other clerical workers; C27012: Sales workers: Manufacturing and wholesale trade; C27013: Sales workers: Retail trade; C27014: Sales workers: Other sales workers;}
   \item{\textbf{1980}:  DIB001: Managerial and professional speciality occupations: Executive, administrative, and managerial occupations (codes 3-37); DIB002: Managerial and professional speciality occupations: Professional speciality occupations (codes 43-199); DIB003: Technical, sales and administrative support occupations: Technicians and related support occupations (codes 203-235); DIB004: Technical, sales and administrative support occupations: Sales occupations (codes 243-285); DIB005: Technical, sales and administrative support occupations: Administrative support occupations, including clerical (codes 303-389);}
   \item{\textbf{1990}:  E4Q001: Managerial and professional specialty occupations (000-202): Executive, administrative, and managerial occupations (000-042); E4Q002: Managerial and professional specialty occupations (000-202): Professional specialty occupations (043-202); E4Q003: Technical, sales, and administrative support occupations (203-402): Technicians and related support occupations (203-242); E4Q004: Technical, sales, and administrative support occupations (203-402): Sales occupations (243-302); E4Q005: Technical, sales, and administrative support occupations (203-402): Administrative support occupations, including clerical (303-402);}
   \item{\textbf{2000}:  GMJ001: Male - Management, professional, and related occupations; GMJ003: Male - Sales and office occupations; GMJ007: Female - Management, professional, and related occupations; GMJ009: Female - Sales and office occupations;}
   \item{\textbf{2010}:  JQ5E003: Male: Management, business, science, and arts occupations; JQ5E027: Male: Sales and office occupations; JQ5E039: Female: Management, business, science, and arts occupations; JQ5E063: Female: Sales and office occupations;}
\end{itemize}

\paragraph{Laborers}
\begin{itemize}
   \item{\textbf{1970}:  C27018: Craftsmen, foremen, and kindred workers: Automobile mechanics and body repairmen; C27019: Craftsmen, foremen, and kindred workers: Mechanics and repairmen, except auto; C27020: Craftsmen, foremen, and kindred workers: Machinists; C27021: Craftsmen, foremen, and kindred workers: Metal craftsmen, except mechanics and machinists; C27022: Craftsmen, foremen, and kindred workers: Carpenters; C27023: Craftsmen, foremen, and kindred workers: Construction craftsmen, except carpenters; C27024: Craftsmen, foremen, and kindred workers: Other craftsmen; C27025: Operatives, except transport: Durable goods, manufacturing; C27026: Operatives, except transport: Nondurable goods, manufacturing; C27027: Operatives, except transport: Nonmanufacturing industries; C27028: Transport equipment operatives: Truck drivers; C27029: Transport equipment operatives: Other transport equipment operatives; C27030: Laborers, except farm: Construction laborers; C27031: Laborers, except farm: Freight, stock, and material handlers; C27032: Laborers, except farm: Other laborers, except farm; C27033: Farmers and farm managers; C27034: Farm laborers and farm foremen: Farm laborers, unpaid family workers; C27035: Farm laborers and farm foremen: Farm laborers, except unpaid, farm foremen;}
   \item{\textbf{1980}:  DIB009: Farming, forestry and fishing occupations (codes 473-499); DIB010: Precision production, craft and repair occupations (codes 503-699); DIB011: Operators, fabricators, and laborers: Machine operators, assemblers and inspectors (codes 703-799); DIB012: Operators, fabricators, and laborers: Transportation and material moving occupations (codes 803-859); DIB013: Operators, fabricators, and laborers: Handlers, equipment cleaners, helpers and laborers (codes 863-889);}
   \item{\textbf{1990}:  E4Q009: Farming, forestry, and fishing occupations (473-502); E4Q010: Precision production, craft, and repair occupations (503-702); E4Q011: Operators, fabricators, and laborers (703-902): Machine operators, assemblers, and inspectors (703-802); E4Q012: Operators, fabricators, and laborers (703-902): Transportation and material moving occupations (803-863); E4Q013: Operators, fabricators, and laborers (703-902): Handlers, equipment cleaners, helpers, and laborers (864-902);}
   \item{\textbf{2000}:  GMJ004: Male - Farming, fishing, and forestry occupations; GMJ005: Male - Construction, extraction, and maintenance occupations; GMJ006: Male - Production, transportation, and material moving occupations; GMJ010: Female - Farming, fishing, and forestry occupations; GMJ011: Female - Construction, extraction, and maintenance occupations; GMJ012: Female - Production, transportation, and material moving occupations;}
   \item{\textbf{2010}:  JQ5E030: Male: Natural resources, construction, and maintenance occupations; JQ5E034: Male: Production, transportation, and material moving occupations; JQ5E066: Female: Natural resources, construction, and maintenance occupations; JQ5E070: Female: Production, transportation, and material moving occupations;}
\end{itemize}

\paragraph{Service}
\begin{itemize}
   \item{\textbf{1970}:  C27036: Service workers, except private household: Cleaning service workers; C27037: Service workers, except private household: Food service workers; C27038: Service workers, except private household: Health service workers; C27039: Service workers, except private household: Personal service workers; C27040: Service workers, except private household: Protective service workers; C27041: Service workers, except private household: Service workers, except private household; C27042: Private household workers;}
   \item{\textbf{1980}:  DIB006: Service Occupations: Private household occupations (codes 403-407); DIB007: Service Occupations: Protective service occupations (codes 413-427); DIB008: Service Occupations: Service occupations, except household and protective (codes 433-469);}
   \item{\textbf{1990}:  E4Q006: Service occupations (403-472): Private household occupations (403-412); E4Q007: Service occupations (403-472): Protective service occupations (413-432); E4Q008: Service occupations (403-472): Service occupations, except protective and household (433-472);}
   \item{\textbf{2000}:  GMJ002: Male - Service occupations; GMJ008: Female - Service occupations;}
   \item{\textbf{2010}:  JQ5E019: Male: Service occupations; JQ5E055: Female: Service occupations;}
\end{itemize}

\clearpage
\section{Overview of the results for all used regions}

\subsection{Vancouver}
\begin{center}
    \includegraphics[width=\linewidth]{39a.png}
    \includegraphics[width=\linewidth]{39b.png}
\end{center}

\subsection{Toronto}
\begin{center}
    \includegraphics[width=\linewidth]{40a.png}
    \includegraphics[width=\linewidth]{40b.png}
\end{center}

\subsection{Montreal}
\begin{center}
    \includegraphics[width=\linewidth]{41a.png}
    \includegraphics[width=\linewidth]{41b.png}
\end{center}

\subsection{Calgary}
\begin{center}
	\includegraphics[width=\linewidth]{1a.png}
	\includegraphics[width=\linewidth]{1b.png}
\end{center}

\subsection{Edmonton}
\begin{center}
	\includegraphics[width=\linewidth]{2a.png}
	\includegraphics[width=\linewidth]{2b.png}
\end{center}

\subsection{Halifax}
\begin{center}
	\includegraphics[width=\linewidth]{3a.png}
	\includegraphics[width=\linewidth]{3b.png}
\end{center}

\subsection{Ottawa}
\begin{center}
	\includegraphics[width=\linewidth]{4a.png}
	\includegraphics[width=\linewidth]{4b.png}
\end{center}

\subsection{Quebec City}
\begin{center}
	\includegraphics[width=\linewidth]{5a.png}
	\includegraphics[width=\linewidth]{5b.png}
\end{center}

\subsection{Regina}
\begin{center}
	\includegraphics[width=\linewidth]{6a.png}
	\includegraphics[width=\linewidth]{6b.png}
\end{center}

\subsection{Saskatoon}
\begin{center}
	\includegraphics[width=\linewidth]{7a.png}
	\includegraphics[width=\linewidth]{7b.png}
\end{center}

\subsection{St John}
\begin{center}
	\includegraphics[width=\linewidth]{8a.png}
	\includegraphics[width=\linewidth]{8b.png}
\end{center}

\subsection{Winnipeg}
\begin{center}
	\includegraphics[width=\linewidth]{9a.png}
	\includegraphics[width=\linewidth]{9b.png}
\end{center}

\subsection{LA County}
\begin{center}
	\includegraphics[width=\linewidth]{10a.png}
	\includegraphics[width=\linewidth]{10b.png}
\end{center}
\clearpage

\subsection{Staten Island}
\begin{center}
	\includegraphics[width=\linewidth]{11a.png}
	\includegraphics[width=\linewidth]{11b.png}
\end{center} \clearpage

\subsection{Manhattan}
\begin{center}
	\includegraphics[width=\linewidth]{12a.png}
	\includegraphics[width=\linewidth]{12b.png}
\end{center} \clearpage

\subsection{Queens}
\begin{center}
	\includegraphics[width=\linewidth]{13a.png}
	\includegraphics[width=\linewidth]{13b.png}
\end{center} \clearpage

\subsection{The Bronx}
\begin{center}
	\includegraphics[width=\linewidth]{14a.png}
	\includegraphics[width=\linewidth]{14b.png}
\end{center} \clearpage

\subsection{Brooklyn}
\begin{center}
	\includegraphics[width=\linewidth]{15a.png}
	\includegraphics[width=\linewidth]{15b.png}
\end{center} \clearpage

\subsection{Chicago}
\begin{center}
	\includegraphics[width=\linewidth]{16a.png}
	\includegraphics[width=\linewidth]{16b.png}
\end{center} \clearpage

\subsection{Atlanta}
\begin{center}
	\includegraphics[width=\linewidth]{17a.png}
	\includegraphics[width=\linewidth]{17b.png}
\end{center} \clearpage

\subsection{Austin}
\begin{center}
	\includegraphics[width=\linewidth]{18a.png}
	\includegraphics[width=\linewidth]{18b.png}
\end{center} \clearpage

\subsection{Baltimore}
\begin{center}
	\includegraphics[width=\linewidth]{19a.png}
	\includegraphics[width=\linewidth]{19b.png}
\end{center} \clearpage

\subsection{Boston}
\begin{center}
	\includegraphics[width=\linewidth]{20a.png}
	\includegraphics[width=\linewidth]{20b.png}
\end{center} \clearpage

\subsection{Columbus}
\begin{center}
	\includegraphics[width=\linewidth]{21a.png}
	\includegraphics[width=\linewidth]{21b.png}
\end{center} \clearpage

\subsection{Dallas}
\begin{center}
	\includegraphics[width=\linewidth]{22a.png}
	\includegraphics[width=\linewidth]{22b.png}
\end{center} \clearpage

\subsection{Denver}
\begin{center}
	\includegraphics[width=\linewidth]{23a.png}
	\includegraphics[width=\linewidth]{23b.png}
\end{center} \clearpage

\subsection{Fargo}
\begin{center}
	\includegraphics[width=\linewidth]{24a.png}
	\includegraphics[width=\linewidth]{24b.png}
\end{center} \clearpage

\subsection{Houston}
\begin{center}
	\includegraphics[width=\linewidth]{25a.png}
	\includegraphics[width=\linewidth]{25b.png}
\end{center} \clearpage

\subsection{Indianapolis}
\begin{center}
	\includegraphics[width=\linewidth]{26a.png}
	\includegraphics[width=\linewidth]{26b.png}
\end{center} \clearpage

\subsection{Jacksonville}
\begin{center}
	\includegraphics[width=\linewidth]{27a.png}
	\includegraphics[width=\linewidth]{27b.png}
\end{center} \clearpage

\subsection{Kansas City}
\begin{center}
	\includegraphics[width=\linewidth]{28a.png}
	\includegraphics[width=\linewidth]{28b.png}
\end{center} \clearpage

\subsection{Las Vegas}
\begin{center}
	\includegraphics[width=\linewidth]{29a.png}
	\includegraphics[width=\linewidth]{29b.png}
\end{center} \clearpage

\subsection{Miami}
\begin{center}
	\includegraphics[width=\linewidth]{30a.png}
	\includegraphics[width=\linewidth]{30b.png}
\end{center} \clearpage

\subsection{Minneapolis}
\begin{center}
	\includegraphics[width=\linewidth]{31a.png}
	\includegraphics[width=\linewidth]{31b.png}
\end{center} \clearpage

\subsection{Philadelphia}
\begin{center}
	\includegraphics[width=\linewidth]{32a.png}
	\includegraphics[width=\linewidth]{32b.png}
\end{center} \clearpage

\subsection{Phoenix}
\begin{center}
	\includegraphics[width=\linewidth]{33a.png}
	\includegraphics[width=\linewidth]{33b.png}
\end{center} \clearpage

\subsection{Salt Lake City}
\begin{center}
	\includegraphics[width=\linewidth]{34a.png}
	\includegraphics[width=\linewidth]{34b.png}
\end{center} \clearpage

\subsection{San Antonio}
\begin{center}
	\includegraphics[width=\linewidth]{35a.png}
	\includegraphics[width=\linewidth]{35b.png}
\end{center} \clearpage

\subsection{San Francisco}
\begin{center}
	\includegraphics[width=\linewidth]{36a.png}
	\includegraphics[width=\linewidth]{36b.png}
\end{center} \clearpage

\subsection{Seattle}
\begin{center}
	\includegraphics[width=\linewidth]{37a.png}
	\includegraphics[width=\linewidth]{37b.png}
\end{center} \clearpage

\subsection{Washington DC}
\begin{center}
	\includegraphics[width=\linewidth]{38a.png}
	\includegraphics[width=\linewidth]{38b.png}
\end{center}

\end{document}

### 2018-08-13 backend/auxplots.py

@@ -1,0 +1,139 @@
+import numpy as np
+import matplotlib.pyplot as plt
+from matplotlib.collections import PatchCollection
+#from mpl_toolkits.basemap import Basemap
+from mpl_toolkits.mplot3d import Axes3D #3d doesn't work without this
+from matplotlib.patches import Polygon
+from shapely.geometry.polygon import orient
+import matplotlib.cm as cmx
+import matplotlib.colors as colors
+from random import random 
+
+def plot3dWS(G,psi,ds,dsconf,showLabels=False):    
+    years=[]
+
+    cm=plt.get_cmap('viridis') 
+    cNorm  = colors.Normalize(vmin=min(psi.values()), vmax=max(psi.values()))
+    scalarMap = cmx.ScalarMappable(norm=cNorm, cmap=cm)
+
+    fig = plt.figure()
+    ax = fig.add_subplot(111, projection='3d')
+    
+    for e in G.edges():
+
+        c1=G.node[e[0]]['pos']
+        c2=G.node[e[1]]['pos']
+        lw=0.1+1*(1-ds.distance(e[0],e[1],dsconf))
+        if (c1[2]==c2[2]):
+            c='black'
+        else:
+            c='red'
+        ax.plot(xs=[c1[0],c2[0]],ys=[c1[1],c2[1]],zs=[c1[2],c2[2]],c=c,linewidth=lw)
+        if (showLabels):
+            ax.text( (c1[0]+c2[0])/2.0,
+                    (c1[1]+c2[1])/2.0,
+                    (c1[0]+c2[0])/2.0,
+                    "{0}".format(ds.distance(e[0],e[1],dsconf)))
+
+    c=[]
+    xs=[]
+    ys=[]
+    zs=[]
+    for n in G.nodes():
+        xs.append(G.node[n]['pos'][0])
+        ys.append(G.node[n]['pos'][1])
+        zs.append(G.node[n]['pos'][2])
+        c.append(scalarMap.to_rgba(psi[n]))
+        if (showLabels):
+            ax.text(xs[-1], ys[-1], zs[-1],' '.join(['{0}'.format(ds.getValue(n,dsconf))])) 
+
+    ax.scatter(xs, ys, zs, c=c, marker='o',alpha=1)
+    
+
+    plt.show()
+
+def plotPolys(polylist,title=''):
+    fig = plt.figure()
+    plt.title(title)
+    plt.set_cmap('tab20c')
[…]

### 2018-08-13 backend/clearCache.py

@@ -1,0 +1,27 @@
+from shutil import rmtree
+from os.path import exists
+from os import makedirs, remove
+from glob import glob
+import json
+import sys
+
+with open('conf.json','r') as fin:
+    cities=json.load(fin)
+
+erase=(len(sys.argv)>1)
+
+
+for city in cities:
+    cache=city['folder']+'/cache'
+
+    if (not exists(cache)):
+        makedirs(cache)
+    else:
+        if (erase):
+            rmtree(cache)
+            makedirs(cache)
+        else:        
+            for f in glob(cache+'/*.json'):
+                remove(f)
+
+

### 2018-08-13 backend/dataStore.py

@@ -1,0 +1,218 @@
+from scipy.stats import skew
+from os.path import exists
+import numpy as np
+from util import geoJsons
+from copy import deepcopy
+from scipy.spatial.distance import sqeuclidean
+
+def _norm(X):
+    if (len(X)==1):
+        return(X)
+    sX=np.sum(X)
+    if (sX>0):
+        ret=X/sX
+    else:
+        ret=np.zeros_like(X)
+    return(ret)
+def _varDist(v1, v2):
+    if len(v1)==1:
+        return(sqeuclidean(v1,v2))
+    else:
+        return(1-np.sum(np.sqrt(np.multiply(v1,v2))))
+
+
+class dataStore(object):
+    def __init__(self):
+        self._avFilters=[{'id':0, 'name':'Original data', 'func':(lambda x:x)},
+                        {'id':1, 'name':'Skewness (+1)', 'func':(lambda x:skew(x)+1)},
+                        ]     # available filters 
+        self._avVars=[]       # variables included
+        self._nodes=[]        # list of nodes (year,rid)
+        self._rawdata=dict()  # raw data
+        self._normdata=dict() # normalized data
+        self._tabdata=None    # tabular version of the data - faster knn
+        self._rowByVarID={}   # row ranges for each var
+        self._varById={}      # variable name (key) by varID
+        self._distCache={}    # cache of computed distances
+        self._NodeByJSID={} # to enable talk with the react frontend - stable ordering
+        self._JSIDByNode={}
+
+    def avVars(self):
+        return(self._avVars)
+    def getVarName(self,varInd):
+        return(self._avVars[varInd]['name'])
+    def getFilterName(self,fInd):
+        return(self._avFilters[fInd]['name'])
+    def getVarLabels(self,varInd):
+        return(self._avVars[varInd]['labels'])    
+    def getVarShortLabels(self,varInd):
+        return(self._avVars[varInd]['short'])    
+    def avFilters(self):
+        return(self._avFilters)
+    def getRaw(self,n):
+        return(list(self._rawdata[n].values()))
+    def getPopulation(self,n):
+        return(self._rawdata[n]['Population']['values'][0])
+    def _buildtabdata(self):
+        lenVars=0        
+        for V in self.avVars():
+            clen=len(self._normdata[self._nodes[0]][V['name']]['values'])
+            self._rowByVarID[V['id']]=(lenVars,lenVars+clen)
[…]

### 2018-08-13 backend/genBasegraph.py

@@ -1,0 +1,86 @@
+import sys
+import networkx as nx
+from util import indexedPols,geoJsons
+from shapely.geometry import shape   
+import geojson
+#from auxplots import plotPolys
+#import matplotlib.pylab as plt
+#from mpl_toolkits.mplot3d import Axes3D
+
+
+
+if __name__ == '__main__':
+    if (len(sys.argv)) < 3:
+        print(".py geojsons.zip display.gj (basegraph.gp)")
+        exit(-1)
+
+    if (len(sys.argv)==3):
+        outName='basegraph.gp'
+    else:
+        outName=sys.argv[3]
+    geo = {}
+    print('Make sure you are using 4269!')
+    with geoJsons(sys.argv[1]) as gj:
+        for year,cgj in gj:
+            print(year)
+            geo[year]=indexedPols()
+            for feat in cgj['features']:
+                geo[year].insertJSON(G=feat['geometry'],props={'CT_ID':feat['properties']['CT_ID'],})
+    years = sorted(geo.keys())
+
+    print('loading display')
+    with open(sys.argv[2],'r') as fgjin:
+        displayGJ=geojson.load(fgjin)
+
+    didByN=dict()
+    AreaByDid=dict()
+    for feat in displayGJ['features']:
+        did=feat['properties']['display_id']
+        AreaByDid[did]=feat['properties']['area']
+        for n in feat['properties']['nodes']:
+            n=tuple(n)
+            if n not in didByN:
+                didByN[n]=[]
+            didByN[n].append(did)
+
+
+    print('starting spatial edges')
+    B = nx.Graph()
+
+
+    
+    for year in years:
+        for fid, geom in geo[year].iterIDGeom('CT_ID'):
+            n=(year,fid)
+
+            B.add_node(n) 
+            if ('display_ids' not in B.node[n]):
+                if (n not in didByN):
+                    B.node[n]['display_ids']=[]
+                    B.node[n]['areas']=[]
+                else:
+                    B.node[n]['display_ids']=didByN[n]
+                    B.node[n]['areas']=[AreaByDid[x] for x in didByN[n]]
+
+                representative=geom.representative_point()
+                B.node[n]['pos']=[representative.x, representative.y,year]
+            #spatial edges                
+            for polID in geo[year].search(geom.buffer(1e-3)):
+                nid=geo[year].getProperty(polID,'CT_ID')
+                if (fid != nid):
[…]

### 2018-08-13 backend/genCities.py

@@ -1,0 +1,112 @@
+"""This file reads cities.json, cuts the data accordingly (from my personal
+dropbox, sorry), and generates the appropriate folders and conf.json to run the
+server"""
+import sys
+from geojson import load
+from subprocess import call
+from shapely.geometry import Polygon
+from util import indexedPols, zipJsons
+from os.path import exists
+from os import makedirs
+import string
+import json
+from shapely.geometry import shape
+from varConf import getVariables
+
+baseFiles=dict()
+baseFiles['US']='/home/diasf/Dropbox/USCensus/us.data.gj.zip'
+baseFiles['CA']='/home/diasf/Dropbox/USCensus/CA/ca.data.gj.zip'
+baseFiles['NY']='/home/diasf/workspace/ugp/analysis/CTEvo/backend/cities/US_New_York_City/geojsons.zip'
+baseFiles['LA']='/home/diasf/workspace/ugp/analysis/CTEvo/backend/cities/US_Los_Angeles/geojsons.zip'
+
+def doCut(zipname,cutPolygon,destZip):
+    with zipJsons(zipname) as gj:
+        for year,cgj in gj:
+            print(year)
+            pols=indexedPols()
+            for feat in cgj['features']:
+                pols.insertJSON(G=feat['geometry'],props=feat['properties'])
+            pols.keepOnly(pols.bbSearch(cutPolygon))
+            # pols.keepOnly(pols.search(cutPolygon))
+            pols.keepOnly(pols.containedIn(cutPolygon))
+            pols.saveGeoJson('{0}.gj'.format(year))
+    call('zip -9 -m {0} *.gj'.format(destZip),shell=True)
+
+
+if __name__ == '__main__':
+    if len(sys.argv)!=2:
+        print('.py cities.json')
+        exit(-1)
+    with open(sys.argv[1],'r') as fin:
+        cities=load(fin)
+    printable = set(string.printable)
+    conf=[]
+    makeBuff=''
+    cities['features'].append({'properties':{'name':'Synthetic','kind':'SY'},'geometry':[]})
+
+    allNames=[]
+    allFolders=[]
+
+    for city in cities['features']:
+        cName=city['properties']['name']
+        kind=city['properties']['kind']
+
+        if (kind!='SY'):
+            pol=shape(city['geometry'])
+
+        print(cName)
+
[…]

### 2018-08-13 backend/genDisplayGJ.py

@@ -1,0 +1,86 @@
+from util import geoJsons
+from util import indexedPols,split    
+from shapely.geometry import shape   
+from shapely.ops import unary_union
+import sys
+from auxplots import plotPolys
+
+
+if __name__ == '__main__':
+    if (len(sys.argv))!=3:
+        print(".py <geojsons.zip> OUTPUT.gj")
+        exit(-1)
+
+    areaT=1e-9
+    print('Make sure you are using 4269!')
+                
+#    years=sorted(geo.keys())
+    resPols=indexedPols()
+    print('Display geojson')
+    
+    with geoJsons(sys.argv[1]) as gj:
+        for year,cgj in gj:
+            print('\n\n')
+            print(year)
+            toInsert=[]
+            last=[]
+            for fcount,feat in enumerate(cgj['features']):
+                if (len(feat['geometry']['coordinates'])==0):
+                    continue
+                A=shape(feat['geometry']).buffer(0)
+                if (A.is_empty):
+                    continue
+                I=resPols.search(A.buffer(-1e-6).buffer(0))
+                
+                nodesA=[(year,feat['properties']['CT_ID']),]
+                if (not I):
+                    resPols.insert(A,props={'nodes':nodesA, 'area':A.area})
+                    continue
+
+                UB=[]
+                for nii in I:
+                    B=resPols.remove(nii)
+                    _,AinterB,bp=split(A,B['geometry'])
+                    if (AinterB is not None) and (not AinterB.is_empty):
+                        resPols.insert(AinterB,props={'nodes':nodesA+B['properties']['nodes'], 'area':AinterB.area},areaThreshold=areaT)
+                    if (bp is not None) and (not bp.is_empty):
+                        resPols.insert(bp,props={'nodes':B['properties']['nodes'], 'area':bp.area},areaThreshold=areaT)
+
+                    UB.append(B['geometry'])
+
+                if (UB):
+                    try:
+                        UB=unary_union(UB)
+                    except:
+                        UB=[x.buffer(1e-6) for x in UB]
+                        UB=unary_union([x for x in UB if (not x.is_empty)])
+                        UB=UB.buffer(-1e-6)
+                        
+
+                    try:
+                        UB=UB.buffer(2e-5).buffer(-2e-5) #removes slits                            
+                    except:
+                        UB=UB.buffer(2e-5).buffer(-2e-5).buffer(0) #removes slits                            
+
+                    try:
+                        newA=A.difference(UB)
[…]

### 2018-08-13 backend/genSynt.py

@@ -1,0 +1,204 @@
+import networkx as nx
+from geojson import Feature, Polygon
+from json import dump
+from numpy.random import normal
+import numpy as np
+import matplotlib.pylab as plt
+import sys
+
+def _fix(p,M):
+    return(int(round( (p/100) *M)))
+
+if len(sys.argv)!=2:
+    print('.py outFolder')
+    exit(-1)
+outFolder=sys.argv[1]
+
+np.random.seed(783)
+
+centers = {1:(1,1),2:(6,6), 3:(3,3), 4:(4,6), 5:(1,9), 6:(6,2), 7:(9,1), 8:(4.5,4.5) }
+
+def getSample(c):
+    X=normal(loc=10*centers[c][0],scale=4)
+    Y=normal(loc=10*centers[c][1],scale=4)
+    return([(X,100-X),(Y,100-Y)])
+
+WW=20
+WH=20
+world=np.zeros((WW,WH,5))
+
+
+world[:_fix(50,WW),:,:]=1
+world[_fix(50,WW):,:,:]=2
+for x in range(_fix(70,WW),_fix(75,WW)):
+    world[x,np.arange((x+1)%2,WW,8),2]=3
+    
+
+
+# world[:5,:,:]=1
+# world[5:,:,:]=2
+
+# world[:]=4
+# for i in range(world.shape[2]):
+#     plt.figure()
+#     plt.imshow(world[:,:,i].T,origin="lower")
+#     plt.colorbar()
+#     # plt.savefig('{0}_synt.png'.format(i+1))
+# plt.show()
+
+
+
+pols=dict()
+points=dict()
+ids=np.zeros_like(world)
+clusters=dict()
+revID=dict()
+for y in range(world.shape[2]):
+    pols[y]=[]
+    points[y]=[]
+    clusters[y]=dict()
+    revID[y]=dict()
+    print('year',y)
+    for i in range(world.shape[0]):
+        for j in range(world.shape[1]):
+            coords=[(i,j),(i+1,j),(i+1,j+1),(i,j+1)]
+            iC=[]
+            for c in coords:
+                if (c not in points[y]):
+                    iC.append(len(points[y]))
+                    points[y].append(c)
+                else:
+                    iC.append(points[y].index(c))
+            ids[i,j,y]=int(len(pols[y]))
+            clusters[y][ids[i,j,y]]=world[i,j,y]
+            revID[y][len(pols[y])]=(i,j)
[…]

### 2018-08-13 backend/hierMWW.py

@@ -1,0 +1,178 @@
+import numpy as np
+import networkx as nx
+from collections import deque
+#import matplotlib.pylab as plt
+from numpy.linalg import norm
+
+from time import time
+from fibonacci_heap_mod import Fibonacci_heap
+
+
+def removeLayer(G, layer):
+    for n in G.nodes():
+        if layer in G.node[n]:
+            del(G.node[n][layer])
+    for e in G.edges():
+        if (layer in G[e[0]][e[1]]):
+            del(G[e[0]][e[1]][layer])
+
+
+def Fminus(G, n, layer):
+    return(min([e[2][layer] for e in G.edges(n, data=True)]))
+
+
+def stream(G, psi, x, layer):
+    L = [x, ]
+    lp = deque([x, ])
+    while (len(lp) > 0):
+        y = lp.popleft()
+        breadth_first = True
+        allZ = [z for z in G.neighbors(y) if (
+            z not in L) and G[y][z][layer] == Fminus(G, y, layer)]
+        if (not allZ):
+            break
+        for z in allZ:
+            if (not breadth_first):
+                break
+            if (psi[z] >= 0):
+                return([L, psi[z]])
+            elif Fminus(G, z, layer) < Fminus(G, y, layer):
+                L.append(z)
+                lp.clear()
+                lp.append(z)
+                breadth_first = False
+            else:
+                L.append(z)
+                lp.append(z)
+    return (L, -1)
+
+
+def watershed(G, layer):
+    psi = dict()
+    for n in G.nodes():
+        psi[n] = -1
+    nb_labs = 0
+    for x in G.nodes():
+        if (psi[x] == -1):
+            [L, lab] = stream(G, psi, x, layer)
+            if (lab == -1):
+                for y in L:
+                    psi[y] = nb_labs
+                nb_labs += 1
+            else:
+                for y in L:
+                    psi[y] = lab
+    return(psi)
+
+
+def hierMWW(D, Gr, NbyInd, dsconf, maxClusters=20):
+
+    IndByN = {NbyInd[i]: i for i in NbyInd}
+
+    T=np.percentile(D,25)
+
+    G = Gr.copy()
+    lName = 'data'
+    for e in G.edges():
+        i1 = IndByN[e[0]]
+        i2 = IndByN[e[1]]
+        cD = D[i1, i2]
+        G[e[0]][e[1]][lName] = cD
+
+    psi = watershed(G, lName)
[…]

### 2018-08-13 backend/mmg.py

@@ -1,0 +1,351 @@
+import networkx as nx
+from fibonacci_heap_mod import Fibonacci_heap, merge
+#import matplotlib.pylab as plt
+from collections import deque
+import numpy as np
+
+#from auxplots import plot3dWS
+from time import time
+
+
+def bhat(u1,u2,v1,v2):
+    return((1.0/4) * np.log ( 1.0/4 * (v1/v2 + v2/v1 + 2)) + 1.0/4 * (((u1-u2)**2)/(v1+v2)))
+
+
+# Reference: https://hal.archives-ouvertes.fr/file/index/docid/622505/filename/hal.pdf
+def MakeSet(C,x):
+    C[x]={'rank':0,'parent':x}
+    
+def find(C, x):
+    if (C[x]['parent']!=x):
+        C[x]['parent']=find(C,C[x]['parent'])
+    return(C[x]['parent'])
+
+def Link(C, x, y):
+    xRoot = find(C,x)
+    yRoot = find(C,y)
+ 
+    if xRoot == yRoot:
+        return
+   
+    if C[xRoot]['rank'] < C[yRoot]['rank']:
+        C[xRoot]['parent'] = yRoot
+    elif C[xRoot]['rank'] > C[yRoot]['rank']:
+        C[yRoot]['parent'] = xRoot
+    else:
+        C[yRoot]['parent'] = xRoot    
+        C[xRoot]['rank']   = C[xRoot]['rank'] + 1
+    return(find(C,x))
+
+def meld(C, L, x, y):
+    res=Fibonacci_heap()
+    while (len(L[x])>0):
+        el=L[x].dequeue_min()
+        _,v=el.get_value()
+        if (find(C,v[0])!=find(C,v[1])):
+            res.enqueue(el.get_value(),el.get_priority())
+    while (len(L[y])>0):
+        el=L[y].dequeue_min()
+        _,v=el.get_value()
+        if (find(C,v[0])!=find(C,v[1])):
+            res.enqueue(el.get_value(),el.get_priority())
+    return(res)
+
+def uprooting(G, X, ds, dsconf):
+    C = dict()
+    L = dict()
+    I = dict()
+    for x in G.nodes():
+        MakeSet(C, x)
+        L[x] = Fibonacci_heap()
+
+    for u in G.edges():
+        if (u in X.edges()):
+            I[(min(u), max(u))]=0
+        else:
+            I[(min(u), max(u))]=-1
+
[…]

### 2018-08-13 backend/normGJVars.py

@@ -1,0 +1,74 @@
+""" This script gather the usable variables across the years and
+put them on the same place, with the same names, saving it as an actual geojson"""
+import sys
+from util import zipJsons
+from varConf import getVariables
+from geojson import FeatureCollection, Feature, dump
+import numpy as np
+from os.path import dirname
+from shapely.geometry import shape   
+from util import indexedPols,split    
+from shapely.ops import unary_union
+
+def _newProps(varNames):
+    ret = {}
+    for n in varNames:
+        ret[n] = []
+    return(ret)
+
+
+def _setVals(year, avVars, props):
+    newProps = {'CT_ID': props['CT_ID'], 'variables': []}
+    for v in avVars:
+        newProps['variables'].append({
+            'name'  : v.getName(),
+            'labels': v.getLabels(),
+            'short' : v.getShortLabels(), 
+            'type'  : v.getType(),
+            'values': v.computeValue(props, year)
+        })
+        
+        # if (any([isinstance(x,str) for x in newProps['variables'][-1]['values']])):
+        #     print(newProps['CT_ID'])
+        #     print(newProps['variables'][-1])
+    
+    return(newProps)
+
+
+if __name__ == '__main__':
+    if (len(sys.argv)) != 3:
+        print(".py geoJsons.zip US/CA")
+        exit(-1)
+
+    folder = dirname(sys.argv[1])
+
+    avVars = getVariables(sys.argv[2])
+    print([v.getName() for v in avVars])
+    with zipJsons(sys.argv[1]) as gj:
+        for year, cgj in gj:
+            usedPols=indexedPols()
+
+            if ('labels' in cgj):
+                with open('{0}_labels.txt'.format(year), 'w') as f:
+                    f.write("\n".join(
+                        ['{0}-{1}'.format(k, cgj['labels'][k]) for k in sorted(cgj['labels'].keys())]))
+
+            feats = []
+            for f in cgj['features']:
+                g = shape(f['geometry']).buffer(0)
+                I = usedPols.search(g)
+                if (len(I)>0):
+                    # print('found stuff!', year,len(I),f['properties']['CT_ID'])
+                    UB=unary_union([usedPols.getPolygon(i) for i in I]).buffer(0)
+                    g=g.difference(UB).buffer(0)
+
+                if (not g.is_empty):
+                    feats.append(Feature(geometry=g, properties=_setVals(year, avVars, f['properties'])))
[…]

### 2018-08-13 backend/sortCities.py

@@ -1,0 +1,50 @@
+
+
+'Toronto'
+'Greater Toronto Area'
+'Vancouver'
+
+'Calgary'
+'Edmonton'
+'Halifax'
+'Montreal'
+'Ottawa'
+'Quebec City'
+'Regina'
+'Saskatoon'
+'St John'
+'Winnipeg'
+
+
+'Manhattan'
+'Brooklyn'
+'Queens'
+'The Bronx'
+'Staten Island'
+
+'Chicago'
+'LA County'
+
+'Atlanta'
+'Austin'
+'Baltimore'
+'Boston'
+'Columbus'
+'Dallas'
+'Denver'
+'Fargo'
+'Houston'
+'Indianapolis'
+'Jacksonville'
+'Kansas City'
+'Las Vegas'
+'Miami'
+'Minneapolis'
+'Philadelphia'
+'Phoenix'
+'Salt Lake City'
+'San Antonio'
+'San Francisco'
+'Seattle'
+'Washington DC'
+

### 2018-08-13 backend/srv.py

@@ -1,0 +1,699 @@
+"""Server module for the backend"""
+import json
+import os
+import pickle
+import sys
+from itertools import combinations, combinations_with_replacement
+from os import makedirs
+from os.path import exists
+from time import time
+
+import numpy as np
+from numpy import nan
+
+import cherrypy
+import networkx as nx
+from dataStore import dataStore
+from hierMWW import hierMWW
+from mmg import segmentation
+from networkx.readwrite import json_graph
+from scipy.spatial.distance import pdist, sqeuclidean, squareform
+from sklearn.manifold import TSNE
+from sklearn.neighbors import NearestNeighbors
+
+#import matplotlib.pylab as plt
+
+def cors():
+  if cherrypy.request.method == 'OPTIONS':
+    # preflign request 
+    # see http://www.w3.org/TR/cors/#cross-origin-request-with-preflight-0
+    cherrypy.response.headers['Access-Control-Allow-Methods'] = 'POST'
+    cherrypy.response.headers['Access-Control-Allow-Headers'] = 'content-type'
+    cherrypy.response.headers['Access-Control-Allow-Origin']  = '*'
+    # tell CherryPy no avoid normal handler
+    return True
+  else:
+    cherrypy.response.headers['Access-Control-Allow-Origin'] = '*'
+
+cherrypy.tools.cors = cherrypy._cptools.HandlerTool(cors)
+
+
+#from scipy.stats import wasserstein_distance, entropy
+
+
+def _doStats(V):
+    return(np.nanmedian(V,axis=0),np.nanmin(V,axis=0),np.nanmax(V,axis=0),np.percentile(V,25,axis=0),np.percentile(V,75,axis=0))
+
+def _parD(x):
+    return(squareform(pdist(x, 'cosine')))
+    # dt=np.zeros((x.shape[0],x.shape[0]))
+    # for i in range(x.shape[0]):
+    #     for j in range(i+1,x.shape[0]):
+    #         # dt[i,j]=entropy(x[i,:],x[j,:])
+    #         dt[i,j]=wasserstein_distance(x[i,:],x[j,:])
+    # return(dt+dt.T)
+
+def _GeoMetric(G,nodes):
+
+    nodesByYear=dict()
+    for n in nodes:
+        if (n[0]) not in nodesByYear:
+            nodesByYear[n[0]]=[]
+        nodesByYear[n[0]].append(n)
+
+    allNodesByYear=dict()
[…]

### 2018-08-13 backend/texVars.py

@@ -1,0 +1,34 @@
+from varConf import getVariables
+from subprocess import run, PIPE
+
+vars=getVariables(kind='CA')
+
+for A in vars:
+    print("\\subsubsection{{{0}}}".format(A.getName()))
+    for v in A._vars:
+        print("\\paragraph{{{0}}}".format(v.getName()))
+        print("\\begin{itemize}")
+
+        for y in sorted(v._cols.keys()):
+            line="   \\item{{\\textbf{{{0}}}: ".format(y)
+            for c in v._cols[y]:
+                if (c==''):
+                    continue
+                if (c[0]=='-'):
+                    line=line+" (-) "
+                    cc=c[1:]
+                else:
+                    cc=c
+                res=run(['cat *{1}*.txt | grep {0}'.format(cc,y)], shell=True, stdout=PIPE)
+                if (res.returncode==0):
+                    colName=res.stdout.decode('utf-8').strip().replace("      "," ").replace("     "," ").replace(";",',')
+                    colName=colName[len(cc)+1:].strip()
+                    line=line+" "+colName.replace('$','\\$').replace('%','\\%').replace('&','\\&')+";"
+            line=line+"}"
+            print(line)
+        print("\\end{itemize}\n")
+
+
+
+
+#vars=getVariables(kind='US')

### 2018-08-13 backend/util.py

@@ -1,0 +1,235 @@
+import tempfile
+import zipfile
+import shutil
+import geojson
+from glob import glob
+from os.path import basename
+from rtree import index
+from shapely.geometry import shape, MultiPolygon
+from shapely.ops import snap
+from geojson import Feature, FeatureCollection
+from copy import deepcopy
+import json
+
+def polOnly(geom):
+    if (not geom) or ('Polygon' not in geom.type):
+        if ('Collection' in geom.geom_type):
+          return(MultiPolygon([g for g in geom if 'Polygon' in g.geom_type]))
+
+        return(None)
+    else:
+        return(geom)
+
+def split(AA, BB):
+    A=deepcopy(AA)
+    B=deepcopy(BB)
+    """Splits A and B into A-B, A inter B, and B-A"""
+    if (not A) or (not B):
+        return(A, None, B)
+    try:
+        AmB = A.difference(B)
+        AiB = A.intersection(B)
+        BmA = B.difference(A)
+    except:
+        nA = snap(A, B, 1e-3)
+        AmB = nA.difference(B)
+        AiB = nA.intersection(B)
+        BmA = B.difference(nA)
+    return(polOnly(AmB),
+           polOnly(AiB),
+           polOnly(BmA))
+
+class indexedPols:
+    def __init__(self):
+        self._index = index.Index()
+        self._pols = []
+        self._removed = [] #each polygon has a allocated value - more memory, but faster
+        self._properties = {}
+
+    def __next__(self):
+        for i in range(len(self._pols)):
+            if not self._removed[i]:
+                yield(self._pols[i])
+
+    def __iter__(self):
+        return(next(self))
+
+    # def cleanup(self):
+    #     """Actually removes deleted data - INVALIDATES THE INDEXES!"""
+    #     toRemove = sorted([i for i, v in enumerate(self._removed) if v], reverse=True)
+    #     for i in toRemove:
+    #         self._index.delete(i, self._pols[i].bounds)
+    #         del(self._pols[i])
+    #     self._removed = [False,]*len(self._pols)
+        
+    def _getPols(self, geocollection):
+        gc=geocollection
+        res=[]
+        if (gc.geom_type=='Polygon'):
+            return([gc,])
+        if (gc.geom_type=='MultiPolygon'):
+            return([p for p in gc])
+        if ('Collection' in gc.geom_type):
+            res.extend([self._getPols(p) for p in gc])
+        return(res)
+        
+    def saveGeoJson(self,fname,display_id=False,area=False, saveOnly=[]):
[…]

### 2018-08-13 backend/varConf.py

@@ -1,0 +1,784 @@
+from numpy import isnan
+class oneVariable(object):
+    def __init__(self,vname,sname=None):
+        self._cols=dict()
+        self._mult=dict()
+        if not sname:
+            sname=vname[:5]
+        self.setName(vname,sname)
+    def setName(self,vname,shortname):
+        self._name=vname
+        self._sname=shortname
+    def getName(self):
+        return(self._name)
+    def getShortName(self):
+        return(self._sname)    
+    def getLabels(self):
+        return([self._name,])
+    def getShortLabels(self):
+        return([self._sname,])    
+    def _addColumn(self, year, colName,mult=1):
+        if year not in self._cols:
+            self._cols[year] = []
+            self._mult[year] = []
+
+        self._cols[year].append(colName)
+        self._mult[year].append(mult)
+    def addColumns(self,year, cols,mult=[]):#mult is handy for average income inflation correction
+        if not mult:
+            mult=[1.0,]*len(cols)
+
+        for i,col in enumerate(cols):
+            self._addColumn(year,col,mult[i])
+            
+    def computeValue(self,gjProps,year):
+        ret = 0
+        for i,col in enumerate(self._cols[year]):
+            if (col==''):
+                continue
+            if (col[0]=='-'):
+                s=True
+                col=col[1:]
+            else:
+                s=False
+            if (col not in gjProps):
+                continue
+
+            try:
+                val=self._mult[year][i]*float(gjProps[col])
+            except:
+                return(['NaN',])
+
+            if (isnan(val)):
+                return(['NaN',])                                
+
+            if (s):
+                ret-=val
+            else:
+                ret+=val
+        return([ret,])
+
+class distVariable(object):
+    def __init__(self,dName,vtype='category'):
+        """Kind of variable influences how it is treated afterwards: 'category' or 'sequential' """
+        self._vars=[]
+        self.setName(dName)
+        self._vtype=vtype
+    def getType(self):
+        return(self._vtype)
+    def setName(self,dName):
+        self._name=dName
+    def getName(self):
+        return(self._name)
[…]

### 2018-08-13 sup/cropB.py

@@ -1,0 +1,12 @@
+from scipy.misc import imread, imsave
+import matplotlib.pylab as plt
+from glob import glob
+
+for f in glob('*b.png'):
+    print(f)
+    img = imread(f)
+    if img[849,1836,0]==0:
+        img=img[54:851,192:1838]
+    else:
+        img=img[54:562,192:1838]
+    imsave(f,img)

### 2018-08-13 ui/src/App.js

@@ -1,0 +1,67 @@
+import React, {Component } from 'react';
+import './App.css';
+import TempEvo from './sankey'
+import Overview from './overview'
+import Detail from './detail'
+import { connect } from 'react-redux';
+import { actionCreators, getURL, getData } from './reducers';
+import ConfigPanel from './configPanel'
+import NewMap from './newmap'
+import TrajDet from './trajectoryDetails';
+import TransMat from './transMat';
+
+const mapStateToProps = (state) => ({
+  showLoading: state.showLoading,
+
+});
+
+class App extends Component {
+  render() {
+    let retJSX=[];
+    let {showLoading}=this.props;
+    retJSX.push(<ConfigPanel key='cop' />);
+
+    if (showLoading===true){
+      retJSX.push(<div key="loading" className="loading">Loading, please wait...</div>);
+    }else{
+      retJSX.push(<Overview key='op'/>);
+      retJSX.push(<TempEvo key='tp'/>);
+      retJSX.push(<TransMat key='tm'/>)
+      retJSX.push(<NewMap key='nm'/>);
+      retJSX.push(<TrajDet key='tj'/>);
+      retJSX.push(<Detail key='dp'/>);  
+    }
+    return (
+        <div key='base' className="App">    
+          {retJSX}
+        </div>
+    );
+  }
+
+
+  componentDidMount = () => {    
+    let {dispatch}=this.props;
+    getData(getURL.CityOptions(), function(data) {
+
+      dispatch(actionCreators.SetCityOptions(data))
+      dispatch(actionCreators.SetCity(0));
+      getData(getURL.GeoJSON(0),(gj)=>{
+        dispatch(actionCreators.SetGeoJson(gj));
+      });
+
+      dispatch(actionCreators.ShowLoading(true));
+
+      getData(getURL.Segmentation(0),function(data) {    
+        // console.log(data);
+          dispatch(actionCreators.UpdateMap(data));                                
+          if (data.levelCorr.length>5)
+              dispatch(actionCreators.SetLevel(data.levelCorr.length-5));      
+          else
+              dispatch(actionCreators.SetLevel(data.levelCorr.length-2));          
+        dispatch(actionCreators.ShowLoading(false));
+      });
+    });
+  }  
+}
+
+export default  connect(mapStateToProps)(App);

### 2018-08-13 ui/src/HorBoxSeries.js

@@ -1,0 +1,110 @@
+// Copyright (c) 2016 - 2017 Uber Technologies, Inc.
+//
+// Permission is hereby granted, free of charge, to any person obtaining a copy
+// of this software and associated documentation files (the "Software"), to deal
+// in the Software without restriction, including without limitation the rights
+// to use, copy, modify, merge, publish, distribute, sublicense, and/or sell
+// copies of the Software, and to permit persons to whom the Software is
+// furnished to do so, subject to the following conditions:
+//
+// The above copyright notice and this permission notice shall be included in
+// all copies or substantial portions of the Software.
+//
+// THE SOFTWARE IS PROVIDED "AS IS", WITHOUT WARRANTY OF ANY KIND, EXPRESS OR
+// IMPLIED, INCLUDING BUT NOT LIMITED TO THE WARRANTIES OF MERCHANTABILITY,
+// FITNESS FOR A PARTICULAR PURPOSE AND NONINFRINGEMENT. IN NO EVENT SHALL THE
+// AUTHORS OR COPYRIGHT HOLDERS BE LIABLE FOR ANY CLAIM, DAMAGES OR OTHER
+// LIABILITY, WHETHER IN AN ACTION OF CONTRACT, TORT OR OTHERWISE, ARISING FROM,
+// OUT OF OR IN CONNECTION WITH THE SOFTWARE OR THE USE OR OTHER DEALINGS IN
+// THE SOFTWARE.
+
+import React from 'react';
+
+import {AbstractSeries} from 'react-vis';
+
+const predefinedClassName = 'horbox';
+
+class BoxplotSeries extends AbstractSeries {
+  render() {
+    const {className, data, marginLeft, marginTop, doMinMax, xDomain} = this.props;
+    if (!data) {
+      return null;
+    }
+    let doMM=(doMinMax===undefined)||(doMinMax===true);
+
+    const xFunctor = this._getAttributeFunctor('x');
+    const yFunctor = this._getAttributeFunctor('y');
+    const strokeFunctor = this._getAttributeFunctor('stroke') ||
+      this._getAttributeFunctor('color');
+    const fillFunctor = this._getAttributeFunctor('fill') ||
+      this._getAttributeFunctor('color');
+    const opacityFunctor = this._getAttributeFunctor('opacity');
+
+    let distance;
+    if (data.length>1) {
[…]

### 2018-08-13 ui/src/configPanel.js

@@ -1,0 +1,255 @@
+import React, { Component } from 'react';
+import { connect } from 'react-redux';
+import { actionCreators, getData, getURL, toInt } from './reducers';
+import './configPanel.css';
+
+const mapStateToProps = (state) => ({
+    cityOptions: state.cityOptions,
+    cityID: state.cityID,
+    dsconf: state.dsconf,
+    showConfig: state.showConfig,
+  });
+
+class ConfigPanel extends Component {
+    constructor(props){
+        super(props);
+        this.state= { 
+            k: 5,
+            nextId: 0,
+            conf : undefined,
+        };
+    }
+
+    render() {
+        let {showConfig}=this.props;
+        let {cityOptions}=this.props;
+        let {dispatch}=this.props;
+        let {cityID}=this.props;
+        let {dsconf}=this.props;
+        let tarray;
+
+        let defaultConf=(variables)=>{
+            let conf=[];
+            for (let i=0;i<variables.length;i++){
+                conf.push({id:i,var:variables[i].id,fs:0,w:1,enabled:true});
+            }
+            return(conf);
+        };
+
+        if ((showConfig===undefined) || (!showConfig) || (cityOptions===undefined)){
+            return(null);
+        }
+        else {
+            if (this.state.conf===undefined){
+                if (dsconf===undefined){
+                    this.setState({nextId:cityOptions[cityID].variables.length,conf:defaultConf(cityOptions[cityID].variables)});
+                }else{
+                    let cconf=[];
+                    for (let i=0;i<dsconf.vars.length;i++){
+                        cconf.push({id:i,var:dsconf.vars[i],fs:0,w:dsconf.w[i],enabled:true});
+                    }
+                    this.setState({nextID:dsconf.vars.length, conf:cconf});
+                }                
+                return(null);
+            }
+                
+            let retJSX=[]
+
+            retJSX.push(<div key='city' className="oneVar">
+                            <div className="aroundDiv" style={{border:'none'}}>Geographic region:</div>
+                            <select 
+                                className="dropboxes" 
+                                onChange={(e)=>{
+                                    let curVal=toInt(e.target.value);
+                                    if (cityOptions[cityID].kind!==cityOptions[curVal].kind){
+                                        this.setState({nextId:cityOptions[curVal].variables.length,conf:defaultConf(cityOptions[curVal].variables)});                                    }
+                                    dispatch(actionCreators.SetCity(curVal));
+                                }} >
+                                {cityOptions.map( (e) => {
[…]

### 2018-08-13 ui/src/curConf.js

@@ -1,0 +1,139 @@
+import React, { Component } from 'react';
+import { connect } from 'react-redux';
+import { actionCreators, selModes } from './reducers';
+import './style.css';
+import './curConf.css';
+
+const mapStateToProps = (state) => ({
+    conf: state.conf,
+    dsconf: state.dsconf,
+    level: state.level,
+    showConfig: state.showConfig,
+    simpleColours: state.simpleColours,
+    selMode: state.selMode,
+    nLevels: state.nLevels
+  });
+
+//https://github.com/iconic/open-iconic
+class CurConf extends Component {
+    render() {
+        let {dispatch,simpleColours}=this.props;
+        let {conf,nLevels}=this.props;
+        let {dsconf}=this.props;
+        let {selMode}=this.props;
+        let {showConfig}=this.props;
+        let {level}=this.props;
+        let tJSX=[];
+//                         <li key={'k'}>Augmented edges: {dsconf.k}</li>
+        
+        if (conf!==undefined){
+            tJSX.push(
+                <div key='vconf' 
+                     style={{lineHeight:'1em', textAlign:'left', width:'fit-content',margin:'auto',border:'solid',borderWidth:'thin',borderColor:'lightgray',padding:'10px'}}>
+                     <p style={{fontWeight:'bold', width:'fit-content',height:'fit-content',margin:'auto 0',textAlign:'left'}}>Current configuration:</p> 
+                     <ul 
+                         style={{paddingLeft: '10px', margin:'auto'}}>
+                         {conf.map((d)=>{
+                             return(<li key={d.var}>{d.var} ({d.w})</li>);//: {d.filter} 
+                         })}
+                     </ul> 
+                 </div>);
+         }        
+        let onChange= (d)=> {
+            dispatch(actionCreators.SetLevel(8-parseInt(d.target.value,10)));
+          };
+        
+        tJSX.push(<div style={{textAlign: 'center',border:'solid',borderWidth:'thin',borderColor:'lightgray',padding:'10px'}} key={'confB'}>
+            <button 
+                key={'cButton'}
+                className="configButton" 
+                onClick={(e) => {
+                    dispatch(actionCreators.ShowConfig(!showConfig));
+                }}> 
+                <img src="cog.svg" alt='Configuration' height="24" width="24" style={{verticalAlign:'middle'}}></img>
+            </button>
+            <div>
+                <div key={'simpleCheck'} style={{display:'flex',width:'fit-content',height:'fit-content',margin:'auto'}}>
+                    <input name="varEnable"
[…]

### 2018-08-13 ui/src/detail.js

@@ -1,0 +1,131 @@
+import React, { Component } from 'react';
+import { connect } from 'react-redux';
+import './style.css';
+import { XYPlot, XAxis, YAxis, VerticalBarSeries} from 'react-vis';
+import { actionCreators, getURL } from './reducers';
+import './detail.css';
+
+const mapStateToProps = (state) => ({
+    showDetails: state.showDetails,
+    region: state.region,
+    years: state.years,
+    cityID: state.cityID,
+    details: state.details,
+  });
+
+ function requestDetails(cityID,region) {
+    return function (dispatch) {
+        return fetch(getURL.RegionDetails(cityID,region)).then(
+            details => {
+                details.json().then((d)=> {//promise of a promise. really.
+                    dispatch(actionCreators.UpdateDetails(d));
+                })
+            },
+            error => console.log('ERRORRRRR')
+        );
+    }
+}
+
+class Detail extends Component {
+    componentWillReceiveProps(nextProps){
+        let {region}=this.props;
+        let {dispatch}=this.props;
+        if (region!==nextProps.region)
+            dispatch(actionCreators.UpdateDetails(undefined))
+    }
+    render() {
+        let {region}=this.props;
+        let {details}=this.props;
+        let {showDetails}=this.props;
+        let {years}=this.props;
+        let {dispatch}=this.props;
+        let {cityID}=this.props;
+        let retJSX=[];
+
+        if ((cityID!==undefined)&&(region!==undefined)&&(details===undefined))
+        {
+            dispatch(requestDetails(cityID,region))
+            return(null);
+        }
+        if ((showDetails===undefined)||(showDetails===false)||
+            (region===undefined)||(cityID===undefined)||(details===undefined)||
+            ((details!==undefined)&&(details.length===0))){
+            return(null);
+        }        
+        if ((details!==undefined)&&(years!==undefined))
+        {
+            for (let year of years){
+                let tJSX=[];
+                tJSX.push(<p>{year}</p>);
+
+                if (details.hasOwnProperty(year)){
+                    details[year].sort((a,b)=> {
+                        if ((a.values.length===1)&&(b.values.length>1)){
+                            return(-1);
+                        }
+                        else{
+                            if ((a.values.length>1)&&(b.values.length===1)){
+                                return(1);
+                            }
+                            else{
+                                return(a.name>b.name);
+                            }
+                        }
+                    });
+                    for (let i=0;i<details[year].length;i++){
+                        let d=details[year][i];
[…]

### 2018-08-13 ui/src/glmap.js

@@ -1,0 +1,210 @@
+import React from 'react'
+import PropTypes from 'prop-types'
+import mapboxgl from 'mapbox-gl'
+import bbox from '@turf/bbox';
+import 'mapbox-gl/dist/mapbox-gl.css'
+import './glmap.css'
+
+mapboxgl.accessToken = 'pk.eyJ1IjoiZGlhc2YiLCJhIjoiY2pqc243cW9wNDN6NTNxcm1jYW1jajY2NyJ9.2C0deXZ03NJyH2f51ui4Jg';
+
+
+let Map = class Map extends React.Component {
+  map;
+    
+
+  static propTypes = {
+    data: PropTypes.object.isRequired,
+  };
+
+  componentDidUpdate() {
+    this.setFill();
+  }
+
+  componentWillUpdate(nextProps, nextState) {
+    let bounds;
+    if ((this.moving!==undefined)&&(this.moving)){
+      return;
+    }
+    // console.log(this.props.tids,nextProps.tids);
+    let arrayEQ=(a,b)=>{
+      if ((a===undefined)&&(b===undefined)){
+        return(true);
+      }
+      if ((a===undefined)||(b===undefined)){
+        return(false);
+      }
+      if (a.length!==b.length){
+        return(false)
+      }
+      for (let aa of a){
+        if (!b.includes(aa)){
+          return(false);
+        }
+      }
+      return(true);
+    }
+    // console.log(this.props.tids,nextProps.tids,arrayEQ(this.props.tids,nextProps.tids));
+    if (!arrayEQ(this.props.tids,nextProps.tids)){
+      if ((nextProps.tids!==undefined)&&(nextProps.tids.length>0)){
+        let nextTids={};
+        let tempGJ={type:'FeatureCollection',features:[]}
+        
+        for (let i=0;i<nextProps.tids.length;i++){
+          nextTids[nextProps.tids[i]]=0;
+        } 
+        for (let feat of nextProps.data.features){
+          if (nextTids.hasOwnProperty(feat.properties.tID)){
+            tempGJ.features.push(feat);
+          }
+        }
+        bounds=bbox(tempGJ);
+        this.map.fitBounds([[bounds[0],bounds[1]],
+                            [bounds[2],bounds[3]]]);         
+        this.moving=true;
+      }else{
+        if (nextProps.tids.length===0){
+          bounds=bbox(this.props.data);
+          this.map.fitBounds([[bounds[0],bounds[1]],
+                              [bounds[2],bounds[3]]]);         
[…]

### 2018-08-13 ui/src/graphTest.js

@@ -1,0 +1,51 @@
+import React, { Component } from 'react';
+import { connect } from 'react-redux';
+import './style.css';
+import './sankey.css';
+import { actionCreators, getColour, getClass,} from './reducers';
+import chroma from 'chroma-js';
+import {ForceGraph, ForceGraphNode, ForceGraphLink} from 'react-vis-force';
+
+
+const mapStateToProps = (state) => ({
+    graph: state.graph,
+    colours: state.colours,
+    years: state.years,
+    GraphLabels: state.GraphLabels,
+  });
+
+class GraphTest extends Component {
+    render() {
+        let {graph}=this.props;
+        let {colours}=this.props;
+        let {GraphLabels}=this.props;
+        let {years}=this.props;
+        let fixID=(ID)=>{
+            return(ID[0]+' '+ID[1]);
+        }
+
+        if ((GraphLabels===undefined)||(graph===undefined)||(colours===undefined)||(years===undefined)){
+            return(null)
+        }
+        else {
+            let retJSX=[]
+            console.log(GraphLabels);
+            for (let node of graph.nodes){
+                let c=getColour(colours,GraphLabels[+node.id[0]][node.id[1]].id);
+                let x=years.indexOf(node.year);
+                retJSX.push(<ForceGraphNode node={{ id: fixID(node.id), x:{x}, fixed:true }} fill={c} />)
+            }
+            for (let link of graph.links){
+                retJSX.push(<ForceGraphLink link={{ source: fixID(link.source), target: fixID(link.target) }} />)
+            }
+        return(<ForceGraph zoom simulationOptions={{ height: 1000, width: 1000 }}>            
+                {retJSX}
+               </ForceGraph>);
+
+        }
+
+
+    }
+}
+
+export default connect(mapStateToProps)(GraphTest);

### 2018-08-13 ui/src/index.js

@@ -1,0 +1,21 @@
+import React from 'react';
+import ReactDOM from 'react-dom';
+import './index.css';
+import App from './App';
+import registerServiceWorker from './registerServiceWorker';
+
+import { Provider } from 'react-redux'
+import { createStore, applyMiddleware } from 'redux'
+import thunk from 'redux-thunk';
+import { reducer } from './reducers'
+
+
+
+const store = createStore(reducer,applyMiddleware(thunk))
+const AppWithStore = (
+    <Provider store={store}>
+      <App />
+    </Provider>
+  )
+ReactDOM.render(AppWithStore, document.getElementById('root'));
+registerServiceWorker();

### 2018-08-13 ui/src/newmap.js

@@ -1,0 +1,59 @@
+import React, { Component } from 'react';
+import { connect } from 'react-redux';
+import { actionCreators, toInt} from './reducers';
+import Map from './glmap';
+
+
+const mapStateToProps = (state) => ({
+  gj: state.gj,
+  tID: state.tID,
+  level: state.level,
+  simpleColours: state.simpleColours,
+});
+
+
+class NewMap extends Component {
+  render() {
+    let {level,tID,simpleColours}=this.props;
+    let {dispatch}=this.props;
+    let {gj}=this.props;
+
+
+    let doHighlight = (d) => {
+      let dtid=toInt(d.features["0"].properties.tID);
+      dispatch(actionCreators.SetTID(dtid));
+    }
+
+    
+
+    if (gj!==undefined){
+      return (
+        <Map 
+          data={gj}
+          className='mainMap'
+          level={level}
+          paintProp={simpleColours?'simplified':'colour'}
+          onClick={doHighlight}
+          tids={tID}
+        />
+        //   <Control
+        //     position="bottomleft">
+        //     <div key='clearSels'>            
+        //      <button 
+        //         key={'clearBtn'}
+        //         disabled={!(((selClassYear!==undefined)&&(selClassYear.length>0))||(selTID>=0))}
+        //         onClick={(e) => {
+        //             dispatch(actionCreators.SelectClassYear(undefined));
+        //             dispatch(actionCreators.SetHighlightChain(-1));
+        //         }}>                 
+        //         Clear selection
+        //      </button>
+        //     </div>
+        //   </Control>
+      );
+    }
+    else
+      return(null);
+  }
+}
+export default connect(mapStateToProps)(NewMap);

### 2018-08-13 ui/src/overview.js

@@ -1,0 +1,219 @@
+import React, { Component } from 'react';
+import { connect } from 'react-redux';
+import { getColour, toInt,actionCreators} from './reducers';
+import { XYPlot, XAxis, YAxis } from 'react-vis';
+import './style.css';
+import './overview.css';
+import  BoxplotSeries  from './boxplot';
+import chroma from 'chroma-js';
+import CurConf from './curConf';
+
+const mapStateToProps = (state) => ({
+    colours: state.colours,
+    level: state.level,
+    patt: state.patt,    
+    traj: state.traj,
+  });
+
+class Overview extends Component {
+    constructor(props){
+        super(props);
+        this.state={'extended':[],'full':false};
+    }
+    render() {
+        
+        let {colours,dispatch}=this.props;
+        let {patt,traj}=this.props;
+        let {level}=this.props;
+        let retJSX=[];
+        let pattJSX=[];
+        let tJSX=[]
+        retJSX.push(<CurConf key='cc'/>);
+        let cScale=chroma.scale(chroma.brewer.Viridis).mode('lab').correctLightness().classes(5);            
+
+        if (patt!==undefined){
+            let cMax=[];
+            pattJSX=[];
+            let RedExt=(e)=>{
+                let c=toInt(e.target.getAttribute('data-cluster'));
+                let cExtended=this.state.extended.slice();
+                if (cExtended.includes(c)){
+                    cExtended.splice(cExtended.indexOf(c),1);
+                }else{
+                    cExtended.push(c);
+                }
+                this.setState({extended:cExtended.slice()});
+
+            }
+
+
+            for(let j=0;j<patt[0].variables.length;j++){
+                let tMax=0.1;
+                for(let i=0;i<patt.length;i++){
+                    for (let k=0; k<patt[i].variables[j].length;k++){
+                        tMax=Math.max(tMax,patt[i].variables[j][k].yMax);    
+                    }
+                }
+                cMax.push(tMax);
+            }
+            let sumImportance=((a)=>{
+                let s=0;
+                let val;
+                    for (let i=0;i<a.length;i++){
+                        val=a[i].importance[level];
+                        s=(s>val)?s:val;
+                }
+                return(s);
+            })
+            let selectTIDs=(d)=>{
+                let C=toInt(d.target.getAttribute('data-clusterid'));
+                let tids=[];
+    
+                    for (let i=0;i<traj.length;i++){
+                        for (let j=0;j<traj[i].chain.length;j++){
+                            if (traj[i].chain[j]===C){
[…]

### 2018-08-13 ui/src/reducers.js

@@ -1,0 +1,399 @@
+import chroma from 'chroma-js';
+import clone from '@turf/clone';
+
+const baseURL=()=>{
+    //return('http://142.1.190.188:8080/');
+    //return('http://localhost:8080/');
+    //return('http://138.197.139.94/ct/');//alpha
+    return('http://142.1.190.14/ct/');// clr
+}
+
+export const selModes={
+    SET:'Set',
+    ADD:'Add',
+    REMOVE: 'Remove'
+}
+
+export const types={
+    UPDATE_MAP:'UpdateMap',
+    UPDATE_DETAILS: 'UpdateDetails',
+    SET_GEOJSON:'SetGeoJson',
+    SET_SELMODE: 'SetSelectionMode',    
+    SET_CITY_OPTIONS: 'SetCityOptions',
+    SET_LEVEL: 'SetLevel',
+    SET_REGION: 'SetRegion',
+    SET_CITY: 'SetCity',
+    SHOW_CONFIG: 'ShowConfig',
+    SHOW_LOADING: 'ShowLoading',
+    SHOW_DETAILS: 'ShowDetails',
+    SET_TID: 'SetTID',
+    CLEAR_TID: 'ClearTID',
+    SET_PATH: 'SetPath',
+    SET_COLOUR_OPT: 'SetColourOption'
+};
+// Helper functions to dispatch actions, optionally with payloads
+export const actionCreators = {
+    SetColourOption: opt =>{
+        return {type: types.SET_COLOUR_OPT, payload: opt};
+    },
+    UpdateMap: segData => {
+      return { type: types.UPDATE_MAP, payload: segData };
+    },
+    SetPath: path => {
+      return { type: types.SET_PATH, payload: path};
+    },
+    SetTID: tID =>{
+      return { type: types.SET_TID, payload: tID};
+    },
+    ClearTID: () =>{
+        return { type: types.CLEAR_TID, payload: {}};
+    },
+    UpdateDetails: details => {
+        return {type: types.UPDATE_DETAILS, payload: details};
+    },
+    SetGeoJson: gj => {
+      return { type: types.SET_GEOJSON, payload: gj };
+    },
+    SetCity: cityID => {
+        return {type: types.SET_CITY, payload: cityID};
+    },
+    SetLevel: level => {
+        return {type: types.SET_LEVEL, payload:level};
+    },
+    SetRegion: rid => {
+        return {type: types.SET_REGION, payload: rid};
+    },
+    SetCityOptions: data => {
[…]

### 2018-08-13 ui/src/sankey.js

@@ -1,0 +1,138 @@
+import React, { Component } from 'react';
+import { connect } from 'react-redux';
+import './style.css';
+import { Sankey} from 'react-vis';
+import './sankey.css';
+import {  actionCreators, getColour } from './reducers';
+
+
+const mapStateToProps = (state) => ({
+    gj: state.gj,
+    years: state.years,
+    tID: state.tID,
+    traj: state.traj,
+    colours: state.colours,
+    labels: state.labels,
+    simpleColours: state.simpleColours,
+  });
+
+
+class TempEvo extends Component {
+    constructor(props) {
+        super(props);
+        this.state = {
+           min: 0, 
+           max: 100,
+        };
+      }
+
+    render() {
+        let {colours,simpleColours}=this.props;
+        let {dispatch}=this.props;
+        let {years}=this.props;
+        let {labels}=this.props;
+        let {traj}=this.props;
+        let selTID=this.props.tID;
+        
+        let {gj}=this.props;
+        let retJSX=[]
+        let nodes=[];
+        let links=[];
+
+        const insertGetInd = (k) => {
+            let ind=nodes.findIndex((element)=>{
+                return((element.key.year===k.year) && (element.key.id===k.id));
+            });
+            if (ind<0)
+            {
+                ind=nodes.length;
+                nodes.push({name:'', key:{...k,ind:ind,tID:[k.tID,]}, class:k.id, opacity: 1, color:getColour(colours,k.id)});
+            }else{
+                if (!nodes[ind].key.tID.includes(k.tID)){
+                    nodes[ind].key.tID.push(k.tID);
+                }
+                
+            }
+            return(ind);
+        };
+        const addLinkVal = (L,s,t,v,tID) =>{
+            let tColour;
+            if ((selTID.length>0) && (! selTID.includes(tID))){
+                tColour=traj[tID].faded;
+            }
+            else{
+                if (simpleColours){
+                    tColour=traj[tID].simplified;    
+                }else{
+                    tColour=traj[tID].colour;
+                }
+                
+            }
+            L.push({source:s,target:t,value:v, tID:tID, chain: traj[tID].chain, years: traj[tID].years, color: tColour});
+            return(L);
+        };
+
+        if ((labels!==undefined)&&(gj!==undefined) && (years!==undefined) && (colours!==undefined))
+        {        
+            
+            // let vMax=traj[0].value;
+            // let vMin=traj[0].value;
[…]

### 2018-08-13 ui/src/smmaps.js

@@ -1,0 +1,63 @@
+import React, { Component } from 'react';
+import { connect } from 'react-redux';
+import { actionCreators} from './reducers';
+import './style.css';
+import './trajectoryDetails.css';
+import Map from './glmap';
+
+
+const mapStateToProps = (state) => ({
+    level: state.level,
+    years: state.years,
+    colours: state.colours,
+    tID: state.tID,
+    gj: state.gj,
+    path: state.path,
+    population: state.population,
+    patt: state.patt,
+    traj: state.traj
+  });
+
+
+class SmMaps extends Component {
+    constructor(props){
+        super(props);
+        this.state={bounds:undefined};
+    }
+
+    render(){
+        let {gj,tID,years,dispatch,level}=this.props;
+        let rowJSX=[];
+      let RegionClick = (d) =>{
+          dispatch(actionCreators.ShowDetails(true));
+          dispatch(actionCreators.SetRegion(d.features["0"].properties.display_id))      
+        };
+      
+      for (let y of years){
+          if (gj!==undefined){
+              if (gj.features.length>0){
+                  rowJSX.push(
+                    <Map 
+                        key={'sm'+y}
+                        className={'mapSM'}
+                        data={gj}
+                        bbox={this.state.bounds}
+                        boundsCallback={(bb)=>{this.setState({bounds:bb});}}
+                        tids={tID}
+                        level={level}
+                        paintProp={y.toString()}
+                        onClick={RegionClick}
+                  />);
+              }
+              else{
+                  rowJSX.push(<div className="mapSM"></div>)
+              }
+          }             
+      }
+      return(<div className='mapRow'>{rowJSX}</div>);
+    }
+    
+}
+
+export default connect(mapStateToProps)(SmMaps);
+

### 2018-08-13 ui/src/trajectoryDetails.js

@@ -1,0 +1,373 @@
+import React, { Component } from 'react';
+import { connect } from 'react-redux';
+import { toInt, requestPath} from './reducers';
+import './style.css';
+import './trajectoryDetails.css';
+import { XYPlot, XAxis, YAxis, LineSeries, HorizontalGridLines, VerticalGridLines, AreaSeries  } from 'react-vis';
+import MappleToolTip from 'reactjs-mappletooltip';
+
+import SmMaps from './smmaps';
+
+const mapStateToProps = (state) => ({
+    level: state.level,
+    years: state.years,
+    colours: state.colours,
+    basedata: state.basedata,
+    tID: state.tID,
+    gj: state.gj,
+    path: state.path,
+    globalPath: state.globalPath,
+    population: state.population,
+    patt: state.patt,
+    cityID: state.cityID,
+    NbyTID: state.NbyTID,
+    traj: state.traj,
+    dsconf: state.dsconf
+  });
+
+let arrEQ=(A1,A2)=>{
+    if (A1.length!==A2.length){
+        return(false);
+    }
+    for (let a of A1){
+        if (!A2.includes(a)){
+            return(false);
+        }
+    }
+    return(true);
+}
+
+class TrajDet extends Component {
+    constructor(props){
+        super(props);
+        this.state={extended:[],value:undefined}
+    }
+    
+    componentWillUpdate(nextProps, nextState) {
+        let {dispatch,cityID,NbyTID,traj,dsconf,tID}=nextProps;
+        if ((nextProps.tID.length>0)&&(!arrEQ(this.props.tID,nextProps.tID))){
+            let tn=[];
+            for (let t of tID){
+                tn=tn.concat(NbyTID[traj[t]['tid']]);
+            }
+            dispatch(requestPath(cityID,dsconf.vars,dsconf.fs,tn));    
+        }
+    }
+
+    
+    render () {
+        let {path}=this.props;
+        let {tID}=this.props;
+        let {years}=this.props;
+        let {population}=this.props;
+        let {basedata,globalPath}=this.props;
+        let {gj}=this.props;
+        let retJSX=[];
+        let rowJSX=[];
+        let tJSX=[];
+        let imp={};
+        let FW=1300;
+
+        let RedExt=(e)=>{
+            let c=toInt(e.target.getAttribute('data-cluster'));
+            let cExtended=this.state.extended.slice();
+            if (cExtended.includes(c)){
+                cExtended.splice(cExtended.indexOf(c),1);
+            }else{
[…]

### 2018-08-13 ui/src/transMat.js

@@ -1,0 +1,232 @@
+import React, { Component } from 'react';
+import { connect } from 'react-redux';
+import './style.css';
+import './transMat.css'
+import { actionCreators, getColour, toInt } from './reducers';
+// import MappleToolTip from 'reactjs-mappletooltip';
+
+const mapStateToProps = (state) => ({
+    years: state.years,
+    tID: state.tID,
+    traj: state.traj,
+    colours: state.colours,
+  });
+
+class TransMat extends Component {
+    constructor(props){
+        super(props);
+        this.state={cYear:0};
+    }
+
+    render(){
+        let retJSX=[];
+        let tJSX=[];
+        let {traj,colours,years,dispatch,tID}=this.props;
+        let mat={};
+        let bolds={};
+        let selectTIDsSingle=(d)=>{
+            let C=toInt(d.target.getAttribute('data-clusterid'));
+            let tids=[];
+
+                for (let i=0;i<traj.length;i++){
+                    for (let j=0;j<traj[i].chain.length;j++){
+                        if (traj[i].chain[j]===C){
+                            tids.push(traj[i].tid);
+                            break
+                        }
+                    }
+                }
+            dispatch(actionCreators.SetTID(tids));
+        }
+
+
+        let selectTIDs=(d)=>{
+            let l1=toInt(d.target.getAttribute('data-from'));
+            let l2=toInt(d.target.getAttribute('data-to'));
+            let y=toInt(this.state.cYear);
+            let indY;
+            let tids=[];
+
+            if (y>0){
+                indY=years.indexOf(y);
+                let y2=years[indY+1];
+                for (let i=0;i<traj.length;i++){
+                    let j1=traj[i].years.indexOf(y);
+                    let j2=traj[i].years.indexOf(y2);
+                    if ((j1>=0)&&(j2>=0)){
+                        if ((traj[i].chain[j1]===l1)&&(traj[i].chain[j2]===l2)){
+                            tids.push(traj[i].tid);
+                        }
+                    }
+                }
+            }else{
+                for (let i=0;i<traj.length;i++){
+                    for (let j=0;j<(traj[i].chain.length-1);j++){
+                        if (y===-1){
+                            if ((traj[i].chain.length===years.length)&&(traj[i].chain[0]===l1)&&(traj[i].chain[years.length-1]===l2)){
+                                tids.push(traj[i].tid);
+                                break;
+                            }    
+                        }else{
[…]

### 2018-08-13 service-worker.js
Updates

@@ -1,0 +1,1 @@
+"use strict";var precacheConfig=[["/piccard/index.html","5aa99084ff95382f300a91041444441b"],["/piccard/static/css/main.70a8adca.css","c6ead2567bdb36ea0be062292ba1df52"],["/piccard/static/js/main.3c4bde20.js","eeb4e1a8017bc64af54d022626670aa1"]],cacheName="sw-precache-v3-sw-precache-webpack-plugin-"+(self.registration?self.registration.scope:""),ignoreUrlParametersMatching=[/^utm_/],addDirectoryIndex=function(e,t){var n=new URL(e);return"/"===n.pathname.slice(-1)&&(n.pathname+=t),n.toString()},cleanResponse=function(t){return t.redirected?("body"in t?Promise.resolve(t.body):t.blob()).then(function(e){return new Response(e,{headers:t.headers,status:t.status,statusText:t.statusText})}):Promise.resolve(t)},createCacheKey=function(e,t,n,r){var a=new URL(e);return r&&a.pathname.match(r)||(a.search+=(a.search?"&":"")+encodeURIComponent(t)+"="+encodeURIComponent(n)),a.toString()},isPathWhitelisted=function(e,t){if(0===e.length)return!0;var n=new URL(t).pathname;return e.some(function(e){return n.match(e)})},stripIgnoredUrlParameters=function(e,n){var t=new URL(e);return t.hash="",t.search=t.search.slice(1).split("&").map(function(e){return e.split("=")}).filter(function(t){return n.every(function(e){return!e.test(t[0])})}).map(function(e){return e.join("=")}).join("&"),t.toString()},hashParamName="_sw-precache",urlsToCacheKeys=new Map(precacheConfig.map(function(e){var t=e[0],n=e[1],r=new URL(t,self.location),a=createCacheKey(r,hashParamName,n,/\.\w{8}\./);return[r.toString(),a]}));function setOfCachedUrls(e){return e.keys().then(function(e){return e.map(function(e){return e.url})}).then(function(e){return new Set(e)})}self.addEventListener("install",function(e){e.waitUntil(caches.open(cacheName).then(function(r){return setOfCachedUrls(r).then(function(n){return Promise.all(Array.from(urlsToCacheKeys.values()).map(function(t){if(!n.has(t)){var e=new Request(t,{credentials:"same-origin"});return fetch(e).then(function(e){if(!e.ok)throw new Error("Request for "+t+" returned a response with status "+e.status);return clea
[…]

### 2018-08-13 ui/src/reducers.js
fixing for github

@@ -2,9 +2,7 @@ import chroma from 'chroma-js';
 import clone from '@turf/clone';
 
 const baseURL=()=>{
-    //return('http://142.1.190.188:8080/');
     //return('http://localhost:8080/');
-    //return('http://138.197.139.94/ct/');//alpha
     return('http://142.1.190.14/ct/');// clr
 }
 

### 2018-08-13 paper/method.tex
one past until LA

@@
     \centering 
-    \includegraphics[width=0.95\linewidth]{overview.pdf}
+    \includegraphics[width=0.975\linewidth]{overview.pdf}
     \caption{Overview of the proposed method. A graph is generated combining the
@@
 \subsection{User interface}
+\label{sec:ui}
 The initial interface is illustrated in Figure~\ref{fig:ui}. Since demographic
 data can be nuanced, with intricate interconnections, we decided against
-presenting the interface using synthetic data, considering instead data from the
-Chicago region between 1970 and 2010, using previous published studies as
-corroboration. This region is known for its entrenched racial divide and the
+validating the interface using a synthetic dataset, considering instead data
+from the Chicago region between 1970 and 2010, using previous published studies
+as corroboration. This region is known for its entrenched racial divide and the
 emergence of a \emph{'young urban'} population with a higher education
-level~\cite{Delmelle2016,Delmelle2017}. More details about this scenario are
-presented in Section~\ref{sec:study}.
+level~\cite{Delmelle2016,Delmelle2017}. More details about this dataset are
+presented in Section~\ref{sec:study}, with an workflow example in
+Figure~\ref{fig:chiWorkflow}.
 
@@
 The configuration panel, on top left in Figure~\ref{fig:ui}, displays which
-aspects were considered and their weights (following Equation~\ref{eq:dist}). It
-also includes other configuration options that can be altered without
-re-processing the data, such as the number of clusters and the color option. The
-gear button allows access to the other configuration options that do require
-further processing, such as changing location, aspects, and weights. This panel
-also includes the configuration of the selection mode for the trajectories,
-which allows the user to set, add, or remove the next selected trajectories to
-the current selection. This feature enables the analysis of complex sets of
+aspects were used and their weights (following Equation~\ref{eq:dist}). It also
+includes other configuration options that can be altered without re-processing
+the data, such as the number of clusters and the color option. The gear button
+allows access to the other configuration options that do require further
+processing, such as changing location, aspects, and weights. This panel also
+includes the configuration of the selection mode for the trajectories, which
+allows the user to set, add, or remove the next selected trajectories to the
+current selection. This feature enables the analysis of complex sets of
 trajectories.
@@
 trajectory identification, none of the experts consulted reported confusion.
-These simple changes allow the user to easily understand the composition of the
-cluster and how it relates to the others. Violin plots~\cite{hintze1998violin}
-could also be used, but with increased potential for cluttering the
-representation of the other clusters, hindering the comparisons.
+Indeed, one expert reported confusion regarding the grey rectangles that
+represent the IQRs for other distributions when they are not colored; when that
+variable is not relevant to the characterization of the cluster. These simple
+changes allow the user to easily understand the composition of the cluster and
+how it relates to the others. Violin plots~\cite{hintze1998violin} could also be
+used, providing more information about the shape of the distribution, but with
+increased potential for obstructing the representation of the other clusters.
 
-For instance, the boxplot that summarizes the purple cluster, detailed on
-Figure~\ref{fig:boxplot}, illustrates that this cluster is best defined by the
-proportion of the population with four or more years of college. The user can
-quickly see that this is relevant because the corresponding IQR is colored with
-the highest relevance present in the legend. It is also clear that, while this
-cluster includes CTs that have between 10\% to 90\% of people in this variable,
-approximately, half of them have about 60\% of the population with four or more
-years of college. Since all the other IQRs are well separated, this is a
-defining characteristic of this cluster. By clicking on the colored bar above
-the boxplot, the user can select all trajectories that contain this cluster at
-any point in time.
+For instance, the boxplot that summarizes the purple cluster illustrated in
+Figure~\ref{fig:ui}, detailed on Figure~\ref{fig:boxplot}, illustrates that this
+cluster is best defined by the proportion of the population with four or more
+years of college. The user can quickly see that this is relevant because the
+corresponding IQR is colored with the highest relevance present in the legend.
+It is also clear that, while this cluster includes CTs that have between 10\% to
+90\% of people in this variable, approximately, half of them have about 60\% of
+the population with four or more years of college. Since all the other IQRs are
+well separated, this is a defining characteristic of this cluster. Conversely,
+the proportion of the population with one to three years of college is not
+relevant, as indicated by black fill in the rectangle representing the IQR of
+this cluster, in overlapping position with the rectangles of the other clusters.
+By clicking on the colored bar above the boxplot, the user can select all
+trajectories that contain this cluster at any point in time.
 

### 2018-08-13 paper/study.tex
one past until LA

@@
 details of which census columns were used for each aspect. Income is slightly
-inaccurate, even though we did correct for inflation. We grouped the original ranges into
-three larger ranges, but they do not match precisely.
+inaccurate, even though we did correct for official inflation. We grouped the
+original ranges into three larger ranges, but they do not match precisely.
 
-% \subsection{Chicago}
-% \begin{figure*}
-%     \centering
-%  \includegraphics[width=0.85\linewidth]{newTeaser.pdf}
-%  \caption{Boxplot representation for some of the clusters found, characterized
-%  by White population (orange), high education (purple), and Black population
-%  (green). The trajectories plot represents the associated population, showing an
-%  increase in the purple cluster. The plot on the right compares the intra
-%  quantile ranges for \emph{Four or more years of College}. While there was an
-%  increase in the whole city (gray band), it was sharper in this
-%  region.\label{fig:teaser}}
-% \end{figure*}
+\subsection{Chicago}
+We selected a region loosely following the administrative borders. The
+demographic composition is well explored in the literature, with reports of
+racial divide and gentrification~\cite{Delmelle2016,Delmelle2017,Hwang2014}.
+While the definition of gentrification is still unclear and out of the scope of
+this paper, we associate gentrification with higher education and income levels.
 
-\subsection{Los Angeles}
-We selected a region around the metropolitan area of Los Angeles (LA). This
-region follows urban density, not administrative boundaries. The summary of the
-results using all aspects with equal weights and eight clusters is illustrated
-in Figure~\ref{fig:la}, while the statistical description of the clusters is
+The initial state of the prototype is illustrated in Figure~\ref{fig:ui}, and
+its findings are explained in Section~\ref{sec:ui}, where the racial divide is
+clear. Starting from this initial state, the specific workflow used to identify
+the existence and details of the gentrification process are illustrated in
+Figure~\ref{fig:chiWorkflow}. For the users, the first step is to identify the
+compositions of each cluster from the boxplots, so orange is associated with
+majority of White population, green with majority Black, and purple with higher
+proportion of four years of college or more (high education level). The expanded
+version of the boxplots for the purple cluster shows a higher income level and
+majority of occupations in administrative jobs, therefore the purple cluster
+identifies gentrified regions.
+
+The trajectories plot illustrates the process of gentrification, progressively
+absorbing regions from the orange cluster (White). This corroborates results
+from the literature reporting that Black neighborhoods are less likely to
+gentrify~\cite{Hwang2014}. Moreover, this process is unlikely to be reversed, as
+indicated by the limited number of trajectories leaving the purple stream. Next,
+we select the region that is gentrified in 2010, by clicking on the
+corresponding rectangle in the trajectories plot, updating the information on
+the maps and the details portion of the interface.
+
+The corresponding CTs are highlighted in the maps, where the spatial pattern is
+clear, corresponding exactly to previous findings in the
+literature~\cite{Hwang2014}. Further, we can also identify the regions that
+gentrified earlier, with an stronger purple hue, compared to newer regions,
+where the orange and green colors are still present. This also indicates from
+which clusters they belonged before gentrifying. In the details portion of the
+interface, the order of the aspects was updated to reflect the order of
+relevance considering only the selected region. The most relevant aspect is the
+Education, specifically "Four or more years of college", illustrated in the
+rightmost portion of Figure~\ref{fig:chiWorkflow}, which is increasing for the
+whole city (grey band), but faster and to a higher level in this region (black
+band). Indeed, while the IQR for the city goes from 11\% to 45\%, the IQR for
+this region goes from 55\% to 77\%. However, there are portions of this region
+with significantly lower or higher proportions, as indicated by the dotted black
+lines representing the minimum and maximum for the selected region.
+
+\begin{figure*}
+    \centering
+ \includegraphics[width=0.85\linewidth]{newTeaser.pdf}
+ \caption{Worflow to discover gentrification in Chicago: the purple cluster
+ corresponds to high education / income. Its population is increasing over time,
+ absorbing from the majority White cluster (orange). By selecting the purple
+ cluster in 2010, the region is highlighted in the maps. The proportion of
+ people with 4+ years of college is increasing in the whole city (grey IQRs),
+ but significantly more in this region (black).\label{fig:chiWorkflow}}
+\end{figure*}
+
+\subsection{Toronto}
+
+We considered a region that is approximately the administrative border of the
+current city of Toronto and all seven available aspects with equal weights. The
+results are summarized in Figure~\ref{fig:to}, considering eight clusters.
+
+\begin{figure}
+    \centering 
+    \includegraphics[width=0.95\linewidth]{to.pdf}
+    \caption{Clustering results for Toronto, with eight clusters, including
+    clusters representing Jewish population, high and low income, low education,
+    and Asian immigration.\label{fig:to}}
+\end{figure}
+
+The population with low percentage of University degrees is represented in
+orange, mostly anglophone population in green, Asian immigrants in yellow, high
+percentage of income in the highest bracket in purple, high percentage of Jewish
+people in light green and brown, high percentage of Eastern Non-Christian
+religion in pink, and high concentration of single people in dark gray. From the
+trajectories plot, we can see that Toronto is more dynamic than Chicago, with
+one cluster constantly shrinking. In the 1970s, the city was divided into four
+clusters: low number of university degrees, Jewish population, majority
+anglophones, and high income. Interestingly, the more recent clusters that
+absorbed regions from the orange cluster have similar education profiles and are
+differentiated by other aspects. In this sense, the city is growing diverse,
+changing from a common low education profile to a higher level of education with
+more diversity in religion (pink) and immigration (yellow).
+
+Indeed, the influx of Asian population is visible starting in the 1980s and
+building thereafter, leading to the yellow and pink clusters. While both include
+a high percentage of people born in Asia, the pink is more defined by religion,
+with low percentage of university degrees, and contains the lowest percentage of
+people in the highest income bracket for these clusters; the yellow is less
+defined by religion, and has higher education and income, geographically
+corresponding to the Markham region, know for its Chinese population. A similar
+division also happens for the two Jewish clusters, where the light green cluster
+has lower education and income levels than the brown cluster. The purple cluster
+of high income is somewhat stable. This cluster includes the Bridle Path
+neighborhood, known for its wealthy population, until 2011. In 2011 it was
+classified into the yellow cluster of Asian immigration, since about 35\% of the
+population for this CT were then born in Asia. The income distribution did not
+change, with 85\% of the population with an income of 90k CAD or more.
+
+The most significant indicator of Toronto's dynamism is the presence of grey
+regions on the simplified color map; representing regions associated to three or
+more clusters over this five census period. Using the 'Add' mode for the
+trajectory selection, we select their trajectories, and a subset of the details
+is illustrated in Figure~\ref{fig:toVol}. These regions account for about 5\% of
+Toronto's population. The whole region was classified into the orange cluster in
+1971 (low level of university degrees). By 1991, most of the region was
+classified into the green cluster, representing anglophone population, mostly
+Canadian born, with a higher level of education. As the corresponding plot
+indicates, this trend in increasing education is city-wide, but this region has
+people with better education than most.
+
+\begin{figure}
+    \centering 
+    \includegraphics[width=0.95\linewidth]{toVol.pdf}
+    \caption{Details for some regions of Toronto that were classified into 3 or
+         more clusters over time.\label{fig:toVol}}
+\end{figure}
+
+In 2001, the purple cluster of high income annexes neighboring parts of the
+volatile region, and the Asian born population increases sharply, as illustrated
+by the appearance of the yellow cluster.  This cluster indicates well educated,
+higher income, and about 30\%-50\% Asian born population. By 2011, the yellow
+cluster increased considerably, annexing parts of the high income purple
+cluster, including the neighboring Bridle Path area.
+
+\subsection{Los Angeles}%TODO HERE
+We selected a region around the metropolitan area of Los Angeles (LA), following
+urban density, not administrative boundaries. The summary of the results using
+all aspects with equal weights and eight clusters is illustrated in
+Figure~\ref{fig:la}. The full statistical description of the clusters is
 illustrated in Figure~\ref{fig:laAll}, where the most relevant aspect of each
@@
 
-\subsection{Toronto}
-
-We considered a region that is approximately the administrative border of the
-current city of Toronto and all seven available aspects. The results are
-summarized in Figure~\ref{fig:to}, considering eight clusters.
-
-\begin{figure}
-    \centering 
-    \includegraphics[width=0.95\linewidth]{to.pdf}
-    \caption{Results for Toronto, with eight clusters.\label{fig:to}}
-\end{figure}
-
-The population with low percentage of University degrees is represented in
-orange, mostly anglophone population in green, Asian immigrants in yellow, high
-percentage of income in the highest bracket in purple, high percentage of Jewish
-people in light green and brown, high percentage of Eastern Non-Christian
-religion in pink, and high concentration of single people in dark gray.
-
-From the trajectories plot, we can see that Toronto is more dynamic than both
-Chicago and LA, with one cluster constantly shrinking, and others growing more
-prominent. In the 1970s, the city was divided into four clusters: low number of
-university degrees, Jewish population, majority anglophones, and high income,
-with the latter being rather small. There is some variance in the other
-characteristics of the orange cluster, but none as significant as the education
-level, especially when compared to the other years. While the orange cluster
-persists until 2011, the other clusters present have similar education profiles
-to one another, leading the clustering method to use other aspects to
-characterize them.
-
-Toronto's dynamism also appears in that we see the influx of Asian population
-starting in the 1980s and building thereafter, as represented by the yellow and
-pink clusters. While both include a high percentage of people born in Asia, the
-pink region is more defined by religion, with low percentage of university
-degrees, and contains the lowest percentage  of people in the highest income
-bracket for these clusters; the yellow is less defined by religion, and has
-higher education and income, geographically corresponding to the Markham region,
-know for its Chinese population. A similar division also happens for the two
-Jewish clusters, where the light green cluster has lower education and income
-levels than the brown cluster. The purple cluster of high income is somewhat
-stable. This cluster includes until 2011 the Bridle Path neighborhood, known for
-its wealthy population.  In 2011, however, it was classified into the yellow
-cluster that includes Asian immigration, since about 35\% of the population for
-this CT were born in Asia.  It maintained its income profile, with 85\% with 90k
-CAD or more.
-
-The grey regions of the map are another indication of Toronto's volatility over
-this period. While there is a cluster that is painted dark grey, annotated by a
-red circle in Figure~\ref{fig:to}, the light gray regions on the north part of the
-city indicate volatility. Here the same region was classified into three or more
-clusters for this five census period. Using the 'Add' mode for the trajectory
-selection, we select these volatile regions. A subset of the details is
-illustrated in Figure~\ref{fig:toVol}.
-
-\begin{figure}
-    \centering 
-    \includegraphics[width=0.95\linewidth]{toVol.pdf}
-    \caption{Details for some regions of Toronto that were classified into 3 or
-         more clusters over time.\label{fig:toVol}}
-\end{figure}
-
-These regions account for about 5\% of Toronto's population. The whole region
-was classified into the orange cluster in 1971 (low level of university degrees).
-By 1991, most of the region was classified into the green cluster, representing
-anglophone population, mostly Canadian born, with a higher level of education.
-As the corresponding plot indicates, this trend in increasing education is
-city-wide, but this region has people with better education than most.
-
-In 2001, the purple cluster of high income annexes neighboring parts of the
-volatile region, and the Asian born population increases sharply, as illustrated
-by the appearance of the yellow cluster.  This cluster, once again, indicates
-well educated, higher income, and about 30\%-50\% Asian born population. By
-2011, the yellow cluster increased considerably, annexing parts of the high
-income purple cluster, including the neighboring Bridle Path area.
-
 While Toronto is more dynamic than Los Angeles, possibly due to size
-differences, the volatile regions shown in Figure~\ref{fig:toVol} did not change as
-quickly or dramatically as the ones shown in Figure~\ref{fig:laVol}, which involved
-twice as many people. We found this trend to be related to the countries
-themselves, Canadian cities have larger areas undergoing slow, gradual changes,
-whereas American cities have more general stability, but quicker changes in
-smaller scales. The supplementary material contains brief summaries of all the
-regions accessible in this prototype.
+differences, the volatile regions shown in Figure~\ref{fig:toVol} did not change
+as quickly or dramatically as the ones shown in Figure~\ref{fig:laVol}, which
+involved twice as many people. We found this trend to be related to the
+countries themselves, Canadian cities have larger areas undergoing slow, gradual
+changes, whereas American cities have more general stability, but quicker
+changes in smaller scales. The supplementary material contains brief summaries
+of all the regions accessible in this prototype.
 

### 2018-08-15 paper/discussion.tex
missing the conclusions

@@
 across the years. Matching the fields of the public census can be trivial for
-some aspects, like Age, but challenging for others, like Income. The divulged
-income ranges vary over time, the actual value changes due to inflation and
-other factors, and so on. Moreover, some fields were not considered in earlier
-censuses, such as Race in Canada, or Hispanic population in the USA, hampering its
-use when they are available. We matched some aspects, but a deeper demographic
-analysis would greatly benefit from all available information.
+some aspects (Age), but challenging for others (Income). The divulged income
+ranges vary over time, the actual value changes due to inflation and other
+factors, and so on. Moreover, some fields were not considered in earlier
+censuses, such as Race in Canada, or Hispanic population in the USA, hampering
+its use when they are available. Since this is only a prototype, we matched few
+aspects, but a deeper demographic analysis would greatly benefit from all
+available information.
+
+The limitation on the number of displayed clusters because of the limited number
+of distinguishable colors was significant. While increasing the number of
+clusters would further complicate an already complex analysis, it might be
+warranted for some regions. Color is a fundamental and intuitive tool for
+information representation that can be coherently used across different plots,
+so we opted to use it, even if in a limited way. With eight colors, there was
+overlap between some clusters, the relevance gradient, and the color
+combination.
 

### 2018-08-15 paper/intro.tex
one past until LA
missing the conclusions

@@
 some measurements only make sense over a defined area. Census data is a classic
-example of this format, with datasets available from 1790 onwards for the
+example of this format, with datasets available from 1790 onward for the
 US~\cite{nhgis}. Despite this unmatched temporal availability, longitudinal
@@
 while one is experiencing a process of economic improvement and the other one
-impoverishment. Quite obviously, a single snapshot cannot be used to identify
-gentrification, migration, education changes, or any of the relevant processes
-that happen over time.
+impoverishment. A single snapshot cannot be used to identify gentrification,
+migration, education changes, or any of the relevant processes that happen over
+time.
 
@@
 include illustrative scenarios and our prototype is available at
-\url{http://uoft.me/piccard}, including more than fourty regions in the US and
-Canada.
-
-% We adopt census data as familiar operational
-% context, but the proposed methodology is suitable for most region based
-% measurements.
+\url{http://uoft.me/piccard}, including more than forty regions in the US and
+Canada. The source code is publicly available at
+\url{https://github.com/fabioasdias/piccard}.

### 2018-08-15 paper/method.tex
missing the conclusions

@@
 but non-contiguous, sets of CTs will be classified into two different clusters,
-which can be counterintuitive. To overcome this issue, we \emph{augment} the
+which can be counter-intuitive. To overcome this issue, we \emph{augment} the
 graph with two new edges per node from a nearest neighbors
@@
 more of these content based edges could be used to decrease the impact of the
-spatio-temporal edges, controling the balance between content and topology in
+spatio-temporal edges, controlling the balance between content and topology in
 the result. In practice, the effect is dependent on the data itself, and the
 results are not consistent, or predictable, across different cities. We fixed it
-at two edges because it was the lowest number that empircally led to consistent
+at two edges because it was the lowest number that empirically led to consistent
 clusters, but we believe that this idea warrants further investigation, as an
@@
 that information is available at the larger map. All the maps are linked with
-syncronized navigation, and the use of small multiples allows the exploration of
+synchronized navigation, and the use of small multiples allows the exploration of
 each temporal census individually, and its comparison to the others, with

### 2018-08-15 paper/related.tex
missing the conclusions

@@
 
-Since our problem emcompasses several fields, we divided this section into
-specific subproblems: \emph{longitudinal demographic studies}, describing the
+Since our problem encompasses several fields, we divided this section into
+specific sub problems: \emph{longitudinal demographic studies}, describing the
 traditional approach to perform longitudinal studies; \emph{data
@@
 Census data is used not only to discover demographic
-patterns~\cite{Firebaugh2016}, but to correlate demographic charateristics to
+patterns~\cite{Firebaugh2016}, but to correlate demographic characteristics to
 other measurements~\cite{diez1997neighborhood}. However, longitudinal studies
@@
 countries~\cite{Liu2015,Lee2015,Allen2018}. This result was significant for the
-field, but it also restricts the useable data, since new datasets need to be
+field, but it also restricts the usable data, since new datasets need to be
 similarly processed.
@@
 imagery. Beyond the increased spatial accuracy, this approach does not require
-complex harmonizations when new data is considered. However, demographic data is
+complex harmonization when new data is considered. However, demographic data is
 usually not available in this format, especially from older sources, and the
@@
 would be the watershed cuts algorithm~\cite{Cousty2009}, inspired by the well
-known image processing segmentation and equally prone to oversegmentation.
+known image processing segmentation and equally prone to over segmentation.
 Considering that the processing time is also a relevant factor, we opted for an
@@
 problem~\cite{monmonier1990strategies,andrienko2003exploratory,ferreira2015visual,Zheng2016}.
-Most geographic data is naturally bidimensional and maps work well in this
+Most geographic data is naturally bi dimensional and maps work well in this
 case~\cite{Zheng2016,ward2015interactive}, but the temporal dimension cannot be

### 2018-08-15 paper/study.tex
missing the conclusions

@@
  \includegraphics[width=0.85\linewidth]{newTeaser.pdf}
- \caption{Worflow to discover gentrification in Chicago: the purple cluster
+ \caption{Workflow to discover gentrification in Chicago: the purple cluster
  corresponds to high education / income. Its population is increasing over time,
@@
 
-\subsection{Los Angeles}%TODO HERE
+\subsection{Los Angeles}
 We selected a region around the metropolitan area of Los Angeles (LA), following
-urban density, not administrative boundaries. The summary of the results using
-all aspects with equal weights and eight clusters is illustrated in
-Figure~\ref{fig:la}. The full statistical description of the clusters is
-illustrated in Figure~\ref{fig:laAll}, where the most relevant aspect of each
-cluster is highlighted. From the trajectories plot, we can see that there is a
-large but shrinking cluster, depicted in green, one increasing cluster in
-purple, an almost constant orange cluster, a smaller but increasing pink
-cluster, and three other small clusters. The corresponding map illustrates where
-these clusters are located, and that they are somewhat geographically stable,
-with some movement between the green, orange, and purple clusters.
+urban density. The summary of the results using all aspects with equal weights
+and eight clusters is illustrated in Figure~\ref{fig:la}. The full statistical
+description of the clusters is illustrated in Figure~\ref{fig:laAll}, where the
+most relevant aspect of each cluster is highlighted. From the trajectories plot,
+we can see that there is a large but shrinking cluster, depicted in green, one
+increasing cluster in purple, an almost constant orange cluster, a smaller but
+increasing pink cluster, and three other small clusters. The corresponding map
+illustrates where these clusters are located, and that they are somewhat
+geographically stable, with some movement between the green, orange, and purple
+clusters.
 
@@
 
-From Figure~\ref{fig:laAll}, we can see that the green cluster is characterized by
-a high percentage of White population, low percentage of population in the
+From Figure~\ref{fig:laAll}, we can see that the green cluster is characterized
+by a high percentage of White population, low percentage of population in the
 lowest income bracket, mostly administrative occupations, and about 30\% of the
@@
 cluster is characterized by all its population in the lowest income bracket, low
-education level, with a majority of White population.
+education level, with a majority of White population. Since the extremes in the
+boxplots of the grey cluster are not significantly different, we can also
+surmise that this cluster is either small or homogeneous.
 
@@
 
-Further, the larger clusters, green, orange, and purple, present a significant
-intra-cluster variance in most variables, as indicated by the maximum and
-minimum whiskers of the boxplots. For instance, while fifty percent of the CTs
-in the green cluster have between 20\% and 40\% of people in the lowest income
-bracket, that cluster also includes CTs where none and all the population
-belongs to that bracket. This might indicate that this cluster could be further
-divided, and currently represents different groups of people that are not
-different enough to be separated into clusters in this level of the hierarchy.
+The  green, orange, and purple clusters present a significant intra-cluster
+variance in most variables, as indicated by extreme whiskers of the boxplots.
+While fifty percent of the CTs in the green cluster have between 20\% and 40\%
+of people in the lowest income bracket, that cluster also includes CTs where
+none and all the population belongs to that bracket. This might indicate that
+this cluster represents different groups of people that are not different enough
+to be separated at this level of the hierarchy. Conversely, the light green,
+brown, and dark gray clusters are different enough to be separated into their
+own clusters at this level, despite being small and ephemeral, including only a
+few CTs.
 
-While the larger clusters are expected, the light green, brown, and dark gray
-clusters are more surprising. These clusters are small and ephemeral, including
-only a few CTs, and are peculiarly different enough to be separated into their
-own clusters, at this level of the hierarchy. While these small clusters might
-be interesting enough to warrant further study, they might suppress larger, but
-more subtle, clusters from appearing in this interface,  due to the limited
-number of clusters.
+% While they might be interesting enough to
+% warrant further study, they might suppress larger, but more subtle, clusters
+% from appearing in this interface due to the limited number of clusters.
 
@@
 
-While Toronto is more dynamic than Los Angeles, possibly due to size
+\smallskip While Toronto is more dynamic than Los Angeles, possibly due to size
 differences, the volatile regions shown in Figure~\ref{fig:toVol} did not change
@@
 To assess the proposed method, we contacted academic and industry experts in
-sociology and urban sciences. The supplementary material contains all our
-communication with five experts, identified from A to E.
+sociology and urban sciences. The supplementary material contains the
+communication with five experts, verbatim, identified from A to E.
 
-Their response was positive,  mentioning that the prototype allows them to
-analyze census data without the additional work of obtaining and cleaning the
+Their overall response was positive,  mentioning that the prototype allows them
+to analyze census data without the additional work of obtaining and cleaning the
 data (A, B, E), and it allows the inclusion of geographic visual analysis tools
@@
 highlighted how the method can be used to compare neighborhoods and cities (A).
-These remarks attest that we satisfied the design requirements regarding the
-existing demographic groups and their changes.
+In summary, the proposed methodology can be a viable alternative for the visual
+analytics of evolving demographic data.
 
@@
 side-effects of our effort to increase  representational accuracy, where we
-avoided using simplified labels. Specifically, while identifying clusters by
-their most relevant variables was welcome, the overlap of information from
-different clusters in the boxplot was "a bit confusing" (C). While the map of
-trajectories was mentioned as a "good summary map", how it related to the
-clustering method was unclear (C). While we provide different options on how the
-colors are used on this map, both are sub optimal, reliably representing several
-distinct entities using colors is still an open problem in visualization.
+avoided using simplified representation or labels. Identifying clusters by their
+most relevant variables was welcome, but the overlap of information from
+different clusters in the boxplot was "a bit confusing" (C) when color is not
+present. Further, most clusters can be sufficiently characterized using only the
+most relevant aspect, but this is not generally true.
+
+While the map of trajectories was mentioned as a "good summary map", how it
+related to the clustering method was unclear (C). While we provide different
+options on how the colors are used, both are sub optimal since reliably
+representing several distinct entities using colors is humanly unfeasible.
+Indeed, the number of distinguishable colors was a significant constraint, we
+found indications that more clusters should be used in some cases, even if eight
+clusters is more than what is traditionally considered in these analyses.
+Conversely, increasing the number of clusters would also complicate the
+interpretation of the results.
 
@@
 current implementation can take a few minutes to cluster regions with high
-number of CTs, like Los Angeles or Brooklyn.
+number of CTs, like Los Angeles or Brooklyn. Server-processing reduced the
+amount of data transfered to client, but might increase response time under
+load. We implemented a cache policy that greatly reduces the response time, but
+fully pre-processing the results is not practical due to size of the parameter
+space.
 
@@
 inquired if they found interesting insights, leading them to consider the
-interface, its options, and results as a whole. Therefore the method was
-neutrally compared against other, traditionally normalized results. We interpret
-the fact that most of them were interested in the next steps as an affirmation
-of the accuracy of the method.
+interface, its options, and results as a whole. Most experts found phenomena
+corroborated by the specialized literature, indicating that our methodology
+produces equivalent results, with a fraction of the effort. We interpret the
+fact that most of them were interested in the next steps as an affirmation of
+the accuracy of the method.

### 2018-08-15 sup/sup.tex
one past until LA
missing the conclusions

@@
 
-% We sent feedback requests to several researchers involved in Sociology and Urban
-% sciences. Not all answered before the submission deadline, but we will keep an
-% updated version of this document at:
-
-% \href{http://ug.daniels.utoronto.ca/~diasf/CTEvo/sup.pdf}{\url{http://ug.daniels.utoronto.ca/~diasf/CTEvo/sup.pdf}}
-
-% Last edit: \today
-
 \subsection{Original request}
@@
 \subsection{A - Cary Wu}
-\href{https://soci.ubc.ca/persons/cary-wu/}{\url{https://soci.ubc.ca/persons/cary-wu/}}
-% \subsection{Expert A}
-\emph{Doctoral candidate in Sociology, UBC.}
+%\href{https://soci.ubc.ca/persons/cary-wu/}{\url{https://soci.ubc.ca/persons/cary-wu/}}
+\emph{Doctoral candidate in Sociology at the University of British Columbia, Canada.}
 
@@
 % \href{https://scholar.harvard.edu/ethanfosse}{\url{https://scholar.harvard.edu/ethanfosse}}
-% \subsection{Expert B}
-\emph{Research Associate in  Sociology, Harvard}
+\emph{Research Associate at Princeton University, USA}
 
@@
 \subsection{C - Fernando Calderón Figueroa}
-\emph{Sociology Department, University of Toronto}
+\emph{Doctoral candidate in Sociology at the University of Toronto, Canada}
 
@@
 
-\subsection{E - James Murdoch}%, 
-\emph{Policy Analyst, 2M Research}
+\subsection{E - James Murdoch}
+\emph{Policy Analyst at 2M Research, USA}
 

### 2018-08-16 paper/study.tex
almost done expert

@@
 To assess the proposed method, we contacted academic and industry experts in
-sociology and urban sciences. The supplementary material contains the
-communication with five experts, verbatim, identified from A to E.
+sociology and urban sciences. We focused our inquires on the utility of the
+prototype, and if they found anything interesting on the data using it. The
+supplementary material contains the communication with five experts, verbatim,
+identified from A to E.
 
@@
 
-While the interface was "easy to navigate" (B), it was also considered
+The interface was "easy to navigate" (B), but it was also considered
 "overwhelming" (A), "intimidating" (E), and "tricky to interpret" (C), possible
@@
 most relevant variables was welcome, but the overlap of information from
-different clusters in the boxplot was "a bit confusing" (C) when color is not
+different clusters in the boxplot was "a bit confusing" (C) when color was not
 present. Further, most clusters can be sufficiently characterized using only the
@@
 While the map of trajectories was mentioned as a "good summary map", how it
-related to the clustering method was unclear (C). While we provide different
-options on how the colors are used, both are sub optimal since reliably
+related to the clustering method was unclear (C). The methos includes different
+options on how the colors are used, but both are sub optimal since reliably
 representing several distinct entities using colors is humanly unfeasible.
@@
 number of CTs, like Los Angeles or Brooklyn. Server-processing reduced the
-amount of data transfered to client, but might increase response time under
-load. We implemented a cache policy that greatly reduces the response time, but
-fully pre-processing the results is not practical due to size of the parameter
-space.
+amount of data transfered to client, but it might increase the response time
+under load. We implemented a cache policy that greatly improved the performance,
+but fully pre-processing the results is not practical due to size of the
+parameter space.
 
@@
 providing demographic context. They also mentioned the need to export subsets of
-data, plots, and maps to be used in reports and publications (C, D, E).
-
-More importantly, while these experts were aware that our method does not
-perform geographic normalization, none of them mention it. We did not
-specifically ask if this difference led to unexpected results, but rather
-inquired if they found interesting insights, leading them to consider the
-interface, its options, and results as a whole. Most experts found phenomena
-corroborated by the specialized literature, indicating that our methodology
-produces equivalent results, with a fraction of the effort. We interpret the
-fact that most of them were interested in the next steps as an affirmation of
-the accuracy of the method.
+data, plots, and maps to be used in reports and publications (C, D, E). More
+importantly, while these experts were aware that our method does not perform
+geographic harmonization, none of them mention it. We did not specifically ask
+if this difference led to unexpected results, but rather if they found
+interesting insights. Most experts found phenomena corroborated by the
+specialized literature, indicating that our methodology produces equivalent
+results, with a fraction of the effort. We interpret the fact that most of them
+were interested in the next steps as confirmation of the accuracy of the method.

### 2018-08-16 paper/discussion.tex
beta?

@@
 
-The objective of this work was to leverage a graph based data representation
-with visual tools to allow for the exploration of geographically inconsistent
-census data. While we successfully replicated and corroborated results from the
-literature, this method still has significant limitations.
+Our objective was to leverage a graph based data representation and
+visualization methods for the exploration of geographically inconsistent
+region-based data. While we successfully replicated and corroborated results
+from the literature, this method still has significant limitations.
 
-Removing the geographical normalization/interpolation step greatly reduces the
-amount of work necessary, but the method still requires consistent variables
-across the years. Matching the fields of the public census can be trivial for
+Removing the need for geographical harmonization greatly reduces the amount of
+work necessary to explore demographic data, but the method still requires
+consistent variables across the years. Matching the variables can be trivial for
 some aspects (Age), but challenging for others (Income). The divulged income
-ranges vary over time, the actual value changes due to inflation and other
-factors, and so on. Moreover, some fields were not considered in earlier
-censuses, such as Race in Canada, or Hispanic population in the USA, hampering
-its use when they are available. Since this is only a prototype, we matched few
-aspects, but a deeper demographic analysis would greatly benefit from all
-available information.
+ranges vary over time and the actual values change due to inflation. Some
+variables were not considered in earlier censuses, such as Race in Canada, or
+Hispanic population in the USA, hampering its use when they are available. Since
+this is only a prototype, we matched few aspects, but a proper demographic
+analysis would benefit from all available information.
+
+While using one small map for each year leads to an easier visualization that
+does not require interaction, it does not scale if more than five or six years
+are considered. In this case, it might be interesting to replace the larger map
+considering each year individually, along with a temporal control for
+navigation. Indeed, including more years would likely lead  to a stronger
+mixture of colors in the trajectory map, leading to a predominantly grey hue.
 

### 2018-08-16 paper/method.tex
beta?

@@
 While the partition of the data into different clusters helps the user to
-understand what demographic groups exist and where they are, we are also
-interested in the changes in those groups. To this end, we introduce the concept
-of \emph{trajectories}, which are composed by regions that are classified into
-the same sequence of clusters over the considered period. This enables direct
-access to regions that evolved in the same manner. While the interface provides
-access to data by census tract, the trajectories are the main unit of
-exploration in this work.
+understand what groups exist and where they are, we are also interested in the
+evolution of these groups. We introduce the concept of \emph{trajectories},
+composed by regions classified into the same sequence of clusters over the
+considered period. This enables direct access to regions that evolved in the
+same manner. While the interface provides access to data by individual census
+tract, the trajectories are the main unit of exploration in this work.
 
@@
 level~\cite{Delmelle2016,Delmelle2017}. More details about this dataset are
-presented in Section~\ref{sec:study}, with an workflow example in
-Figure~\ref{fig:chiWorkflow}.
+presented in Section~\ref{sec:study}.
 
@@
 The trajectories overview aims to convey basic information about the
-trajectories present in the data, where they are, and what changes are involved.
-This is done using three sub panels. The first, on the left, contains a Sankey
-diagram illustrating the evolution of the clusters over time. The widths are
-proportional to the population involved, the colors follow a policy detailed in
+trajectories, where they are, and what changes are involved. This is done using
+three sub panels. The first, on the left, contains a Sankey diagram illustrating
+the evolution of the clusters over time. The widths are proportional to the
+population involved, the colors follow a policy detailed in
 Section~\ref{sec:colors}. A stacked graph could also be used to represent the
@@
 
-In our example in Figure~\ref{fig:ui}, we can see that the total population of
-Chicago is decreasing. Additionally, the orange and green clusters contain most
-of the population and are fairly stable over time. The pink cluster is small and
-mostly stable. The purple cluster is increasing, mostly by incorporating areas
-that were previously orange. Since the purple group corresponds to the emergent
+In our example in Figure~\ref{fig:ui}, the total population of Chicago is
+decreasing. Additionally, the orange and green clusters contain most of the
+population and are fairly stable over time. The pink cluster is small and mostly
+stable. The purple cluster is increasing, mostly by incorporating areas that
+were previously orange. Since the purple group corresponds to the emergent
 'young urban' group, this corroborates the findings of
@@
 
-The next sub panel, in the top middle of Figure~\ref{fig:ui}, is a transition
-matrix between the clusters. It indicates the percentage of the population whose
-area changed in each pattern. This kind of table can be found in the related
-literature~\cite{Delmelle2016}, so it is familiar to the advanced users, and it
-not only informs the proportional changes, but allows the selection of the
-corresponding trajectories for further analysis.
+In the next panel, illustrated in the top middle of Figure~\ref{fig:ui}, is a
+transition matrix between the clusters. It indicates the percentage of the
+population whose area changed between each pair of clusters. This kind of table
+can be found in the related literature~\cite{Delmelle2016}, so it is familiar to
+the advanced users. It not only informs the proportional changes, but allows the
+selection of the corresponding trajectories for further analysis.
 

### 2018-08-16 paper/study.tex
beta?

@@
 While the map of trajectories was mentioned as a "good summary map", how it
-related to the clustering method was unclear (C). The methos includes different
+related to the clustering method was unclear (C). The methods includes different
 options on how the colors are used, but both are sub optimal since reliably
@@
 number of CTs, like Los Angeles or Brooklyn. Server-processing reduced the
-amount of data transfered to client, but it might increase the response time
+amount of data transferred to client, but it might increase the response time
 under load. We implemented a cache policy that greatly improved the performance,

### 2018-08-18 paper/letter/letter.tex
sample added

@@
 
-    }
+    \reply{We added/expanded discussion about the choices, including the
+     subsection about color.}}
 
@@
     work, probably due to their writing style. (R2, R4)}
+    \reply{The introduction was re-written to make this clearer.}
 
@@
     method, we could have used mixed regions for instance. We adopted CTs
-    because it is the most finely grained data available including more than
-    population count. Additionally, other census areas do not change as
-    frequently as CTs. }
+    because it is the most finely grained and volatile data available including
+    more than population count. And all the communications with the experts was
+    included in the supplementary material.}
 
@@
     since this is the figure that explains the entire paper.
-}
+
+    \reply{This change wasn't made because we do not introduce the clustering
+    algorithm. For the purposes of this work, any hierarchical clustering would
+    work.}}
 \end{itemize}
@@
     The weakness of the paper is that the motivation and design requirements are not
-    clearly demonstrated, and the design process is not justified.}
+    clearly demonstrated, and the design process is not justified.
+
+    \reply{The intro was re-written and justifications/discussions added for every plot/option.}
+    }
 
@@
     explicitly describe the motivation of the proposed work.
+
+    \reply{The introduction was re-written to make this explicit.}
 
@@
     distance).
-    \reply{TODO}
-    }
+
+    \reply{We did not propose a new clusering method. Any hierarchical
+    clustering would work, with potentially minimal differences in the result.
+    This specific suggestion corresponds to the ClustGeo, an R package that does
+    Ward clustering with two distance matrices to express geographic distance.
+    That work is mentioned and discussed.}}
 
@@
     confusion in this specific aspect. It is not ideal, but we believe it is a
-    acceptable approximation.}}
+    acceptable approximation. Not ideal, but reasonable enough.}}
 
@@
 
-    \reply{Following the guidelines, an application paper doesn't need to
-    include new techniques. Indeed, the objective of this work was to leverage
-    "commonplace" VIS methods to provide a more efficient solution to a common
-    issue in another field. Indeed, we provide a viable alternative to time
-    consuming and challenging problem that was believed to be unavoidable.}}
+    \reply{This is an application paper, not a technique paper. It's objective
+    is to demonstrate how cleverly combining simple, well explored, VIS methods
+    can solve an "unsolvable" problem for demographic analysis. All the panels
+    contain established tools (except for the "enhanced" boxplot, I didn't find
+    anything similar in the literature, but that's hardly the main contribution
+    of the paper).}
 
@@
     fairly known for this, but some were not included because they are not
-    relevant to the current problem. TODO I included more text expliciting
+    relevant to the current problem. \\The introduction and related work were
+    re-written and significantly expanded. I included more text expliciting
     exactly what was accomplished in this work, which hopefully will make that
@@
     \end{itemize}
+    \reply{We removed the design objectives. Indeed they were rather obvious for
+    the context/application. \\
+    The discussion for edge augmentation was extended.\\
+    Figure placement is up to latex, and the publishers.\\
+    The data for a single CT would occupy too much space. It is included in the
+    supplementary material.}
 

### 2018-08-18 sup/sup.tex
alpha
name update
sample added

@@
 \usepackage{hyperref}
+\usepackage{multirow}
 \title{Supplementary material}
@@
 
-\subsection{B - Ethan A. Fosse}
+\subsection{B - Ethan Fosse}
 % \href{https://scholar.harvard.edu/ethanfosse}{\url{https://scholar.harvard.edu/ethanfosse}}
@@
 trajectories in which the population involved in each cluster is nested in
-colours. Once one gets into the data, the changes don’t look so sharp after all
+colours. Once one gets into the data, the changes don't look so sharp after all
 (maybe in education and income). This may not be a problem of the platform as it
@@
 \clearpage
+\section{Data samples}
+\subsection{Canada}
+Data for one CT over time in Toronto.
+\begin{center}
+  \includegraphics[width=\linewidth]{SampleCA.png}
+  \input{tableCA}
+  \end{center}
+
+\clearpage
+
+\subsection{USA}
+Data for one CT over time in Brooklyn.
+\begin{center}
+  \includegraphics[width=\linewidth]{SampleUS.png}
+  \input{tableUS}
+\end{center}
+
+\end{document}
+
+\clearpage
 \section{Variable matching}

### 2018-08-18 sup/getTable.py

@@ -1,0 +1,46 @@
+import json
+import sys
+
+
+if len(sys.argv)!=3:
+    print('.py folder US/CA')
+    exit(-1)
+
+
+which=sys.argv[2]
+
+if (which=='CA'):
+    CT_ID=['CA021065','CA5350065.00','CA535065.00','CA5350065.00','CA5350065.00']
+    years=[1971,1981,1991,2001,2011]
+if (which=='US'):
+    years=[1970,1980,1990,2000,2010]
+    CT_ID=['US36047014700',]*5
+
+savedData={}
+
+for i,y in enumerate(years):
+    savedData[y]={}
+    f=sys.argv[1]+'/{0}_n.gj'.format(y)
+    with open(f,'r') as inGJ:
+        gj=json.load(inGJ)
+        for feat in gj['features']:
+            if feat['properties']['CT_ID']==CT_ID[i]:
+                for v in feat['properties']['variables']:
+                    savedData[y][v['name']]={'labels':v['labels'],'values':v['values']}
+
+newData={}
+for y in savedData:
+    for v in savedData[y]:
+        if (v not in newData):
+            newData[v]={}
+        newData[v][y]=savedData[y][v]
+
+print("\\begin{tabular}{lccccc}")
+for v in newData:
+    print('\t\\hline')
+    print('\t'+v+'&'+'&'.join(['{0}'.format(y) for y in years])+'\\\\')
+    print('\t\\hline')
+    for i,l in enumerate(newData[v][years[0]]['labels']):
+        print('\t',l,'&','&'.join(['{0}'.format(int(newData[v][y]['values'][i])) for y in years]),'\\\\')
+
+print("\\end{tabular}")

### 2018-08-18 sup/sup.tex
online documenttion included
better order

@@
 \usepackage{multirow}
+\usepackage[final]{pdfpages}
+
 \title{Supplementary material}
@@
 
-\section{Expert comments}
-
-\subsection{Original request}
+\section{Original request for feedback}
 As part of our efforts in the Urban Genome Project of the University of Toronto,
@@
 
+\section{Documentation}
+\includepdf[pages=-,pagecommand={},width=1.5\textwidth]{Documentation.pdf}
+
+\section{Expert comments}
 \subsection{A - Cary Wu}
@@
 
-\end{document}
+% \end{document}
 

### 2018-08-18 paper/letter/letter.tex
g2g

@@
 \maketitle
-A previous version of this paper was submitted to IEEE VAST 2018. While the full
-text of the reviews is attached, we would like to overview the changes made,
-leveraging the summary review:
+A previous version of this paper was submitted to IEEE VAST 2018. Before going
+to the reviews, I am personally humbly requesting from this new wave of
+reviewers: \textbf{please be specific}. I can't really do much with \emph{"most
+technical content would benefit from a thorough re-written"} or \emph{"the
+writing style and paper organization is hardly a qualified research paper on
+visualization"}, especially because I'm following the same pattern of my last
+three publications at VAST and TVCG.
+
+The full text of the reviews is attached. We start with the overview of the
+changes using the summary review:
 
@@
     \reply{We did not introduce a new clustering method. The original paper
-    contains such evaluations, we added that explicitly on the text. All the
-    communication with the experts is included, verbatim, in the supplementary
-    material, with the exception of Patrick Adler, which is a member of the
-    Urban Genome Project, but did not collaborate directly in this work.}}
+    contains such evaluations, we added that explicitly on the text. More
+    details about the information available to the experts at the time of their
+    review was added to the respective section, the supplementary material
+    contains the full transcript of our communications.}}
 
@@
 
-    \reply{We added/expanded discussion about the choices, including the
-     subsection about color.}}
+    \reply{We added/expanded the discussion about the visual and
+     representational choices, including a five paragraph subsection about
+     color.}}
 
@@
     work, probably due to their writing style. (R2, R4)}
-    \reply{The introduction was re-written to make this clearer.}
 
     \item{Presentation: no technical contribution clearly stated, writing style not
-    qualified as a good research paper, paper organization can be improved (R2, R4)}
+    qualified as a good research paper, paper organization can be improved (R2, R4)
+
+    \reply{The paper was re-organized to better reflect our contributions and
+     motivation, including a full rewrite of the intro and related work.}}
+
 \end{itemize}
@@
 
-    \reply{The specific region used for the analysis is irrelevant to the
-    method, we could have used mixed regions for instance. We adopted CTs
-    because it is the most finely grained and volatile data available including
-    more than population count. And all the communications with the experts was
-    included in the supplementary material.}
+    \reply{The text was modified to make clear that the specific region used for
+    the analysis is irrelevant to the method. We adopted census data because it
+    is well explored, and we adopted CTs because it is the most finely grained
+    and volatile data available in the censuses that include more than
+    population count.}
 
@@
     algorithm. For the purposes of this work, any hierarchical clustering would
-    work.}}
+    work. The text was modified to make this point clearer.}}
 \end{itemize}
@@
 
-    \reply{The introduction was re-written to make this explicit.}
+    \reply{The introduction was re-written to make the motivation and our contributions explicit.}
 
@@
 
-    \reply{We did not propose a new clusering method. Any hierarchical
-    clustering would work, with potentially minimal differences in the result.
-    This specific suggestion corresponds to the ClustGeo, an R package that does
-    Ward clustering with two distance matrices to express geographic distance.
-    That work is mentioned and discussed.}}
+    \reply{I modified the text to make clear that the clustering method is not
+    new. Indeed, any hierarchical clustering would work. This specific
+    suggestion corresponds to the ClustGeo, an R package that does Ward
+    clustering with two distance matrices to express geographic distance. That
+    work is mentioned and discussed.}}
 
@@
 
-    \reply{Colour assignment was indeed challenge for this work, mostly because
-    it needs to be used, but there isn't a definitive way to actually do it.
-    That was also the reason for artificially limiting the maxium number of
-    clusters to eight. There is indeed overlap between the colours used for
-    relevance and the clusters, mostly because the clusters used virtually all
-    of the colour space. However, we believe that their different meanings are
-    clear, since none of the experts, nor the other reviewers, mentioned any
-    confusion in this specific aspect. It is not ideal, but we believe it is a
-    acceptable approximation. Not ideal, but reasonable enough.}}
+    \reply{The colour discussion was expanded. However, we believe the colour
+    confusion mentioned is not particular problematic since none of the experts,
+    nor the other reviewers, mentioned it. Not ideal, but reasonable enough.}}
 
@@
     still plenty of space, I would suggest the authors to add more details in Section
-    6.}
+    6.
+    \reply{Done.}}
     \end{enumerate}
@@
     case study instead of fragmented videos to explain the functionalities.
-}
+
+    \reply{These videos are part of the documentation, illustrating specific
+    activities.}}
 \end{itemize}
@@
     visualization is rather limited. The authors are also encouraged to re-submit to a
-    GIS venue.
+    GIS venue.}
 
@@
     relevant to the current problem. \\The introduction and related work were
-    re-written and significantly expanded. I included more text expliciting
-    exactly what was accomplished in this work, which hopefully will make that
-    distinction clearer. \\ As the others reviewers mentioned, the tool was
-    available not only for the experts, but for the reviewers as well, on an
-    anonymous site.}
+    re-written and significantly expanded. I included more text explaining
+    exactly what was accomplished in this work, to make that distinction
+    clearer. \\ As the others reviewers mentioned, quite obviously the tool was
+    available to the experts, and for the reviewers as well, on an anonymous
+    site.}
 
@@
     Project, which will cite this publication for the technical details.}}
-
 \end{itemize}

### 2018-08-18 paper/study.tex
g2g

@@
 \section{Expert feedback}
-%This could be a big step toward better understanding of evolving urban morphology.
-To assess the proposed method, we contacted academic and industry experts in
-sociology and urban sciences. We focused our inquires on the utility of the
-prototype, and if they found anything interesting on the data using it. The
-supplementary material contains the communication with five experts, verbatim,
-identified from A to E.
+We contacted academic and industry experts in sociology and urban sciences for
+their appreciation of our methodology. They had access to the prototype tool, a
+descriptive documentation of the features (included in the supplementary
+material), and a sequence of documentation videos illustrating how to perform
+specific tasks. The documentation explains which datasets are used, how the data
+is represented and processed, including that there is no geographic
+harmonization. We focused our inquires on the results obtained, asking if they
+found anything interesting on the data. The message sent and their full response
+is included in the supplementary material. Each of the five experts is
+identified by a letter, from A to E.
 

### 2019-06-13 paper/discussion.tex
new paper version

@@
 \section{Discussion and limitations}
+\label{sec:discussion}
 
-Our objective was to leverage a graph based data representation and
-visualization methods for the exploration of geographically inconsistent
+Our objective was to leverage a network based data representation and
+visualisation methods for the exploration of geographically inconsistent
 region-based data. While we successfully replicated and corroborated results
@@
 
-Removing the need for geographical harmonization greatly reduces the amount of
+Removing the need for geographical harmonisation greatly reduces the amount of
 work necessary to explore demographic data, but the method still requires
@@
 
-While using one small map for each year leads to an easier visualization that
-does not require interaction, it does not scale if more than five or six years
-are considered. In this case, it might be interesting to replace the larger map
-considering each year individually, along with a temporal control for
-navigation. Indeed, including more years would likely lead  to a stronger
-mixture of colors in the trajectory map, leading to a predominantly grey hue.
-
 The limitation on the number of displayed clusters because of the limited number
-of distinguishable colors was significant. While increasing the number of
+of distinguishable colours was significant. While increasing the number of
 clusters would further complicate an already complex analysis, it might be
-warranted for some regions. Color is a fundamental and intuitive tool for
+warranted for some regions. Colour is a fundamental and intuitive tool for
 information representation that can be coherently used across different plots,
-so we opted to use it, even if in a limited way. With eight colors, there was
-overlap between some clusters, the relevance gradient, and the color
+so we opted to use it, even if in a limited way. With eight colours, there was
+overlap between some clusters, the relevance gradient, and the colour
 combination.
 
-Another limitation is the lack of control on how much the geographical
-information will impact the clustering result. While the adopted method met our
-needs for this work, a configurable control would add another dimension to the
-exploration, allowing for more intra-cluster variance to obtain more 'compact'
-clusters. We explored changing the number of content based augmented edges, but
-this proved to be unreliable and hard to interpret. The~\emph{ClustGeo}
-method~\cite{Chavent2017} can be a viable option for this, allowing a graph
-based input and a hierarchical output, combined using a single mixing
-parameter. Alternatively, one could cluster the changes~\cite{bian2018survey}
-instead of the stable states.
-
-There are also technological limitations, such as memory use on the
-visualization client. To allow for changes on the CTs over the years, we use a
-geographic file that contains all possible intersections, which can grow rather
-large if the original city was expansive and contained several CTs, like NYC or
-LA. However, the most significant technological limitation relates to parameters
-that are not immediately interactive, such as the clustering configuration.
-Since the clustering is computationally expensive and performed on the server,
-which allows for cached results, some changes can take a few minutes to be
-considered, removing any possibility of a continuous exploration.
-
-Indeed, the cognitive load on the user is already significant, as we compromised
-simplicity for accuracy. While other works labelled the clusters, as 'young
-urban', 'struggling', and so on~\cite{Delmelle2016,Delmelle2017}, we show the
+The cognitive load on the user is significant, as we compromised simplicity for
+accuracy. While other works labelled the clusters, as 'young urban',
+'struggling', and so on~\citep{Delmelle2016,Delmelle2017}, we show the
 statistical characteristics of the clusters, which are harder to interpret, as
 the data may have subtle nuances that labels would otherwise hide. This also led
-to a crowded interface, mitigated somewhat the use of pop-up panels and collapsible
-sections. For some cities, especially if they are small and stable, the panels
-can appear redundant, but each provide a different way to interact with the
-information that can ease the exploration process for larger and dynamic cities.
+to a crowded interface, mitigated somewhat the use of pop-up panels and
+collapsible sections. For some cities, especially if they are small and stable,
+the panels can appear redundant, but each provide a different way to interact
+with the information that can ease the exploration process for larger and
+dynamic cities.

### 2019-06-13 paper/intro.tex
new paper version

@@
 % General context + motivation
-\IEEEPARstart{U}{rban} sciences are blooming thanks to a renewed interest in
-understanding and improving the urban environment. Visual analytics is following
-this trend, fueled by new public datasets that encompass progressively more of
-our daily lives~\cite{Chen2017}. There is no shortage of methods to explore
-mobility patterns~\cite{VonLandesberger2016}, social media~\cite{Chen2017},
-traffic~\cite{chen2015survey}, and so on, providing experts, planners, policy
-makers, and the general population with deeper insights about their cities.
 
-These new datasets usually contain GPS coordinates for the records, leading to
-\emph{point-based} data. Combined with the corresponding timestamps, this data
-is easily suitable for longitudinal analysis. But most demographic datasets are
-\emph{region-based}, where the measurements are associated with pre-defined
-regions, not only for an additional level of privacy protection, but because
-some measurements only make sense over a defined area. Census data is a classic
-example of this format, with datasets available from 1790 onward for the
-US~\cite{nhgis}. Despite this unmatched temporal availability, longitudinal
-analyses of census data are often restricted in time, especially when smaller
-tabulation areas are considered, such as census tracts (CT) or dissemination
-areas, which evolve to reflect changes in population density, leading to
-geographic inconsistencies across time, and the traditional time-series based
-approach is no longer viable.
+Neighbourhoods have increasingly become a central concept in social research and
+targets for social
+policy~\citep{sampson2012great,galster2019making,stone2015urban,looker2015nation}.
+To be sure, a focus on neighbourhoods extends to the formative period of the
+modern social sciences~\citep{Abbott1997}. Recent interest has at least partly
+been rekindled through newly available longitudinal demographic
+datasets~\citep{Logan2014,nhgis}, convenient computational
+tools~\citep{rey2018spatio}, and new sources of data~\citep{Poorthuis2018}.
 
-However, these analyses are necessary to understand the urban environment.
-Indeed, two different regions can have similar average income for a given year,
-while one is experiencing a process of economic improvement and the other one
-impoverishment. A single snapshot cannot be used to identify gentrification,
-migration, education changes, or any of the relevant processes that happen over
-time.
+These advances have made neighbourhood constructs more tractable to quantitative
+analysis. Yet new challenges have also emerged, especially at the convergence of
+research on neighbourhood effects and neighbourhood dynamics. Neighbourhood
+effects research assumes knowledge about the nature and scope of "the
+neighbourhood" that presumably shapes individual
+outcomes~\citep{Kwan2018,Shelton2019}. Concurrently, researchers note that
+neighbourhoods are not necessarily fixed containers in which other processes
+occur, but themselves dynamically
+evolve~\citep{Delmelle2017,Reades2019,li2018new}. The result is to open up key
+assumptions about neighbourhoods for theoretical and empirical examination: how
+do we appropriately define and compare neighbourhoods at a given time?; how do
+we appropriately define and compare the temporal trajectories of
+neighbourhoods?; and can we do both at once, "fully
+interactionally"~\citep{Abbott1997}: classify neighbourhoods now based on where
+they came from and where they are going?
 
-To overcome these inconsistencies, the traditional approach is the
-\emph{geographical harmonization} of the data, the interpolation of the
-measurements into a common set of
-regions\cite{Logan2014,Hallisey2017,Allen2018}, so that each variable can be
-represented using time-series. This is laborious work that inevitably introduces
-some amount of error~\cite{Logan2016}, even when additional data is
-provided~\cite{eicher2001dasymetric}. Nevertheless, this step is considered
-mandatory in the current literature: \emph{"(...) tract-by-tract comparison is
-not possible unless data from 2000 is interpolated to 2010 boundaries
-(...)"}~\cite{Dmowska2017}.
+In principle, much of the recent research is committed to the proposition that
+neighbourhoods are open and evolving entities. Ironically, its empirical
+practice tends to rely on methods that require fixed geographical regions. This
+requirement is difficult to satisfy, as most longitudinal datasets are based on
+pre-defined tabulation areas that are routinely modified by data collection
+agencies, usually to follow population changes.
 
-The main contribution of this application paper is an visualization-based
-alternative to the geographical harmonization, a combination of established
-graph based processing and information visualization techniques allowing
-tract-by-tract comparison, the identification and visualization of patterns of
-demographic evolution without geographic harmonization, effectively removing one
-of the most challenging problems in longitudinal demographic analysis. We also
-include illustrative scenarios and our prototype is available at
-\url{http://uoft.me/piccard}, including more than forty regions in the US and
-Canada. The source code is publicly available at
-\url{https://github.com/fabioasdias/piccard}.
+The standard approach then is to \emph{geographically harmonise} data. This
+involves interpolating existing measurements into a common set of
+regions~\citep{Logan2014,Hallisey2017,Allen2018}. Recent computational tools
+have somewhat simplified this process~\citep{rey2018spatio}, but it still
+involves non-trivial questions: which geometry to use as target, how to
+apportion the variables, or how to combine data from different sources. Further,
+these question do not necessarily have optimal answers. Indeed, regardless of
+how well this process is performed, it still introduces
+errors~\citep{Logan2016}, even when additional data is
+provided~\citep{eicher2001dasymetric}. Essentially, harmonisation generates
+artificial data points that can potentially lead to inaccurate results, even
+though they are seldom interpreted as such. Nevertheless, because there has been
+no viable alternative, and the results often appear plausible, these concerns
+are generally overlooked. The result is that the harmonisation approach is
+virtually mandatory in the current literature: \emph{"(...) tract-by-tract
+comparison is not possible unless data from 2000 is interpolated to 2010
+boundaries (...)"}~\citep{Dmowska2017}, \emph{"(...) This limits cross-year
+comparison since data are not representative of the same spatial units. (...)"}
+~\citep{Allen2018}.
+
+The main contribution of this paper is a method for longitudinal data processing
+that works with the original data by leveraging a network based representation.
+It enables tract-by-tract comparison and the identification of patterns of
+demographic evolution \emph{without geographic harmonisation}.
+
+To allow a proper examination of our method and its results, we built an online
+interactive system using this representation. It enables users to visualise,
+interpret, and explore trajectories of neighbourhood change. This interface
+helps validate our method, by allowing it to be compared to existing and future
+methods. Further, it is a significant contribution to the research community: it
+provides a vehicle for quickly and easily grasping complex long-term changes,
+experimenting with different parameters to interactively learn from data, and
+making neighbourhood change research publicly transparent. The interface thus
+responds to increasing concerns about reproducibility and transparency, as well
+as ongoing attention to the value of visualisation in scientific research and
+communication.
+
+We start by presenting an intuitive example of our representation in
+Section~\ref{sec:introduction}, then we review the relevant literature on
+longitudinal studies, data representation, clustering, and spatio-temporal
+visualisation in Section~\ref{sec:related}. Our methodology is introduced in
+detail in Section~\ref{sec:method}, along with the included interface.
+Illustrative scenarios for Chicago, Toronto, and Los Angeles are presented in
+Section~\ref{sec:study} and the feedback of five field experts are summarised in
+Section~\ref{sec:expert}. Our prototype system is available at
+\censure{\url{http://uoft.me/piccard}}, including more than forty regions in the
+US and Canada. The source code is publicly available at
+\censure{\url{https://github.com/fabioasdias/piccard}}.\footnote{The editors are
+considering, at our request, an exception to the double-blind requirement to
+allow access to the system. We provided them with the URLs of the system, code,
+and documentation separately.}
+
+\section{Intuition}
+\label{sec:intuition}
+While utterly simple, the network model breaks from the deeply rooted
+traditional tabular paradigm in a significant way. Instead of requiring the data
+as a collection of fixed entities which properties evolve over time, literally
+rows in a table with temporal values as columns, it represents each measurement
+as a separate entity and encodes the evolution of these entities over time.
+
+To ease this cognitive transition, we start with an intuitive example of how it
+works, using a small portion of a fictitious urban region illustrated on the
+left part of Figure~\ref{fig:intuition}. This example includes three different
+times ($t_0,t_1,t_2$), with different aggregation areas identified as letters
+from A to H. For $t_0$, the initial time, we have areas A and B, with small
+houses and a park, respectively. The park remains stable (B, E, and H), but the
+houses are partially replaced by larger buildings (C and F).
+
+This example also illustrates some of the challenges of harmonisation. None of
+these aggregation areas is clearly suitable as an interpolation target. In fact,
+adopting any of these areas as a target would require merging heterogeneous
+regions and/or dividing homogeneous regions. For instance, by choosing the
+regions of $t_1$, A would be split to match regions C and D, which appears to be
+a rather reasonable approximation in this homogeneous artificial example, albeit
+the fallacy of division. Region F would be similarly split to match C, but F
+would be split and merged with G to match D, potentially leading to statistical
+measurements that do not properly represent either region.
+
+Real world measurements are seldom as homogeneous and noise-free as this
+artificial example. By splitting and merging the data to fit arbitrary borders,
+that were not necessarily coherent at the time of the measurement, harmonisation
+increases the distance between measurement and reality.
+
+\begin{figure}
+    \centering 
+    \subfloat{
+        \includegraphics[width=0.475\linewidth]{intuition.png}
+    }
+    \subfloat{
+        \begin{tikzpicture}
+            \node (push) at (-2,0) {};
+            \node [circle, draw=red, thick] (a) at (0,4) {A};
+            \node [circle, draw=green, thick] (b) at (4,4) {B};
+            \node [circle, draw=blue,  thick] (c) at (0,2) {C};
+            \node [circle, draw=red, thick] (d) at (2,2) {D};
+            \node [circle, draw=green, thick] (e) at (4,2) {E};
+            \node [circle, draw=blue, thick] (f) at (0,0) {F};
+            \node [circle, draw=red, thick] (g) at (2,0) {G};
+            \node [circle, draw=green, thick] (h) at (4,0) {H};
+            \draw (a) -- (c);
+            \draw (a) -- (d);
+            \draw (b) -- (e);
+            \draw (c) -- (f);
+            \draw (d) -- (f);
+            \draw (d) -- (g);
+            \draw (e) -- (h);
+        \end{tikzpicture}
+    } \caption{Network based spatio-temporal data representation. \textbf{Left}:
+    Three temporal stages of the evolution of a fictitious urban area, with
+    aggregation areas A to H. \textbf{Right}: Network representation of the
+    aggregation areas where the colours identify similar regions.
+        \label{fig:intuition}}
+\end{figure}
+
+Instead, we propose a network representation. Intuitively, a \emph{network}
+(also called a \emph{graph}) is a collection of entities (nodes) that are
+related to each other (edges). In this case, each different aggregation area is
+represented as a node and we connect nodes that have overlapping geographical
+areas, leading to the network illustrated on the right of
+Figure~\ref{fig:intuition}. By partitioning the network into connected nodes
+that are similar, we are effectively identifying clusters in the spatio-temporal
+data, as illustrated by the colours of the nodes on the right side of
+Figure~\ref{fig:intuition}. Further, the possible evolution paths can also be
+obtained by computing temporal sequences of nodes, in this case: (A, C, F), (A,
+D, F), (A, D, G), and (B, E, H). This representation is also suited for
+geographically consistent regions, as illustrated by the stable park in this
+example, and is therefore a generalisation of the traditional paradigm.
+
+%I'm adding this because Jeff Allen (Farber's guy) struggled with this.
+Note that the edges of this network merely encode that two regions are related.
+This is a binary information, there is no apportionment, no areal measurements,
+no population percentages associated with the edge. Indeed, our method also
+connects regions of the same time that share borders, representing exactly that
+they are neighbouring areas.

### 2019-06-13 paper/method.tex
new paper version

@@
-\section{Visualizing the demographic spatio-temporal evolution}
-Beyond the objective of allowing the study of inconsistent data, our method
-includes incremental developments in most steps of the analysis, from data
-representation to the visualization method for the clusters.
+\section{Visualising the demographic spatio-temporal evolution}
+\label{sec:method}
 Figure~\ref{fig:overview} presents an overview of the processing steps of the
-proposed method.
+proposed method, illustrating how the nodes of the network are used to represent
+the regions. The following sections elaborate this figure and explain the main
+features of the interface we built to visualise and explore the evolution of
+neighbourhoods on the basis of our proposed method.
 
@@
     \includegraphics[width=0.975\linewidth]{overview.pdf}
-    \caption{Overview of the proposed method. A graph is generated combining the
+    \caption{Overview of the proposed method. A network is generated combining the
         original census data, encoding the changing geographical information.
-        The graph is partitioned into an hierarchy~\cite{markus2017}. The
+        The network is partitioned into an hierarchy~\citep{markus2017}. The
         characteristics and evolution of the clusters are then visually
@@
 \subsection{Census methodology and data representation}
+\info{I removed any specifics of the experiments (CTs, which variables) from here because this section is theory}
 
 Census data is disseminated in a tabulated form for aggregation areas: whole
-country, state/province, metropolitan region, and so on. To provide as much
-detail as possible, we focus on the smallest region with available data:
-\emph{census tracts} (CT). They are usually defined to maintain the anonymity of
-the population, leading to a population count in the order of thousands in
-densely populated areas. Physical barriers are usually adopted as borders, so
-these regions can change because of new roads, construction or removal of high
-density buildings, and so on. Some census entities also consider demographic
-characteristics, aiming to establish the CTs as a cohesive unit. Therefore, CTs
-are the least geographically stable tabulation area.
+country, state/province, metropolitan region, and so on. To allow for a more
+meaningful comparison of the data, we aggregated related variables (e.g. White,
+Black, Asian, Other) into what we called an \emph{aspect} (e.g. Race). The
+aspects are represented using normalised histograms. This normalisation is
+crucial for direct comparison. In essence, it is a generalisation of the
+standard method of comparing percentages, since each aggregation area has a
+different total population.
 
-Each CT is associated with a series of variables, with counts derived from the
-census questionnaires, covering several aspects of the demographic
-characteristics of the population. Some questions allow for multiple choices or
-open answers, that are then tabulated into the most frequent categories. Since
-the census is often used to direct government initiatives, which variables are
-measured/disseminated is dependent on administrative interests, the general
-understanding of the population, and current customs. For instance, income is
-disseminated with a finer tabulation in the lower portion than on the higher.
-
-To match these variables over time and allow for direct comparison across
-different census years, we aggregated similar ones (e.g. White, Black, Asian,
-Other) into \emph{aspects} (e.g. Race), encoding the distribution of that facet
-of the population. In this convention, we refer to the composing parts of an
-aspect as a \emph{part} or the traditional \emph{variable}. Internally, the
-aspects are represented using normalized histograms. This normalized
-representation is crucial for the comparison between inconsistent regions.
-
-In our graph based representation, each CT of each census year is represented as
-a node, and edges are placed between nodes if the corresponding CTs share
-geographic borders in the same year. Further, edges are placed between nodes if
-the corresponding CTs belong to sequential years and there is geographical
-overlap between them. This approach leads to a single graph representing the
-whole spatio-temporal space of the data. Our objective then becomes to identify
-partitions of this graph such that the nodes of each partition are more similar
-between themselves than to the other nodes. This representation is not the only
-option, nor unique, but it allows the use of existing graph-based methods for
-the other steps.
+Each area of each census year is represented as a node, and edges are placed
+between nodes if the corresponding regions share geographic borders in the same
+year. Further, edges are placed between nodes if the corresponding regions
+belong to sequential years and there is geographical overlap between them. This
+approach leads to a single network representing the whole spatio-temporal space
+of the data. Our objective then becomes to identify partitions of this network
+such that the nodes of each partition are more similar between themselves than
+to the other nodes.
 
 \subsection{Geographic content clustering}
-To partition the graph we must first establish a distance function between the
-nodes, measuring the data similarity. This similarity is then associated with
-the edges, leading to a weighted dynamic graph. Every node has a collection of
-histograms, each representing the distribution of certain aspect in the
-population.
+Having tied the regions together into a network, we can now partition it to
+identify similar sets of regions. We start by adopting a distance function
+between the nodes, measuring the difference between the data of the regions.
+This value is then associated with the edges, leading to a weighted dynamic
+network. Every node has a collection of histograms, each representing the
+distribution of certain aspect in the population.
 
-Let $\G=(\V,\E)$ be a graph, where
+Let $\G=(\V,\E)$ be a network, where
 $\V=\{\vertice_1,\vertice_2,\dots,\vertice_n\}$ is the set of nodes and
@@
 Once the distances are associated to the edges, we use watershed
-cuts~\cite{Cousty2009} to create an initial clustering, which is then refined
-into a hierarchy using the Sorted Maximal Matching (SMM)~\cite{markus2017} with
+cuts~\citep{Cousty2009} to create an initial clustering, which is then refined
+into a hierarchy using the Sorted Maximal Matching (SMM)~\citep{markus2017} with
 median linkage. The initial watershed step is performed to create an initial
-clustering and reduce the running time of the SMM. For completeness, we briefly
-review this method, but we refer the reader to the original
-paper~\cite{markus2017} for more details, including a complete performance
+clustering and reduce the running time of the SMM. We introduced one new
+parameter to this method: a maximum distance threshold for the merges, to avoid
+the early merging of outliers. We refer the reader to the original
+paper~\citep{markus2017} for more details, including a complete performance
 evaluation using several metrics.
 
-We included two application-specific parameters: the maximum number of clusters
-to be shown and a distance threshold. Contrarily to the original SMM, which
-merges all clusters in all steps, we only merge two clusters where the distance
-is above the threshold after we reach the maximum number of displayed clusters.
-Without this restriction, significantly different clusters would be merged
-early, leading to increased intra cluster variance and the disappearance of
-small outlier regions. Further, after the maximum number of clusters is reached,
-we create one step of the hierarchy for each merge, leading to a binary
-partition tree. In this structure we can directly access a result with an
-arbitrary number of clusters.
+Each resulting cluster is contiguous in the network. This means that two
+similar, but non-contiguous, sets of areas will be classified into two different
+clusters, which can be counter-intuitive. To overcome this issue, we
+\emph{augment} the network with two new edges per node from a nearest neighbours
+graph~\citep{scikit-learn} using only the distances between the histograms.
+These edges connect nodes with similar content, if they are not already
+connected, providing a path for the algorithm to group similar nodes.
 
-Each resulting cluster is contiguous in the graph. This means that two similar,
-but non-contiguous, sets of CTs will be classified into two different clusters,
-which can be counter-intuitive. To overcome this issue, we \emph{augment} the
-graph with two new edges per node from a nearest neighbors
-graph~\cite{scikit-learn} using only the distances between the histograms. These
-edges connect nodes with similar content, if they are not already connected,
-providing a path for the algorithm to group similar nodes. Theoretically, adding
-more of these content based edges could be used to decrease the impact of the
-spatio-temporal edges, controlling the balance between content and topology in
-the result. In practice, the effect is dependent on the data itself, and the
-results are not consistent, or predictable, across different cities. We fixed it
-at two edges because it was the lowest number that empirically led to consistent
-clusters, but we believe that this idea warrants further investigation, as an
-alternative to mixing parameters in the distance metric~\cite{Chavent2017}.
+\subsection{Cluster characterisation and variable relevance}
+\label{sec:relevance}
+A crucial step in understanding neighbourhood change is to characterise the
+evolving clusters.  The composition of each cluster is represented here by
+simple statistical measures, considering each aspect separately. We compute the
+minimum, maximum, median, 25\%, and 75\% quantiles for each variable of each
+aspect for all clusters in the hierarchy. While interpreting these values is
+more complex than interpreting just the average, they provide far more
+information about the underlying distribution.
 
-\subsection{Cluster characterization and variable relevance}
-\label{sec:relevance}
-The composition of each cluster is determined by simple statistical measures,
-considering each aspect separately. We compute the minimum, maximum, median,
-25\%, and 75\% quantiles for each part of each aspect for all clusters in the
-hierarchy. While interpreting these values is more complex than interpreting
-just the average, they provide far more information about the underlying
-distribution.
-
-We also use these statistical measurements to discover what characterizes each
+We also use these statistical measurements to discover what characterises each
 cluster, that is, what makes it different from the others.  We define the
-\emph{relevance} of a part of an aspect based on the distance between the
+\emph{relevance} of a variable of an aspect based on the distance between the
 interquantile ranges (IQR) of the clusters in the same hierarchical level. If
 the IQRs overlap for all clusters, that variable is not relevant to the
-characterization of the cluster, but if the IQRs are distant, it means that this
+characterisation of the cluster, but if the IQRs are distant, it means that this
 specific range of values is something that only occurs in this cluster.
+Examining IQRs therefore provides users a straightforward visual method for
+determining what variables most clearly define a given cluster.
 
@@
 understand what groups exist and where they are, we are also interested in the
-evolution of these groups. We introduce the concept of \emph{trajectories},
-composed by regions classified into the same sequence of clusters over the
-considered period. This enables direct access to regions that evolved in the
-same manner. While the interface provides access to data by individual census
-tract, the trajectories are the main unit of exploration in this work.
-
-\subsection{Colors}
-\label{sec:colors}
-As illustrated by Figure~\ref{fig:ui} and further explored in the next
-subsection, our proposed interface heavily relies on color to express
-cluster-related information. We adopted this convention because colors can be
-used in all our visual tools in a coherent manner. However, this also introduced
-significant challenges. The first is the limit on the number of clusters that
-can be visually represented.  We limited the number of clusters to eight because
-this was the largest number of colors that we could reliably use, derived from
-the 8-class Dark2 set from ColorBrewer~\cite{ColorBrewer}.
-
-While we can reasonably limit the number of clusters, there are far more
-possible trajectories. And the color associated with each trajectory should bear
-some resemblance to the clusters included in it. Therefore, we were left with a
-conundrum: \emph{Should we associate each trajectory with a unique color, which
-the user probably cannot distinguish, or should we use a reduced set of colors
-and associate the same color to different trajectories?} Since there are
-advantages and disadvantages for each of those options, we adopted both. The
-user can control which color policy is used via a checkbox in the configuration
-panel, on the top left of the interface.
-
-By default, the interface adopts a simplified color scheme, where a trajectory
-is painted in the same color of a cluster if the regions were associated to that
-cluster for \emph{all} times; in a slightly less saturated version of that color
-if the regions were associated to that cluster for \emph{the simple majority} of
-the time, and gray otherwise. In this mode, the colors will mostly represent
-stability, immediately identifying the regions that were consistently associated
-with each cluster. It also easily identifies volatile regions, painted gray.
-
-When this simplified color scheme is disabled, each trajectory will be painted
-using the average of the colors of the involved clusters, in the LAB color
-space. In this mode, the map becomes more similar to a heatmap, where stronger
-presence of a color indicates more temporal affinity to the cluster. Volatile
-regions will also tend to be displayed in gray, as the average of three or more
-colors.
-
-\begin{figure}
-    \centering 
-    \includegraphics[width=0.95\linewidth]{colours.pdf}
-    \caption{Different color schemes for Chicago with four clusters. Left:
-    simplified, right: average color.\label{fig:colour}}
-\end{figure}
-
-While both approaches will use more than eight colors, in practice this is not
-as significant because most cities can be explained using less than eight
-clusters.  In fact, articles in the literature usually employ from two to five,
-which are fairly stable across time. For the more dynamic scenarios, user
-interaction can be used to alleviate the shortcomings of both approaches.
+evolution of these groups. To examine this process of evolution directly, we
+introduce the concept of \emph{trajectories}. Trajectories are composed by
+regions classified into the same sequence of clusters over the considered
+period. This enables direct access to regions that evolved in the same manner.
+While individual census tracts remain interesting, these trajectories are the
+main unit of exploration in this work.
 
@@
 \label{sec:ui}
-The initial interface is illustrated in Figure~\ref{fig:ui}. Since demographic
-data can be nuanced, with intricate interconnections, we decided against
-validating the interface using a synthetic dataset, considering instead data
-from the Chicago region between 1970 and 2010, using previous published studies
-as corroboration. This region is known for its entrenched racial divide and the
-emergence of a \emph{'young urban'} population with a higher education
-level~\cite{Delmelle2016,Delmelle2017}. More details about this dataset are
-presented in Section~\ref{sec:study}.
+To validate and explore the results of our methodology, we built a user
+interface, illustrated in Figure~\ref{fig:ui}, considering census tract (CT)
+level data from the Chicago region between 1970 and 2010. This region is known
+for its entrenched racial divide and the emergence of a \emph{'young urban'}
+population with a higher education level~\citep{Delmelle2016,Delmelle2017}. More
+details are presented in Section~\ref{sec:study}.
 
@@
     \centering 
-    \includegraphics[width=0.95\linewidth]{ui.pdf}
+    \includegraphics[width=0.975\linewidth]{ui.pdf}
     \caption{Initial interface of our method showing the demographic evolution of Chicago. 
@@
 
+As illustrated by Figure~\ref{fig:ui}, our proposed interface heavily relies on
+colour to express cluster-related information. We adopted this convention
+because colours can be used in all our visual tools in a coherent manner.
+However, there is a limit on the number of distinct colours that can be used. We
+limited the number of clusters to eight because this was the largest number of
+colours that we could reliably and accessibly use, derived from the 8-class
+Dark2 set from ColorBrewer~\citep{ColorBrewer}.
+
 The configuration panel, on top left in Figure~\ref{fig:ui}, displays which
@@
 includes other configuration options that can be altered without re-processing
-the data, such as the number of clusters and the color option. The gear button
+the data, such as the number of clusters and the colour option. The gear button
 allows access to the other configuration options that do require further
-processing, such as changing location, aspects, and weights. This panel also
-includes the configuration of the selection mode for the trajectories, which
-allows the user to set, add, or remove the next selected trajectories to the
-current selection. This feature enables the analysis of complex sets of
-trajectories.
+processing, such as changing location, aspects, and weights.
 
@@
 detailed in Section~\ref{sec:relevance}. The \emph{View all} button opens a new
-panel where all aspects are represented, while the chevron at the side of the
-color lets the user expand each cluster separately. While the standard approach
-to represent cluster characteristics is to use parallel
-coordinates~\cite{johansson2005revealing,guo2006visualization}, this
-representation occupies screen space proportional to the number of variables and
-can get cluttered with a higher number of clusters, or when the clusters are not
-well defined for multiple variables. To save space and leverage the familiarity
-scientists have with statistical tools, we opted to use boxplots to properly
-convey the distribution of each variable in the current cluster. However, a
-simple boxplot would not include information about the other clusters, forcing
-the user to mentally compare them to find what is relevant.
+panel where all aspects are included, while the chevron at the side lets the
+user expand each cluster separately.
 
-We adopt an \emph{enhanced} version of the traditional boxplot, which includes
-the minimum, maximum, 25\% and 75\% quantiles for the current cluster, but also
+We adopted an \emph{enhanced} version of the traditional boxplot, which includes
 the IQRs for the other clusters, in slightly larger and faded black rectangles.
-We also color the current IQR according to its relevance. While there might be
-some degree of similarity between the color schema for relevance and for
-trajectory identification, none of the experts consulted reported confusion.
-Indeed, one expert reported confusion regarding the grey rectangles that
-represent the IQRs for other distributions when they are not colored; when that
-variable is not relevant to the characterization of the cluster. These simple
-changes allow the user to easily understand the composition of the cluster and
-how it relates to the others. Violin plots~\cite{hintze1998violin} could also be
-used, providing more information about the shape of the distribution, but with
-increased potential for obstructing the representation of the other clusters.
-
-For instance, the boxplot that summarizes the purple cluster illustrated in
-Figure~\ref{fig:ui}, detailed on Figure~\ref{fig:boxplot}, illustrates that this
-cluster is best defined by the proportion of the population with four or more
-years of college. The user can quickly see that this is relevant because the
-corresponding IQR is colored with the highest relevance present in the legend.
-It is also clear that, while this cluster includes CTs that have between 10\% to
-90\% of people in this variable, approximately, half of them have about 60\% of
-the population with four or more years of college. Since all the other IQRs are
-well separated, this is a defining characteristic of this cluster. Conversely,
-the proportion of the population with one to three years of college is not
-relevant, as indicated by black fill in the rectangle representing the IQR of
-this cluster, in overlapping position with the rectangles of the other clusters.
-By clicking on the colored bar above the boxplot, the user can select all
-trajectories that contain this cluster at any point in time.
+We also colour the current IQR according to its relevance.  For instance, the
+boxplot that summarises the purple cluster illustrated in Figure~\ref{fig:ui},
+detailed on Figure~\ref{fig:boxplot}, illustrates that this cluster is best
+defined by the proportion of the population with four or more years of college.
+The user can quickly see that this is relevant because the corresponding IQR is
+coloured with the highest relevance present in the legend. It is also clear
+that, while this cluster includes CTs that have between 10\% to 90\% of people
+in this variable, approximately, half of them have about 60\% of the population
+with four or more years of college. Since all the other IQRs are well separated,
+this is a defining characteristic of this cluster. Conversely, the proportion of
+the population with one to three years of college is not relevant, as indicated
+by black fill in the rectangle representing the IQR of this cluster, in
+overlapping position with the rectangles of the other clusters. By clicking on
+the coloured bar above the boxplot, the user can select all trajectories that
+contain this cluster at any point in time.
 
@@
     \centering 
-    \includegraphics[width=0.9\linewidth]{boxplot.pdf}
+    \includegraphics[width=0.45\linewidth]{boxplot.pdf}
     \caption{Enhanced boxplot of the clusters' characteristics allows a quick
@@
 \end{figure}
-
-The other clusters identified on Figure~\ref{fig:ui} correspond to higher
-concentration of people that identify as Black in the green cluster, people that
-identify as "Asian, Hawaiian, other pacific islander" in the pink cluster, and
-people that identify as White in the orange cluster. From these plots, it is
-clear that the city is indeed racially divided~\cite{Delmelle2016}, with several
-CTs that are almost exclusively occupied by people of the same racial category.
 
@@
 the evolution of the clusters over time. The widths are proportional to the
-population involved, the colors follow a policy detailed in
-Section~\ref{sec:colors}. A stacked graph could also be used to represent the
-proportions of each cluster~\cite{Valdivia2015} with less clutter, since the
-transitions between clusters would not be represented. However, this is only
-viable if more temporal steps are available, making the plot smoother. Another
-option to remove clutter is to remove portions of this plot when trajectories
-are selected, but this would change the layout and compromise the user's mental
-map.
-
-In our example in Figure~\ref{fig:ui}, the total population of Chicago is
-decreasing. Additionally, the orange and green clusters contain most of the
-population and are fairly stable over time. The pink cluster is small and mostly
-stable. The purple cluster is increasing, mostly by incorporating areas that
-were previously orange. Since the purple group corresponds to the emergent
-'young urban' group, this corroborates the findings of
-Delmelle~\cite{Delmelle2016,Delmelle2017}. This diagram can also be used to
-select specific trajectories, by clicking on the bands, or all trajectories that
-contain a specific cluster at a specific time, by clicking on the rectangles.
+population involved. In our example in Figure~\ref{fig:ui}, the orange and green
+clusters contain most of the population and are fairly stable over time. The
+pink cluster is small and mostly stable. The purple cluster is increasing,
+mostly by incorporating areas that were previously orange. Since the purple
+group corresponds to the emergent 'young urban' group, this corroborates the
+findings of Delmelle~\citep{Delmelle2016,Delmelle2017}, showing that our
+network-based method can recover results from the traditional data processing
+approach.
 
 In the next panel, illustrated in the top middle of Figure~\ref{fig:ui}, is a
-transition matrix between the clusters. It indicates the percentage of the
+transition matrix between the clusters. It indicates a rounded percentage of the
 population whose area changed between each pair of clusters. This kind of table
-can be found in the related literature~\cite{Delmelle2016}, so it is familiar to
-the advanced users. It not only informs the proportional changes, but allows the
-selection of the corresponding trajectories for further analysis.
-
-Contrary to the trajectories plot, this representation is more Markovian, where
-only the current and next state are considered. This panel also enables easier
-access to trajectories with specific changes, by clicking on the corresponding
-percentage values. The combo box allows the user to refine the transitions, from
-'Any', which includes all transitions between years, to specific transitions, to
-changes from the first year to the last year. In this example, approximately
-99\% of the population in areas classified as green were also in areas
-classified as green in the next year, while 1\% changed to purple at some point
-and another 1\% to orange. The total is over 100\% due to rounding errors.
-Regions changed from the orange to the green cluster for 6\% of its population,
-1\% to pink, and 9\% to purple. This further corroborates the fact that most of
-the growth of the purple cluster came from the orange cluster. Additionally, the
-lack of transitions is also relevant, for instance, no CT changed from majority
-of Black population (green) to Asian population (pink), and no CT with
-significant Asian population had significant increase in education levels
-(purple).
+can be found in the related literature~\citep{Delmelle2016}, so it is familiar
+to the advanced users. It not only informs the proportional changes, but allows
+the selection of the corresponding trajectories for further analysis.
 
 The panel in the top right of Figure~\ref{fig:ui} is a map of the region under
-analysis, summarizing the geographical evolution of the clusters over time.  The
-colors are derived from the clusters involved in each trajectory as detailed in
-Section~\ref{sec:colors}, which are consistent across the linked views.
+analysis, summarising the geographical evolution of the clusters over time.  The
+colours are derived from the clusters involved in each trajectory, which are
+consistent across the linked views.
 
@@
 maps, depicting the clusters at each year, and the stacked bar plots that
-summarize the overall composition of these regions. Some finer localization
-information is lost using small multiples, such as small border changes, but
-that information is available at the larger map. All the maps are linked with
-synchronized navigation, and the use of small multiples allows the exploration of
-each temporal census individually, and its comparison to the others, with
-minimal interaction.
-
-In this example, the maps show the transition from orange to green and purple in
-several regions over time. Clicking on a region in these maps will bring up a
-new panel with the original census data of this specific region. The actual
-population numbers are below the maps, and they confirm the notion provided by
-the Sankey diagram that the total population is indeed decreasing.
+summarise the overall composition of these regions. In this example, the maps
+show the transition from orange to green and purple in several regions over
+time. Clicking on a region in these maps will bring up a new panel with the
+original census data of this specific region. The actual population numbers are
+below the maps.
 
@@
 rectangle corresponds to the average percentage of that variable over the
-considered period. We chose stacked bar plots to represent the composition of
-the regions because they can accurately and succinctly inform the proportions of
-each aspect, without any interaction. In this case, about half of the people in
-Chicago in the considered period are married, and the percentage that are
-Widowers or Divorced is roughly similar. About half of the population work in
-Administrative jobs, a third never completed high-school, approximately half
-have gross family income below 50,000USD per year. The vast majority identify as
-white. Placing the mouse over one of the bars will open a small panel with the
-temporal evolution of that specific variable, and clicking on the chevron on the
-right side expands the corresponding aspect, showing details of the temporal
-evolution of each variable and also the corresponding IQRs for the whole city.
+considered period. In this case, about half of the people in Chicago in the
+considered period are married, and the percentage that are Widowers or Divorced
+is roughly similar. About half of the population work in Administrative jobs, a
+third never completed high-school, approximately half have gross family income
+below 50,000USD per year. The vast majority identify as white. Placing the mouse
+over one of the bars will open a small panel with the temporal evolution of that
+specific variable, and clicking on the chevron on the right side expands the
+corresponding aspect, showing details of the temporal evolution of each variable
+and also the corresponding IQRs for the whole city.

### 2019-06-13 paper/related.tex
new paper version

--- 2018-08-13 paper/related.tex
@@
 \section{Related Work}
+\label{sec:related}
 
-Since our problem emcompasses several fields, we divided this section into
-specific subproblems: \emph{longitudinal demographic studies}, describing the
-traditional approach to perform longitudinal studies; \emph{data
-representation}, exploring how evolving geographic data can be represented for
-processing; \emph{Data clustering}, briefly reviewing existing clustering
-methods; and \emph{cluster characterization}, exploring how the clusters can be
-visually summarized.
+Since our problem encompasses several fields, we divide this section into
+specific sub problems: \emph{longitudinal demographic studies}, describing the
+traditional tabular approach to longitudinal studies; \emph{data
+representation}, elaborating how evolving geographic data can be represented for
+processing; \emph{data clustering}, briefly reviewing existing clustering
+methods; and \emph{cluster characterisation}, articulating how clusters can be
+visually summarised.
 
@@
 Census data is used not only to discover demographic
-patterns~\cite{Firebaugh2016}, but to correlate demographic charateristics to
-other measurements~\cite{diez1997neighborhood}. However, longitudinal studies
-are rare: \emph{"(...) One of the most challenging and fascinating areas in
-spatial statistics is the synthesis of spatial data collected at different
-spatial scales(...)"}~\cite{gotway2002combining}.
+patterns~\citep{Firebaugh2016}, but to correlate demographic characteristics to
+other measurements~\citep{diez1997neighborhood}. However, longitudinal studies
+are rare, because they are difficult : \emph{"(...) One of the most challenging
+and fascinating areas in spatial statistics is the synthesis of spatial data
+collected at different spatial scales(...)"}~\citep{gotway2002combining}. While
+census tract level data is readily available for the US since at least
+1910~\citep{nhgis}, most studies consider the period between 1970 and 2010,
+using pre-harmonised data from the Longitudinal Tract Data
+Base~\citep{Logan2014}. Despite its inherent
+errors~\citep{Logan2016,Hallisey2017}, this dataset has become the standard
+source for longitudinal demographic data at the neighbourhood scale, with
+similar efforts appearing in other countries~\citep{Liu2015,Lee2015,Allen2018}.
+These datasets have been highly significant for the field. Yet they also limit
+the universe of  data that can be used to study neighbourhood change, since any
+new datasets would need to be similarly processed in order to be rendered
+compatible with these sources.
 
-While CT level data is readily available for the US since 1910~\cite{nhgis},
-most studies consider the period between 1970 and 2010, using pre-harmonized
-data~\cite{Logan2014,nhgis}. Despite the inherent
-errors~\cite{Logan2016,Hallisey2017}, this dataset became the standard source
-for longitudinal demographic data, with similar efforts appearing in other
-countries~\cite{Liu2015,Lee2015,Allen2018}. This result was significant for the
-field, but it also restricts the useable data, since new datasets need to be
-similarly processed.
+Another option considers the use of grid
+data~\citep{Dmowska2017,Dmowska2018,stepinski2019imperfect}. Beyond the
+increased spatial accuracy, this approach does not require complex harmonisation
+when new data is considered, if the grids are compatible. However, demographic
+data is usually not available in this format, especially from older sources.
+Additionally, the conversion from tabulation areas can introduce significant
+errors.
 
-Another option considers the use of grid data~\cite{Dmowska2017,Dmowska2018},
-where small rectangular areas are used, in an approach similar to satellite
-imagery. Beyond the increased spatial accuracy, this approach does not require
-complex harmonizations when new data is considered. However, demographic data is
-usually not available in this format, especially from older sources, and the
-conversion from tabulation areas can introduce significant errors.
-
-In the proposed methodology, we avoid the harmonization by considering each
-measurement using its actual geographic region. It does not require the regions
-to be consistent across time because they are already represented as different
-entities.
+Given these challenges, it is worth considering new alternatives. In this work,
+we propose a novel methodology that entirely avoids the problems of geographical
+harmonisation, considering each measurement using its actual geographic region.
+It does not require regions to be consistent across time because they are
+naturally represented as different entities.
 
 \subsection{Data representation}
-Most data is represented in tabular form, where the rows and columns have
-coherent definitions. For example, consider a table with rain measurements over
-time, with the rows representing different locations and the columns different
-times. This representation can also be interpreted as a collection of
-time-series, one for each location. Geographic data followed this format, only
-including an additional field that describes the associated geographic area.
-Following the example, the data would now represent the amount of rain for a
-given region and time. As long each region remains the same, the data is
-coherent and can be interpreted again as a collection of time-series.
 
-In the proposed method, we remove the requirement for consistency in the
-measurement regions by leveraging a graph-based representation, where each
-region in time corresponds to a different node. Instead of a collection of
-time-series, the data is represented as a dynamic graph. Graph based
-representation of geographic information is fairly well explored in the
-literature, as a basis for topological methods for event
-detection~\cite{Doraiswamy2014}, leveraging signal processing on
-graphs~\cite{shuman2013emerging,sandryhaila2013discrete} to find patterns and
-outliers~\cite{Valdivia2015,Dias2015,Alce2018}. Graphs are well suited to
+Network based representation of geographic information is fairly well explored
+in the literature, as a basis for topological methods for event
+detection~\citep{Doraiswamy2014}, leveraging signal processing on
+graphs~\citep{shuman2013emerging,sandryhaila2013discrete} to find patterns and
+outliers~\citep{Valdivia2015,Dias2015,Alce2018}. Networks are well suited to
 represent trajectories as
-well~\cite{VonLandesberger2016,Huang2016,chen2015survey}, allowing the use of
-graph visualization methods~\cite{Vehlow2015,Beck2014}.
+well~\citep{VonLandesberger2016,Huang2016,chen2015survey}, allowing the use of
+graph visualisation methods~\citep{Vehlow2015,Beck2014}. Our proposed method
+builds upon this literature. We leverage a network-based representation that
+removes the requirement for consistency in the measurement regions. Each region
+in time corresponds to a different node. Instead of a collection of time-series,
+the data is represented as a dynamic network.
 
-Graphs were used to represent census data for clustering purposes
-before~\cite{Dias2015,Setiadi2017}, but these works did not explore temporal
-evolution, where graphs are particular powerful as they allow a natural
-representation of inconsistent regions, with both spatial and temporal
-connections. Note that there are other possible representations that have
-similar properties, but we adopted graphs to allow the use of the existing
+Networks have been used to represent census data for clustering purposes
+~\citep{Dias2015,Setiadi2017}, but these works did not explore temporal
+evolution, where they are particularly powerful. Networks allow a natural
+representation of these inconsistent regions, with both spatial and temporal
+connections. There are other possible representations that have similar
+properties, but we adopted networks to allow the use of the vast existing
 literature and methods.
@@
 \subsection{Data clustering}
+
 Data clustering is one of the elementary processes for data analysis,
 simplifying the data into a smaller number of homogeneous sets that can be
-interpreted in the same way. While there is no shortage of contributions for
-this problem~\cite{Fahad2014}, most applications still rely on
-k-means~\cite{jain2010data,Delmelle2016} and, to a lesser extent, Self
-Organizing Maps~\cite{Delmelle2017,Ling2016}.
+interpreted in the same way. There is no shortage of contributions for this
+problem~\citep{Fahad2014}, but most neighbourhood related applications still
+rely on k-means~\citep{jain2010data,Delmelle2016} and, to a lesser extent, Self
+Organising Maps~\citep{Delmelle2017,Ling2016}.
 
 However, a method for geographic data analysis should not ignore the geographic
-component of the data. One straightforward option, for agglomerative
-methods~\cite{han2001spatial}, is to consider only nearby clusters for
-merging~\cite{Chavent2017}, which can also be done for
-k-means~\cite{soor2018extending}. Alternatively, the spatial distance could be
-directly added to the inter-cluster metric~\cite{Chavent2017} via a mixing
-parameter, which adds flexibility to the method, but introduces the problem of
-finding the correct application-dependent values.
+component of the data, and we extend research exploring ways to incorporate it.
+One straightforward option for agglomerative methods~\citep{han2001spatial} is
+to consider only nearby clusters for merging~\citep{Chavent2017}, which can also
+be done for k-means~\citep{soor2018extending}. Alternatively, the spatial
+distance could be directly added to the inter-cluster metric~\citep{Chavent2017}
+via a mixing parameter, which adds flexibility to the method, but introduces the
+problem of finding the correct application-dependent values.
 
 Indeed, one crucial step in most clustering algorithms is the definition of the
-number of clusters. We sidestep this problem by considering hierarchical
-methods~\cite{soille2012morphological}, where the result is not a partition of
-the data, but a tree of partitions. This approach is interesting for interactive
-methods, because it allows the user to change the number of displayed clusters
-with minimal processing. Since our data is represented as a graph, one option
-would be the watershed cuts algorithm~\cite{Cousty2009}, inspired by the well
-known image processing segmentation and equally prone to oversegmentation.
-Considering that the processing time is also a relevant factor, we opted for an
-heuristic variation of the maximum weighted matching algorithm called
-\emph{sorted maximal matching}~\cite{markus2017}, which merges clusters based on
-the weights of the edges between pairs of clusters.
+number of clusters. We avoid this problem by considering hierarchical
+methods~\citep{soille2012morphological}, where the result is not a partition of
+the data, but a tree of partitions, similar to a dendogram. This approach is
+interesting for interactive methods, because it allows the user to change the
+number of clusters on the fly. Since our data is represented as a network, we
+opted for an heuristic variation of the maximum weighted matching algorithm
+called \emph{sorted maximal matching}~\citep{markus2017}, which merges clusters
+based on the weights of the edges between pairs of clusters.
 
-\subsection{Cluster characterization}
-Visually representing evolving spatial data is a challenging old
-problem~\cite{monmonier1990strategies,andrienko2003exploratory,ferreira2015visual,Zheng2016}.
+\subsection{Cluster characterisation}
+While visualisation has gained prominence as a crucial component of scientific
+discovery, justification, and communication \todo{Add Cites: maybe Gelman?
+Tufte?}, visually representing evolving spatial data is a challenging old
+problem~\citep{monmonier1990strategies,andrienko2003exploratory,ferreira2015visual,Zheng2016}.
+
 Most geographic data is naturally bidimensional and maps work well in this
-case~\cite{Zheng2016,ward2015interactive}, but the temporal dimension cannot be
-so naturally represented. One straightforward option is to leverage
-tridimensional plots~\cite{andrienko2014visualization,Tominski2012a}, but this
-can lead to visual obstructions or scaling problems unless a tridimensional
-display device is used. Animation can also be explored in some specific
-cases~\cite{buschmann2014real}, but it is not a general approach. Glyphs can
-also be used~\cite{seebacher2017visual,Andrienko2017}, but this may lead to
-cluttering when many small regions are present. A simpler, well adopted, option
-is to display a map that corresponds to a subset of the temporal information,
-allowing the user to change the time with an associated
-control~\cite{Chen2017,Valdivia2015,Alce2018,Doraiswamy2014}. Small multiples
-can be used~\cite{VonLandesberger2016}, but only when there are few temporal
+case~\citep{Zheng2016,ward2015interactive}, but the additional temporal
+dimension cannot be so naturally represented. One straightforward option is to
+leverage tridimensional plots~\citep{andrienko2014visualization,Tominski2012a},
+but this can lead to visual obstructions or scaling problems unless a
+tridimensional display device is used. A simpler, well adopted,
+option is to display a map that corresponds to a subset of the temporal
+information, allowing the user to change the time with an associated
+control~\citep{Chen2017,Valdivia2015,Alce2018,Doraiswamy2014}. Small multiples
+can be used~\citep{VonLandesberger2016}, but only when there are few temporal
 snapshots. However, none of these options is suitable to represent many
@@
 Using data clustering, we can represent the region's cluster instead of all the
-its variables~\cite{Alce2018,Valdivia2015,VonLandesberger2016}. While this
+its variables~\citep{Alce2018,Valdivia2015,VonLandesberger2016}. While this
 simplifies the geographic portion of the visualization, it introduces the
-problem of how to summarize the contents of each cluster. One traditional
-approach is to use parallel coordinates
-plot~\cite{ferreira2015urbane,LI2018,johansson2005revealing,guo2006visualization},
-but these they can get cluttered representing similar clusters over several
+problem of how to summarise the contents of each cluster. One traditional
+approach is to use parallel coordinates plot~\citep{ferreira2015urbane}, but
+these they can get cluttered representing similar clusters over several
 variables. Further, for demographic applications, the clusters are usually
-strongly characterized by a small subset of
-values~\cite{Delmelle2016,Delmelle2017}. Therefore, in the proposed method, we
-identify the variables that are most relevant to the characterization of each
+strongly characterised by a small subset of
+values~\citep{Delmelle2016,Delmelle2017}. Therefore, in the proposed method, we
+identify the variables that are most relevant to the characterisation of each
 cluster. The distribution of values on that variable is then represented using a

### 2019-06-13 paper/study.tex
new paper version

@@
 
-We used decennial census data from the United States~\cite{nhgis} and
+In this section we present three illustrative scenarios, using decennial census
+data from the United States~\citep{nhgis} and
 Canada\footnote{\url{http://datacentre.chass.utoronto.ca/census/}}, tabulated by
-CTs,  from 1970 to 2010. The prototype allows access to 40 regions, 28 in the US
-and 12 in Canada. Due to the high number of CTs, New York City was split into
-its boroughs.
+CTs,  from 1970 to 2010. The prototype interface allows access to 41 regions, 29
+in the US and 12 in Canada. New York City was split into its boroughs to avoid
+memory crashes on the client browser due to the high number of CTs.  We used
+five aspects for the USA: Education level, Family income, Marital status,
+Occupation, and Race; and seven for Canada: Age, Education level, Home language,
+Household Income, Marital status, Occupation, Place of birth, and Religion.
 
-We used five aspects for the USA: Education level, Family income, Marital
-status, Occupation, and Race; and seven for Canada: Age, Education level, Home
-language, Household Income, Marital status, Occupation, Place of birth, and
-Religion.  While our method does not require geographic harmonization, it
-requires matching variables over time. The supplementary material contains the
-details of which census columns were used for each aspect. Income is slightly
-inaccurate, even though we did correct for official inflation. We grouped the
-original ranges into three larger ranges, but they do not match precisely.
+While our method does not require geographic harmonisation, it requires matching
+the variables over time. The supplementary material contains the details of
+which census columns were used for each aspect. Income is slightly inaccurate,
+even though we did correct for the official inflation. We grouped the original
+ranges into three larger ranges, but they do not match precisely.
+
+These results are meant to demonstrate the utility of the interface for
+understanding the evolutionary dynamics of urban neighbourhoods. They also show
+the face validity of the results generated by our novel network-based approach.
 
 \subsection{Chicago}
-We selected a region loosely following the administrative borders. The
-demographic composition is well explored in the literature, with reports of
-racial divide and gentrification~\cite{Delmelle2016,Delmelle2017,Hwang2014}.
-While the definition of gentrification is still unclear and out of the scope of
-this paper, we associate gentrification with higher education and income levels.
+Our first scenario examines Chicago, focusing on a region loosely following the
+City's administrative borders. Its demographic composition is well explored in
+the literature, with reports of racial divide and
+gentrification~\citep{Delmelle2016,Delmelle2017,Hwang2014}. Our goal is
+therefore to confirm that our novel method can reproduce existing findings.
 
-The initial state of the prototype is illustrated in Figure~\ref{fig:ui}, and
-its findings are explained in Section~\ref{sec:ui}, where the racial divide is
-clear. Starting from this initial state, the specific workflow used to identify
-the existence and details of the gentrification process are illustrated in
-Figure~\ref{fig:chiWorkflow}. For the users, the first step is to identify the
-compositions of each cluster from the boxplots, so orange is associated with
-majority of White population, green with majority Black, and purple with higher
-proportion of four years of college or more (high education level). The expanded
-version of the boxplots for the purple cluster shows a higher income level and
-majority of occupations in administrative jobs, therefore the purple cluster
-identifies gentrified regions.
+The initial state of the prototype is illustrated in Figure~\ref{fig:ui}. The
+specific workflow used to identify the existence and details of the
+gentrification process is illustrated in Figure~\ref{fig:chiWorkflow}. The first
+step is to identify the compositions of each cluster from the boxplots, so
+orange is associated with majority of White population, green with majority
+Black, and purple with higher proportion of four years of college or more (high
+education level). The expanded version of the boxplots for the purple cluster
+shows a higher income level and majority of occupations in administrative jobs,
+therefore the purple cluster identifies gentrified regions.
 
@@
 absorbing regions from the orange cluster (White). This corroborates results
-from the literature reporting that Black neighborhoods are less likely to
-gentrify~\cite{Hwang2014}. Moreover, this process is unlikely to be reversed, as
+from the literature reporting that Black neighbourhoods are less likely to
+gentrify~\citep{Hwang2014}. Moreover, this process is unlikely to be reversed, as
 indicated by the limited number of trajectories leaving the purple stream. Next,
@@
 
-The corresponding CTs are highlighted in the maps, where the spatial pattern is
-clear, corresponding exactly to previous findings in the
-literature~\cite{Hwang2014}. Further, we can also identify the regions that
-gentrified earlier, with an stronger purple hue, compared to newer regions,
-where the orange and green colors are still present. This also indicates from
-which clusters they belonged before gentrifying. In the details portion of the
-interface, the order of the aspects was updated to reflect the order of
-relevance considering only the selected region. The most relevant aspect is the
-Education, specifically "Four or more years of college", illustrated in the
-rightmost portion of Figure~\ref{fig:chiWorkflow}, which is increasing for the
-whole city (grey band), but faster and to a higher level in this region (black
-band). Indeed, while the IQR for the city goes from 11\% to 45\%, the IQR for
-this region goes from 55\% to 77\%. However, there are portions of this region
-with significantly lower or higher proportions, as indicated by the dotted black
-lines representing the minimum and maximum for the selected region.
+The corresponding regions are highlighted in the maps, where the spatial pattern
+is clear, corresponding exactly to previous findings in the literature based
+upon harmonisation~\citep{Hwang2014}. Further, we can also identify the regions
+that gentrified earlier on the small maps that depict the involved regions over
+time. Since the most relevant aspect is Education, specifically "Four or more
+years of college", we can expand the details of this aspect, as illustrated in
+the rightmost portion of Figure~\ref{fig:chiWorkflow}, which is increasing for
+the whole city (grey band), but faster and to a higher level in this region
+(black band).
 
@@
     \centering
- \includegraphics[width=0.85\linewidth]{newTeaser.pdf}
+ \includegraphics[width=0.9\linewidth]{newTeaser.pdf}
  \caption{Workflow to discover gentrification in Chicago: the purple cluster
@@
 
-We considered a region that is approximately the administrative border of the
-current city of Toronto and all seven available aspects with equal weights. The
-results are summarized in Figure~\ref{fig:to}, considering eight clusters.
+We consider a region that is approximately the administrative border of the
+current city of Toronto, using all seven available aspects with equal weights.
+The results are summarised in Figure~\ref{fig:to}, considering eight clusters.
 
@@
     \centering 
-    \includegraphics[width=0.95\linewidth]{to.pdf}
+    \includegraphics[width=0.85\linewidth]{to.pdf}
     \caption{Clustering results for Toronto, with eight clusters, including
@@
 people in light green and brown, high percentage of Eastern Non-Christian
-religion in pink, and high concentration of single people in dark gray. From the
+religion in pink, and high concentration of single people in dark grey. From the
 trajectories plot, we can see that Toronto is more dynamic than Chicago, with
@@
 
-Indeed, the influx of Asian population is visible starting in the 1980s and
+Indeed, the growing Asian population is visible starting in the 1980s and
 building thereafter, leading to the yellow and pink clusters. While both include
@@
 has lower education and income levels than the brown cluster. The purple cluster
-of high income is somewhat stable. This cluster includes the Bridle Path
-neighborhood, known for its wealthy population, until 2011. In 2011 it was
+of high income is somewhat stable. Until 2011 the cluster included the Bridle Path
+neighbourhood, known for its wealthy population. In 2011 it was
 classified into the yellow cluster of Asian immigration, since about 35\% of the
@@
 The most significant indicator of Toronto's dynamism is the presence of grey
-regions on the simplified color map; representing regions associated to three or
-more clusters over this five census period. Using the 'Add' mode for the
-trajectory selection, we select their trajectories, and a subset of the details
-is illustrated in Figure~\ref{fig:toVol}. These regions account for about 5\% of
-Toronto's population. The whole region was classified into the orange cluster in
-1971 (low level of university degrees). By 1991, most of the region was
-classified into the green cluster, representing anglophone population, mostly
-Canadian born, with a higher level of education. As the corresponding plot
-indicates, this trend in increasing education is city-wide, but this region has
-people with better education than most.
+regions on the map. These represent regions associated to three or more clusters
+over this five census period. Using the 'Add' mode for the trajectory selection,
+we select their trajectories, and a subset of the details is illustrated in
+Figure~\ref{fig:toVol}. These regions account for about 5\% of Toronto's
+population. The whole region was classified into the orange cluster in 1971 (low
+level of university degrees). By 1991, most of the region was classified into
+the green cluster, representing anglophone population, mostly Canadian born,
+with a higher level of education. As the corresponding plot indicates, this
+trend in increasing education is city-wide, but this region has people with
+better education than most.
 
@@
     \centering 
-    \includegraphics[width=0.95\linewidth]{toVol.pdf}
+    \includegraphics[width=0.75\linewidth]{toVol.pdf}
     \caption{Details for some regions of Toronto that were classified into 3 or
@@
 
-In 2001, the purple cluster of high income annexes neighboring parts of the
+In 2001, the purple cluster of high income annexes neighbouring parts of the
 volatile region, and the Asian born population increases sharply, as illustrated
@@
 cluster increased considerably, annexing parts of the high income purple
-cluster, including the neighboring Bridle Path area.
+cluster, including the neighbouring Bridle Path area.
+
+The geographical borders of the clusters obtained using our method are similar
+to the regions presented by previous studies considering
+Toronto~\citep{hulchanski2007three}, but our interface provides a deeper insight
+into their demographic composition, since we consider more data than solely
+Average Income, which appears to be a good proxy variable nonetheless. This
+scenario showcases the ability of our method and interface to capture and
+understand the sources of urban volatility.
 
 \subsection{Los Angeles}
-We selected a region around the metropolitan area of Los Angeles (LA), following
-urban density. The summary of the results using all aspects with equal weights
-and eight clusters is illustrated in Figure~\ref{fig:la}. The full statistical
-description of the clusters is illustrated in Figure~\ref{fig:laAll}, where the
-most relevant aspect of each cluster is highlighted. From the trajectories plot,
-we can see that there is a large but shrinking cluster, depicted in green, one
-increasing cluster in purple, an almost constant orange cluster, a smaller but
-increasing pink cluster, and three other small clusters. The corresponding map
-illustrates where these clusters are located, and that they are somewhat
-geographically stable, with some movement between the green, orange, and purple
-clusters.
+
+We selected for this scenario a region around the metropolitan area of Los
+Angeles (LA), following urban density. The summary of the results using all
+aspects with equal weights and eight clusters is illustrated in
+Figure~\ref{fig:la}. The full statistical description of the clusters is
+illustrated in Figure~\ref{fig:laAll}, where the most relevant aspect of each
+cluster is highlighted. From the trajectories plot, we can see that there is a
+large but shrinking cluster, depicted in green, one increasing cluster in
+purple, an almost constant orange cluster, a smaller but increasing pink
+cluster, and three other small clusters. The corresponding map illustrates where
+these clusters are located, and that they are somewhat geographically stable,
+with some movement between the green, orange, and purple clusters.
 
@@
     \centering 
-    \includegraphics[width=0.95\linewidth]{LA.pdf}
+    \includegraphics[width=0.975\linewidth]{LA.pdf}
     \caption{Result for Los Angeles with 8 clusters, including three small and
-        ephemeral clusters. Cluster characterization is displayed
+        ephemeral clusters. Cluster characterisation is displayed
         in Figure~\ref{fig:laAll}.\label{fig:la}}
@@
 
-From Figure~\ref{fig:laAll}, we can see that the green cluster is characterized
+From Figure~\ref{fig:laAll}, we can see that the green cluster is characterised
 by a high percentage of White population, low percentage of population in the
@@
 population with four or more years of college. The orange cluster is
-characterized by a high percentage of Black population, with few people in the
+characterised by a high percentage of Black population, with few people in the
 highest range of income and education. The purple cluster corresponds to a high
 concentration of "Other" in race, which includes Hispanic for this dataset, high
-concentration of Laborers, and low education and income. The pink cluster
+concentration of Labourers, and low education and income. The pink cluster
 contains a high percentage of Asian population and about 30\% of the population
@@
 working administrative jobs. The brown cluster represent a majority of single
-population, working administrative jobs with mostly low income. The dark gray
-cluster is characterized by all its population in the lowest income bracket, low
+population, working administrative jobs with mostly low income. The dark grey
+cluster is characterised by all its population in the lowest income bracket, low
 education level, with a majority of White population. Since the extremes in the
@@
     \includegraphics[width=0.95\linewidth]{laAll.pdf}
-    \caption{Full characterization of the eight clusters found for LA. The red
+    \caption{Full characterisation of the eight clusters found for LA. The red
      rectangles indicate the most relevant aspects for each
@@
 to be separated at this level of the hierarchy. Conversely, the light green,
-brown, and dark gray clusters are different enough to be separated into their
+brown, and dark grey clusters are different enough to be separated into their
 own clusters at this level, despite being small and ephemeral, including only a
 few CTs.
-
-% While they might be interesting enough to
-% warrant further study, they might suppress larger, but more subtle, clusters
-% from appearing in this interface due to the limited number of clusters.
 
@@
     \centering 
-    \includegraphics[width=0.95\linewidth]{laBH.pdf}
+    \includegraphics[width=0.75\linewidth]{laBH.pdf}
     \caption{Top: Geographic changes in the majority Black population cluster
-    (orange) and Laborers cluster (green). Bottom: Income evolution for this
-    region (black) and the whole city (gray).\label{fig:labh}}
+    (orange) and Labourers cluster (green). Bottom: Income evolution for this
+    region (black) and the whole city (grey).\label{fig:labh}}
 \end{figure}
@@
     \centering 
-    \includegraphics[width=0.95\linewidth]{laVol.pdf}
+    \includegraphics[width=0.75\linewidth]{laVol.pdf}
     \caption{Details for a volatile region contained in the area of
@@
 
+If Toronto illustrates a relatively smooth process of whole-sale
+diversification, Los Angeles shows the ability of our system and method to
+uncover more intensive smaller-scale processes of neighbourhood change. This
+capacity to examine and classify different types of change operating at
+different scales demonstrates the flexibility of the method and tool for
+advancing the more open-ended notion of "neighbourhood" envisioned by recent
+neighbourhood research.
+
 \section{Expert feedback}
-We contacted academic and industry experts in sociology and urban sciences for
-their appreciation of our methodology. They had access to the prototype tool, a
-descriptive documentation of the features (included in the supplementary
-material), and a sequence of documentation videos illustrating how to perform
-specific tasks. The documentation explains which datasets are used, how the data
-is represented and processed, including that there is no geographic
-harmonization. We focused our inquires on the results obtained, asking if they
-found anything interesting on the data. The message sent and their full response
-is included in the supplementary material. Each of the five experts is
-identified by a letter, from A to E.
+\label{sec:expert}
+As our method and tool are novel to the field, and somewhat exotic,  we
+subjected them to the critical scrutiny of experts. We contacted academic and
+industry experts in sociology and urban sciences to solicit their evaluation of
+our methodology. They had access to the prototype tool, a descriptive
+documentation of the features (included in the supplementary material), and a
+sequence of documentation videos illustrating how to perform specific tasks. The
+documentation explains which datasets are used and how the data is represented
+and processed, noting explicitly that there is no geographic harmonisation. We
+focused our inquires on the results obtained, asking if they found anything
+interesting in the data. The message sent and their full response is included in
+the supplementary material. Each of the five experts is identified by a letter,
+from A to E.
 
-Their overall response was positive,  mentioning that the prototype allows them
-to analyze census data without the additional work of obtaining and cleaning the
-data (A, B, E), and it allows the inclusion of geographic visual analysis tools
-in their research process (D). It enables the users to tell different stories
-about neighborhoods/cities and their changes (A), visualize the relationship
-between key urban variables over time (D), offering a quick way to identify
-particular neighborhoods that one may be interested in studying more in depth
-around a particular issue or efficiently understanding the context of an area
-(E).  Indeed, the experts identified gentrification processes in Manhattan (B)
-and Dallas (E), reinforced a hypothesis for occupational clustering (D), and
-highlighted how the method can be used to compare neighborhoods and cities (A).
-In summary, the proposed methodology can be a viable alternative for the visual
-analytics of evolving demographic data.
+The overall overall response of the experts was positive,  mentioning that the
+prototype allows them to analyse census data without the additional work of
+obtaining and cleaning the data (A, B, E), and it allows the inclusion of
+geographic visual analysis tools in their research process (D). It enables the
+users to tell different stories about neighbourhoods/cities and their changes
+(A), visualise the relationship between key urban variables over time (D),
+offering a quick way to identify particular neighbourhoods that one may be
+interested in studying more in depth around a particular issue or efficiently
+understanding the context of an area (E).  Indeed, the experts identified
+gentrification processes in Manhattan (B) and Dallas (E), reinforced a
+hypothesis for occupational clustering (D), and highlighted how the method can
+be used to compare neighbourhoods and cities (A). In summary, their view was that
+the proposed methodology can be a viable alternative for the visual analytics of
+evolving demographic data.
 
@@
 most relevant variables was welcome, but the overlap of information from
-different clusters in the boxplot was "a bit confusing" (C) when color was not
-present. Further, most clusters can be sufficiently characterized using only the
+different clusters in the boxplot was "a bit confusing" (C) when colour was not
+present. Further, most clusters can be sufficiently characterised using only the
 most relevant aspect, but this is not generally true.
@@
 While the map of trajectories was mentioned as a "good summary map", how it
-related to the clustering method was unclear (C). The methods includes different
-options on how the colors are used, but both are sub optimal since reliably
-representing several distinct entities using colors is humanly unfeasible.
-Indeed, the number of distinguishable colors was a significant constraint, we
+related to the clustering method was unclear (C). The methods include different
+options on how the colours are used, but both are works in progress since reliably
+representing several distinct entities using colours is humanly unfeasible.
+Indeed, the number of distinguishable colours was a significant constraint, we
 found indications that more clusters should be used in some cases, even if eight
@@
 importantly, while these experts were aware that our method does not perform
-geographic harmonization, none of them mention it. We did not specifically ask
+geographic harmonisation, none of them mention it. We did not specifically ask
 if this difference led to unexpected results, but rather if they found
 interesting insights. Most experts found phenomena corroborated by the
-specialized literature, indicating that our methodology produces equivalent
+specialised literature, indicating that our methodology produces equivalent
 results, with a fraction of the effort. We interpret the fact that most of them

### 2019-06-13 backend/dataStore.py

@@ -1,17 +1,18 @@
+import json
+from glob import glob
+from os.path import basename, exists, join
+from random import choice
+from uuid import uuid4
+
 import matplotlib.pylab as plt
-from os.path import exists, join
+import networkx as nx
 import numpy as np
 import pandas as pd
-import json
-import networkx as nx
-from os.path import join, basename, exists
-from glob import glob
-from hierarchies import compareHierarchies
-from uuid import uuid4
-from clustering import ComputeClustering
-from random import choice
-from sklearn.preprocessing import minmax_scale
 from scipy.stats import skew
+from sklearn.preprocessing import minmax_scale
+
+from clustering import ComputeClustering
+from hierarchies import compareHierarchies
 
 MAX_CACHE = 10
 

### 2019-06-13 backend/srv.py

@@ -160,7 +160,7 @@ def _mapHiers(ds: dataStore, aspects: list, thresholds: list):
                    'y':points[a][n[1]]}, ]
                  for n in M if n[0] == g]
 
-        for i, info in enumerate(full_info_aspects[1:]):
+        for info in full_info_aspects[1:]:
             g = info['geom']
             a = info['id']
             to_add = []
@@ -170,7 +170,7 @@ def _mapHiers(ds: dataStore, aspects: list, thresholds: list):
                 last = {**current[-1]}
                 if last['geom'] == g:  # same geometry:
                     to_add.append(current+[last, ])
-                    to_add[-1][-1]['x'] = i+1
+                    to_add[-1][-1]['x'] = a
                     to_add[-1][-1]['y'] = points[a][last['id']]
                     unused.discard(last['id'])
                 else:
@@ -181,12 +181,12 @@ def _mapHiers(ds: dataStore, aspects: list, thresholds: list):
                         unused.discard(op[1])
                         to_add.append(current+[{'geom': g,
                                                 'id': op[1],
-                                                'x':i+1,
+                                                'x':a,
                                                 'y':points[a][op[1]]}, ])
             # puts in the paths that didn't start at the first aspect
             paths = to_add+[[{'geom': g,
                               'id': x,
-                              'x': i+1,
+                              'x': a,
                               'y': points[a][x]}, ] for x in unused]
 
         # plt.figure()
@@ -210,7 +210,7 @@ def _mapHiers(ds: dataStore, aspects: list, thresholds: list):
         #     nx.draw_networkx_edge_labels(MM,pos=pos,edge_labels={e:'{0}'.format(round(MM[e[0]][e[1]]['area'],2)) for e in MM.edges()})
         #     plt.show()
 
-    return({'clustering': cl, 'evolution': paths})
+    return({'clustering': cl, 'evolution': paths, 'aspects': full_info_aspects})
 
 
 @cherrypy.expose

### 2019-06-13 paper/related.tex
new reference

@@
 While visualisation has gained prominence as a crucial component of scientific
-discovery, justification, and communication \todo{Add Cites: maybe Gelman?
-Tufte?}, visually representing evolving spatial data is a challenging old
+discovery, justification, and communication~\cite{tufte1998visual}, visually
+representing evolving spatial data is a challenging old
 problem~\citep{monmonier1990strategies,andrienko2003exploratory,ferreira2015visual,Zheng2016}.

### 2019-06-13 sup/sup.tex
new reference

@@
 
+\newcommand{\censure}[1]{}
 \begin{document}
@@
 \section{Original request for feedback}
-As part of our efforts in the Urban Genome Project of the University of Toronto,
-we are trying to build infrastructure to help understand recurrent and
-distinctive patterns of urban change.
+As part of our efforts in \censure{the Urban Genome} Project of the
+\censure{University of Toronto}, we are trying to build infrastructure to help
+understand recurrent and distinctive patterns of urban change.
 
@@
 
-This prototype is part of an article, to be submitted to the next IEEE VIS
-conference, and we are humbly requesting feedback on how effective it actually
-is.
+This prototype is part of an article, and we are humbly requesting feedback on
+how effective it actually is.
 
@@
 
-\href{http://ug.daniels.utoronto.ca/~diasf/CTEvo/doc.html}{\url{http://ug.daniels.utoronto.ca/~diasf/CTEvo/doc.html}}
+\censure{\href{http://ug.daniels.utoronto.ca/~diasf/CTEvo/doc.html}{\url{http://ug.daniels.utoronto.ca/~diasf/CTEvo/doc.html}}}
 
@@
 
-\href{https://www.youtube.com/watch?v=OZFUA0ThEaY&list=PLBd6KRI4PG-RcxAX35OfHd1DKl3qWfXRV}{youtube}
+\censure{\href{https://www.youtube.com/watch?v=OZFUA0ThEaY&list=PLBd6KRI4PG-RcxAX35OfHd1DKl3qWfXRV}{youtube}}
 
@@
 
-\href{http://ug.daniels.utoronto.ca/~diasf/CTEvo/}{\url{http://ug.daniels.utoronto.ca/~diasf/CTEvo/}}
+\censure{\href{http://ug.daniels.utoronto.ca/~diasf/CTEvo/}{\url{http://ug.daniels.utoronto.ca/~diasf/CTEvo/}}}
 
@@
 \section{Expert comments}
-\subsection{A - Cary Wu}
-%\href{https://soci.ubc.ca/persons/cary-wu/}{\url{https://soci.ubc.ca/persons/cary-wu/}}
-\emph{Doctoral candidate in Sociology at the University of British Columbia, Canada.}
+
+\subsection{A\censure{ - Cary Wu}}
+\censure{\emph{Doctoral candidate in Sociology at the University of British Columbia, Canada.}}
 
@@
 
-\subsection{B - Ethan Fosse}
+\subsection{B\censure{ - Ethan Fosse}}
 % \href{https://scholar.harvard.edu/ethanfosse}{\url{https://scholar.harvard.edu/ethanfosse}}
-\emph{Research Associate at Princeton University, USA}
+\censure{\emph{Research Associate at Princeton University, USA}}
 
@@
 
-\subsection{C - Fernando Calderón Figueroa}
-\emph{Doctoral candidate in Sociology at the University of Toronto, Canada}
+\subsection{C \censure{- Fernando Calderón Figueroa}}
+\censure{\emph{Doctoral candidate in Sociology at the University of Toronto, Canada}}
 
@@
 
-\subsection{D - Patrick Adler}
-\emph{Research Associate at the Martin Prosperity Institute, Canada.}
+\subsection{D\censure{ - Patrick Adler}}
+\censure{\emph{Research Associate at the Martin Prosperity Institute, Canada.}}
 
@@
 
-\subsection{E - James Murdoch}
-\emph{Policy Analyst at 2M Research, USA}
+\subsection{E\censure{ - James Murdoch}}
+\censure{\emph{Policy Analyst at 2M Research, USA}}
 

### 2019-06-14 paper/study.tex
more text around the results

@@
 the literature, with reports of racial divide and
-gentrification~\citep{Delmelle2016,Delmelle2017,Hwang2014}. Our goal is
-therefore to confirm that our novel method can reproduce existing findings.
+gentrification~\citep{Delmelle2016,Delmelle2017,Hwang2014}, so we expect our
+results to contain stable regions where the Race aspect is relevant, and some
+degree of population change, with increasing income and education levels.
 
 The initial state of the prototype is illustrated in Figure~\ref{fig:ui}. The
-specific workflow used to identify the existence and details of the
-gentrification process is illustrated in Figure~\ref{fig:chiWorkflow}. The first
-step is to identify the compositions of each cluster from the boxplots, so
+first step is to identify the compositions of each cluster from the boxplots, so
 orange is associated with majority of White population, green with majority
@@
 
-The trajectories plot illustrates the process of gentrification, progressively
-absorbing regions from the orange cluster (White). This corroborates results
-from the literature reporting that Black neighbourhoods are less likely to
-gentrify~\citep{Hwang2014}. Moreover, this process is unlikely to be reversed, as
-indicated by the limited number of trajectories leaving the purple stream. Next,
-we select the region that is gentrified in 2010, by clicking on the
+The trajectories plot illustrates the process of gentrification, also
+illustrated in Figure~\ref{fig:chiWorkflow}, progressively absorbing regions
+from the orange cluster (White). This corroborates results from the literature
+reporting that Black neighbourhoods are less likely to
+gentrify~\citep{Hwang2014}. Moreover, this process appears to be unidirectional,
+as indicated by the limited number of trajectories leaving the purple stream.
+Next, we select the region that is gentrified in 2010, by clicking on the
 corresponding rectangle in the trajectories plot, updating the information on
@@
 current city of Toronto, using all seven available aspects with equal weights.
-The results are summarised in Figure~\ref{fig:to}, considering eight clusters.
+While Chicago was fairly stable, Toronto is known to be a more dynamic and
+diverse city, with significant and increasing immigrant
+population~\citep{hulchanski2007three}. Toronto is also known for a stable and
+well defined Jewish community~\citep{Shahar2016}. Therefore, we expect the
+combination of stable and dynamic regions on the results, with Place of Birth,
+Home Language, and Religion identified as relevant aspects. The results are
+summarised in Figure~\ref{fig:to}, considering eight clusters.
 
@@
 We selected for this scenario a region around the metropolitan area of Los
-Angeles (LA), following urban density. The summary of the results using all
-aspects with equal weights and eight clusters is illustrated in
-Figure~\ref{fig:la}. The full statistical description of the clusters is
-illustrated in Figure~\ref{fig:laAll}, where the most relevant aspect of each
-cluster is highlighted. From the trajectories plot, we can see that there is a
-large but shrinking cluster, depicted in green, one increasing cluster in
-purple, an almost constant orange cluster, a smaller but increasing pink
-cluster, and three other small clusters. The corresponding map illustrates where
-these clusters are located, and that they are somewhat geographically stable,
-with some movement between the green, orange, and purple clusters.
+Angeles (LA), following urban density. While Chicago and Toronto demonstrated
+the abilities of our method for stable and dynamic cities, LA is considerably
+larger, both in population and area. Since our method identifies only the eight
+most distinct groups, we expect that some groups not to be identified,
+especially if their characterisation is similar to another groups. Further, our
+system does not include the census variables related to Hispanic heritage, which
+were only included in the newer censuses. However, from the literature, we
+expect to see some clusters where Race is an important aspect, including White,
+Black, and Asian~\citep{Reibel2003}.
+
+The summary of the results using all aspects with equal weights and eight
+clusters is illustrated in Figure~\ref{fig:la}. The full statistical description
+of the clusters is illustrated in Figure~\ref{fig:laAll}, where the most
+relevant aspect of each cluster is highlighted. From the trajectories plot, we
+can see that there is a large but shrinking cluster, depicted in green, one
+increasing cluster in purple, an almost constant orange cluster, a smaller but
+increasing pink cluster, and three other small clusters. The corresponding map
+illustrates where these clusters are located, and that they are somewhat
+geographically stable, with some movement between the green, orange, and purple
+clusters.
 
@@
 
+These results are visually similar the analysis presented by
+Dellmelle~\citep{Delmelle2016}, what was there identified as 'persistently
+struggling' corresponds to the purple cluster in Figure~\ref{fig:la}; some
+portions of the 'stable elite' cluster, near Inglewood, correspond to the high
+income, high education yellow cluster. The other groups identified by that study
+were clumped by our system into the green cluster, which would be further
+divided if more than eight clusters were to be used.
+
+If Toronto illustrates a relatively smooth process of whole-sale
+diversification, Los Angeles shows the ability of our system and method to
+uncover more intensive smaller-scale processes of neighbourhood change. This
+capacity to examine and classify different types of change operating at
+different scales demonstrates the flexibility of the method and tool for
+advancing the more open-ended notion of "neighbourhood" envisioned by recent
+neighbourhood research.
+
 \smallskip While Toronto is more dynamic than Los Angeles, possibly due to size
@@
 of all the regions accessible in this prototype.
-
-If Toronto illustrates a relatively smooth process of whole-sale
-diversification, Los Angeles shows the ability of our system and method to
-uncover more intensive smaller-scale processes of neighbourhood change. This
-capacity to examine and classify different types of change operating at
-different scales demonstrates the flexibility of the method and tool for
-advancing the more open-ended notion of "neighbourhood" envisioned by recent
-neighbourhood research.
 

### 2019-06-17 paper/intro.tex
GEAN SI submitted version

@@
 While utterly simple, the network model breaks from the deeply rooted
-traditional tabular paradigm in a significant way. Instead of requiring the data
-as a collection of fixed entities which properties evolve over time, literally
-rows in a table with temporal values as columns, it represents each measurement
-as a separate entity and encodes the evolution of these entities over time.
+traditional tabular paradigm in a significant way.
+The traditional method requires the data to be treated as a collection of fixed entities with properties that evolve over time -- rows in a table with temporal values as columns.
+By contrast, our method represents each measurement as a separate entity and encodes the evolution of these entities over time.
 
-To ease this cognitive transition, we start with an intuitive example of how it
+To ease the cognitive transition to a new paradigm, we start with an intuitive example of how the method
 works, using a small portion of a fictitious urban region illustrated on the
@@
 
-This example also illustrates some of the challenges of harmonisation. None of
+This example illustrates some of the challenges of harmonisation. None of
 these aggregation areas is clearly suitable as an interpolation target. In fact,
@@
 regions of $t_1$, A would be split to match regions C and D, which appears to be
-a rather reasonable approximation in this homogeneous artificial example, albeit
-the fallacy of division. Region F would be similarly split to match C, but F
+a rather reasonable approximation in this homogeneous artificial example (even if it commits the fallacy of division). Region F would be similarly split to match C, but F
 would be split and merged with G to match D, potentially leading to statistical
@@
 Note that the edges of this network merely encode that two regions are related.
-This is a binary information, there is no apportionment, no areal measurements,
+This is binary information, there is no apportionment, no areal measurements,
 no population percentages associated with the edge. Indeed, our method also
@@
 they are neighbouring areas.
+
+In the following, we argue that this network representation allows us to study neighbourhood change in a way that does not require an prior interpolation.

### 2019-06-17 paper/study.tex
GEAN SI submitted version

@@
 diverse city, with significant and increasing immigrant
-population~\citep{hulchanski2007three}. Toronto is also known for a stable and
-well defined Jewish community~\citep{Shahar2016}. Therefore, we expect the
-combination of stable and dynamic regions on the results, with Place of Birth,
+population~\citep{hulchanski2007three,Fong2011}, especially Asian~\citep{Fong2003}. 
+Toronto is also known for a stable and well defined Jewish community~\citep{Harold2018, Fong2011}.
+Therefore, we expect the combination of stable and dynamic regions on the results, with Place of Birth,
 Home Language, and Religion identified as relevant aspects. The results are
@@
 to the regions presented by previous studies considering
-Toronto~\citep{hulchanski2007three}, but our interface provides a deeper insight
+Toronto~\citep{hulchanski2007three}. However, our interface provides a deeper insight
 into their demographic composition, since we consider more data than solely
@@
 Angeles (LA), following urban density. While Chicago and Toronto demonstrated
-the abilities of our method for stable and dynamic cities, LA is considerably
+the abilities of our method to examine stable and dynamic cities, LA is considerably
 larger, both in population and area. Since our method identifies only the eight
-most distinct groups, we expect that some groups not to be identified,
+most distinct groups, we expect that some groups will not be identified,
 especially if their characterisation is similar to another groups. Further, our
@@
 
-These results are visually similar the analysis presented by
-Dellmelle~\citep{Delmelle2016}, what was there identified as 'persistently
+These results are visually similar to the analysis presented by
+Dellmelle~\citep{Delmelle2016}.  What was there identified as 'persistently
 struggling' corresponds to the purple cluster in Figure~\ref{fig:la}; some

### 2019-06-19 backend/dataStore.py
parallelcoordinates first draft

@@ -115,7 +115,8 @@ class dataStore(object):
             tableFile = join(self._data_folder, '{0}.proj'.format(aspectID))
             self._projection[aspectID] = pd.read_csv(tableFile, sep='\t')
             self._projection[aspectID] = self._projection[aspectID].set_index(
-                [self._info[aspectID]['index'], ])
+                [self._info[aspectID]['index'], ]).to_dict()['1D']
+
 
     def aspects(self) -> list:
         aspects = glob(join(self._data_folder, '*.info.json'))
@@ -244,7 +245,7 @@ class dataStore(object):
             Returns the pre-computed proj projection of the 'id' point from aspectID.
         """
         self._check_and_read(aspectID, proj=True)
-        if id in self._projection[aspectID].index:
-            return(list(self._projection[aspectID].loc[id])[0])
+        if id in self._projection[aspectID]:
+            return(self._projection[aspectID][id])
         else:
             return(None)

### 2019-06-19 backend/hierarchies.py

@@ -113,7 +113,7 @@ def _mergeAll(ds: dict, aspects: list, level:str='level') -> nx.Graph:
     return(F)
 
 @memory.cache(ignore=['ds'])
-def mapHierarchies(ds: dict, aspects: list, level:str='level') -> dict:
+def mapHierarchies(ds: dict, aspects: list, level:str='level', bbox:list=None) -> dict:
     """
     ds: datastore from srv.py (conf['ds']) 
     aspects:  list of aspects

### 2019-06-19 backend/srv.py

@@ -47,9 +47,9 @@ def cors():
         cherrypy.response.headers['Access-Control-Allow-Origin'] = '*'
 
 
-# @memory.cache(ignore=['ds'])
-def _mapHiers(ds: dataStore, aspects: list, thresholds: list):
-    Gs = mapHierarchies(ds, aspects)
+@memory.cache(ignore=['ds'])
+def _mapHiers(ds: dataStore, aspects: list, thresholds: list, bbox:list=None):
+    Gs = mapHierarchies(ds, aspects, bbox)
 
     print('got merged')
 
@@ -82,6 +82,11 @@ def _mapHiers(ds: dataStore, aspects: list, thresholds: list):
                          for a in aspects]
 
     full_info_aspects = sorted(full_info_aspects, key=lambda x: x['year'])
+    #preparations for the domains in ParallelCoordinates (react-vis)
+    for i in range(len(full_info_aspects)):
+        full_info_aspects[i]['order'] = i
+        full_info_aspects[i]['visible'] = True
+
     geoms = sorted(list(Gs.keys()))
 
     for threshold in thresholds:
@@ -149,16 +154,19 @@ def _mapHiers(ds: dataStore, aspects: list, thresholds: list):
             g = info['geom']
             points[a] = {}
             for cc in cc2n[g]:
-                points[a][cc] = np.median(
-                    [x for x in [ds.getProjection(a, id) for id in cc2n[g][cc]] if (x is not None)])
+                vals = [ds.getProjection(a, id) for id in cc2n[g][cc]]
+                vals = [x for x in vals if (x is not None)]
+                if vals:
+                    points[a][cc] = np.nanmedian(vals)
 
         a = full_info_aspects[0]['id']
         g = full_info_aspects[0]['geom']
         paths = [[{'geom': n[0],
                    'id':n[1],
-                   'x':0,
+                   'x':a,
                    'y':points[a][n[1]]}, ]
-                 for n in M if n[0] == g]
+                 for n in M if n[0] == g
+                 if n[1] in points[a]]
 
         for info in full_info_aspects[1:]:
             g = info['geom']
@@ -169,48 +177,38 @@ def _mapHiers(ds: dataStore, aspects: list, thresholds: list):
                 current = paths.pop(0)
                 last = {**current[-1]}
[…]

### 2019-06-19 ui/src/App.js

@@ -13,6 +13,7 @@ const mapStateToProps = (state) => ({
   aspects: state.aspects,  
   geometry: state.geometry,
   clustering: state.clustering,
+  temporal: state.temporal
 });
 
 // //https://dev.to/gaels/an-alternative-to-handle-global-state-in-react-the-url--3753
@@ -61,16 +62,16 @@ class App extends Component {
           }}>Upload data</button>
           </div>);  
         retJSX.push(<div style={{display:'flex'}}>
-                      {/* <TempEvo 
-                        similarity={this.props.similarity}
+                      <TempEvo 
+                        data={this.props.temporal}
                         key='tp'
-                      /> */}
-                      <MapboxMap 
+                      />
+                      {/* <MapboxMap 
                         geometries={availableGeometries}
                         paintProp={'GISJOIN'}
                         cmap={this.props.clustering}
                         detail={(this.props.aspects.length===1)?1:0}
-                      />
+                      /> */}
                     </div>);
       }
     }
@@ -93,8 +94,6 @@ class App extends Component {
         this.setState({availableAspects:data});                
         dispatch(requestClustering(data));
       });
-
-
     this.setState({showLoading:false})
   }   
 }

### 2019-06-19 ui/src/reducers.js

@@ -27,16 +27,22 @@ export const reducer = (state={aspects:[]}, action)=>{
         case types.SELECT_GEOMETRY:
             return({...state, geometry:payload});
         case types.UPDATE_CLUSTERING:
-            return({...state, similarity:payload.similarity, clustering:payload.clustering})
+            return({...state, temporal:{aspects:payload.aspects,evolution:payload.evolution}, clustering:payload.clustering})
         default:
             return(state);
     }
 }
 
-export function requestClustering(aspects) {
+export function requestClustering(aspects,bbox) {
+    let args={aspects:aspects,bbox:bbox};
+    if (bbox!==undefined){
+        args.bbox=bbox;
+    }
+
     return function (dispatch) {
         return fetch(getURL.MapHierarchies(),  {
-            body: JSON.stringify({aspects:aspects}), // must match 'Content-Type' header
+            // must match 'Content-Type' header
+            body: JSON.stringify(args), 
             cache: 'no-cache', // *default, cache, reload, force-cache, only-if-cached
             credentials: 'same-origin', // include, *omit
             headers: {
@@ -50,6 +56,7 @@ export function requestClustering(aspects) {
           }).then(
             data => {
                 data.json().then((d)=> {//promise of a promise. really.
+                    console.log('hier full results',d);
                     dispatch(actionCreators.SelectAspects(aspects));
                     dispatch(actionCreators.UpdateClustering(d));
                 })

### 2019-06-19 ui/src/tempEvo.js

@@ -2,13 +2,12 @@ import React, { Component } from 'react';
 import { connect } from 'react-redux';
 import './style.css';
 import './tempEvo.css';
-import {XYPlot, MarkSeries, XAxis, YAxis, Hint} from 'react-vis';
+import {ParallelCoordinates, LabelSeries} from 'react-vis';
 import { actionCreators, requestClustering } from './reducers';
 
 
 
 const mapStateToProps = (state) => ({
-    aspects: state.aspects,
   });
 
 
@@ -20,46 +19,51 @@ class TempEvo extends Component {
 
     render() {
         let retJSX=[];
-        let {similarity}=this.props;
-        let {dispatch}=this.props;
-        if (similarity!==undefined){
+        // let {dispatch}=this.props;
+        if (this.props.data!==undefined){
+            let {aspects,evolution}=this.props.data;
+            evolution=evolution.slice(0,200);
+            let domains=aspects.filter((d)=>{
+                return(d.visible);
+            }).map((d)=>{
+                return({'name':d.id, domain:[0,1], getValue: (e)=> e[d.id]});
+            }).sort((a,b)=>{
+                return(a.order-b.order);
+            });
             retJSX.push(<div>
-                <XYPlot
-                    width={380}
-                    height={500}
-                    xDomain={[0,1]}
-                    >
-                    <XAxis
-                        // hideTicks
-                    />
-                    <YAxis
-                        tickFormat={v => String(v)}
-                    />
-                    <MarkSeries
-                        // color="darkgray"
-                        stroke="lightgray"
-                        data={similarity}
-                        onValueMouseOver={(datapoint, event)=>{
-                            this.setState({tooltip:datapoint});
-                          }}
-                        onValueMouseOut={(event)=>{
-                            this.setState({tooltip:undefined})
-                        }}
-                        onValueClick={(datapoint, e)=>{
-                            console.log('updating click', datapoint);
-                            if (e.event.shiftKey){
-                                dispatch(requestClustering([datapoint.id,]))
-                            }
-                            dispatch(actionCreators.SelectGeometry(datapoint.geometry));
-                        }}
-                    />
-                    {(this.state.tooltip!==undefined)?
-                        <Hint value={this.state.tooltip}>
-                            <div style={{background: 'black'}}>
-                                <p>{this.state.tooltip.name}</p>
-                        </div>                    
-                    </Hint>:null}
-                </XYPlot>
+                <ParallelCoordinates
[…]

### 2019-06-19 backend/dataStore.py
3857, Update button, bounding box filtered hier

@@ -10,7 +10,7 @@ import numpy as np
 import pandas as pd
 from scipy.stats import skew
 from sklearn.preprocessing import minmax_scale
-
+from rtree.index import Rtree
 from clustering import ComputeClustering
 from hierarchies import compareHierarchies
 
@@ -126,7 +126,7 @@ class dataStore(object):
         hiers = glob(join(self._data_folder, '*.gp'))
         return([basename(x)[:-3] for x in hiers])
 
-    def getHierarchy(self, aspectID: str, level: str = 'level') -> nx.Graph():
+    def getHierarchy(self, aspectID: str, level: str = 'level', bbox:list=None) -> nx.Graph():
         """
             Returns the graph with the hierarchy of aspectID (same ID as the aspect that created it)
         """
@@ -142,7 +142,16 @@ class dataStore(object):
             self._hiers[aspectID] = G
         else:
             G = self._hiers[aspectID]
-        return(G)
+        geom = self.getGeometry(aspectID)
+        if bbox is None:
+            return(G)
+        else:
+            I = Rtree(join(self._geometry_folder,geom+'.rt'))
+            hits=list(I.intersection(bbox, objects=True))
+            newG=nx.subgraph(G,[(geom,item.object) for item in hits])
+            print('newG',len(G),len(newG),len(newG.edges()))
+            return(newG)            
+
 
     def getCrossGeometry(self, g1: str, g2: str) -> nx.Graph():
         """

### 2019-06-19 backend/genBasegraph.py

@@ -2,9 +2,11 @@ import sys
 import networkx as nx
 import fiona
 from util import indexedPols
-from shapely.geometry import shape   
+from rtree.index import Rtree
+from shapely.geometry import shape, Point
 import geojson
 import json
+import pickle
 
 if __name__ == '__main__':
 
@@ -24,10 +26,10 @@ if __name__ == '__main__':
         for feat in source:
             geo.insertJSON(G=feat['geometry'],props={'Geom_ID':feat['properties'][field],})
 
-
-    print('starting spatial edges')
     B = nx.Graph()
+    I = Rtree(outName.replace('.gp','.rt'))
     
+    count=0
     for fid, geom in geo.iterIDGeom('Geom_ID'):
         n=(baseGeometry,fid)
 
@@ -35,6 +37,9 @@ if __name__ == '__main__':
         if ('pos' not in B.node[n]):
             representative=geom.representative_point()
             B.node[n]['pos']=[representative.x, representative.y,baseGeometry]
+            I.insert(count,representative.bounds,obj=n[1])
+            count+=1
+
         #spatial edges                
         for polID in geo.search(geom.buffer(1e-3)):
             nid=geo.getProperty(polID,'Geom_ID')

### 2019-06-19 backend/hierarchies.py

@@ -94,10 +94,10 @@ def _hierMerge(G1: nx.Graph, G2: nx.Graph, X: nx.Graph = None, level: str = 'lev
     return(H)
 
 
-def _mergeAll(ds: dict, aspects: list, level:str='level') -> nx.Graph:
+def _mergeAll(ds: dict, aspects: list, level:str='level', bbox:list=None) -> nx.Graph:
     F = None
     for a in aspects:
-        G = ds.getHierarchy(a)
+        G = ds.getHierarchy(a,bbox=bbox)
 
         if F is None:
             F = G
@@ -136,7 +136,7 @@ def mapHierarchies(ds: dict, aspects: list, level:str='level', bbox:list=None) -
     geoms = sorted(aspectsByGeometry.keys())
     merged = []
     for g in geoms:
-        merged.append(_mergeAll(ds, aspectsByGeometry[g], level=level))
+        merged.append(_mergeAll(ds, aspectsByGeometry[g], level=level, bbox=bbox))
 
     # holds the resulting hierarchy on each original geometry
     # - This could be faster if the partial merges were reused

### 2019-06-19 backend/srv.py

@@ -33,6 +33,23 @@ if not exists(cachedir):
     makedirs(cachedir)
 memory = Memory(cachedir, verbose=0)
 
+def _bbox_create_buffer(bbox:dict)->list:
+    minX=bbox['_sw']['lng']
+    minY=bbox['_sw']['lat']
+    maxX=bbox['_ne']['lng']
+    maxY=bbox['_ne']['lat']
+
+    p1=np.array([minX,minY])
+    p2=np.array([maxX,maxY])
+    # p1n=(p1-p2)*1.05+p2 #~10% a^2+b^2=c^2, so that number gets ^2
+    # p2n=(p2-p1)*1.05+p1
+    p1n=p1
+    p2n=p2
+    return([np.floor(p1n[0]),
+            np.floor(p1n[1]),
+            np.ceil(p2n[0]),
+            np.ceil(p2n[1])])
+
 
 def cors():
     if cherrypy.request.method == 'OPTIONS':
@@ -49,7 +66,7 @@ def cors():
 
 @memory.cache(ignore=['ds'])
 def _mapHiers(ds: dataStore, aspects: list, thresholds: list, bbox:list=None):
-    Gs = mapHierarchies(ds, aspects, bbox)
+    Gs = mapHierarchies(ds, aspects, bbox=bbox)
 
     print('got merged')
 
@@ -118,14 +135,16 @@ def _mapHiers(ds: dataStore, aspects: list, thresholds: list, bbox:list=None):
                             M.node[source]['area'][g2] = 0
 
                         for nn in X.neighbors((g1, n)):
-                            intersection = X[(g1, n)][nn]['intersection']
-                            M.node[source]['area'][g2] += intersection
-                            target = (g2, cl[g2][nn[1]][-1])
-                            if not target in M:
-                                M.add_node(target)
-                            if not M.has_edge(source, target):
-                                M.add_edge(source, target, area=0)
-                            M[source][target]['area'] += intersection
+                            #only goes in if the target is inside the bbox
[…]

### 2019-06-19 ui/src/App.js

@@ -13,7 +13,8 @@ const mapStateToProps = (state) => ({
   aspects: state.aspects,  
   geometry: state.geometry,
   clustering: state.clustering,
-  temporal: state.temporal
+  temporal: state.temporal,
+  bbox: state.bbox
 });
 
 // //https://dev.to/gaels/an-alternative-to-handle-global-state-in-react-the-url--3753
@@ -39,7 +40,7 @@ class App extends Component {
   render() {
     let retJSX=[];
     let {showLoading,showUploadPanel,availableGeometries}=this.state;
-    
+    let {dispatch}=this.props;
     if (showLoading===true){
       retJSX.push(<div key="loading" className="loading">Loading, please wait...</div>);
     }else{
@@ -60,18 +61,24 @@ class App extends Component {
           <button onClick={(e)=>{
             this.setState({showUploadPanel:true});
           }}>Upload data</button>
+          <button onClick={(e)=>{
+            dispatch(requestClustering(this.props.aspects,this.props.bbox));
+          }}>Update</button>
           </div>);  
-        retJSX.push(<div style={{display:'flex'}}>
+        retJSX.push(<div >
+                      <MapboxMap 
+                        geometries={availableGeometries}
+                        paintProp={'GISJOIN'}
+                        cmap={this.props.clustering}
+                        bboxCallback={(box)=>{
+                          console.log(box);
+                          dispatch(actionCreators.UpdateBBOX(box));
+                        }}
+                      />
                       <TempEvo 
                         data={this.props.temporal}
                         key='tp'
                       />
-                      {/* <MapboxMap 
-                        geometries={availableGeometries}
-                        paintProp={'GISJOIN'}
-                        cmap={this.props.clustering}
-                        detail={(this.props.aspects.length===1)?1:0}
-                      /> */}
                     </div>);
       }
     }
@@ -85,7 +92,6 @@ class App extends Component {
   componentDidMount() {    
     let {dispatch}=this.props;
       getData(getURL.AvailableGeometries(), (data)=> {
-        console.log('geometries received', data, data[0]['name']);
         dispatch(actionCreators.SelectGeometry(data[0]['name']));
         this.setState({availableGeometries:data});                
       });

### 2019-06-19 ui/src/glmap.js

@@ -1,6 +1,4 @@
 import React from 'react';
-// import PropTypes from 'prop-types'
-
 import mapboxgl from 'mapbox-gl';
 import 'mapbox-gl/dist/mapbox-gl.css';
 import './glmap.css';
@@ -137,6 +135,11 @@ let MapboxMap = class MapboxMap extends React.Component {
         );
       this.map.addControl(geomControl);
     });
+    this.map.on('moveend',()=>{
+      if (this.props.bboxCallback!==undefined){
+        this.props.bboxCallback(this.map.getBounds());
+      }
+    });    
   }
 
   setFill(colours){
@@ -165,10 +168,9 @@ let MapboxMap = class MapboxMap extends React.Component {
     if (colours!==undefined){
       for (let layer of this.props.geometries){
         if (this.map.getLayer('l_'+layer.year)!==undefined){
-          console.log('detail level',this.props.detail);
           this.map.setPaintProperty('l_'+layer.year, 
           'fill-color', 
-              ['let', 'detail', this.props.detail, exp]
+              ['let', 'detail', 0, exp]
           );
           // This may work or not, but runs out of memory...
           // this.map.setPaintProperty('l_'+layer.year, 

### 2019-06-19 ui/src/reducers.js

@@ -4,6 +4,7 @@ export const types={
     SELECT_GEOMETRY: 'SelectGeometry',
     SELECT_ASPECTS: 'SelectAspects',
     UPDATE_CLUSTERING: 'UpdateClustering',
+    UPDATE_BBOX: 'UpdateBBOX',
 };
 // Helper functions to dispatch actions, optionally with payloads
 export const actionCreators = {
@@ -15,6 +16,9 @@ export const actionCreators = {
     },
     UpdateClustering: (clsim) => {
         return({type: types.UPDATE_CLUSTERING, payload:clsim})
+    },
+    UpdateBBOX: (box) => {
+        return({type: types.UPDATE_BBOX, payload:box});
     }
   };
 
@@ -28,6 +32,8 @@ export const reducer = (state={aspects:[]}, action)=>{
             return({...state, geometry:payload});
         case types.UPDATE_CLUSTERING:
             return({...state, temporal:{aspects:payload.aspects,evolution:payload.evolution}, clustering:payload.clustering})
+        case types.UPDATE_BBOX:
+            return({...state, bbox:payload})
         default:
             return(state);
     }
@@ -38,7 +44,7 @@ export function requestClustering(aspects,bbox) {
     if (bbox!==undefined){
         args.bbox=bbox;
     }
-
+    console.log('request', args)
     return function (dispatch) {
         return fetch(getURL.MapHierarchies(),  {
             // must match 'Content-Type' header

### 2019-06-19 backend/srv.py
trying to match the viewport

@@ -41,10 +41,10 @@ def _bbox_create_buffer(bbox:dict)->list:
 
     p1=np.array([minX,minY])
     p2=np.array([maxX,maxY])
-    # p1n=(p1-p2)*1.05+p2 #~10% a^2+b^2=c^2, so that number gets ^2
-    # p2n=(p2-p1)*1.05+p1
-    p1n=p1
-    p2n=p2
+    p1n=(p1-p2)*1.05+p2 #~10% a^2+b^2=c^2, so that number gets ^2
+    p2n=(p2-p1)*1.05+p1
+    # p1n=p1
+    # p2n=p2
     return([np.floor(p1n[0]),
             np.floor(p1n[1]),
             np.ceil(p2n[0]),

### 2019-08-08 paper/intro.tex
easier reviews

@@
 While utterly simple, the network model breaks from the deeply rooted
-traditional tabular paradigm in a significant way.
-The traditional method requires the data to be treated as a collection of fixed entities with properties that evolve over time -- rows in a table with temporal values as columns.
-By contrast, our method represents each measurement as a separate entity and encodes the evolution of these entities over time.
+traditional tabular paradigm in a significant way. The traditional method
+requires the data to be treated as a collection of fixed entities with
+properties that evolve over time -- rows in a table with temporal values as
+columns. By contrast, our method represents each measurement as a separate
+entity and encodes the evolution of these entities over time.
 
-To ease the cognitive transition to a new paradigm, we start with an intuitive example of how the method
-works, using a small portion of a fictitious urban region illustrated on the
-left part of Figure~\ref{fig:intuition}. This example includes three different
-times ($t_0,t_1,t_2$), with different aggregation areas identified as letters
-from A to H. For $t_0$, the initial time, we have areas A and B, with small
-houses and a park, respectively. The park remains stable (B, E, and H), but the
-houses are partially replaced by larger buildings (C and F).
+To ease the cognitive transition to a new paradigm, we start with an intuitive
+example of how the method works, using a small portion of a fictitious urban
+region illustrated on the left part of Figure~\ref{fig:intuition}. This example
+includes three different times ($t_0,t_1,t_2$), with different aggregation areas
+identified as letters from A to H. For $t_0$, the initial time, we have areas A
+and B, with small houses and a park, respectively. The park remains stable (B,
+E, and H), but the houses are partially replaced by larger buildings (C and F).
 
-This example illustrates some of the challenges of harmonisation. None of
-these aggregation areas is clearly suitable as an interpolation target. In fact,
+This example illustrates some of the challenges of harmonisation. None of these
+aggregation areas is clearly suitable as an interpolation target. In fact,
 adopting any of these areas as a target would require merging heterogeneous
@@
 regions of $t_1$, A would be split to match regions C and D, which appears to be
-a rather reasonable approximation in this homogeneous artificial example (even if it commits the fallacy of division). Region F would be similarly split to match C, but F
-would be split and merged with G to match D, potentially leading to statistical
-measurements that do not properly represent either region.
+a rather reasonable approximation in this homogeneous artificial example (even
+if it commits the fallacy of division). Region F would be similarly split to
+match C, but F would be split and merged with G to match D, potentially leading
+to statistical measurements that do not properly represent either region.
 
@@
             \draw (a) -- (c);
+            \draw (a) -- (b);
             \draw (a) -- (d);
@@
             \draw (c) -- (f);
+            \draw (c) -- (d);
             \draw (d) -- (f);
+            \draw (d) -- (e);
             \draw (d) -- (g);
             \draw (e) -- (h);
+            \draw (f) -- (g);
+            \draw (g) -- (h);
         \end{tikzpicture}
@@
 data, as illustrated by the colours of the nodes on the right side of
-Figure~\ref{fig:intuition}. Further, the possible evolution paths can also be
-obtained by computing temporal sequences of nodes, in this case: (A, C, F), (A,
+Figure~\ref{fig:intuition}. Further, all the possible paths of change can be
+obtained by computing sequences of nodes over time, in this case: (A, C, F), (A,
 D, F), (A, D, G), and (B, E, H). This representation is also suited for
@@
 
-In the following, we argue that this network representation allows us to study neighbourhood change in a way that does not require an prior interpolation.
+In the following, we argue that this network representation allows us to study
+neighbourhood change in a way that does not require an prior interpolation.

### 2019-08-08 paper/method.tex
GEAN SI submitted version
easier reviews

@@
 \subsection{Census methodology and data representation}
-\info{I removed any specifics of the experiments (CTs, which variables) from here because this section is theory}
 
@@
 year. Further, edges are placed between nodes if the corresponding regions
-belong to sequential years and there is geographical overlap between them. This
-approach leads to a single network representing the whole spatio-temporal space
-of the data. Our objective then becomes to identify partitions of this network
-such that the nodes of each partition are more similar between themselves than
-to the other nodes.
+belong to sequential years and \revision{their geometries intersect. To avoid
+spurious connections caused by geometry fluctuations, one of the geometries is
+slightly shrunk before the intersection, using a buffer of -1e6. More
+importantly, while weights will be associated with these edges before they are
+processed, they are not derived from the geometry, but from the data. The actual
+intersection area is not considered in this representation. } This approach
+leads to a single network representing the whole spatio-temporal space of the
+data. Our objective then becomes to identify partitions of this network such
+that the nodes of each partition are more similar between themselves than to the
+other nodes.
 
@@
 paper~\citep{markus2017} for more details, including a complete performance
-evaluation using several metrics.
+evaluation using several metrics. \revision{We chose this algorithm because it
+is simple and easily customisable, and while different algorithms will lead to
+different results, our methodology should work with any hierarchical clustering
+algorithm.}
 
@@
 connected, providing a path for the algorithm to group similar nodes.
+\revision{The regions connected by those edges will be merged on the first
+stages of the clustering, since they are very similar, leaving the remaining
+steps of the hierarchy to be determined only by the geographical edges. We
+explored with different numbers of augmentation edges, but the results were not
+consistent, since the distribution of the edges is data dependent. Adding two
+edges per node was the least number of edges that led to stable and consistent
+results in the scenarios available in our prototype. Since the problem of
+balancing the data space with the geographical space is relevant for
+geographical data analysis, this methodology potentially warrants further
+exploration, beyond the scope of this work.}
 
@@
 \emph{relevance} of a variable of an aspect based on the distance between the
-interquantile ranges (IQR) of the clusters in the same hierarchical level. If
+interquartile ranges (IQR) of the clusters in the same hierarchical level. If
 the IQRs overlap for all clusters, that variable is not relevant to the
@@
 evolution of these groups. To examine this process of evolution directly, we
-introduce the concept of \emph{trajectories}. Trajectories are composed by
-regions classified into the same sequence of clusters over the considered
-period. This enables direct access to regions that evolved in the same manner.
-While individual census tracts remain interesting, these trajectories are the
-main unit of exploration in this work.
+introduce the concepts\revision{of \emph{temporal paths} and \emph{trajectories}.
+
+We call a temporal path any sequence of nodes in our representation network such
+that the temporal information associated with the nodes only increases. For
+instance, in Figure~\ref{fig:intuition}, the sequences ACF, ADF, ADG, and BEH
+are temporal paths. With harmonised data, the time-series of to each region
+would form a temporal path, each node would be connected only to its older and
+newer versions, belonging to only one temporal path, as illustrated by the path
+BEF. Since our data is not harmonised, more connections are allowed and each
+node can belong to an arbitrary number of paths.
+
+Semantically, this is a generalisation of the idea of geographical time-series,
+because each temporal path is one possible option for the data to change over
+time. Returning to Figure~\ref{fig:intuition}, the paths ACF, ADF, and ADG all
+start on the same homogeneous region, but evolve differently over time. In other
+words, this network encodes the information that portions of the region A
+changed to form regions C and D, but we do not know specifically which parts,
+nor we need to, since the same region can belong to several temporal paths.
+Interestingly, when interpreted in this framework, geographical harmonisation is
+a method to split and/or merge nodes so each belongs to a single temporal path.
+
+Since each node in the sequence that forms the temporal paths has an associated
+cluster, we can classify the paths based on the sequence of clusters. We call
+each unique sequence of clusters present in this result a \emph{trajectory}.
+Regions on the same trajectory had the same sequence of clusters, therefore had
+similar temporal evolution.}
 

### 2019-08-08 paper/related.tex
easier reviews

@@
 Base~\citep{Logan2014}. Despite its inherent
-errors~\citep{Logan2016,Hallisey2017}, this dataset has become the standard
-source for longitudinal demographic data at the neighbourhood scale, with
-similar efforts appearing in other countries~\citep{Liu2015,Lee2015,Allen2018}.
-These datasets have been highly significant for the field. Yet they also limit
-the universe of  data that can be used to study neighbourhood change, since any
-new datasets would need to be similarly processed in order to be rendered
-compatible with these sources.
+errors~\citep{Logan2016,Hallisey2017}, this dataset has become \revision{widely
+adopted, along with the Neighborhood Change Database~\cite{ncdb},} as source for
+longitudinal demographic data at the neighbourhood scale, with similar efforts
+appearing in other countries~\citep{Liu2015,Lee2015,Allen2018}. These datasets
+have been highly significant for the field. Yet they also limit the universe of
+data that can be used to study neighbourhood change, since any new datasets
+would need to be similarly processed in order to be rendered compatible with
+these sources.
 
@@
 data~\citep{Dmowska2017,Dmowska2018,stepinski2019imperfect}. Beyond the
-increased spatial accuracy, this approach does not require complex harmonisation
-when new data is considered, if the grids are compatible. However, demographic
-data is usually not available in this format, especially from older sources.
-Additionally, the conversion from tabulation areas can introduce significant
-errors.
+potentially increased spatial precision, this approach does not require complex
+harmonisation when new data is considered, if the grids are compatible. However,
+demographic data is usually not available in this format, especially from older
+sources. Additionally, the conversion from tabulation areas can introduce
+significant errors.
 

### 2019-08-08 paper/study.tex
easier reviews

@@
 
-In this section we present three illustrative scenarios, using decennial census
+In this section we present two illustrative scenarios, using decennial census
 data from the United States~\citep{nhgis} and
 Canada\footnote{\url{http://datacentre.chass.utoronto.ca/census/}}, tabulated by
-CTs,  from 1970 to 2010. The prototype interface allows access to 41 regions, 29
+CTs,  from 1970 to 2010. \revision{Accidentaly, the US data for 2010 is actually
+not from the decennial census, but from the ACS 2006-2010. Further, the regions
+selected do not correspond to any pre-defined regions (metros, cities, census
+areas), but to arbitrary regions defined around a location of interest. We
+selected a minimal set of aspects for each country, and they are not similar to
+each other, out of convenience, since the variable matching was a manual
+process. These caveats would likely compromise any serious attempt on a
+comprehensive demographic study, but this is not the objective of this
+experiment. We aim only to test and validate our methodology, comparing our
+results to pre-existing literature insights. Indeed, our results were perfectly
+aligned to several other studies, despite these methodological missteps, which
+could arguably be an indicator of the robustness of our method.}
+
+The prototype interface allows access to 41 regions, 29
 in the US and 12 in Canada. New York City was split into its boroughs to avoid
@@
 diverse city, with significant and increasing immigrant
-population~\citep{hulchanski2007three,Fong2011}, especially Asian~\citep{Fong2003}. 
-Toronto is also known for a stable and well defined Jewish community~\citep{Harold2018, Fong2011}.
-Therefore, we expect the combination of stable and dynamic regions on the results, with Place of Birth,
+population~\citep{hulchanski2007three,Fong2011}, especially
+Asian~\citep{Fong2003}. Toronto is also known for a stable and well defined
+Jewish community~\citep{Harold2018, Fong2011}. Therefore, we expect a
+combination of stable and dynamic regions on the results, with Place of Birth,
 Home Language, and Religion identified as relevant aspects. The results are
@@
 
-\subsection{Los Angeles}
+% \subsection{Los Angeles}
 
-We selected for this scenario a region around the metropolitan area of Los
-Angeles (LA), following urban density. While Chicago and Toronto demonstrated
-the abilities of our method to examine stable and dynamic cities, LA is considerably
-larger, both in population and area. Since our method identifies only the eight
-most distinct groups, we expect that some groups will not be identified,
-especially if their characterisation is similar to another groups. Further, our
-system does not include the census variables related to Hispanic heritage, which
-were only included in the newer censuses. However, from the literature, we
-expect to see some clusters where Race is an important aspect, including White,
-Black, and Asian~\citep{Reibel2003}.
+% We selected for this scenario a region around the metropolitan area of Los
+% Angeles (LA), following urban density. While Chicago and Toronto demonstrated
+% the abilities of our method to examine stable and dynamic cities, LA is considerably
+% larger, both in population and area. Since our method identifies only the eight
+% most distinct groups, we expect that some groups will not be identified,
+% especially if their characterisation is similar to another groups. Further, our
+% system does not include the census variables related to Hispanic heritage, which
+% were only included in the newer censuses. However, from the literature, we
+% expect to see some clusters where Race is an important aspect, including White,
+% Black, and Asian~\citep{Reibel2003}.
 
-The summary of the results using all aspects with equal weights and eight
-clusters is illustrated in Figure~\ref{fig:la}. The full statistical description
-of the clusters is illustrated in Figure~\ref{fig:laAll}, where the most
-relevant aspect of each cluster is highlighted. From the trajectories plot, we
-can see that there is a large but shrinking cluster, depicted in green, one
-increasing cluster in purple, an almost constant orange cluster, a smaller but
-increasing pink cluster, and three other small clusters. The corresponding map
-illustrates where these clusters are located, and that they are somewhat
-geographically stable, with some movement between the green, orange, and purple
-clusters.
+% The summary of the results using all aspects with equal weights and eight
+% clusters is illustrated in Figure~\ref{fig:la}. The full statistical description
+% of the clusters is illustrated in Figure~\ref{fig:laAll}, where the most
+% relevant aspect of each cluster is highlighted. From the trajectories plot, we
+% can see that there is a large but shrinking cluster, depicted in green, one
+% increasing cluster in purple, an almost constant orange cluster, a smaller but
+% increasing pink cluster, and three other small clusters. The corresponding map
+% illustrates where these clusters are located, and that they are somewhat
+% geographically stable, with some movement between the green, orange, and purple
+% clusters.
 
-\begin{figure}
-    \centering 
-    \includegraphics[width=0.975\linewidth]{LA.pdf}
-    \caption{Result for Los Angeles with 8 clusters, including three small and
-        ephemeral clusters. Cluster characterisation is displayed
-        in Figure~\ref{fig:laAll}.\label{fig:la}}
-\end{figure}
+% \begin{figure}
+%     \centering 
+%     \includegraphics[width=0.975\linewidth]{LA.pdf}
+%     \caption{Result for Los Angeles with 8 clusters, including three small and
+%         ephemeral clusters. Cluster characterisation is displayed
+%         in Figure~\ref{fig:laAll}.\label{fig:la}}
+% \end{figure}
 
-From Figure~\ref{fig:laAll}, we can see that the green cluster is characterised
-by a high percentage of White population, low percentage of population in the
-lowest income bracket, mostly administrative occupations, and about 30\% of the
-population with four or more years of college. The orange cluster is
-characterised by a high percentage of Black population, with few people in the
-highest range of income and education. The purple cluster corresponds to a high
-concentration of "Other" in race, which includes Hispanic for this dataset, high
-concentration of Labourers, and low education and income. The pink cluster
-contains a high percentage of Asian population and about 30\% of the population
-with four or more years of college. The light green cluster contains very few
-people in the lower income bracket, mostly White population, with the highest
-percentage of population with four or more years of college, working
-administrative jobs, and a high concentration of singles. The yellow cluster
-represents Black population, with higher level of education and income, mostly
-working administrative jobs. The brown cluster represent a majority of single
-population, working administrative jobs with mostly low income. The dark grey
-cluster is characterised by all its population in the lowest income bracket, low
-education level, with a majority of White population. Since the extremes in the
-boxplots of the grey cluster are not significantly different, we can also
-surmise that this cluster is either small or homogeneous.
+% From Figure~\ref{fig:laAll}, we can see that the green cluster is characterised
+% by a high percentage of White population, low percentage of population in the
+% lowest income bracket, mostly administrative occupations, and about 30\% of the
+% population with four or more years of college. The orange cluster is
+% characterised by a high percentage of Black population, with few people in the
+% highest range of income and education. The purple cluster corresponds to a high
+% concentration of "Other" in race, which includes Hispanic for this dataset, high
+% concentration of Labourers, and low education and income. The pink cluster
+% contains a high percentage of Asian population and about 30\% of the population
+% with four or more years of college. The light green cluster contains very few
+% people in the lower income bracket, mostly White population, with the highest
+% percentage of population with four or more years of college, working
+% administrative jobs, and a high concentration of singles. The yellow cluster
+% represents Black population, with higher level of education and income, mostly
+% working administrative jobs. The brown cluster represent a majority of single
+% population, working administrative jobs with mostly low income. The dark grey
+% cluster is characterised by all its population in the lowest income bracket, low
+% education level, with a majority of White population. Since the extremes in the
+% boxplots of the grey cluster are not significantly different, we can also
+% surmise that this cluster is either small or homogeneous.
 
-\begin{figure*}
-    \centering 
-    \includegraphics[width=0.95\linewidth]{laAll.pdf}
-    \caption{Full characterisation of the eight clusters found for LA. The red
-     rectangles indicate the most relevant aspects for each
-     cluster.\label{fig:laAll}}
-\end{figure*}
+% \begin{figure*}
+%     \centering 
+%     \includegraphics[width=0.95\linewidth]{laAll.pdf}
+%     \caption{Full characterisation of the eight clusters found for LA. The red
+%      rectangles indicate the most relevant aspects for each
+%      cluster.\label{fig:laAll}}
+% \end{figure*}
 
-The  green, orange, and purple clusters present a significant intra-cluster
-variance in most variables, as indicated by extreme whiskers of the boxplots.
-While fifty percent of the CTs in the green cluster have between 20\% and 40\%
-of people in the lowest income bracket, that cluster also includes CTs where
-none and all the population belongs to that bracket. This might indicate that
-this cluster represents different groups of people that are not different enough
-to be separated at this level of the hierarchy. Conversely, the light green,
-brown, and dark grey clusters are different enough to be separated into their
-own clusters at this level, despite being small and ephemeral, including only a
-few CTs.
+% The  green, orange, and purple clusters present a significant intra-cluster
+% variance in most variables, as indicated by extreme whiskers of the boxplots.
+% While fifty percent of the CTs in the green cluster have between 20\% and 40\%
+% of people in the lowest income bracket, that cluster also includes CTs where
+% none and all the population belongs to that bracket. This might indicate that
+% this cluster represents different groups of people that are not different enough
+% to be separated at this level of the hierarchy. Conversely, the light green,
+% brown, and dark grey clusters are different enough to be separated into their
+% own clusters at this level, despite being small and ephemeral, including only a
+% few CTs.
 
-The orange area in the map in Figure~\ref{fig:la} presents movement, indicated by
-the presence of green and purple tones mixed with the orange, which may warrant
-further exploration. By clicking on the orange bar above the boxplot, we select
-all trajectories that contain the orange cluster. The corresponding details are
-illustrated in Figure~\ref{fig:labh}. This shows a location change, where the
-orange cluster is progressively replaced by the purple cluster on its east side,
-and in turn expanding to the west. Interestingly, the population increased, the
-racial profile changed, but the distribution of income was reasonably stable,
-with a higher amount of the population in the lowest income range and very few
-people in the highest income range. Indeed, the income  difference is
-significant when compared to the city-wide distribution.
+% The orange area in the map in Figure~\ref{fig:la} presents movement, indicated by
+% the presence of green and purple tones mixed with the orange, which may warrant
+% further exploration. By clicking on the orange bar above the boxplot, we select
+% all trajectories that contain the orange cluster. The corresponding details are
+% illustrated in Figure~\ref{fig:labh}. This shows a location change, where the
+% orange cluster is progressively replaced by the purple cluster on its east side,
+% and in turn expanding to the west. Interestingly, the population increased, the
+% racial profile changed, but the distribution of income was reasonably stable,
+% with a higher amount of the population in the lowest income range and very few
+% people in the highest income range. Indeed, the income  difference is
+% significant when compared to the city-wide distribution.
 
-\begin{figure}
-    \centering 
-    \includegraphics[width=0.75\linewidth]{laBH.pdf}
-    \caption{Top: Geographic changes in the majority Black population cluster
-    (orange) and Labourers cluster (green). Bottom: Income evolution for this
-    region (black) and the whole city (grey).\label{fig:labh}}
-\end{figure}
+% \begin{figure}
+%     \centering 
+%     \includegraphics[width=0.75\linewidth]{laBH.pdf}
+%     \caption{Top: Geographic changes in the majority Black population cluster
+%     (orange) and Labourers cluster (green). Bottom: Income evolution for this
+%     region (black) and the whole city (grey).\label{fig:labh}}
+% \end{figure}
 
-A portion of this region is classified into the green cluster in 2010,
-indicating a majority white population. To further understand that change, we
-clear the current selection, and select all regions that changed from orange in
-1970 to green in 2010, using the transition matrix. A portion of the resulting
-region, near the Florence-Graham region, is depicted in Figure~\ref{fig:laVol},
-along with the temporal evolution of Race. Despite this difference, the other
-aspects are similar to the ones from the region in Figure~\ref{fig:labh}, with
-slightly lower income and education profiles. While the racial aspect changed
-considerably, the economic and educational aspects stayed the same.
+% A portion of this region is classified into the green cluster in 2010,
+% indicating a majority white population. To further understand that change, we
+% clear the current selection, and select all regions that changed from orange in
+% 1970 to green in 2010, using the transition matrix. A portion of the resulting
+% region, near the Florence-Graham region, is depicted in Figure~\ref{fig:laVol},
+% along with the temporal evolution of Race. Despite this difference, the other
+% aspects are similar to the ones from the region in Figure~\ref{fig:labh}, with
+% slightly lower income and education profiles. While the racial aspect changed
+% considerably, the economic and educational aspects stayed the same.
 
-\begin{figure}
-    \centering 
-    \includegraphics[width=0.75\linewidth]{laVol.pdf}
-    \caption{Details for a volatile region contained in the area of
-         Figure~\ref{fig:labh}. This region went from Black to Hispanic to
-        White.\label{fig:laVol}}
-\end{figure}
+% \begin{figure}
+%     \centering 
+%     \includegraphics[width=0.75\linewidth]{laVol.pdf}
+%     \caption{Details for a volatile region contained in the area of
+%          Figure~\ref{fig:labh}. This region went from Black to Hispanic to
+%         White.\label{fig:laVol}}
+% \end{figure}
 
-These results are visually similar to the analysis presented by
-Dellmelle~\citep{Delmelle2016}.  What was there identified as 'persistently
-struggling' corresponds to the purple cluster in Figure~\ref{fig:la}; some
-portions of the 'stable elite' cluster, near Inglewood, correspond to the high
-income, high education yellow cluster. The other groups identified by that study
-were clumped by our system into the green cluster, which would be further
-divided if more than eight clusters were to be used.
+% These results are visually similar to the analysis presented by
+% Dellmelle~\citep{Delmelle2016}.  What was there identified as 'persistently
+% struggling' corresponds to the purple cluster in Figure~\ref{fig:la}; some
+% portions of the 'stable elite' cluster, near Inglewood, correspond to the high
+% income, high education yellow cluster. The other groups identified by that study
+% were clumped by our system into the green cluster, which would be further
+% divided if more than eight clusters were to be used.
 
-If Toronto illustrates a relatively smooth process of whole-sale
-diversification, Los Angeles shows the ability of our system and method to
-uncover more intensive smaller-scale processes of neighbourhood change. This
-capacity to examine and classify different types of change operating at
-different scales demonstrates the flexibility of the method and tool for
-advancing the more open-ended notion of "neighbourhood" envisioned by recent
-neighbourhood research.
+% If Toronto illustrates a relatively smooth process of whole-sale
+% diversification, Los Angeles shows the ability of our system and method to
+% uncover more intensive smaller-scale processes of neighbourhood change. This
+% capacity to examine and classify different types of change operating at
+% different scales demonstrates the flexibility of the method and tool for
+% advancing the more open-ended notion of "neighbourhood" envisioned by recent
+% neighbourhood research.
 
@@
 if this difference led to unexpected results, but rather if they found
-interesting insights. Most experts found phenomena corroborated by the
-specialised literature, indicating that our methodology produces equivalent
-results, with a fraction of the effort. We interpret the fact that most of them
-were interested in the next steps as confirmation of the accuracy of the method.
+interesting insights.

### 2019-08-08 paper/method.tex
one last review

@@
 
+%TODO THIS IS WRONG!
 Each aspect is represented by a stacked bar plot, where the width of each
 rectangle corresponds to the average percentage of that variable over the
-considered period. In this case, about half of the people in Chicago in the
-considered period are married, and the percentage that are Widowers or Divorced
+considered period. In this case, about half of the people in this region, in the
+considered period, are married, and the percentage that are Widowers or Divorced
 is roughly similar. About half of the population work in Administrative jobs, a

### 2019-08-10 paper/intro.tex
more fixes, regionalisation ideas

@@
 
-These advances have made neighbourhood constructs more tractable to quantitative
-analysis. Yet new challenges have also emerged, especially at the convergence of
-research on neighbourhood effects and neighbourhood dynamics. Neighbourhood
-effects research assumes knowledge about the nature and scope of "the
-neighbourhood" that presumably shapes individual
+\revision{In this work, we consider neighbourhoods as \emph{formal
+regions}~\citep{montello2003regions}, geographically continous areas with
+similar data characteristics.} These advances have made neighbourhood constructs
+more tractable to quantitative analysis. Yet new challenges have also emerged,
+especially at the convergence of research on neighbourhood effects and
+neighbourhood dynamics. Neighbourhood effects research assumes knowledge about
+the nature and scope of "the neighbourhood" that presumably shapes individual
 outcomes~\citep{Kwan2018,Shelton2019}. Concurrently, researchers note that
@@
 
-Instead, we propose a network representation. Intuitively, a \emph{network}
-(also called a \emph{graph}) is a collection of entities (nodes) that are
-related to each other (edges). In this case, each different aggregation area is
-represented as a node and we connect nodes that have overlapping geographical
-areas, leading to the network illustrated on the right of
-Figure~\ref{fig:intuition}. By partitioning the network into connected nodes
-that are similar, we are effectively identifying clusters in the spatio-temporal
-data, as illustrated by the colours of the nodes on the right side of
-Figure~\ref{fig:intuition}. Further, all the possible paths of change can be
-obtained by computing sequences of nodes over time, in this case: (A, C, F), (A,
-D, F), (A, D, G), and (B, E, H). This representation is also suited for
-geographically consistent regions, as illustrated by the stable park in this
-example, and is therefore a generalisation of the traditional paradigm.
+Instead, we propose a network-based representation. A \emph{network} (also
+called a \emph{graph}) is a collection of entities (nodes) that are related to
+each other (edges). In this case, each different aggregation area is represented
+as a node and we connect nodes that have overlapping geographical areas\revision{ in
+different times or are neighbours in the same time}, leading to the network
+illustrated on the right of Figure~\ref{fig:intuition}. By partitioning the
+network into connected nodes that are similar, we are effectively identifying
+clusters in the spatio-temporal data, as illustrated by the colours of the nodes
+on the right side of Figure~\ref{fig:intuition}. Further, all the possible paths
+of change can be obtained by computing sequences of nodes over time, in this
+case: (A, C, F), (A, D, F), (A, D, G), and (B, E, H). This representation is
+also suited for geographically consistent regions, as illustrated by the stable
+park in this example, and is therefore a generalisation of the traditional
+paradigm.
 

### 2019-08-10 paper/related.tex
more fixes, regionalisation ideas

@@
 
-\subsection{Data clustering}
+\subsection{Data clustering and regionalisation}
 
@@
 interpreted in the same way. There is no shortage of contributions for this
-problem~\citep{Fahad2014}, but most neighbourhood related applications still
-rely on k-means~\citep{jain2010data,Delmelle2016} and, to a lesser extent, Self
-Organising Maps~\citep{Delmelle2017,Ling2016}.
+problem~\citep{Fahad2014}\revision{, since variations of it appear in almost all
+scientific fields. Most neighbourhood related applications still rely on
+k-means~\citep{jain2010data,Delmelle2016}, the Louvain method for community
+detection~\citep{blondel2008Louvain,Thomas2012}, or, to a lesser extent, Self
+Organising Maps~\citep{Delmelle2017,Ling2016,Dani2013_som_maxp}.}
 
-However, a method for geographic data analysis should not ignore the geographic
-component of the data, and we extend research exploring ways to incorporate it.
-One straightforward option for agglomerative methods~\citep{han2001spatial} is
-to consider only nearby clusters for merging~\citep{Chavent2017}, which can also
-be done for k-means~\citep{soor2018extending}. Alternatively, the spatial
-distance could be directly added to the inter-cluster metric~\citep{Chavent2017}
-via a mixing parameter, which adds flexibility to the method, but introduces the
-problem of finding the correct application-dependent values.
+\revision{In geography, this problem is known as
+\emph{regionalisation}~\citep{montello2003regions}, a rather old problem that
+has been throughly explored, leveraging different mathematical tools, including
+discrete topology~\citep{brantingham1978topological} and discrete
+geometry~\citep{assunccao2006efficient}. Indeed, network-based methods are among
+the current state-of-the-art~\citep{Guo2008Redcap,maxp}. However,
+\emph{temporal} regionalisation is significantly less explored, especially in a
+demographic context, arguably due to the difficulties in dealing with
+unharmonised longitudinal data.
 
-Indeed, one crucial step in most clustering algorithms is the definition of the
-number of clusters. We avoid this problem by considering hierarchical
-methods~\citep{soille2012morphological}, where the result is not a partition of
-the data, but a tree of partitions, similar to a dendogram. This approach is
-interesting for interactive methods, because it allows the user to change the
-number of clusters on the fly. Since our data is represented as a network, we
-opted for an heuristic variation of the maximum weighted matching algorithm
-called \emph{sorted maximal matching}~\citep{markus2017}, which merges clusters
-based on the weights of the edges between pairs of clusters.
+Since we adopted a network-based data representation and our objectives include
+an interactive interface, we opted for an heuristic variation of the maximum
+weighted matching algorithm called \emph{sorted maximal
+matching}~\citep{markus2017}, because of its simplicity, customisability, and
+fast computation times. This algorithm merges clusters based on the weights of
+the edges between pairs of clusters, creating an hierarchy over the data,
+allowing the user to change the number of clusters without reprocessing the
+data. Changing the clustering algorithm would lead to different results, but any
+hierarchical network clustering method can be used in our framework. Indeed, our
+core contribution is a different interpretation of the data representation, not
+a new clustering or regionalisation method.}
 
@@
 its variables~\citep{Alce2018,Valdivia2015,VonLandesberger2016}. While this
-simplifies the geographic portion of the visualization, it introduces the
+simplifies the geographic portion of the visualisation, it introduces the
 problem of how to summarise the contents of each cluster. One traditional

### 2019-08-10 paper/study.tex
more fixes, regionalisation ideas

@@
 
-\smallskip While Toronto is more dynamic than Los Angeles, possibly due to size
-differences, the volatile regions shown in Figure~\ref{fig:toVol} did not change
-as quickly or dramatically as the ones shown in Figure~\ref{fig:laVol}, which
-involved twice as many people. We found this trend to be related to the
-countries themselves, Canadian cities have larger areas undergoing slow, gradual
-changes, whereas American cities have more general stability, but quicker
-changes in smaller scales. The supplementary material contains brief summaries
-of all the regions accessible in this prototype.
+% \smallskip While Toronto is more dynamic than Los Angeles, possibly due to size
+% differences, the volatile regions shown in Figure~\ref{fig:toVol} did not change
+% as quickly or dramatically as the ones shown in Figure~\ref{fig:laVol}, which
+% involved twice as many people. We found this trend to be related to the
+% countries themselves, Canadian cities have larger areas undergoing slow, gradual
+% changes, whereas American cities have more general stability, but quicker
+% changes in smaller scales. The supplementary material contains brief summaries
+% of all the regions accessible in this prototype.
 

### 2019-08-10 paper/intro.tex
more regionalization

@@
 regions}~\citep{montello2003regions}, geographically continous areas with
-similar data characteristics.} These advances have made neighbourhood constructs
-more tractable to quantitative analysis. Yet new challenges have also emerged,
+similar data characteristics, which these advances made more tractable to
+approach in a data-driven fashion.} Yet new challenges have also emerged,
 especially at the convergence of research on neighbourhood effects and

### 2019-08-10 paper/related.tex
more regionalization

@@
 problem~\citep{Fahad2014}\revision{, since variations of it appear in almost all
-scientific fields. Most neighbourhood related applications still rely on
-k-means~\citep{jain2010data,Delmelle2016}, the Louvain method for community
-detection~\citep{blondel2008Louvain,Thomas2012}, or, to a lesser extent, Self
-Organising Maps~\citep{Delmelle2017,Ling2016,Dani2013_som_maxp}.}
+scientific fields.}
 
@@
 demographic context, arguably due to the difficulties in dealing with
-unharmonised longitudinal data.
+unharmonised longitudinal data.   Recent neighbourhood related applications rely
+on k-means~\citep{jain2010data,Delmelle2016}, the Louvain method for community
+detection~\citep{blondel2008Louvain,Thomas2012}, or, to a lesser extent, Self
+Organising Maps~\citep{Delmelle2017,Ling2016,Dani2013_som_maxp}, but do not
+consider longitudinal data, or harmonised data.
 

### 2019-08-12 backend/dataStore.py
raw values for the bottom plot (so it's proper percentages over the bars/mouseover plots)

@@ -118,7 +118,7 @@ class dataStore(object):
             for n in nodes:
                 res[-1][outIbyN[n],:]=f(self._tabdata[self._IDbyN[n],ind[v]])
         return(res,outNbyInd)
-    def getValue(self, n, dsconf=None):
+    def getValue(self, n, dsconf=None, normalized=True):
         """If dsconf==None, returns all variables, original data, in the avVars order"""
         if (dsconf):
             ivars=dsconf['ivars']
@@ -126,11 +126,14 @@ class dataStore(object):
         else:
             ivars=[v['id'] for v in self._avVars]
             fs=[0,]*len(ivars)
-
         ret=[]
-        for i in range(len(ivars)):
-            r=self._rowByVarID[ivars[i]]
-            ret.append(self._avFilters[fs[i]]['func'](self._tabdata[self._IDbyN[n],r[0]:r[1]]))
+        if normalized:
+            for i in range(len(ivars)):
+                r=self._rowByVarID[ivars[i]]
+                ret.append(self._tabdata[self._IDbyN[n],r[0]:r[1]])
+        else:
+            for v in [self._avVars[i]['name'] for i in ivars]:
+                ret.append(np.array(self._rawdata[n][v]['values']))
         return(ret)
     def distance(self, node1, node2, dsconf):
         ivars=dsconf['ivars']

### 2019-08-12 backend/srv.py

@@ -3,23 +3,24 @@ import json
 import os
 import pickle
 import sys
+from collections import defaultdict
 from itertools import combinations, combinations_with_replacement
 from os import makedirs
 from os.path import exists
 from time import time
 
-import numpy as np
-from numpy import nan
-
 import cherrypy
 import networkx as nx
-from dataStore import dataStore
-from hierMWW import hierMWW
-from mmg import segmentation
+import numpy as np
 from networkx.readwrite import json_graph
+from numpy import nan
 from scipy.spatial.distance import pdist, sqeuclidean, squareform
 from sklearn.manifold import TSNE
 from sklearn.neighbors import NearestNeighbors
+
+from dataStore import dataStore
+from hierMWW import hierMWW
+from mmg import segmentation
 
 # import matplotlib.pylab as plt
 
@@ -146,42 +147,34 @@ def _TemporalDifferences(curCity, dsconf, H, nodes=None):
     
     years = curCity['years']
 
-    states = dict()
-    pop = dict()
+    states = defaultdict(lambda : defaultdict(list))
+    raw = defaultdict(lambda : defaultdict(list)) #year, aspect
+    pop = defaultdict(int)
 
     indVars={v:i for i,v in enumerate(dsconf['ivars'])}
 
     for n in G.nodes():
         V=ds.getValue(n,dsconf)
-        if (n[0] not in states):
-            states[n[0]]=dict()
-            pop[n[0]]=0
-            
-        pop[n[0]]+=ds.getPopulation(n)
-
-
+        y=n[0]
+        pop[y]+=ds.getPopulation(n)
+        rawVals = ds.getValue(n,dsconf,normalized=False)
+        
         for i in dsconf['ivars']:
-            if (i not in states[n[0]]):
-                states[n[0]][i]=[]
-            states[n[0]][i].append(V[indVars[i]])
+            raw[y][i].append(rawVals[indVars[i]])
+            states[y][i].append(V[indVars[i]])
             
-    smin=dict()
-    smax=dict()
-    smed=dict()
-    sq1=dict()
-    sq3=dict()
+    smin=defaultdict(dict)
+    smax=defaultdict(dict)
+    smed=defaultdict(dict)
+    sq1=defaultdict(dict)
+    sq3=defaultdict(dict)
+    rawPercentages=defaultdict(dict)
 
     for y in states:
-        smin[y]=dict()
-        smax[y]=dict()
-        smed[y]=dict()
-        sq1[y]=dict()
-        sq3[y]=dict()
[…]

### 2019-08-12 paper/method.tex
more fixes, regionalisation ideas
interface change (percentage population)

--- 2019-08-08 paper/method.tex
@@
 evolution of these groups. To examine this process of evolution directly, we
-introduce the concepts\revision{of \emph{temporal paths} and \emph{trajectories}.
+introduce the concepts\revision{ of \emph{temporal paths} and \emph{trajectories}.
 
@@
 Each aspect is represented by a stacked bar plot, where the width of each
-rectangle corresponds to the average percentage of that variable over the
-considered period. In this case, about half of the people in Chicago in the
-considered period are married, and the percentage that are Widowers or Divorced
-is roughly similar. About half of the population work in Administrative jobs, a
-third never completed high-school, approximately half have gross family income
-below 50,000USD per year. The vast majority identify as white. Placing the mouse
-over one of the bars will open a small panel with the temporal evolution of that
-specific variable, and clicking on the chevron on the right side expands the
-corresponding aspect, showing details of the temporal evolution of each variable
-and also the corresponding IQRs for the whole city.
+rectangle corresponds to \revision{the percentage of population in that variable
+over the considered period. In this case, a little less than half of the
+population are married, and the percentage that are Widowers or Divorced is
+roughly similar. About half of the population work in Administrative jobs, a
+third never completed high-school, approximately fourty percentq have gross
+family income below 50,000USD per year. About sixty percent identify as white.
+Placing the mouse over one of the bars will open a small panel with the temporal
+evolution of the population percentage of that specific variable, and clicking
+on the chevron on the right side expands the corresponding aspect, showing
+census level statistics, with details of the temporal evolution of each variable
+and also the corresponding IQRs for the whole city. Those values are potentially
+different, for instance, while the overall population of this region is about
+sixty percent White, it includes regions with wildly different percentages, so
+the average percentage of White population over the CTs is a little over fourty
+percent.}

### 2019-08-14 paper/intro.tex
first pass

--- 2019-08-10 paper/intro.tex
@@
 
-\revision{In this work, we consider neighbourhoods as \emph{formal
-regions}~\citep{montello2003regions}, geographically continous areas with
-similar data characteristics.} These advances have made neighbourhood constructs
-more tractable to quantitative analysis. Yet new challenges have also emerged,
-especially at the convergence of research on neighbourhood effects and
-neighbourhood dynamics. Neighbourhood effects research assumes knowledge about
-the nature and scope of "the neighbourhood" that presumably shapes individual
-outcomes~\citep{Kwan2018,Shelton2019}. Concurrently, researchers note that
-neighbourhoods are not necessarily fixed containers in which other processes
-occur, but themselves dynamically
-evolve~\citep{Delmelle2017,Reades2019,li2018new}. The result is to open up key
-assumptions about neighbourhoods for theoretical and empirical examination: how
-do we appropriately define and compare neighbourhoods at a given time?; how do
-we appropriately define and compare the temporal trajectories of
-neighbourhoods?; and can we do both at once, "fully
+Yet new challenges have also emerged, especially at the convergence of research
+on neighbourhood effects and neighbourhood dynamics. Neighbourhood effects
+research assumes knowledge about the nature and scope of "the neighbourhood"
+that presumably shapes individual outcomes~\citep{Kwan2018,Shelton2019}.
+Concurrently, researchers note that neighbourhoods are not necessarily fixed
+containers in which other processes occur, but themselves dynamically
+evolve~\citep{Delmelle2017,Reades2019,li2018new}. \revision{In this work, we
+consider neighbourhoods as \emph{formal regions}~\citep{montello2003regions},
+geographically continous areas with similar data characteristics, which these
+advances made more tractable to approach in a data-driven fashion.} The result
+is to open up key assumptions about neighbourhoods for theoretical and empirical
+examination: how do we appropriately define and compare neighbourhoods at a
+given time?; how do we appropriately define and compare the temporal
+trajectories of neighbourhoods?; and can we do both at once, "fully
 interactionally"~\citep{Abbott1997}: classify neighbourhoods now based on where
@@
 
-The standard approach then is to \emph{geographically harmonise} data. This
+The standard approach then is to \emph{geographically harmonise} the data. This
 involves interpolating existing measurements into a common set of
@@
 provided~\citep{eicher2001dasymetric}. Essentially, harmonisation generates
-artificial data points that can potentially lead to inaccurate results, even
-though they are seldom interpreted as such. Nevertheless, because there has been
-no viable alternative, and the results often appear plausible, these concerns
-are generally overlooked. The result is that the harmonisation approach is
-virtually mandatory in the current literature: \emph{"(...) tract-by-tract
+\emph{artificial data points} that can potentially lead to inaccurate results,
+even though they are seldom interpreted as such. Nevertheless, because there has
+been no viable alternative, and the results often appear plausible, these
+concerns are generally overlooked. The result is that the harmonisation approach
+is virtually mandatory in the current literature: \emph{"(...) tract-by-tract
 comparison is not possible unless data from 2000 is interpolated to 2010
@@
 Section~\ref{sec:introduction}, then we review the relevant literature on
-longitudinal studies, data representation, clustering, and spatio-temporal
-visualisation in Section~\ref{sec:related}. Our methodology is introduced in
-detail in Section~\ref{sec:method}, along with the included interface.
-Illustrative scenarios for Chicago, Toronto, and Los Angeles are presented in
+longitudinal studies, data representation, clustering and regionalisation, and
+spatio-temporal visualisation in Section~\ref{sec:related}. Our methodology is
+introduced in detail in Section~\ref{sec:method}, along with the included
+interface. Illustrative scenarios for Chicago and Toronto are presented in
 Section~\ref{sec:study} and the feedback of five field experts are summarised in
@@
 
-To ease the cognitive transition to a new paradigm, we start with an intuitive
+To ease the cognitive transition to this paradigm, we start with an intuitive
 example of how the method works, using a small portion of a fictitious urban
@@
 
-This example illustrates some of the challenges of harmonisation. None of these
-aggregation areas is clearly suitable as an interpolation target. In fact,
-adopting any of these areas as a target would require merging heterogeneous
+The aggregation areas of none of these years is clearly suitable as an
+interpolation target. Adopting any of them would require merging heterogeneous
 regions and/or dividing homogeneous regions. For instance, by choosing the
@@
 Note that the edges of this network merely encode that two regions are related.
-This is binary information, there is no apportionment, no areal measurements,
-no population percentages associated with the edge. Indeed, our method also
-connects regions of the same time that share borders, representing exactly that
-they are neighbouring areas.
+This is binary information, there is no apportionment, no areal measurements, no
+population percentages, or weights of any kind associated with the edge. Indeed,
+our method also connects regions of the same time that share borders,
+representing exactly that they are neighbouring areas.
 
-In the following, we argue that this network representation allows us to study
-neighbourhood change in a way that does not require an prior interpolation.
+In the following, we demonstrate that this network representation allows us to
+study neighbourhood change in a way that does not require prior interpolation.

### 2019-08-14 paper/method.tex
first pass

@@
 proposed method, illustrating how the nodes of the network are used to represent
-the regions. The following sections elaborate this figure and explain the main
-features of the interface we built to visualise and explore the evolution of
-neighbourhoods on the basis of our proposed method.
+the regions. The following sections elaborate on this figure and explain the
+main features of the interface we built to visualise and explore the evolution
+of neighbourhoods on the basis of our proposed method.
 
@@
 Having tied the regions together into a network, we can now partition it to
-identify similar sets of regions. We start by adopting a distance function
-between the nodes, measuring the difference between the data of the regions.
-This value is then associated with the edges, leading to a weighted dynamic
-network. Every node has a collection of histograms, each representing the
-distribution of certain aspect in the population.
+identify similar sets of regions. \revision{In essence, we are performing
+temporal regionalisation by applying a regionalisation method over this
+spatio-temporal network.}
+
+We start by adopting a distance function between the nodes, measuring the
+difference between the data of the regions. This value is then associated with
+the edges, leading to a weighted dynamic network. Every node has a collection of
+histograms, each representing the distribution of certain aspect in the
+population.
 
@@
 evaluation using several metrics. \revision{We chose this algorithm because it
-is simple and easily customisable, and while different algorithms will lead to
-different results, our methodology should work with any hierarchical clustering
-algorithm.}
+is fast, simple, and easily customisable, but our methodology should work with
+any hierarchical network clustering algorithm.}
 
@@
 \revision{The regions connected by those edges will be merged on the first
-stages of the clustering, since they are very similar, leaving the remaining
+stages of the clustering, since the nodes they connect are, by definition, as
+similar as possible\footnote{This assertion relies on the euclidean distance and
+its relationship with the space where k-nn operates.}, leaving the remaining
 steps of the hierarchy to be determined only by the geographical edges. We
-explored with different numbers of augmentation edges, but the results were not
-consistent, since the distribution of the edges is data dependent. Adding two
-edges per node was the least number of edges that led to stable and consistent
-results in the scenarios available in our prototype. Since the problem of
-balancing the data space with the geographical space is relevant for
-geographical data analysis, this methodology potentially warrants further
-exploration, beyond the scope of this work.}
+experimented with different numbers of augmentation edges, but the results were
+not consistent, since the distribution of the edges is data dependent. Adding
+two edges per node was the smallest number of edges that led to stable and
+consistent results in the scenarios available in our prototype. Since the
+problem of balancing the data space with the geographical space is relevant for
+geographical data analysis, this idea potentially warrants further exploration,
+beyond the scope of this work.}
 

### 2019-08-14 paper/related.tex
first pass

@@
 representation}, elaborating how evolving geographic data can be represented for
-processing; \emph{data clustering}, briefly reviewing existing clustering
-methods; and \emph{cluster characterisation}, articulating how clusters can be
-visually summarised.
+processing; \emph{data clustering and regionalisation}, briefly reviewing
+existing data clustering methods,\revision{ geographic constraints and
+regionalisation}; and \emph{cluster characterisation}, articulating how clusters
+can be visually summarised.
 
@@
 detection~\citep{Doraiswamy2014}, leveraging signal processing on
-graphs~\citep{shuman2013emerging,sandryhaila2013discrete} to find patterns and
+networks~\citep{shuman2013emerging,sandryhaila2013discrete} to find patterns and
 outliers~\citep{Valdivia2015,Dias2015,Alce2018}. Networks are well suited to
@@
 well~\citep{VonLandesberger2016,Huang2016,chen2015survey}, allowing the use of
-graph visualisation methods~\citep{Vehlow2015,Beck2014}. Our proposed method
+network visualisation methods~\citep{Vehlow2015,Beck2014}. Our proposed method
 builds upon this literature. We leverage a network-based representation that
-removes the requirement for consistency in the measurement regions. Each region
-in time corresponds to a different node. Instead of a collection of time-series,
-the data is represented as a dynamic network.
+removes the rigidity in the measurement regions. Each region in time corresponds
+to a different node. Instead of a collection of time-series, the data is
+represented as a dynamic network.
 
@@
 detection~\citep{blondel2008Louvain,Thomas2012}, or, to a lesser extent, Self
-Organising Maps~\citep{Delmelle2017,Ling2016,Dani2013_som_maxp}, but do not
-consider longitudinal data, or harmonised data.
+Organising Maps~\citep{Delmelle2017,Ling2016,Dani2013_som_maxp}.
 
@@
 matching}~\citep{markus2017}, because of its simplicity, customisability, and
-fast computation times. This algorithm merges clusters based on the weights of
-the edges between pairs of clusters, creating an hierarchy over the data,
-allowing the user to change the number of clusters without reprocessing the
-data. Changing the clustering algorithm would lead to different results, but any
-hierarchical network clustering method can be used in our framework. Indeed, our
-core contribution is a different interpretation of the data representation, not
-a new clustering or regionalisation method.}
+fast computation times. This algorithm operates on weighted networks, where a
+distance metric is between the nodes is associated with the edges. We adopted a
+distance based on the data associated with each node. The algorithm merges
+clusters based on these distances, creating an hierarchy instead of a fixed
+number of clusters. This hierarchical result is rendered by the interface,
+allowing the user to change the number of clusters without reprocessing.
+Changing the clustering algorithm would lead to different results, but any
+hierarchical network clustering method, and distance metric, can be used in our
+framework. }
 

### 2019-08-14 paper/study.tex
first pass

@@
 
-In this section we present two illustrative scenarios, using decennial census
-data from the United States~\citep{nhgis} and
+In this section we present two illustrative scenarios, usingq census data from
+the United States~\citep{nhgis} and
 Canada\footnote{\url{http://datacentre.chass.utoronto.ca/census/}}, tabulated by
-CTs,  from 1970 to 2010. \revision{Accidentaly, the US data for 2010 is actually
-not from the decennial census, but from the ACS 2006-2010. Further, the regions
-selected do not correspond to any pre-defined regions (metros, cities, census
-areas), but to arbitrary regions defined around a location of interest. We
-selected a minimal set of aspects for each country, and they are not similar to
-each other, out of convenience, since the variable matching was a manual
-process. These caveats would likely compromise any serious attempt on a
-comprehensive demographic study, but this is not the objective of this
-experiment. We aim only to test and validate our methodology, comparing our
-results to pre-existing literature insights. Indeed, our results were perfectly
-aligned to several other studies, despite these methodological missteps, which
-could arguably be an indicator of the robustness of our method.}
+CTs,  from 1970 to 2010.
 
@@
 
-These results are meant to demonstrate the utility of the interface for
-understanding the evolutionary dynamics of urban neighbourhoods. They also show
-the face validity of the results generated by our novel network-based approach.
+\revision{Accidentaly, the US data for 2010 is actually not from the decennial
+census, but from the ACS 2006-2010. Further, the regions selected do not
+correspond to any pre-defined regions (metros, cities, census areas), but to
+arbitrary regions defined around a location of interest. We selected a minimal
+set of aspects for each country, and they are not similar to each other, out of
+convenience, since the variable matching was a manual process. These caveats
+would likely compromise any serious attempt on a comprehensive demographic
+study, but this is not the objective of this experiment. These results are meant
+to demonstrate the utility of the interface for understanding the evolutionary
+dynamics of urban neighbourhoods. They also show the face validity of the
+results generated by our novel network-based approach. Indeed, our results were
+perfectly aligned with several other studies, despite these methodological
+missteps, which could arguably be an indicator of its robustness.}
 
@@
 
-We consider a region that is approximately the administrative border of the
-current city of Toronto, using all seven available aspects with equal weights.
-While Chicago was fairly stable, Toronto is known to be a more dynamic and
-diverse city, with significant and increasing immigrant
+We consider a region that corresponds approximately to the administrative border
+of the current city of Toronto, using all seven available aspects with equal
+weights. While Chicago was fairly stable, Toronto is known to be a more dynamic
+and diverse city, with significant and increasing immigrant
 population~\citep{hulchanski2007three,Fong2011}, especially
@@
 has lower education and income levels than the brown cluster. The purple cluster
-of high income is somewhat stable. Until 2011 the cluster included the Bridle Path
-neighbourhood, known for its wealthy population. In 2011 it was
-classified into the yellow cluster of Asian immigration, since about 35\% of the
-population for this CT were then born in Asia. The income distribution did not
-change, with 85\% of the population with an income of 90k CAD or more.
+of high income is somewhat stable. Until 2011 the cluster included the Bridle
+Path neighbourhood, known for its wealthy population. In 2011 it was classified
+into the yellow cluster of Asian immigration, since about 40\% of the population
+for this CT were then born in Asia. The income distribution did not change, with
+85\% of the population with an income of 90k CAD or more.
 
@@
 amount of data transferred to client, but it might increase the response time
-under load. We implemented a cache policy that greatly improved the performance,
-but fully pre-processing the results is not practical due to size of the
-parameter space.
+under load. We implemented a caching policy that greatly improved the
+performance, but fully pre-processing the results is not practical due to size
+of the parameter space.
 

### 2019-08-24 paper/intro.tex
from overleaf

@@
 containers in which other processes occur, but themselves dynamically
-evolve~\citep{Delmelle2017,Reades2019,li2018new}. \revision{In this work, we
-consider neighbourhoods as \emph{formal regions}~\citep{montello2003regions},
-geographically continous areas with similar data characteristics, which these
-advances made more tractable to approach in a data-driven fashion.} The result
-is to open up key assumptions about neighbourhoods for theoretical and empirical
-examination: how do we appropriately define and compare neighbourhoods at a
-given time?; how do we appropriately define and compare the temporal
-trajectories of neighbourhoods?; and can we do both at once, "fully
+evolve~\citep{Delmelle2017,Reades2019,li2018new}.  The result is to open up key
+assumptions about neighbourhoods for theoretical and empirical examination: how
+do we appropriately define and compare neighbourhoods at a given time?; how do
+we appropriately define and compare the temporal trajectories of
+neighbourhoods?; and can we do both at once, "fully
 interactionally"~\citep{Abbott1997}: classify neighbourhoods now based on where
@@
 practice tends to rely on methods that require fixed geographical regions. This
-requirement is difficult to satisfy, as most longitudinal datasets are based on
-pre-defined tabulation areas that are routinely modified by data collection
-agencies, usually to follow population changes.
+requirement is not trivial to satisfy in general, as most longitudinal datasets
+are based on pre-defined tabulation areas that are routinely modified by data
+collection agencies, usually to follow population changes.
 
@@
 involves interpolating existing measurements into a common set of
-regions~\citep{Logan2014,Hallisey2017,Allen2018}. Recent computational tools
-have somewhat simplified this process~\citep{rey2018spatio}, but it still
-involves non-trivial questions: which geometry to use as target, how to
-apportion the variables, or how to combine data from different sources. Further,
-these question do not necessarily have optimal answers. Indeed, regardless of
-how well this process is performed, it still introduces
-errors~\citep{Logan2016}, even when additional data is
+regions~\citep{Logan2014,Hallisey2017,Allen2018}. \revision{1.18}{Recent
+computational tools have somewhat simplified this
+process~\citep{rey2018spatio}}, but it still involves non-trivial questions:
+which geometry to use as target, how to apportion the variables, or how to
+combine data from different sources. Further, these question do not necessarily
+have optimal answers. Indeed, regardless of how well this process is performed,
+it still introduces errors~\citep{Logan2016}, even when additional data is
 provided~\citep{eicher2001dasymetric}. Essentially, harmonisation generates
@@
 been no viable alternative, and the results often appear plausible, these
-concerns are generally overlooked. The result is that the harmonisation approach
-is virtually mandatory in the current literature: \emph{"(...) tract-by-tract
+concerns are overlooked. The result is that the harmonisation approach is
+virtually mandatory in the current literature: \emph{"(...) tract-by-tract
 comparison is not possible unless data from 2000 is interpolated to 2010
@@
 
-The main contribution of this paper is a method for longitudinal data processing
-that works with the original data by leveraging a network based representation.
-It enables tract-by-tract comparison and the identification of patterns of
-demographic evolution \emph{without geographic harmonisation}.
+\unsure{CHECK blue text below}
+\revision{1.1}{In some cases, harmonisation might seem methodologically
+adequate, if a study is focused on a small subregion, fixed in time, limiting
+and modifying the data to fit that specific region of interest sounds not only
+reasonable, but recommended, albeit the creation of artificial data points.
+However, that is a simplistic view that ignores interactions beyond those
+borders: each different portion of the area of interest may belong to larger
+homogeneous processes beyond these arbitrarily defined borders. This effect,
+along with the creation of artificial points, is caused by the geometry mismatch
+between the data and the region of interest. Instead of forcing one set of
+borders into another, our method sidesteps this issue by changing how the data
+is represented and interpreted.
+
+The main contribution of this paper is a method for longitudinal geographical
+data representation and processing that works with the original data by
+leveraging a network based representation. It enables direct comparison and
+temporal regionalisation without the need to match arbitrarily defined borders.
+Even in the case of a small, temporally fixed, region of interest, our approach
+leads to a better understanding of its dynamics by exploring the related
+data-driven homogeneous regions. }
 
@@
 
-Real world measurements are seldom as homogeneous and noise-free as this
-artificial example. By splitting and merging the data to fit arbitrary borders,
-that were not necessarily coherent at the time of the measurement, harmonisation
-increases the distance between measurement and reality.
+\revision{2.6.4}{While didactic, this example is not realistic because each
+building is visible in Figure~\ref{fig:intuition}, so we know each individual of
+the tabulation area, which is never the case. Further, real measurements are
+seldom as homogeneous and noise-free as this artificial example.  In this
+example, one possible harmonisation would split region A  in two to provide
+different ancestors to C ($A_C$) and D ($A_D$). Then $A_D$ would not be
+considered when analysing the evolution of C, nor $A_C$ for $D$. This is
+inaccurate, because $A$ was the original state for both evolutions, and we
+should not assume that these processes are disconnected.} Either way, by
+splitting and merging the data to fit arbitrary borders, the harmonisation
+process increases the distance between the data and reality by creating
+artificial data points.
 
@@
 each other (edges). In this case, each different aggregation area is represented
-as a node and we connect nodes that have overlapping geographical areas\revision{ in
+as a node and we connect nodes that have overlapping geographical areas\revision{1.2}{ in
 different times or are neighbours in the same time}, leading to the network
@@
 representing exactly that they are neighbouring areas.
-
-In the following, we demonstrate that this network representation allows us to
-study neighbourhood change in a way that does not require prior interpolation.

### 2019-08-24 paper/method.tex
from overleaf

@@
     \includegraphics[width=0.975\linewidth]{overview.pdf}
-    \caption{Overview of the proposed method. A network is generated combining the
-        original census data, encoding the changing geographical information.
-        The network is partitioned into an hierarchy~\citep{markus2017}. The
-        characteristics and evolution of the clusters are then visually
-        represented.
+    \caption{Overview of the proposed method. 
+        \textbf{(a)} A network is generated from the original data,
+        encoding the changing geographical information. 
+        \textbf{(b)} The network is partitioned into an hierarchy. 
+        \textbf{(c)} The characteristics and evolution of the clusters are then
+        visually represented.
         \label{fig:overview}}
@@
 year. Further, edges are placed between nodes if the corresponding regions
-belong to sequential years and \revision{their geometries intersect. To avoid
-spurious connections caused by geometry fluctuations, one of the geometries is
-slightly shrunk before the intersection, using a buffer of -1e6. More
-importantly, while weights will be associated with these edges before they are
-processed, they are not derived from the geometry, but from the data. The actual
-intersection area is not considered in this representation. } This approach
-leads to a single network representing the whole spatio-temporal space of the
-data. Our objective then becomes to identify partitions of this network such
-that the nodes of each partition are more similar between themselves than to the
-other nodes.
+belong to sequential years and \revision{1.6/2.6.3}{their geometries intersect.
+To avoid spurious connections caused by small fluctuations, one of the
+geometries is slightly shrunk before the intersection, using a buffer of -1e6.
+While the result would not change considerably by adding more connections
+between nearby regions, these extra connections would greatly increase the
+computational cost. More importantly, while weights will be associated with
+these edges before they are processed, they are not derived from the geometry,
+but from the data. The actual intersection area is not considered in this
+representation. } This approach leads to a single network representing the whole
+spatio-temporal space of the data.
 
@@
 Having tied the regions together into a network, we can now partition it to
-identify similar sets of regions. \revision{In essence, we are performing
-temporal regionalisation by applying a regionalisation method over this
+identify similar sets of regions. \revision{2.3}{In essence, we are performing
+temporal regionalisation by applying a clustering method over this
 spatio-temporal network.}
@@
 cuts~\citep{Cousty2009} to create an initial clustering, which is then refined
-into a hierarchy using the Sorted Maximal Matching (SMM)~\citep{markus2017} with
-median linkage. The initial watershed step is performed to create an initial
-clustering and reduce the running time of the SMM. We introduced one new
-parameter to this method: a maximum distance threshold for the merges, to avoid
-the early merging of outliers. We refer the reader to the original
-paper~\citep{markus2017} for more details, including a complete performance
-evaluation using several metrics. \revision{We chose this algorithm because it
-is fast, simple, and easily customisable, but our methodology should work with
-any hierarchical network clustering algorithm.}
+into a hierarchy using the Sorted Maximal Matching (SMM)~\citep{markus2017}. The
+initial watershed step is performed to create an initial clustering and reduce
+the running time of the SMM. We introduced one new parameter to this method: a
+maximum distance threshold for the merges, to avoid the early merging of
+outliers. We refer the reader to the original paper~\citep{markus2017} for more
+details, including a complete performance evaluation using several metrics.
+\revision{1.5/2.2}{We chose this algorithm because it is fast,
+simple, and easily customisable, but our methodology should work with any
+hierarchical network clustering algorithm. However, we caution against the use
+of single-linkage methods due to their tendency to form chain-like clusters with
+elements that are similar pairwise, but significantly different on the ends.
+That is also the reason we did not use a Watershed hierarchy
+directly~\citep{najman2011equivalence}.}
 
@@
 connected, providing a path for the algorithm to group similar nodes.
-\revision{The regions connected by those edges will be merged on the first
+\revision{2.4}{The regions connected by those edges will be merged on the first
 stages of the clustering, since the nodes they connect are, by definition, as
@@
 \emph{relevance} of a variable of an aspect based on the distance between the
-interquartile ranges (IQR) of the clusters in the same hierarchical level. If
+\revision{1.8}{interquartile} ranges (IQR) of the clusters in the same hierarchical level. If
 the IQRs overlap for all clusters, that variable is not relevant to the
@@
 
+To allow for an easier visualisation of these ideas, we represent it using an
+\emph{enhanced} version of the traditional boxplot, which includes the IQRs for
+the other clusters, in slightly larger and faded black rectangles. We also
+colour the current IQR according to its relevance, as illustrated in
+Figure~\ref{fig:boxplot}. In this example, this cluster is best defined by the
+proportion of the population with four or more years of college. The user can
+quickly see that this is relevant because the corresponding IQR is coloured with
+the highest relevance present in the legend. It is also clear that, while this
+cluster includes CTs that have between 10\% to 90\% of people in this variable,
+approximately, half of them have about 60\% of the population with four or more
+years of college. Since all the other IQRs are well separated, this is a
+defining characteristic of this cluster. Conversely, the proportion of the
+population with one to three years of college is not relevant, as indicated by
+black fill in the rectangle representing the IQR of this cluster, in overlapping
+position with the rectangles of the other clusters.
+
 \subsection{Clusters and trajectories}
@@
 evolution of these groups. To examine this process of evolution directly, we
-introduce the concepts\revision{ of \emph{temporal paths} and \emph{trajectories}.
+introduce the concepts\revision{2.6.4}{ of \emph{temporal paths} and \emph{trajectories}.
 
@@
 instance, in Figure~\ref{fig:intuition}, the sequences ACF, ADF, ADG, and BEH
-are temporal paths. With harmonised data, the time-series of to each region
-would form a temporal path, each node would be connected only to its older and
-newer versions, belonging to only one temporal path, as illustrated by the path
-BEF. Since our data is not harmonised, more connections are allowed and each
-node can belong to an arbitrary number of paths.
+are temporal paths. With harmonised data, as illustrated by the path BEF, the
+time-series of each region would form a temporal path, each node would be
+connected only to its older and newer versions, and would belong to only one
+temporal path. Since our data is not harmonised, more connections are allowed
+and each node can belong to an arbitrary number of paths.
 
@@
 words, this network encodes the information that portions of the region A
-changed to form regions C and D, but we do not know specifically which parts,
-nor we need to, since the same region can belong to several temporal paths.
+changed to form regions C and D, but we do not know specifically which parts.
+Nor do we need to, since the same region can belong to several temporal paths.
 Interestingly, when interpreted in this framework, geographical harmonisation is
@@
 
-\subsection{User interface}
+\subsection{Overview of the prototype interface}
 \label{sec:ui}
@@
     \includegraphics[width=0.975\linewidth]{ui.pdf}
-    \caption{Initial interface of our method showing the demographic evolution of Chicago. 
-        \textbf{(a)}: Configuration panel with the current clustering parameters and controls.
-        \textbf{(b)}: Cluster overview illustrating the most relevant aspect for each cluster. 
-        \textbf{(c)}: Trajectories overview and the general evolution of the population, geographical information, and how it changed. 
-        \textbf{(d)}: Details of the selected trajectories, including precise geographic locations, population numbers, and the composition of the aspects.\label{fig:ui}}
+    \caption{Initial interface of our method showing the demographic evolution of Chicago and identifying the objectives of each plot.
+        \textbf{(what)}: Cluster overview illustrating the most relevant aspect for each cluster. 
+        \textbf{(where)}: Maps illustrating the geographical location, both as an overall summary and for each year.
+        \textbf{(how)}: Tabular and visual summmarisation of how the regions classification changed over time.
+        \textbf{(who)}: Demographic details for the currently selected data.
+        \label{fig:ui}}
 \end{figure*}
 
-As illustrated by Figure~\ref{fig:ui}, our proposed interface heavily relies on
-colour to express cluster-related information. We adopted this convention
-because colours can be used in all our visual tools in a coherent manner.
-However, there is a limit on the number of distinct colours that can be used. We
-limited the number of clusters to eight because this was the largest number of
-colours that we could reliably and accessibly use, derived from the 8-class
-Dark2 set from ColorBrewer~\citep{ColorBrewer}.
+As illustrated by Figure~\ref{fig:ui}\unsure{CHECK image/ sec 4.5}, our proposed
+interface heavily relies on colour to express cluster-related information. We
+adopted this convention because colours can be used in all our visual tools in a
+coherent manner. However, there is a limit on the number of distinct colours
+that can be used. We limited the number of clusters to eight because this was
+the largest number of colours that we could reliably and accessibly use, derived
+from the 8-class Dark2 set from ColorBrewer~\citep{ColorBrewer}.
 
-The configuration panel, on top left in Figure~\ref{fig:ui}, displays which
-aspects were used and their weights (following Equation~\ref{eq:dist}). It also
-includes other configuration options that can be altered without re-processing
-the data, such as the number of clusters and the colour option. The gear button
-allows access to the other configuration options that do require further
-processing, such as changing location, aspects, and weights.
+The configuration panel, on top left in Figure~\ref{fig:ui}, displays the
+processing configuration of the visualised results: which aspects were used and
+their weights (following Equation~\ref{eq:dist}). It also includes other
+configuration options that can be altered without re-processing the data, such
+as the number of clusters and the colour option. The gear button allows access
+to the other configuration options that do require further processing, such as
+changing location, aspects, and weights.
 
-The cluster overview panel, on the bottom left in Figure~\ref{fig:ui}, displays
-a brief summary of each cluster, based on the distance between the IQRs, as
-detailed in Section~\ref{sec:relevance}. The \emph{View all} button opens a new
-panel where all aspects are included, while the chevron at the side lets the
-user expand each cluster separately.
-
-We adopted an \emph{enhanced} version of the traditional boxplot, which includes
-the IQRs for the other clusters, in slightly larger and faded black rectangles.
-We also colour the current IQR according to its relevance.  For instance, the
-boxplot that summarises the purple cluster illustrated in Figure~\ref{fig:ui},
-detailed on Figure~\ref{fig:boxplot}, illustrates that this cluster is best
-defined by the proportion of the population with four or more years of college.
-The user can quickly see that this is relevant because the corresponding IQR is
-coloured with the highest relevance present in the legend. It is also clear
-that, while this cluster includes CTs that have between 10\% to 90\% of people
-in this variable, approximately, half of them have about 60\% of the population
-with four or more years of college. Since all the other IQRs are well separated,
-this is a defining characteristic of this cluster. Conversely, the proportion of
-the population with one to three years of college is not relevant, as indicated
-by black fill in the rectangle representing the IQR of this cluster, in
-overlapping position with the rectangles of the other clusters. By clicking on
-the coloured bar above the boxplot, the user can select all trajectories that
-contain this cluster at any point in time.
+The cluster overview panel, marked as \emph{what} in Figure~\ref{fig:ui},
+displays the most relevant aspects for each cluster, based on the distance
+between the IQRs, as detailed in Section~\ref{sec:relevance}.\revision{2.5}{This
+feature allows the user to quickly understand what exists in this region and
+which aspect makes each cluster unique, without going into excessive detail. In
+Chicago's example, we expect race to be relevant in general, and indeed, it is
+recognised as the most relevant aspect for three of the four most distinct
+clusters, with the fourth being education level. As we increase the number of
+clusters, they get more nuanced, requiring more than one aspect for their proper
+characterisation, as each cluster is divided into its composing sub-clusters.
+The \emph{View all} button opens a new panel where all aspects are included,
+while the chevron at the side lets the user expand each cluster separately,
+allowing the exploration of more details on demand.}
 
@@
 
-The trajectories overview aims to convey basic information about the
-trajectories, where they are, and what changes are involved. This is done using
-three sub panels. The first, on the left, contains a Sankey diagram illustrating
-the evolution of the clusters over time. The widths are proportional to the
-population involved. In our example in Figure~\ref{fig:ui}, the orange and green
-clusters contain most of the population and are fairly stable over time. The
-pink cluster is small and mostly stable. The purple cluster is increasing,
-mostly by incorporating areas that were previously orange. Since the purple
-group corresponds to the emergent 'young urban' group, this corroborates the
-findings of Delmelle~\citep{Delmelle2016,Delmelle2017}, showing that our
-network-based method can recover results from the traditional data processing
-approach.
+With the knowledge of what clusters are present, we can explore where they are
+through the maps, marked as \emph{where} in Figure~\ref{fig:ui}. While the maps
+on the lower part of the interface show the location of the clusters at each
+time, the larger map shows a summarised version, based on the trajectories of
+each region. \revision{1.9}{In simplified mode the colour corresponds to the
+cluster present in that trajectory for more than half the temporal samples, and
+is grey otherwise. With simplified colours disabled, the colours are the average
+of the colours of the clusters in that trajectory in RGB colour
+space~\footnote{\url{https://gka.github.io/chroma.js/\#chroma-average}}.}
 
-In the next panel, illustrated in the top middle of Figure~\ref{fig:ui}, is a
-transition matrix between the clusters. It indicates a rounded percentage of the
-population whose area changed between each pair of clusters. This kind of table
-can be found in the related literature~\citep{Delmelle2016}, so it is familiar
-to the advanced users. It not only informs the proportional changes, but allows
-the selection of the corresponding trajectories for further analysis.
+\revision{2.5}{Each trajectory represents a different temporal path through the
+clusters. Therefore the trajectories encode how the demographic characteristics
+changed over time. These changes are summarised by the two sections identified
+as \emph{how} in Figure~\ref{fig:ui}. In the left panel, there is a Sankey
+diagram where each ribbon represents a different trajectory. The widths are
+proportional to the population involved.} In our example in Figure~\ref{fig:ui},
+the orange and green clusters contain most of the population and are fairly
+stable over time. The pink cluster is small and mostly stable. The purple
+cluster is increasing, mostly by incorporating areas that were previously
+orange. Since the purple group corresponds to the emergent 'young urban' group,
+this corroborates the findings of Delmelle~\citep{Delmelle2016,Delmelle2017},
+showing that our network-based method can recover results from the traditional
+data processing approach.
 
-The panel in the top right of Figure~\ref{fig:ui} is a map of the region under
-analysis, summarising the geographical evolution of the clusters over time.  The
-colours are derived from the clusters involved in each trajectory, which are
-consistent across the linked views.
+\revision{2.5}{In the next panel, illustrated in the top middle of
+Figure~\ref{fig:ui}, is a transition matrix between the clusters. It indicates a
+rounded percentage of regions whose area changed between each pair of
+clusters. In this example, we can see that all clusters are fairly stable, with
+more than ninety percent of identity transitions.} This kind of table can be
+found in the related literature~\citep{Delmelle2016}, so it is familiar to the
+advanced users. It not only informs the proportional changes, but allows the
+selection of specific changes for further analysis.
 
 The bottom part of the interface contains the details for the selected
-trajectories, or for the whole city if nothing is selected, as illustrated in
-Figure~\ref{fig:ui}. This panel contains two main regions: the small multiple
-maps, depicting the clusters at each year, and the stacked bar plots that
-summarise the overall composition of these regions. In this example, the maps
-show the transition from orange to green and purple in several regions over
-time. Clicking on a region in these maps will bring up a new panel with the
-original census data of this specific region. The actual population numbers are
-below the maps.
+trajectories, or for the whole city if nothing is selected, marked as \emph{who}
+in Figure~\ref{fig:ui}. The stacked bar plots summarise the overall demographic
+composition of these regions. In this example, the maps show the transition from
+orange to green and purple in several regions over time. Each aspect is
+represented by a stacked bar plot, where the width of each rectangle corresponds
+to \revision{1.10}{the percentage of population in that variable over the
+considered period. In this case, a little less than half of the population are
+married, and the percentage that are Widowers or Divorced is roughly similar.
+About half of the population work in Administrative jobs, a third never
+completed high-school, approximately forty percent have gross family income
+below 50,000USD per year. About sixty percent identify as white. Placing the
+mouse over one of the bars will open a small panel with the temporal evolution
+of the population percentage of that specific variable, and clicking on the
+chevron on the right side expands the corresponding aspect, showing census tract
+level statistics, with details of the temporal evolution of each variable and
+also the corresponding IQRs for the whole city. Those values differ from the
+total population percentage, while the overall population of this region is
+about sixty percent White, the average percentage of White population over the
+CTs is a little over forty percent.}
 
-Each aspect is represented by a stacked bar plot, where the width of each
-rectangle corresponds to \revision{the percentage of population in that variable
-over the considered period. In this case, a little less than half of the
-population are married, and the percentage that are Widowers or Divorced is
-roughly similar. About half of the population work in Administrative jobs, a
-third never completed high-school, approximately fourty percentq have gross
-family income below 50,000USD per year. About sixty percent identify as white.
-Placing the mouse over one of the bars will open a small panel with the temporal
-evolution of the population percentage of that specific variable, and clicking
-on the chevron on the right side expands the corresponding aspect, showing
-census level statistics, with details of the temporal evolution of each variable
-and also the corresponding IQRs for the whole city. Those values are potentially
-different, for instance, while the overall population of this region is about
-sixty percent White, it includes regions with wildly different percentages, so
-the average percentage of White population over the CTs is a little over fourty
-percent.}
+More importantly, this interface is fully interactive, allowing for a
+progressive exploration of the data. Clicking on a cluster in the cluster
+overview panel and the other panels will filter the results to consider only
+that cluster, highlighting the geographic regions and involved changes, and
+updating population count and demographic composition. If the user selects a
+region from the larger map, all identical trajectories are considered, and the
+whole interface changes to allow for further study of that specific region.
+Clicking on a region in small maps will bring up a new panel with the original
+census data of this specific region. Further, these selectors can be combined in
+different ways, enabling complex queries with instantaneous results.

### 2019-08-24 paper/related.tex
from overleaf

@@
 processing; \emph{data clustering and regionalisation}, briefly reviewing
-existing data clustering methods,\revision{ geographic constraints and
+existing data clustering methods,\revision{2.1}{ geographic constraints and
 regionalisation}; and \emph{cluster characterisation}, articulating how clusters
@@
 1910~\citep{nhgis}, most studies consider the period between 1970 and 2010,
-using pre-harmonised data from the Longitudinal Tract Data
-Base~\citep{Logan2014}. Despite its inherent
-errors~\citep{Logan2016,Hallisey2017}, this dataset has become \revision{widely
-adopted, along with the Neighborhood Change Database~\cite{ncdb},} as source for
-longitudinal demographic data at the neighbourhood scale, with similar efforts
-appearing in other countries~\citep{Liu2015,Lee2015,Allen2018}. These datasets
-have been highly significant for the field. Yet they also limit the universe of
-data that can be used to study neighbourhood change, since any new datasets
-would need to be similarly processed in order to be rendered compatible with
-these sources.
+\revision{1.3}{using pre-harmonised data from the Longitudinal Tract Data
+Base~\citep{Logan2014} or the Neighborhood Change Database~\citep{ncdb}, albeit
+their inherent errors~\citep{Logan2016,Hallisey2017}.} Similar harmonisation
+efforts appear in other countries~\citep{Liu2015,Lee2015,Allen2018}. These
+datasets have been highly significant for the field. Yet they also limit the
+universe of data that can be used to study neighbourhood change, since any new
+datasets would need to be similarly processed in order to be rendered compatible
+with these sources.
 
@@
 data~\citep{Dmowska2017,Dmowska2018,stepinski2019imperfect}. Beyond the
-potentially increased spatial precision, this approach does not require complex
-harmonisation when new data is considered, if the grids are compatible. However,
-demographic data is usually not available in this format, especially from older
-sources. Additionally, the conversion from tabulation areas can introduce
-significant errors.
+potentially increased spatial\revision{1.4}{ precision}, this approach does not
+require complex harmonisation when new data is considered, if the grids are
+compatible. However, demographic data is usually not available in this format,
+especially from older sources. Additionally, the conversion from tabulation
+areas can introduce significant errors.
 
@@
 interpreted in the same way. There is no shortage of contributions for this
-problem~\citep{Fahad2014}\revision{, since variations of it appear in almost all
+problem~\citep{Fahad2014}\revision{}{, since variations of it appear in almost all
 scientific fields.}
 
-\revision{In geography, this problem is known as
+\revision{2.1}{In geography, this problem is known as
 \emph{regionalisation}~\citep{montello2003regions}, a rather old problem that
-has been throughly explored, leveraging different mathematical tools, including
+has been thoroughly explored, leveraging different mathematical tools, including
 discrete topology~\citep{brantingham1978topological} and discrete
@@
 detection~\citep{blondel2008Louvain,Thomas2012}, or, to a lesser extent, Self
-Organising Maps~\citep{Delmelle2017,Ling2016,Dani2013_som_maxp}.
+Organising Maps~\citep{Delmelle2017,Ling2016,Dani2013_som_maxp}.}
 
-Since we adopted a network-based data representation and our objectives include
+\revision{1.5/2.2}{Since we adopted a network-based data representation and our objectives include
 an interactive interface, we opted for an heuristic variation of the maximum
@@
 fast computation times. This algorithm operates on weighted networks, where a
-distance metric is between the nodes is associated with the edges. We adopted a
+distance metric between the nodes is associated with the edges. We adopted a
 distance based on the data associated with each node. The algorithm merges
-clusters based on these distances, creating an hierarchy instead of a fixed
+clusters based on these distances, creating a hierarchy instead of a fixed
 number of clusters. This hierarchical result is rendered by the interface,
@@
 While visualisation has gained prominence as a crucial component of scientific
-discovery, justification, and communication~\cite{tufte1998visual}, visually
+discovery, justification, and communication~\citep{tufte1998visual}, visually
 representing evolving spatial data is a challenging old
@@
 approach is to use parallel coordinates plot~\citep{ferreira2015urbane}, but
-these they can get cluttered representing similar clusters over several
+they can get cluttered when representing similar clusters over several
 variables. Further, for demographic applications, the clusters are usually

### 2019-08-24 paper/study.tex
from overleaf

@@
 
-In this section we present two illustrative scenarios, usingq census data from
+In this section we present two illustrative scenarios, using census data from
 the United States~\citep{nhgis} and
@@
 
-\revision{Accidentaly, the US data for 2010 is actually not from the decennial
-census, but from the ACS 2006-2010. Further, the regions selected do not
-correspond to any pre-defined regions (metros, cities, census areas), but to
-arbitrary regions defined around a location of interest. We selected a minimal
-set of aspects for each country, and they are not similar to each other, out of
-convenience, since the variable matching was a manual process. These caveats
-would likely compromise any serious attempt on a comprehensive demographic
-study, but this is not the objective of this experiment. These results are meant
-to demonstrate the utility of the interface for understanding the evolutionary
-dynamics of urban neighbourhoods. They also show the face validity of the
-results generated by our novel network-based approach. Indeed, our results were
-perfectly aligned with several other studies, despite these methodological
-missteps, which could arguably be an indicator of its robustness.}
+\unsure{CHECK}\revision{1.12.(1-5)}{These results are meant to demonstrate the
+utility of the interface for understanding the evolutionary dynamics of urban
+neighbourhoods. They also show the face validity of the results generated by our
+novel network-based approach. However, the US data for 2010 is  not from the
+decennial census, but from the ACS 2006-2010, compromising the temporal
+stability of the data. Further, the regions selected do not correspond to any
+pre-defined regions (metros, cities, census areas), but to somewhat arbitrary
+regions defined around a location of interest, making direct comparisons harder.
+Since  variable matching is a  manual process, we selected a minimal set of
+aspects for each country, which are not necessarily similar to each other, or
+similar to studies in the current literature. These factors would seriously
+hinder any study using the existing methods, but our results were well-aligned
+with several other studies, illustrating the robustness of our approach to
+imperfect conditions. }
 
 \subsection{Chicago}
-Our first scenario examines Chicago, focusing on a region loosely following the
-City's administrative borders. Its demographic composition is well explored in
-the literature, with reports of racial divide and
+Our first scenario examines Chicago, using an arbitrary region larger than the
+administrative borders. Its demographic composition is well explored in the
+literature, with reports of racial divide and
 gentrification~\citep{Delmelle2016,Delmelle2017,Hwang2014}, so we expect our
@@
 reporting that Black neighbourhoods are less likely to
-gentrify~\citep{Hwang2014}. Moreover, this process appears to be unidirectional,
-as indicated by the limited number of trajectories leaving the purple stream.
-Next, we select the region that is gentrified in 2010, by clicking on the
-corresponding rectangle in the trajectories plot, updating the information on
-the maps and the details portion of the interface.
+gentrify~\citep{Hwang2014}. The transition matrix provides more details, showing
+that around one percent of all regions that became purple at any time
+transitioned from the green cluster (Black population), while nine percent were
+previouly associated with the orange cluster (White). Next, we select the region
+that is gentrified in 2010, by clicking on the corresponding rectangle in the
+trajectories plot, updating the information on the maps and the details portion
+of the interface.
 
@@
 Asian~\citep{Fong2003}. Toronto is also known for a stable and well defined
-Jewish community~\citep{Harold2018, Fong2011}. Therefore, we expect a
+Jewish community~\citep{Harold2018, Fong2011}. Therefore, we expect \revision{1.14}{a}
 combination of stable and dynamic regions on the results, with Place of Birth,
@@
 to the regions presented by previous studies considering
-Toronto~\citep{hulchanski2007three}. However, our interface provides a deeper insight
-into their demographic composition, since we consider more data than solely
-Average Income, which appears to be a good proxy variable nonetheless. This
-scenario showcases the ability of our method and interface to capture and
+Toronto~\citep{hulchanski2007three}. However, our interface provides a deeper
+insight into their demographic composition, since we consider more data than
+solely Average Income, which appears to be a good proxy variable nonetheless.
+This scenario showcases the ability of our method and interface to capture and
 understand the sources of urban volatility.
-
-% \subsection{Los Angeles}
-
-% We selected for this scenario a region around the metropolitan area of Los
-% Angeles (LA), following urban density. While Chicago and Toronto demonstrated
-% the abilities of our method to examine stable and dynamic cities, LA is considerably
-% larger, both in population and area. Since our method identifies only the eight
-% most distinct groups, we expect that some groups will not be identified,
-% especially if their characterisation is similar to another groups. Further, our
-% system does not include the census variables related to Hispanic heritage, which
-% were only included in the newer censuses. However, from the literature, we
-% expect to see some clusters where Race is an important aspect, including White,
-% Black, and Asian~\citep{Reibel2003}.
-
-% The summary of the results using all aspects with equal weights and eight
-% clusters is illustrated in Figure~\ref{fig:la}. The full statistical description
-% of the clusters is illustrated in Figure~\ref{fig:laAll}, where the most
-% relevant aspect of each cluster is highlighted. From the trajectories plot, we
-% can see that there is a large but shrinking cluster, depicted in green, one
-% increasing cluster in purple, an almost constant orange cluster, a smaller but
-% increasing pink cluster, and three other small clusters. The corresponding map
-% illustrates where these clusters are located, and that they are somewhat
-% geographically stable, with some movement between the green, orange, and purple
-% clusters.
-
-% \begin{figure}
-%     \centering 
-%     \includegraphics[width=0.975\linewidth]{LA.pdf}
-%     \caption{Result for Los Angeles with 8 clusters, including three small and
-%         ephemeral clusters. Cluster characterisation is displayed
-%         in Figure~\ref{fig:laAll}.\label{fig:la}}
-% \end{figure}
-
-% From Figure~\ref{fig:laAll}, we can see that the green cluster is characterised
-% by a high percentage of White population, low percentage of population in the
-% lowest income bracket, mostly administrative occupations, and about 30\% of the
-% population with four or more years of college. The orange cluster is
-% characterised by a high percentage of Black population, with few people in the
-% highest range of income and education. The purple cluster corresponds to a high
-% concentration of "Other" in race, which includes Hispanic for this dataset, high
-% concentration of Labourers, and low education and income. The pink cluster
-% contains a high percentage of Asian population and about 30\% of the population
-% with four or more years of college. The light green cluster contains very few
-% people in the lower income bracket, mostly White population, with the highest
-% percentage of population with four or more years of college, working
-% administrative jobs, and a high concentration of singles. The yellow cluster
-% represents Black population, with higher level of education and income, mostly
-% working administrative jobs. The brown cluster represent a majority of single
-% population, working administrative jobs with mostly low income. The dark grey
-% cluster is characterised by all its population in the lowest income bracket, low
-% education level, with a majority of White population. Since the extremes in the
-% boxplots of the grey cluster are not significantly different, we can also
-% surmise that this cluster is either small or homogeneous.
-
-% \begin{figure*}
-%     \centering 
-%     \includegraphics[width=0.95\linewidth]{laAll.pdf}
-%     \caption{Full characterisation of the eight clusters found for LA. The red
-%      rectangles indicate the most relevant aspects for each
-%      cluster.\label{fig:laAll}}
-% \end{figure*}
-
-% The  green, orange, and purple clusters present a significant intra-cluster
-% variance in most variables, as indicated by extreme whiskers of the boxplots.
-% While fifty percent of the CTs in the green cluster have between 20\% and 40\%
-% of people in the lowest income bracket, that cluster also includes CTs where
-% none and all the population belongs to that bracket. This might indicate that
-% this cluster represents different groups of people that are not different enough
-% to be separated at this level of the hierarchy. Conversely, the light green,
-% brown, and dark grey clusters are different enough to be separated into their
-% own clusters at this level, despite being small and ephemeral, including only a
-% few CTs.
-
-% The orange area in the map in Figure~\ref{fig:la} presents movement, indicated by
-% the presence of green and purple tones mixed with the orange, which may warrant
-% further exploration. By clicking on the orange bar above the boxplot, we select
-% all trajectories that contain the orange cluster. The corresponding details are
-% illustrated in Figure~\ref{fig:labh}. This shows a location change, where the
-% orange cluster is progressively replaced by the purple cluster on its east side,
-% and in turn expanding to the west. Interestingly, the population increased, the
-% racial profile changed, but the distribution of income was reasonably stable,
-% with a higher amount of the population in the lowest income range and very few
-% people in the highest income range. Indeed, the income  difference is
-% significant when compared to the city-wide distribution.
-
-% \begin{figure}
-%     \centering 
-%     \includegraphics[width=0.75\linewidth]{laBH.pdf}
-%     \caption{Top: Geographic changes in the majority Black population cluster
-%     (orange) and Labourers cluster (green). Bottom: Income evolution for this
-%     region (black) and the whole city (grey).\label{fig:labh}}
-% \end{figure}
-
-% A portion of this region is classified into the green cluster in 2010,
-% indicating a majority white population. To further understand that change, we
-% clear the current selection, and select all regions that changed from orange in
-% 1970 to green in 2010, using the transition matrix. A portion of the resulting
-% region, near the Florence-Graham region, is depicted in Figure~\ref{fig:laVol},
-% along with the temporal evolution of Race. Despite this difference, the other
-% aspects are similar to the ones from the region in Figure~\ref{fig:labh}, with
-% slightly lower income and education profiles. While the racial aspect changed
-% considerably, the economic and educational aspects stayed the same.
-
-% \begin{figure}
-%     \centering 
-%     \includegraphics[width=0.75\linewidth]{laVol.pdf}
-%     \caption{Details for a volatile region contained in the area of
-%          Figure~\ref{fig:labh}. This region went from Black to Hispanic to
-%         White.\label{fig:laVol}}
-% \end{figure}
-
-% These results are visually similar to the analysis presented by
-% Dellmelle~\citep{Delmelle2016}.  What was there identified as 'persistently
-% struggling' corresponds to the purple cluster in Figure~\ref{fig:la}; some
-% portions of the 'stable elite' cluster, near Inglewood, correspond to the high
-% income, high education yellow cluster. The other groups identified by that study
-% were clumped by our system into the green cluster, which would be further
-% divided if more than eight clusters were to be used.
-
-% If Toronto illustrates a relatively smooth process of whole-sale
-% diversification, Los Angeles shows the ability of our system and method to
-% uncover more intensive smaller-scale processes of neighbourhood change. This
-% capacity to examine and classify different types of change operating at
-% different scales demonstrates the flexibility of the method and tool for
-% advancing the more open-ended notion of "neighbourhood" envisioned by recent
-% neighbourhood research.
-
-% \smallskip While Toronto is more dynamic than Los Angeles, possibly due to size
-% differences, the volatile regions shown in Figure~\ref{fig:toVol} did not change
-% as quickly or dramatically as the ones shown in Figure~\ref{fig:laVol}, which
-% involved twice as many people. We found this trend to be related to the
-% countries themselves, Canadian cities have larger areas undergoing slow, gradual
-% changes, whereas American cities have more general stability, but quicker
-% changes in smaller scales. The supplementary material contains brief summaries
-% of all the regions accessible in this prototype.
 
@@
 current implementation can take a few minutes to cluster regions with high
-number of CTs, like Los Angeles or Brooklyn. Server-processing reduced the
+number of CTs, like Los Angeles or Brooklyn. Server-side processing reduced the
 amount of data transferred to client, but it might increase the response time
-under load. We implemented a caching policy that greatly improved the
-performance, but fully pre-processing the results is not practical due to size
-of the parameter space.
+under load. We implemented a caching policy, but fully pre-processing the
+results is not practical due to size of the parameter space.
 

### 2019-08-24 paper/intro.tex
almost done

@@
 \revision{1.1}{In some cases, harmonisation might seem methodologically
-adequate, if a study is focused on a small subregion, fixed in time, limiting
+adequate. If a study is focused on a small subregion that is fixed in time, limiting
 and modifying the data to fit that specific region of interest sounds not only
-reasonable, but recommended, albeit the creation of artificial data points.
-However, that is a simplistic view that ignores interactions beyond those
+reasonable, but recommended, even if it comes at the cost of producing artificial data points. However, this view is limited in that it ignores interactions beyond the fixed
 borders: each different portion of the area of interest may belong to larger
@@
 borders into another, our method sidesteps this issue by changing how the data
-is represented and interpreted.
+is represented and interpreted. It therefore allows researchers to examine changes in both the composition and extent of local areas.
 

### 2019-08-24 paper/method.tex
almost done

@@
 
+The main functionalities of the interface allow users to examine four
+inter-related questions central to understanding local area dynamics. 1)
+\emph{What} do the clusters consist in?; 2) \emph{Where} are they located; 3)
+\emph{How} do they change?; and 4) \emph{Who} occupy them? We discuss each in
+turn.
+
 The cluster overview panel, marked as \emph{what} in Figure~\ref{fig:ui},
@@
 also the corresponding IQRs for the whole city. Those values differ from the
-total population percentage, while the overall population of this region is
+total population percentage. While the overall population of this region is
 about sixty percent White, the average percentage of White population over the

### 2019-08-24 paper/study.tex
almost done

@@
 Occupation, and Race; and seven for Canada: Age, Education level, Home language,
-Household Income, Marital status, Occupation, Place of birth, and Religion.
-
-While our method does not require geographic harmonisation, it requires matching
-the variables over time. The supplementary material contains the details of
-which census columns were used for each aspect. Income is slightly inaccurate,
-even though we did correct for the official inflation. We grouped the original
-ranges into three larger ranges, but they do not match precisely.
+Household Income, Marital status, Occupation, Place of birth, and Religion. The supplementary material contains the details of
+which census columns were used for each aspect.
 
@@
 neighbourhoods. They also show the face validity of the results generated by our
-novel network-based approach. However, the US data for 2010 is  not from the
-decennial census, but from the ACS 2006-2010, compromising the temporal
+novel network-based approach. While our method does not require geographic harmonisation, it does however require matching
+the variables over time. This is (currently) a painstaking, manual process, and the current version of the interface remains somewhat limited. For example, we selected a minimal set of
+aspects for each country, which are not necessarily similar to each other, or
+similar to studies in the current literature. Income moreover is slightly inaccurate,
+even though we did correct for official inflation. We grouped the original
+ranges into three larger ranges, but they do not match precisely.
+In addition, the US data labelled "2010" is  not from the decennial census, but from the ACS 2006-2010, somewhat compromising the temporal
 stability of the data. Further, the regions selected do not correspond to any
@@
 regions defined around a location of interest, making direct comparisons harder.
-Since  variable matching is a  manual process, we selected a minimal set of
-aspects for each country, which are not necessarily similar to each other, or
-similar to studies in the current literature. These factors would seriously
-hinder any study using the existing methods, but our results were well-aligned
+Taken together, these factors would no doubt require caveats for 
+a substantive study using the current prototype version of the interface. Nevertheless, in the present context we find that the interface results are well-aligned
 with several other studies, illustrating the robustness of our approach to
@@
 The initial state of the prototype is illustrated in Figure~\ref{fig:ui}. The
-first step is to identify the compositions of each cluster from the boxplots, so
-orange is associated with majority of White population, green with majority
-Black, and purple with higher proportion of four years of college or more (high
-education level). The expanded version of the boxplots for the purple cluster
-shows a higher income level and majority of occupations in administrative jobs,
-therefore the purple cluster identifies gentrified regions.
+first step is to identify \textit{what} the clusters consist in. To do so, we
+examine the compositions of each cluster from the boxplots. Orange is associated
+with majority of White population, green with majority Black, and purple with
+higher proportion of four years of college or more (high education level). The
+expanded version of the boxplots for the purple cluster shows a higher income
+level and majority of occupations in administrative jobs. The purple cluster
+therefore could be described as gentrified regions.
 
-The trajectories plot illustrates the process of gentrification, also
-illustrated in Figure~\ref{fig:chiWorkflow}, progressively absorbing regions
-from the orange cluster (White). This corroborates results from the literature
-reporting that Black neighbourhoods are less likely to
+The trajectories plot shows \textit{how} clusters have changed over time.  For
+example, it illustrates the process of gentrification, also illustrated in
+Figure~\ref{fig:chiWorkflow}, with the purple cluster progressively absorbing
+regions from the orange cluster (White). This corroborates results from the
+literature reporting that Black neighbourhoods are less likely to
 gentrify~\citep{Hwang2014}. The transition matrix provides more details, showing
@@
 
-The corresponding regions are highlighted in the maps, where the spatial pattern
-is clear, corresponding exactly to previous findings in the literature based
-upon harmonisation~\citep{Hwang2014}. Further, we can also identify the regions
-that gentrified earlier on the small maps that depict the involved regions over
-time. Since the most relevant aspect is Education, specifically "Four or more
-years of college", we can expand the details of this aspect, as illustrated in
-the rightmost portion of Figure~\ref{fig:chiWorkflow}, which is increasing for
-the whole city (grey band), but faster and to a higher level in this region
-(black band).
+The maps show \textit{where} these processes are occurring, by highlighting the
+corresponding regions. The spatial pattern is clear, corresponding exactly to
+previous findings in the literature based upon harmonisation~\citep{Hwang2014}.
+Further, we can also identify the regions that gentrified earlier on the small
+maps that depict the involved regions over time. Finally, we can examine
+\textit{who} is driving these changes. Since the most relevant aspect is
+Education, specifically "Four or more years of college", we can expand the
+details of this aspect, as illustrated in the rightmost portion of
+Figure~\ref{fig:chiWorkflow}, which is increasing for the whole city (grey
+band), but faster and to a higher level in this region (black band).
+
+\unsure{CHECK - maybe add what/how/where/who to figure 5?}
 
@@
 
-The population with low percentage of University degrees is represented in
-orange, mostly anglophone population in green, Asian immigrants in yellow, high
-percentage of income in the highest bracket in purple, high percentage of Jewish
-people in light green and brown, high percentage of Eastern Non-Christian
-religion in pink, and high concentration of single people in dark grey. From the
-trajectories plot, we can see that Toronto is more dynamic than Chicago, with
-one cluster constantly shrinking. In the 1970s, the city was divided into four
-clusters: low number of university degrees, Jewish population, majority
-anglophones, and high income. Interestingly, the more recent clusters that
-absorbed regions from the orange cluster have similar education profiles and are
-differentiated by other aspects. In this sense, the city is growing diverse,
-changing from a common low education profile to a higher level of education with
-more diversity in religion (pink) and immigration (yellow).
+We again are guided by the four questions embedded in the interface: what, how,
+where, and who. The boxplots address what clusters defined Toronto over this
+period. The population with low percentage of University degrees is represented
+in orange, mostly anglophone population in green, Asian immigrants in yellow,
+high percentage of income in the highest bracket in purple, high percentage of
+Jewish people in light green and brown, high percentage of Eastern Non-Christian
+religion in pink, and high concentration of single people in dark grey.
 
-Indeed, the growing Asian population is visible starting in the 1980s and
-building thereafter, leading to the yellow and pink clusters. While both include
-a high percentage of people born in Asia, the pink is more defined by religion,
-with low percentage of university degrees, and contains the lowest percentage of
-people in the highest income bracket for these clusters; the yellow is less
-defined by religion, and has higher education and income, geographically
-corresponding to the Markham region, know for its Chinese population. A similar
-division also happens for the two Jewish clusters, where the light green cluster
-has lower education and income levels than the brown cluster. The purple cluster
-of high income is somewhat stable. Until 2011 the cluster included the Bridle
-Path neighbourhood, known for its wealthy population. In 2011 it was classified
-into the yellow cluster of Asian immigration, since about 40\% of the population
-for this CT were then born in Asia. The income distribution did not change, with
-85\% of the population with an income of 90k CAD or more.
+From the trajectories plot, we can see how the city changed. It shows that
+Toronto is more dynamic than Chicago, with one cluster constantly shrinking. In
+the 1970s, the city was divided into four clusters: low number of university
+degrees, Jewish population, majority anglophones, and high income.
+Interestingly, the more recent clusters that absorbed regions from the orange
+cluster have similar education profiles and are differentiated by other aspects.
+In this sense, the city is growing diverse, changing from a common low education
+profile to a higher level of education with more diversity in religion (pink)
+and immigration (yellow). Indeed, the growing Asian population is visible
+starting in the 1980s and building thereafter, leading to the yellow and pink
+clusters. While both include a high percentage of people born in Asia, the pink
+is more defined by religion, with low percentage of university degrees, and
+contains the lowest percentage of people in the highest income bracket for these
+clusters; the yellow is less defined by religion, and has higher education and
+income, geographically corresponding to the Markham region, know for its Chinese
+population. A similar division also happens for the two Jewish clusters, where
+the light green cluster has lower education and income levels than the brown
+cluster. The purple cluster of high income is somewhat stable. Until 2011 the
+cluster included the Bridle Path neighbourhood, known for its wealthy
+population. In 2011 it was classified into the yellow cluster of Asian
+immigration, since about 40\% of the population for this CT were then born in
+Asia. The income distribution did not change, with 85\% of the population with
+an income of 90k CAD or more.
 
-The most significant indicator of Toronto's dynamism is the presence of grey
-regions on the map. These represent regions associated to three or more clusters
-over this five census period. Using the 'Add' mode for the trajectory selection,
-we select their trajectories, and a subset of the details is illustrated in
+Examining where and by whom these changes occurred reveals additional
+information about the city's underlying dynamics. The most significant indicator
+of Toronto's dynamism is the presence of grey regions on the map. These
+represent regions associated to three or more clusters over this five census
+period. Using the 'Add' mode for the trajectory selection, we select their
+trajectories, and a subset of the details is illustrated in
 Figure~\ref{fig:toVol}. These regions account for about 5\% of Toronto's
@@
 trend in increasing education is city-wide, but this region has people with
-better education than most.
+better education than most. In 2001, the purple cluster of high income annexes
+neighbouring parts of the volatile region, and the Asian born population
+increases sharply, as illustrated by the appearance of the yellow cluster.  This
+cluster indicates well educated, higher income, and about 30\%-50\% Asian born
+population. By 2011, the yellow cluster increased considerably, annexing parts
+of the high income purple cluster, including the neighbouring Bridle Path area.
 
@@
 \end{figure}
-
-In 2001, the purple cluster of high income annexes neighbouring parts of the
-volatile region, and the Asian born population increases sharply, as illustrated
-by the appearance of the yellow cluster.  This cluster indicates well educated,
-higher income, and about 30\%-50\% Asian born population. By 2011, the yellow
-cluster increased considerably, annexing parts of the high income purple
-cluster, including the neighbouring Bridle Path area.
 

### 2019-08-27 backend/srv.py
year/geom selection map working

@@ -30,7 +30,7 @@ from upload import gatherInfoJsons, processUploadFolder
 cachedir = './cache/'
 if not exists(cachedir):
     makedirs(cachedir)
-memory = Memory(cachedir, verbose=0)
+memory = Memory(cachedir, verbose=20000)
 
 # same as in clustering.py! - There is probably a better way...
 NBINS = 100
@@ -90,8 +90,7 @@ def cors():
         cherrypy.response.headers['Access-Control-Allow-Origin'] = '*'
 
 
-@memory.cache(ignore=['ds'])
-# @profile
+@memory.cache(ignore=['ds',])
 def _mapHiers(ds: dataStore, aspects: list, nClusters: int = 10, threshold: float = 0.75, bbox: list = None):
 
     Gs = mapHierarchies(ds, aspects, bbox=bbox)
@@ -445,7 +444,9 @@ def _mapHiers(ds: dataStore, aspects: list, nClusters: int = 10, threshold: floa
         for g in cl[y]:
             maxCC = max([maxCC, max(cl[y][g].values())])
 
-    return({'clustering': {**(cl[1970]), **(cl[2010]), **(cl[2015])},
+    # print(cl.keys())
+
+    return({'clustering': dict(cl), #{**(cl[1970]), **(cl[2010]), **(cl[2015])}
             'evolution': evo,
             'hist': {'aspect': aspect_hist, 'cc': cc_hist},
             'aspects': full_info_aspects,

### 2019-08-27 ui/src/glmap.js

@@ -15,15 +15,22 @@ const mapStateToProps = (state) => ({
 mapboxgl.accessToken = 'pk.eyJ1IjoiZGlhc2YiLCJhIjoiY2pqc243cW9wNDN6NTNxcm1jYW1jajY2NyJ9.2C0deXZ03NJyH2f51ui4Jg';
 
 class GeomControl {
-  constructor(geoms,callback){
+  constructor(clustering, callback){
     this.callbackfcn=callback;
-    this.geoms=geoms;
+    this.options=[];
+    let years=Object.keys(clustering);
+    for (let iy=0;iy<years.length;iy++){
+      let curGeoms=Object.keys(clustering[years[iy]]);
+      for (let ig=0;ig<curGeoms.length;ig++){
+        this.options.push({id:this.options.length,name:years[iy]+'-'+curGeoms[ig],year:years[iy],geom:curGeoms[ig]});
+      }
+    }
+
   }
   onAdd(map){
     this.map = map;
 
     this.container = document.createElement('div');
-    // this.container.className = 'geom-control';
     this.container.className = 'mapboxgl-ctrl';
 
 
@@ -32,14 +39,14 @@ class GeomControl {
     sel.onchange=this.callbackfcn;
     // document.body.appendChild(x);
 
-    for (let i=0;i<this.geoms.length;i++){
+    for (let i=0;i<this.options.length;i++){
       var op = document.createElement("option");
-      op.setAttribute("value", this.geoms[i].name);
-      var text = document.createTextNode(this.geoms[i].name);
+      op.setAttribute("value", this.options[i].id);
+      var text = document.createTextNode(this.options[i].name);
       op.appendChild(text);
-      sel.appendChild(op)  
+      sel.appendChild(op);  
     }
-    this.container.appendChild(sel)
+    this.container.appendChild(sel);
     // document.getElementById("mySelect").appendChild(z);
   
     // this.container.textContent = 'My custom control';
@@ -84,7 +91,7 @@ let MapboxMap = class MapboxMap extends React.Component {
   map;
   constructor(props){
     super(props);
-    this.state={loaded:false, selected:undefined, bbox:undefined}
+    this.state={loaded:false, selected:undefined, bbox:undefined};
   }
   
   componentDidUpdate(props,state) {
@@ -94,20 +101,36 @@ let MapboxMap = class MapboxMap extends React.Component {
     let {dispatch}=this.props;
     
[…]

### 2019-08-27 ui/src/reducers.js

@@ -27,7 +27,7 @@ export const actionCreators = {
     }
   };
 
-export const reducer = (state={aspects:[], selectedPaths:[], loading:true}, action)=>{
+export const reducer = (state={aspects:[], selectedPaths:[], years:[], loading:true}, action)=>{
     const { type, payload } = action;
     console.log(state, type, payload);
     switch (type){

### 2019-08-27 ui/src/glmap.js
dropdown working

@@ -76,7 +76,7 @@ class UpdateButton {
     btn.setAttribute("id", "myButton");
     btn.innerHTML="Update";
     btn.onclick=this.cb;
-    this.container.appendChild(btn)
+    this.container.appendChild(btn);
     this.container.onclick = this.cb;
     return(this.container);
   }
@@ -101,23 +101,20 @@ let MapboxMap = class MapboxMap extends React.Component {
     let {dispatch}=this.props;
     
     if ((cmaps!==undefined)&&(geometries!==undefined)&&(geometries.length>0)){
-      if (this.menu!==undefined){
-        this.map.removeControl(this.menu);
-      }
-      let ygControl = new GeomControl(cmaps, //this.props.geometries,
-          (d)=>{
-            this.setState({selected:{year:this.menu.options[d.target.value].year, geom:this.menu.options[d.target.value].geom}});
-          }
-        );
-      this.menu=ygControl;
-      this.map.addControl(ygControl);    
-      
+      if (this.map.menu===undefined){
+        let ygControl = new GeomControl(cmaps, //this.props.geometries,
+            (d)=>{
+              this.setState({selected:this.map.menu.options[d.target.value]});
+            }
+          );
+        this.map.menu=ygControl;
+        this.map.addControl(ygControl);    
+      }      
 
       
       if (selected===undefined){
-        this.setState({selected:{year:this.menu.options[0].year, geom:this.menu.options[0].geom}});
+        this.setState({selected:{year:this.map.menu.options[0].year, geom:this.map.menu.options[0].geom}});
       } else {
-        console.log((cmaps[selected.year]).hasOwnProperty(selected.geom));
         if ((cmaps.hasOwnProperty(selected.year)) && ((cmaps[selected.year]).hasOwnProperty(selected.geom))){
 
           if (
@@ -218,6 +215,7 @@ let MapboxMap = class MapboxMap extends React.Component {
   }
 
   setFill(colours){
+    console.log('setfill',this.state.selected);
     let exp=['case',
               [
                 'has',

### 2019-08-28 backend/dataStore.py
removing kmeans

@@ -230,7 +230,7 @@ class dataStore(object):
                 G.node[n]['data'] = np.empty((ncols,))
                 G.node[n]['data'][:] = np.nan
 
-        G = ComputeClustering(G, 'data')
+        G = ComputeClustering(G, 'data', k=2)
         nx.write_gpickle(G, join(self._data_folder, aspect+'.gp'))
         # self._update_distances()
 

### 2019-08-28 backend/srv.py

@@ -20,7 +20,7 @@ from joblib import Memory
 from networkx.readwrite import json_graph
 from scipy.spatial.distance import cosine, euclidean
 # from scipy.stats import wasserstein_distance,
-from sklearn.cluster import KMeans
+# from sklearn.cluster import KMeans
 from tqdm import tqdm
 
 from dataStore import dataStore
@@ -60,7 +60,7 @@ def _mergePaths(p1: dict, p2: dict, id: int):
             ret[k] = (p1[k]+p2[k])/2
     return(ret)
 
-
+# {'_sw': {'lng': -97.36262426260146, 'lat': 24.36091074100848}, '_ne': {'lng': -65.39166177011971, 'lat': 33.61501866716327}}
 def _bbox_create_buffer(bbox: dict) -> list:
     minX = bbox['_sw']['lng']
     minY = bbox['_sw']['lat']
@@ -136,47 +136,47 @@ def _mapHiers(ds: dataStore, aspects: list, nClusters: int = 10, threshold: floa
         # # # -------------------------------------------
         # # # Merging disconnected similar clusters
 
-        aspects_in_this_year_geom = [a 
-                                for a in full_info_aspects 
-                                if (a['geom'] == g) and (a['year'] == y)]
-        nDims = [ds.getDimension(a['id']) for a in aspects_in_this_year_geom]
-        singleVar = [x == 1 for x in nDims]
-        # fixes the histogram size for scalar variables
-        nDims = [x if x > 1 else NBINS for x in nDims]
-        M = np.zeros((current_number_ccs, np.sum(nDims)))
-        for i, aspect in enumerate(aspects_in_this_year_geom):
-            a = aspect['id']
-            y = aspect['year']
+        # aspects_in_this_year_geom = [a 
+        #                         for a in full_info_aspects 
+        #                         if (a['geom'] == g) and (a['year'] == y)]
+        # nDims = [ds.getDimension(a['id']) for a in aspects_in_this_year_geom]
+        # singleVar = [x == 1 for x in nDims]
+        # # fixes the histogram size for scalar variables
+        # nDims = [x if x > 1 else NBINS for x in nDims]
+        # M = np.zeros((current_number_ccs, np.sum(nDims)))
+        # for i, aspect in enumerate(aspects_in_this_year_geom):
+        #     a = aspect['id']
+        #     y = aspect['year']
 
-            if i == 0:
-                start = 0
-            else:
-                start = sum(nDims[:i])
[…]

### 2019-08-30 paper/intro.tex
getting ready to send

@@
 
-\unsure{CHECK blue text below}
 \revision{1.1}{In some cases, harmonisation might seem methodologically
@@
 
-To allow a proper examination of our method and its results, we built an online
+To allow a proper examination of our method and its results, we built a prototype online
 interactive system using this representation. It enables users to visualise,
-interpret, and explore trajectories of neighbourhood change. This interface
+interpret, and explore trajectories of neighbourhood change. This prototype
 helps validate our method, by allowing it to be compared to existing and future

### 2019-08-30 paper/method.tex
getting ready to send

@@
         \textbf{(where)}: Maps illustrating the geographical location, both as an overall summary and for each year.
-        \textbf{(how)}: Tabular and visual summmarisation of how the regions classification changed over time.
+        \textbf{(how)}: Tabular and visual summary of how the regions classification changed over time.
         \textbf{(who)}: Demographic details for the currently selected data.
@@
 
-As illustrated by Figure~\ref{fig:ui}\unsure{CHECK image/ sec 4.5}, our proposed
-interface heavily relies on colour to express cluster-related information. We
-adopted this convention because colours can be used in all our visual tools in a
-coherent manner. However, there is a limit on the number of distinct colours
-that can be used. We limited the number of clusters to eight because this was
-the largest number of colours that we could reliably and accessibly use, derived
-from the 8-class Dark2 set from ColorBrewer~\citep{ColorBrewer}.
+As illustrated by Figure~\ref{fig:ui}, our proposed interface heavily
+relies on colour to express cluster-related information. We adopted
+this convention because colours can be used in all our visual tools in
+a coherent manner. However, there is a limit on the number of distinct
+colours that can be used. We limited the number of clusters to eight
+because this was the largest number of colours that we could reliably
+and accessibly use, derived from the 8-class Dark2 set from
+ColorBrewer~\citep{ColorBrewer}.
 

### 2019-08-30 paper/study.tex
getting ready to send

@@
 
-The prototype interface allows access to 41 regions, 29
-in the US and 12 in Canada. New York City was split into its boroughs to avoid
-memory crashes on the client browser due to the high number of CTs.  We used
-five aspects for the USA: Education level, Family income, Marital status,
-Occupation, and Race; and seven for Canada: Age, Education level, Home language,
-Household Income, Marital status, Occupation, Place of birth, and Religion. The supplementary material contains the details of
-which census columns were used for each aspect.
+The prototype interface allows access to 41 regions, 29 in the US and
+12 in Canada. New York City was split into its boroughs to avoid
+memory crashes on the client browser due to the high number of CTs.
+We used five aspects for the USA: Education level, Family income,
+Marital status, Occupation, and Race; and seven for Canada: Age,
+Education level, Home language, Household Income, Marital status,
+Occupation, Place of birth, and Religion. The supplementary material
+contains the details of which census columns were used for each
+aspect.
 
-\unsure{CHECK}\revision{1.12.(1-5)}{These results are meant to demonstrate the
-utility of the interface for understanding the evolutionary dynamics of urban
-neighbourhoods. They also show the face validity of the results generated by our
-novel network-based approach. While our method does not require geographic harmonisation, it does however require matching
-the variables over time. This is (currently) a painstaking, manual process, and the current version of the interface remains somewhat limited. For example, we selected a minimal set of
-aspects for each country, which are not necessarily similar to each other, or
-similar to studies in the current literature. Income moreover is slightly inaccurate,
-even though we did correct for official inflation. We grouped the original
-ranges into three larger ranges, but they do not match precisely.
-In addition, the US data labelled "2010" is  not from the decennial census, but from the ACS 2006-2010, somewhat compromising the temporal
-stability of the data. Further, the regions selected do not correspond to any
-pre-defined regions (metros, cities, census areas), but to somewhat arbitrary
-regions defined around a location of interest, making direct comparisons harder.
-Taken together, these factors would no doubt require caveats for 
-a substantive study using the current prototype version of the interface. Nevertheless, in the present context we find that the interface results are well-aligned
-with several other studies, illustrating the robustness of our approach to
-imperfect conditions. }
+\revision{1.12.(1-5)}{These results are meant to demonstrate the
+  utility of the interface for understanding the evolutionary dynamics
+  of urban neighbourhoods. They also show the face validity of the
+  results generated by our novel network-based approach. While our
+  method does not require geographic harmonisation, it does however
+  require matching the variables over time. This is (currently) a
+  painstaking, manual process, and the current version of the
+  interface remains somewhat limited. For example, we selected a
+  minimal set of aspects for each country, which are not necessarily
+  similar to each other, or similar to studies in the current
+  literature. Income moreover is slightly inaccurate, even though we
+  did correct for official inflation. We grouped the original ranges
+  into three larger ranges, but they do not match precisely. In
+  addition, the US data labelled "2010" is not from the decennial
+  census, but from the ACS 2006-2010, somewhat compromising the
+  temporal stability of the data. Further, the regions selected do not
+  correspond to any pre-defined regions (metros, cities, census
+  areas), but to somewhat arbitrary regions defined around a location
+  of interest, making direct comparisons harder.  Taken together,
+  these factors would no doubt require caveats for a substantive study
+  using the current prototype version of the interface. Nevertheless,
+  in the present context we find that the interface results are
+  well-aligned with several other studies, illustrating the robustness
+  of our approach to imperfect conditions. }
 
@@
 transitioned from the green cluster (Black population), while nine percent were
-previouly associated with the orange cluster (White). Next, we select the region
+previously associated with the orange cluster (White). Next, we select the region
 that is gentrified in 2010, by clicking on the corresponding rectangle in the
@@
 band), but faster and to a higher level in this region (black band).
-
-\unsure{CHECK - maybe add what/how/where/who to figure 5?}
 

### 2019-10-10 paper/intro.tex
last paper review piccard 1.0 GEAN

@@
 involves interpolating existing measurements into a common set of
-regions~\citep{Logan2014,Hallisey2017,Allen2018}. \revision{1.18}{Recent
+regions~\citep{Logan2014,Hallisey2017,Allen2018}. Recent
 computational tools have somewhat simplified this
-process~\citep{rey2018spatio}}, but it still involves non-trivial questions:
+process~\citep{rey2018spatio}, but it still involves non-trivial questions:
 which geometry to use as target, how to apportion the variables, or how to
@@
 
-\revision{1.1}{In some cases, harmonisation might seem methodologically
-adequate. If a study is focused on a small subregion that is fixed in time, limiting
-and modifying the data to fit that specific region of interest sounds not only
-reasonable, but recommended, even if it comes at the cost of producing artificial data points. However, this view is limited in that it ignores interactions beyond the fixed
+In some cases, harmonisation might seem methodologically adequate. If a study is
+focused on a small subregion that is fixed in time, limiting and modifying the
+data to fit that specific region of interest sounds not only reasonable, but
+recommended, even if it comes at the cost of producing artificial data points.
+However, this view is limited in that it ignores interactions beyond the fixed
 borders: each different portion of the area of interest may belong to larger
@@
 borders into another, our method sidesteps this issue by changing how the data
-is represented and interpreted. It therefore allows researchers to examine changes in both the composition and extent of local areas.
+is represented and interpreted. It therefore allows researchers to examine
+changes in both the composition and extent of local areas.
 
@@
 leads to a better understanding of its dynamics by exploring the related
-data-driven homogeneous regions. }
+data-driven homogeneous regions.
 
@@
 
-\revision{2.6.4}{While didactic, this example is not realistic because each
-building is visible in Figure~\ref{fig:intuition}, so we know each individual of
-the tabulation area, which is never the case. Further, real measurements are
-seldom as homogeneous and noise-free as this artificial example.  In this
-example, one possible harmonisation would split region A  in two to provide
-different ancestors to C ($A_C$) and D ($A_D$). Then $A_D$ would not be
-considered when analysing the evolution of C, nor $A_C$ for $D$. This is
-inaccurate, because $A$ was the original state for both evolutions, and we
-should not assume that these processes are disconnected.} Either way, by
-splitting and merging the data to fit arbitrary borders, the harmonisation
-process increases the distance between the data and reality by creating
-artificial data points.
+While didactic, this example is not realistic because each building is visible
+in Figure~\ref{fig:intuition}, so we know each individual of the tabulation
+area, which is never the case. Further, real measurements are seldom as
+homogeneous and noise-free as this artificial example.  In this example, one
+possible harmonisation would split region A  in two to provide different
+ancestors to C ($A_C$) and D ($A_D$). Then $A_D$ would not be considered when
+analysing the evolution of C, nor $A_C$ for $D$. This is inaccurate, because $A$
+was the original state for both evolutions, and we should not assume that these
+processes are disconnected. Either way, by splitting and merging the data to
+fit arbitrary borders, the harmonisation process increases the distance between
+the data and reality by creating artificial data points.
 
@@
 each other (edges). In this case, each different aggregation area is represented
-as a node and we connect nodes that have overlapping geographical areas\revision{1.2}{ in
-different times or are neighbours in the same time}, leading to the network
+as a node and we connect nodes that have overlapping geographical areas in
+different times or are neighbours in the same time, leading to the network
 illustrated on the right of Figure~\ref{fig:intuition}. By partitioning the

### 2019-10-10 paper/method.tex
last paper review piccard 1.0 GEAN

@@
     \centering 
-    \includegraphics[width=0.975\linewidth]{overview.pdf}
+    \includegraphics[width=0.95\linewidth]{overview.pdf}
     \caption{Overview of the proposed method. 
@@
 year. Further, edges are placed between nodes if the corresponding regions
-belong to sequential years and \revision{1.6/2.6.3}{their geometries intersect.
+belong to sequential years and their geometries intersect.
 To avoid spurious connections caused by small fluctuations, one of the
-geometries is slightly shrunk before the intersection, using a buffer of -1e6.
-While the result would not change considerably by adding more connections
-between nearby regions, these extra connections would greatly increase the
-computational cost. More importantly, while weights will be associated with
-these edges before they are processed, they are not derived from the geometry,
-but from the data. The actual intersection area is not considered in this
-representation. } This approach leads to a single network representing the whole
-spatio-temporal space of the data.
+geometries is slightly shrunk before the intersection, using a buffer of -1e-6
+degrees. While the result would not change considerably by adding more
+connections between nearby regions, these extra connections would greatly
+increase the computational cost. More importantly, while weights will be
+associated with these edges before they are processed, they are not derived from
+the geometry, but from the data. The actual intersection area is not considered
+in this representation.  This approach leads to a single network representing
+the whole spatio-temporal space of the data.
 
@@
 Having tied the regions together into a network, we can now partition it to
-identify similar sets of regions. \revision{2.3}{In essence, we are performing
+identify similar sets of regions. In essence, we are performing
 temporal regionalisation by applying a clustering method over this
-spatio-temporal network.}
+spatio-temporal network.
 
@@
 outliers. We refer the reader to the original paper~\citep{markus2017} for more
-details, including a complete performance evaluation using several metrics.
-\revision{1.5/2.2}{We chose this algorithm because it is fast,
-simple, and easily customisable, but our methodology should work with any
-hierarchical network clustering algorithm. However, we caution against the use
-of single-linkage methods due to their tendency to form chain-like clusters with
-elements that are similar pairwise, but significantly different on the ends.
-That is also the reason we did not use a Watershed hierarchy
-directly~\citep{najman2011equivalence}.}
+details, including a complete performance evaluation using several metrics. We
+chose this algorithm because it is fast, simple, and easily customisable, but
+our methodology should work with any hierarchical network clustering algorithm.
+However, we caution against the use of single-linkage methods due to their
+tendency to form chain-like clusters with elements that are similar pairwise,
+but significantly different on the ends. That is also the reason we did not use
+a Watershed hierarchy directly~\citep{najman2011equivalence}.
 
@@
 These edges connect nodes with similar content, if they are not already
-connected, providing a path for the algorithm to group similar nodes.
-\revision{2.4}{The regions connected by those edges will be merged on the first
-stages of the clustering, since the nodes they connect are, by definition, as
-similar as possible\footnote{This assertion relies on the euclidean distance and
-its relationship with the space where k-nn operates.}, leaving the remaining
-steps of the hierarchy to be determined only by the geographical edges. We
+connected, providing a path for the algorithm to group similar nodes. The
+regions connected by those edges will be merged on the first stages of the
+clustering, since the nodes they connect are, by definition, as similar as
+possible\footnote{This assertion relies on the euclidean distance and its
+relationship with the space where k-nn operates.}, leaving the remaining steps
+of the hierarchy to be determined only by the geographical edges. We
 experimented with different numbers of augmentation edges, but the results were
@@
 geographical data analysis, this idea potentially warrants further exploration,
-beyond the scope of this work.}
+beyond the scope of this work.
 
@@
 \emph{relevance} of a variable of an aspect based on the distance between the
-\revision{1.8}{interquartile} ranges (IQR) of the clusters in the same hierarchical level. If
+interquartile ranges (IQR) of the clusters in the same hierarchical level. If
 the IQRs overlap for all clusters, that variable is not relevant to the
@@
 evolution of these groups. To examine this process of evolution directly, we
-introduce the concepts\revision{2.6.4}{ of \emph{temporal paths} and \emph{trajectories}.
+introduce the concepts of \emph{temporal paths} and \emph{trajectories}.
 
-We call a temporal path any sequence of nodes in our representation network such
-that the temporal information associated with the nodes only increases. For
-instance, in Figure~\ref{fig:intuition}, the sequences ACF, ADF, ADG, and BEH
-are temporal paths. With harmonised data, as illustrated by the path BEF, the
-time-series of each region would form a temporal path, each node would be
-connected only to its older and newer versions, and would belong to only one
-temporal path. Since our data is not harmonised, more connections are allowed
-and each node can belong to an arbitrary number of paths.
+From network theory, a \emph{path} is a sequence of nodes of the network that
+are connected pairwise by edges. As each node has associated temporal
+information, we call \emph{temporal} any path such that the temporal information
+increases along it. For instance, in Figure~\ref{fig:intuition}, the paths ACF,
+ADF, ADG, and BEH are temporal paths, but not HEB, CDE, nor FDA. With harmonised
+data, as illustrated by the path BEH, the time-series of each region would form
+a temporal path, each node would be connected only to its older and newer
+versions, and would belong to only one temporal path. Since our data is not
+harmonised, more connections are allowed and each node can belong to an
+arbitrary number of paths.
 
@@
 Regions on the same trajectory had the same sequence of clusters, therefore had
-similar temporal evolution.}
+similar temporal evolution.
 
@@
     \centering 
-    \includegraphics[width=0.975\linewidth]{ui.pdf}
+    \includegraphics[width=0.95\linewidth]{ui.pdf}
     \caption{Initial interface of our method showing the demographic evolution of Chicago and identifying the objectives of each plot.
@@
 displays the most relevant aspects for each cluster, based on the distance
-between the IQRs, as detailed in Section~\ref{sec:relevance}.\revision{2.5}{This
+between the IQRs, as detailed in Section~\ref{sec:relevance}. This
 feature allows the user to quickly understand what exists in this region and
@@
 while the chevron at the side lets the user expand each cluster separately,
-allowing the exploration of more details on demand.}
+allowing the exploration of more details on demand.
 
@@
     \centering 
-    \includegraphics[width=0.45\linewidth]{boxplot.pdf}
+    \includegraphics[width=0.425\linewidth]{boxplot.pdf}
     \caption{Enhanced boxplot of the clusters' characteristics allows a quick
@@
 time, the larger map shows a summarised version, based on the trajectories of
-each region. \revision{1.9}{In simplified mode the colour corresponds to the
+each region. In simplified mode the colour corresponds to the
 cluster present in that trajectory for more than half the temporal samples, and
@@
 of the colours of the clusters in that trajectory in RGB colour
-space~\footnote{\url{https://gka.github.io/chroma.js/\#chroma-average}}.}
+space~\footnote{\url{https://gka.github.io/chroma.js/\#chroma-average}}.
 
-\revision{2.5}{Each trajectory represents a different temporal path through the
-clusters. Therefore the trajectories encode how the demographic characteristics
-changed over time. These changes are summarised by the two sections identified
-as \emph{how} in Figure~\ref{fig:ui}. In the left panel, there is a Sankey
-diagram where each ribbon represents a different trajectory. The widths are
-proportional to the population involved.} In our example in Figure~\ref{fig:ui},
-the orange and green clusters contain most of the population and are fairly
-stable over time. The pink cluster is small and mostly stable. The purple
-cluster is increasing, mostly by incorporating areas that were previously
-orange. Since the purple group corresponds to the emergent 'young urban' group,
-this corroborates the findings of Delmelle~\citep{Delmelle2016,Delmelle2017},
-showing that our network-based method can recover results from the traditional
-data processing approach.
+Each trajectory represents a different temporal path through the clusters.
+Therefore the trajectories encode how the demographic characteristics changed
+over time. These changes are summarised by the two sections identified as
+\emph{how} in Figure~\ref{fig:ui}. In the left panel, there is a Sankey diagram
+where each ribbon represents a different trajectory. Since the associated
+regions have different population numbers, the width of each ribbon is
+proportional to the smaller of the populations involved, pairwise. This
+approximation works well in general, but leads to small inconsistencies in some
+specific cases. For instance, in the results for Miami, which is available in
+the supplementary material, the combined width of the ribbons decreases
+slightly, despite an increase in the population numbers.
 
-\revision{2.5}{In the next panel, illustrated in the top middle of
-Figure~\ref{fig:ui}, is a transition matrix between the clusters. It indicates a
-rounded percentage of regions whose area changed between each pair of
-clusters. In this example, we can see that all clusters are fairly stable, with
-more than ninety percent of identity transitions.} This kind of table can be
-found in the related literature~\citep{Delmelle2016}, so it is familiar to the
-advanced users. It not only informs the proportional changes, but allows the
-selection of specific changes for further analysis.
+In our example in Figure~\ref{fig:ui}, the orange and green clusters contain
+most of the population and are fairly stable over time. The pink cluster is
+small and mostly stable. The purple cluster is increasing, mostly by
+incorporating areas that were previously orange. Since the purple group
+corresponds to the emergent 'young urban' group, this corroborates the findings
+of Delmelle~\citep{Delmelle2016,Delmelle2017}, showing that our network-based
+method can recover results from the traditional data processing approach.
+
+In the next panel, illustrated in the top middle of Figure~\ref{fig:ui}, is a
+transition matrix between the clusters. It indicates a rounded percentage of
+regions whose area changed between each pair of clusters. In this example, we
+can see that all clusters are fairly stable, with more than ninety percent of
+identity transitions. This kind of table can be found in the related
+literature~\citep{Delmelle2016}, so it is familiar to the advanced users. It not
+only informs the proportional changes, but allows the selection of specific
+changes for further analysis.
 
@@
 represented by a stacked bar plot, where the width of each rectangle corresponds
-to \revision{1.10}{the percentage of population in that variable over the
-considered period. In this case, a little less than half of the population are
-married, and the percentage that are Widowers or Divorced is roughly similar.
-About half of the population work in Administrative jobs, a third never
-completed high-school, approximately forty percent have gross family income
-below 50,000USD per year. About sixty percent identify as white. Placing the
-mouse over one of the bars will open a small panel with the temporal evolution
-of the population percentage of that specific variable, and clicking on the
-chevron on the right side expands the corresponding aspect, showing census tract
-level statistics, with details of the temporal evolution of each variable and
-also the corresponding IQRs for the whole city. Those values differ from the
-total population percentage. While the overall population of this region is
-about sixty percent White, the average percentage of White population over the
-CTs is a little over forty percent.}
+to the percentage of population in that variable over the considered period. In
+this case, a little less than half of the population are married, and the
+percentage that are Widowers or Divorced is roughly similar. About half of the
+population work in Administrative jobs, a third never completed high-school,
+approximately forty percent have gross family income below 50,000USD per year.
+About sixty percent identify as white. Placing the mouse over one of the bars
+will open a small panel with the temporal evolution of the population percentage
+of that specific variable, and clicking on the chevron on the right side expands
+the corresponding aspect, showing census tract level statistics, with details of
+the temporal evolution of each variable and also the corresponding IQRs for the
+whole city. Those values differ from the total population percentage. While the
+overall population of this region is about sixty percent White, the average
+percentage of White population over the CTs is a little over forty percent.
 

### 2019-10-10 paper/related.tex
last paper review piccard 1.0 GEAN

@@
 processing; \emph{data clustering and regionalisation}, briefly reviewing
-existing data clustering methods,\revision{2.1}{ geographic constraints and
-regionalisation}; and \emph{cluster characterisation}, articulating how clusters
+existing data clustering methods, geographic constraints and
+regionalisation; and \emph{cluster characterisation}, articulating how clusters
 can be visually summarised.
@@
 1910~\citep{nhgis}, most studies consider the period between 1970 and 2010,
-\revision{1.3}{using pre-harmonised data from the Longitudinal Tract Data
-Base~\citep{Logan2014} or the Neighborhood Change Database~\citep{ncdb}, albeit
-their inherent errors~\citep{Logan2016,Hallisey2017}.} Similar harmonisation
+using pre-harmonised data from the Longitudinal Tract Data
+Base~\citep{Logan2014} or the Neighborhood Change Database~\citep{ncdb}, despite
+their inherent errors~\citep{Logan2016,Hallisey2017}. Similar harmonisation
 efforts appear in other countries~\citep{Liu2015,Lee2015,Allen2018}. These
@@
 data~\citep{Dmowska2017,Dmowska2018,stepinski2019imperfect}. Beyond the
-potentially increased spatial\revision{1.4}{ precision}, this approach does not
-require complex harmonisation when new data is considered, if the grids are
-compatible. However, demographic data is usually not available in this format,
-especially from older sources. Additionally, the conversion from tabulation
-areas can introduce significant errors.
+potentially increased spatial precision, this approach does not require complex
+harmonisation when new data is considered, if the grids are compatible. However,
+demographic data is usually not available in this format, especially from older
+sources. Additionally, the conversion from tabulation areas can introduce
+significant errors.
 
@@
 interpreted in the same way. There is no shortage of contributions for this
-problem~\citep{Fahad2014}\revision{}{, since variations of it appear in almost all
-scientific fields.}
+problem~\citep{Fahad2014}, since variations of it appear in almost all
+scientific fields.
 
-\revision{2.1}{In geography, this problem is known as
+In geography, this problem is known as
 \emph{regionalisation}~\citep{montello2003regions}, a rather old problem that
@@
 detection~\citep{blondel2008Louvain,Thomas2012}, or, to a lesser extent, Self
-Organising Maps~\citep{Delmelle2017,Ling2016,Dani2013_som_maxp}.}
+Organising Maps~\citep{Delmelle2017,Ling2016,Dani2013_som_maxp}.
 
-\revision{1.5/2.2}{Since we adopted a network-based data representation and our objectives include
+Since we adopted a network-based data representation and our objectives include
 an interactive interface, we opted for an heuristic variation of the maximum
@@
 hierarchical network clustering method, and distance metric, can be used in our
-framework. }
+framework.
 

### 2019-10-10 paper/study.tex
last paper review piccard 1.0 GEAN

@@
 
-\revision{1.12.(1-5)}{These results are meant to demonstrate the
-  utility of the interface for understanding the evolutionary dynamics
-  of urban neighbourhoods. They also show the face validity of the
-  results generated by our novel network-based approach. While our
-  method does not require geographic harmonisation, it does however
-  require matching the variables over time. This is (currently) a
-  painstaking, manual process, and the current version of the
-  interface remains somewhat limited. For example, we selected a
-  minimal set of aspects for each country, which are not necessarily
-  similar to each other, or similar to studies in the current
-  literature. Income moreover is slightly inaccurate, even though we
-  did correct for official inflation. We grouped the original ranges
-  into three larger ranges, but they do not match precisely. In
-  addition, the US data labelled "2010" is not from the decennial
-  census, but from the ACS 2006-2010, somewhat compromising the
-  temporal stability of the data. Further, the regions selected do not
-  correspond to any pre-defined regions (metros, cities, census
-  areas), but to somewhat arbitrary regions defined around a location
-  of interest, making direct comparisons harder.  Taken together,
-  these factors would no doubt require caveats for a substantive study
-  using the current prototype version of the interface. Nevertheless,
-  in the present context we find that the interface results are
-  well-aligned with several other studies, illustrating the robustness
-  of our approach to imperfect conditions. }
+These results are meant to demonstrate the utility of the interface for
+understanding the evolutionary dynamics of urban neighbourhoods. They also show
+the face validity of the results generated by our novel network-based approach.
+While our method does not require geographic harmonisation, it does however
+require matching the variables over time. This is (currently) a painstaking,
+manual process, and the current version of the interface remains somewhat
+limited. For example, we selected a minimal set of aspects for each country,
+which are not necessarily similar to each other, or similar to studies in the
+current literature. Income moreover is slightly inaccurate, even though we did
+correct for official inflation. We grouped the original ranges into three larger
+ranges, but they do not match precisely. In addition, the US data labelled
+"2010" is not from the decennial census, but from the ACS 2006-2010, somewhat
+compromising the temporal stability of the data. Further, the regions selected
+do not correspond to any pre-defined regions (metros, cities, census areas), but
+to somewhat arbitrary regions defined around a location of interest, making
+direct comparisons harder.  Taken together, these factors would no doubt require
+caveats for a substantive study using the current prototype version of the
+interface. Nevertheless, in the present context we find that the interface
+results are well-aligned with several other studies, illustrating the robustness
+of our approach to imperfect conditions.
 
@@
 Asian~\citep{Fong2003}. Toronto is also known for a stable and well defined
-Jewish community~\citep{Harold2018, Fong2011}. Therefore, we expect \revision{1.14}{a}
+Jewish community~\citep{Harold2018, Fong2011}. Therefore, we expect a
 combination of stable and dynamic regions on the results, with Place of Birth,
@@
     \centering 
-    \includegraphics[width=0.75\linewidth]{toVol.pdf}
+    \includegraphics[width=0.725\linewidth]{toVol.pdf}
     \caption{Details for some regions of Toronto that were classified into 3 or
@@
 hypothesis for occupational clustering (D), and highlighted how the method can
-be used to compare neighbourhoods and cities (A). In summary, their view was that
-the proposed methodology can be a viable alternative for the visual analytics of
-evolving demographic data.
-
-The interface was "easy to navigate" (B), but it was also considered
-"overwhelming" (A), "intimidating" (E), and "tricky to interpret" (C), possible
-side-effects of our effort to increase  representational accuracy, where we
-avoided using simplified representation or labels. Identifying clusters by their
-most relevant variables was welcome, but the overlap of information from
-different clusters in the boxplot was "a bit confusing" (C) when colour was not
-present. Further, most clusters can be sufficiently characterised using only the
-most relevant aspect, but this is not generally true.
+be used to compare neighbourhoods and cities (A). In summary, their view was
+that the proposed methodology can be a viable alternative for the visual
+analytics of evolving demographic data. The interface was "easy to navigate"
+(B), but it was also considered "overwhelming" (A), "intimidating" (E), and
+"tricky to interpret" (C), possible side-effects of our effort to increase
+representational accuracy, where we avoided using simplified representation or
+labels. Identifying clusters by their most relevant variables was welcome, but
+the overlap of information from different clusters in the boxplot was "a bit
+confusing" (C) when colour was not present. Further, most clusters can be
+sufficiently characterised using only the most relevant aspect, but this is not
+generally true.
 

### 2019-10-10 backend/srv.py

@@ -114,24 +114,24 @@ def _mapHiers(ds: dataStore, aspects: list, nClusters: int = 10, threshold: floa
     # -----------------------------------
     # cutting the hierarchies
 
-    for yg in tqdm(Gs):
-        y, g = yg
+    # for yg in tqdm(Gs):
+    #     y, g = yg
 
-        base_number_ccs = nx.number_connected_components(Gs[yg])
-        Gs[yg].remove_edges_from([e[:2]
-                                  for e in Gs[yg].edges(data='level')
-                                  if e[2] >= threshold])
-        current_number_ccs = nx.number_connected_components(Gs[yg])
-        while (current_number_ccs-base_number_ccs) < nClusters:
-            E = sorted([e for e in Gs[yg].edges(data='level')],
-                       key=lambda e: e[2], reverse=True)
-            Gs[yg].remove_edges_from(
-                E[:nClusters-(base_number_ccs-base_number_ccs)])
-            current_number_ccs = nx.number_connected_components(Gs[yg])
+    #     base_number_ccs = nx.number_connected_components(Gs[yg])
+    #     Gs[yg].remove_edges_from([e[:2]
+    #                               for e in Gs[yg].edges(data='level')
+    #                               if e[2] >= threshold])
+    #     current_number_ccs = nx.number_connected_components(Gs[yg])
+    #     while (current_number_ccs-base_number_ccs) < nClusters:
+    #         E = sorted([e for e in Gs[yg].edges(data='level')],
+    #                    key=lambda e: e[2], reverse=True)
+    #         Gs[yg].remove_edges_from(
+    #             E[:nClusters-(base_number_ccs-base_number_ccs)])
+    #         current_number_ccs = nx.number_connected_components(Gs[yg])
 
-        for cc, nodes in enumerate(nx.connected_components(Gs[yg])):
-            for n in nodes:
-                cl[y][g][n[1]] = cc
+    #     for cc, nodes in enumerate(nx.connected_components(Gs[yg])):
+    #         for n in nodes:
+    #             cl[y][g][n[1]] = cc
 
         # # # -------------------------------------------
         # # # Merging disconnected similar clusters

### 2019-10-13 paper/method.tex
not working

@@
 proportional to the smaller of the populations involved, pairwise. This
-approximation works well in general, but leads to small inconsistencies in some
-specific cases. For instance, in the results for Miami, which is available in
-the supplementary material, the combined width of the ribbons decreases
-slightly, despite an increase in the population numbers.
+approximation is necessary to interpolate two different population numbers into
+one and it works well in general, but it can lead to inconsistencies in some
+specific cases, where some regions can be counted more than once. For instance,
+in the results for Miami, which is available in the supplementary material, the
+combined width of the ribbons decreases over time, despite an increase in the
+population numbers. However, the population numbers presented in the bottom part
+of the interface are correct.
 

### 2019-10-13 backend/dataStore.py

@@ -1,17 +1,18 @@
 ##
-# THIS WILL CRUMBLE UNDER LOAD! 
+# THIS WILL CRUMBLE UNDER LOAD!
 #
 # With more than MAX_CACHE simultaneous users, the cache might erase the entry
-# between _check_read and the actual read/action. 
+# between _check_read and the actual read/action.
 #
 # Not fixing it because this whole file needs to be overhauled for GUDR support.
 # Otherwise, move the whole database to a proper DBMS.
 ##
 
-import json
 from glob import glob
+import json
 from os.path import basename, exists, join
 from random import choice
+from typing import Optional
 from uuid import uuid4
 
 import matplotlib.pylab as plt
@@ -58,8 +59,7 @@ def _check_level(G: nx.Graph):
                     used[y] = True
                     # if this edge is more distant than anything in G.neighbors(x), there is no viable path
                     if G[last_node][y]['level'] < cutoff:
-                        to_do.append(
-                            p+[(y, max([maxDistance, G[last_node][y]['level']]))])
+                        to_do.append(p+[(y, max([maxDistance, G[last_node][y]['level']]))])
 
 
 def _getMaxLevel(G: nx.Graph, level: str = 'level') -> int:
@@ -107,7 +107,8 @@ class dataStore(object):
                             unused.discard(op)
                             to_add.append((current+[op, ]))
                         continue
-                done_paths.append(current) #if this line runs, (last \not\in X) or (options=\emptyset)
+                # if this line runs, (last \not\in X) or (options=\emptyset)
+                done_paths.append(current)
 
             # puts in the paths that start in this aspect
             paths = to_add+[[(-1, -1), ]*i+[n, ] for n in unused]+done_paths
@@ -270,16 +271,18 @@ class dataStore(object):
         self._check_and_read(aspectID)
         return([self._info[aspectID]['descriptions'][x] for x in self._info[aspectID]['columns']])
 
-    def getData(self, aspectID: str, id: str, normalized=False) -> list:
+    def getData(self, aspectID: str, id: str, normalized=False) -> Optional[list]:
         self._check_and_read(aspectID, data=True)
         if id in self._data[aspectID].index:
             vals = list(self._data[aspectID].loc[id])
             if normalized:
-                if len(vals)>1:
+                if len(vals) > 1:
                     with np.errstate(divide='ignore', invalid='ignore'):
[…]

### 2019-10-13 backend/hierarchies.py

@@ -1,14 +1,12 @@
 from collections import defaultdict
-from itertools import combinations
-from os.path import join
 
 import matplotlib.pylab as plt
 import networkx as nx
 from joblib import Memory
-from numpy import median
-from tqdm import tqdm
 
-from upload import gatherInfoJsons_AsDict
+from tqdm import tqdm
+from numpy import median
+# from dataStore import dataStore
 
 cachedir = './cache/'
 memory = Memory(cachedir, verbose=0)
@@ -25,7 +23,7 @@ def _plotHier(H, threshold: float = 0.5):
         data='level') if e[2] >= threshold], edge_color='red')
 
 
-def compareHierarchies(ds: dict, a1, a2, g1: str = None, g2: str = None, level: str = 'level') -> float:
+def compareHierarchies(ds, a1, a2, g1: str = None, g2: str = None, level: str = 'level') -> float:
     """
         Computes a distance between hierarchies/aspects a1 and a2.
 
@@ -97,12 +95,29 @@ def _hierMerge(G1: nx.Graph, G2: nx.Graph, X: nx.Graph, level: str = 'level') ->
             H[e[0]][e[1]][level] = max(
                 [H[e[0]][e[1]][level], ]+[x[2] for x in G2.subgraph(otherside).edges(data=level)])
 
+    # plt.hist([e[2] for e in G1.edges(data='level')])
+    # plt.axis([0, 1, 0, len(G1.edges())])
+    # plt.title('G1')
+
+    # plt.figure()
+    # plt.hist([e[2] for e in G2.edges(data='level')])
+    # plt.axis([0, 1, 0, len(G2.edges())])
+    # plt.title('G2')
+
+    # plt.figure()
+    # plt.hist([e[2] for e in H.edges(data='level')])
+    # plt.axis([0, 1, 0, len(H.edges())])
+    # plt.title('H')
+
+    # plt.show()
+
+
     return(H)
 
 # @profile
 
 
-def _mergeAll(ds: dict, aspects: list, level: str = 'level', bbox: list = None) -> nx.Graph:
+def _mergeAll(ds, aspects: list, level: str = 'level', bbox: list = None) -> nx.Graph:
     F = None
     for a in aspects:
         G = ds.getHierarchy(a, bbox=bbox)
@@ -117,9 +132,9 @@ def _mergeAll(ds: dict, aspects: list, level: str = 'level', bbox: list = None) 
 
 
 @memory.cache(ignore=['ds'])
[…]

### 2019-10-13 backend/srv.py

@@ -1,26 +1,24 @@
 """Server module for the backend"""
+from collections import defaultdict
+from itertools import product
 import json
 import os
-import pickle
-import sys
-import tempfile
-from collections import defaultdict
-from itertools import combinations
 from os import makedirs
 from os.path import exists, join
+import pickle
 from random import sample
+import sys
+import tempfile
 from time import time
 
 import cherrypy
+from joblib import Memory
 import matplotlib.pylab as plt
 import networkx as nx
+from networkx.readwrite import json_graph
 import numpy as np
 import pandas as pd
-from joblib import Memory
-from networkx.readwrite import json_graph
 from scipy.spatial.distance import cosine, euclidean
-# from scipy.stats import wasserstein_distance,
-# from sklearn.cluster import KMeans
 from tqdm import tqdm
 
 from dataStore import dataStore
@@ -61,6 +59,8 @@ def _mergePaths(p1: dict, p2: dict, id: int):
     return(ret)
 
 # {'_sw': {'lng': -97.36262426260146, 'lat': 24.36091074100848}, '_ne': {'lng': -65.39166177011971, 'lat': 33.61501866716327}}
+
+
 def _bbox_create_buffer(bbox: dict) -> list:
     minX = bbox['_sw']['lng']
     minY = bbox['_sw']['lat']
@@ -90,10 +90,100 @@ def cors():
         cherrypy.response.headers['Access-Control-Allow-Origin'] = '*'
 
 
-@memory.cache(ignore=['ds',])
-def _mapHiers(ds: dataStore, aspects: list, nClusters: int = 10, threshold: float = 0.75, bbox: list = None):
+def _convert_name_node(n: tuple, y: int) -> tuple:
+    """ (geom,id) -> (year, geom, id)"""
+    return((y, n[0], n[1]))
+
+
+def _convert_name_graph(G: nx.Graph, y: int) -> nx.Graph:
+    """ Adds y to the names of the graphs nodes (geom,id) -> (year, geom, id)"""
+    ret = nx.Graph()
+    for n in G:
+        ret.add_node(_convert_name_node(n, y))
+    for e in G.edges():
+        n1 = _convert_name_node(e[0], y)
+        n2 = _convert_name_node(e[1], y)
+        ret.add_edge(n1, n2)
+        if 'level' in G[e[0]][e[1]]:
+            ret[n1][n2]['level'] = G[e[0]][e[1]]['level']
[…]

### 2019-10-26 paper/intro.tex
final version paper piccard 1.0

@@
 Section~\ref{sec:expert}. Our prototype system is available at
-\censure{\url{http://uoft.me/piccard}}, including more than forty regions in the
-US and Canada. The source code is publicly available at
-\censure{\url{https://github.com/fabioasdias/piccard}}.\footnote{The editors are
-considering, at our request, an exception to the double-blind requirement to
-allow access to the system. We provided them with the URLs of the system, code,
-and documentation separately.}
+\url{http://uoft.me/piccard}, including more than forty regions in the US and
+Canada. The source code is publicly available at
+\url{https://github.com/fabioasdias/piccard}.
 
@@
     \subfloat{
-        \begin{tikzpicture}
+        \begin{tikzpicture}[scale=0.75, every node/.style={scale=0.75}]
             \node (push) at (-2,0) {};

### 2019-12-10 paper/discussion.tex
merged the published version of the paper with the actual code

@@
 
-Removing the need for geographical harmonisation greatly reduces the amount of
-work necessary to explore demographic data, but the method still requires
-consistent variables across the years. Matching the variables can be trivial for
-some aspects (Age), but challenging for others (Income). The divulged income
-ranges vary over time and the actual values change due to inflation. Some
-variables were not considered in earlier censuses, such as Race in Canada, or
-Hispanic population in the USA, hampering its use when they are available. Since
-this is only a prototype, we matched few aspects, but a proper demographic
-analysis would benefit from all available information.
+We removed the need for geographical harmonisation, but the method still
+requires consistent variables across the years. Matching the variables can be
+trivial for some aspects (Age), but challenging for others (Income). The
+divulged income ranges vary over time and the actual values change due to
+inflation. Since this is only a prototype, we matched few aspects, but a proper
+demographic analysis would benefit from all available information.
 

### 2019-12-10 paper/method.tex
merged the published version of the paper with the actual code

@@
 year. Further, edges are placed between nodes if the corresponding regions
-belong to sequential years and their geometries intersect.
-To avoid spurious connections caused by small fluctuations, one of the
-geometries is slightly shrunk before the intersection, using a buffer of -1e-6
-degrees. While the result would not change considerably by adding more
-connections between nearby regions, these extra connections would greatly
-increase the computational cost. More importantly, while weights will be
-associated with these edges before they are processed, they are not derived from
-the geometry, but from the data. The actual intersection area is not considered
-in this representation.  This approach leads to a single network representing
-the whole spatio-temporal space of the data.
+belong to sequential years and their geometries intersect. To avoid spurious
+connections caused by small fluctuations, one of the geometries is slightly
+shrunk before the intersection, using a buffer of -1e-6 degrees, or about 10cm
+at the equator. As long as this buffer is not large enough to remove significant
+overlaps, its exact value will not have a significant impact on the results,
+since they also rely on the data similarity, and it could safely be set to zero.
+However, this would lead to extra connections that would increase the
+computational cost. More importantly, while weights will be associated with
+these edges before they are processed, they are not derived from the geometry,
+but from the data. The actual intersection area is not considered in this
+representation.  This approach leads to a single network representing the whole
+spatio-temporal space of the data.
 
@@
 information, we call \emph{temporal} any path such that the temporal information
-increases along it. For instance, in Figure~\ref{fig:intuition}, the paths ACF,
-ADF, ADG, and BEH are temporal paths, but not HEB, CDE, nor FDA. With harmonised
-data, as illustrated by the path BEH, the time-series of each region would form
-a temporal path, each node would be connected only to its older and newer
-versions, and would belong to only one temporal path. Since our data is not
-harmonised, more connections are allowed and each node can belong to an
-arbitrary number of paths.
+increases along it, representing information advancing in time. For instance, in
+Figure~\ref{fig:intuition}, the paths ACF, ADF, ADG, and BEH are temporal paths,
+but not HEB, CDE, nor FDA. With harmonised data, as illustrated by the path BEH,
+the time-series of each region would form a temporal path, each node would be
+connected only to its older and newer versions, and would belong to only one
+temporal path. Since our data is not harmonised, more connections are allowed
+and each node can belong to an arbitrary number of paths.
 
